### 2015-07-05 write_up/cricket.tex
Reorg

Unfairly Random?

In many cricket matches, it is claimed that there is a clear advantage to bowling (batting) first. The advantage is pointed to by commentators, and by captains of the competing teams in the pre-toss interview. And sometimes in the post-match interview.

The opportunity to bowl or bat first is decided by a coin toss. While this method of deciding on who is advantaged is fair on average, the system isn’t fair in any one game. At first glance, the imbalance seems inevitable. After all someone has to bat first. One can, however, devise a baseball like system where short innings are interspersed. If that violates the nature of the game too much, one can easily create pitches that don’t deteriorate appreciably over the course of a game. Or, one can come up with an estimate of the advantage and adjust scores accordingly (something akin to an adjustment issued when matches are shortened due to rain).

But before we move to seriously consider these solutions, one may ask about the evidence.

Data are from nearly five thousand one-day international matches.

The team that wins the toss wins the match approximately 49.3\% of the times. With 5335 matches, we cannot rule out that the true proportion is 50\%. Thus, counter to intuition, the effect of winning the toss is, on average, at best minor. This may be so because it is impossible to predict well in advance the advantage of bowling or batting first. Or it may simply be because teams are bad at predicting it, perhaps because they use bad heuristics.

### 2015-07-05 download_data.py

@@ -1,43 +1,0 @@
-'''
-Download Cricket Data
-Last Edited: 05.27.15
-
-@author: Gaurav Sood
-
-'''
-
-#!/usr/bin/env python
-# -*- coding: utf-8 -*-
-
-import requests
-import math
-import csv
-import sys
-import time
-import os
-import simplejson
-import unicodedata
-from BeautifulSoup import BeautifulSoup, SoupStrainer
-
-for match_type in ['odi', 'test', 't20i', 't20', 'list%20a', 'first%20class']:
-    results = []
-    r = requests.get('http://search.espncricinfo.com/ci/content/match/search.html?all=1;page=0;search=' + match_type)
-    soup = BeautifulSoup(r.text)
-    last_match = int(soup.findAll('span', attrs={'class':'PaginationNmbrs'})[-1].text)
-    last_page = int(math.ceil(float(last_match)/float(20)))
-    for i in range(0, last_page):
-        time.sleep(2)
-        results_page = requests.get("http://search.espncricinfo.com/ci/content/match/search.html?search={0};all=1;page={1}".format(match_type, i))
-        soupy = BeautifulSoup(results_page.text)
-        for new_host in soupy.findAll('a', {'class' : 'srchPlyrNmTxt'}):
-            try:
-                new_host = new_host['href']
-            except:
-                continue
-            new_host = unicodedata.normalize('NFKD', new_host).encode('ascii','ignore')
-            #print(type(str.split(new_host)[3]))
-            print str.split(new_host, "/")[4].split('.')[0]
-            results.append(str.split(new_host, "/")[4].split('.')[0])
-
-    with open("matches-{0}.json".format(match_type), "wb") as f:
-        simplejson.dump(results, f)

### 2015-07-05 parse_cric.py

@@ -1,31 +1,0 @@
-'''
-Parse Downloaded Cricket Data
-
-'''
-
-import time
-import csv
-import simplejson
-from espncricinfo.match import Match
-
-headers = ["url", "team1", "team2", "win_toss", "bat_or_bowl", "outcome", "win_game", "date", "day_n_night", "ground", "rain", "duckworth_lewis", "match_id", "type_of_match"]
-
-results = simplejson.loads(open('matches-odi.json').read())
-
-##################################START PROCESSING DATA#########################################
-with open("final_output.csv", "wb") as csvfile:
-    writer = csv.writer(csvfile)
-    writer.writerow(headers)
-    for result in results:
-        time.sleep(2)
-        m = Match(int(result))
-        if m.match_json()['match_status'] == 'forthcoming':
-            continue
-        if m.rain_rule() == 'D/L method':
-            duckworth_lewis = 1
-        else:
-            duckworth_lewis = 0
-        writer.writerow([m.team_1()['team_name'], m.team_2()['team_name'], m.toss_winner(), m.toss_decision(), m.result(), m.match_winner(), m.date(), m.lighting(), m.ground_name(), None, duckworth_lewis, result, m.match_class()])
-
-##################################FINISHED#########################################
-print "DONE. Wrote output to %s" %(FINAL_OUTPUT_FILE,)

### 2015-07-05 scripts/02_parse_cric.py

@@ -1,0 +1,31 @@
+'''
+Parse Downloaded Cricket Data
+
+'''
+
+import time
+import csv
+import simplejson
+from espncricinfo.match import Match
+
+headers = ["url", "team1", "team2", "win_toss", "bat_or_bowl", "outcome", "win_game", "date", "day_n_night", "ground", "rain", "duckworth_lewis", "match_id", "type_of_match"]
+
+results = simplejson.loads(open('matches-odi.json').read())
+
+##################################START PROCESSING DATA#########################################
+with open("final_output.csv", "wb") as csvfile:
+    writer = csv.writer(csvfile)
+    writer.writerow(headers)
+    for result in results:
+        time.sleep(2)
+        m = Match(int(result))
+        if m.match_json()['match_status'] == 'forthcoming':
+            continue
+        if m.rain_rule() == 'D/L method':
+            duckworth_lewis = 1
+        else:
+            duckworth_lewis = 0
+        writer.writerow([m.team_1()['team_name'], m.team_2()['team_name'], m.toss_winner(), m.toss_decision(), m.result(), m.match_winner(), m.date(), m.lighting(), m.ground_name(), None, duckworth_lewis, result, m.match_class()])
+
+##################################FINISHED#########################################
+print "DONE. Wrote output to %s" %(FINAL_OUTPUT_FILE,)

### 2015-07-05 scripts/03_cricket.R
shell of script file

@@ -1,0 +1,31 @@
+"
+Unfairly Random? Impact of Winning the Coin Toss on Probability of Winning
+
+"
+
+# setwd
+setwd("cricket-stats")
+
+# Read in data
+cricket <- read.csv("data/final_output.csv")
+
+# Recode
+
+# Figs
+library(ggplot2)
+
+# Win by Format
+ggplot(cricket, aes()) + 
+
+ggsave("../figs/winbyType")
+
+# Win by Day/Night
+ggplot(cricket, aes()) + 
+
+ggsave("../figs/winbyDayNight")
+
+# Win by DL
+ggplot(cricket, aes()) + 
+
+ggsave("../figs/winbyDL")
+

### 2015-07-09 write_up/cricket.tex
reorg
initial graphs

@@
-Unfairly Random?
+\documentclass[11pt]{article}
+\usepackage{color}
+\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+\definecolor{dark-red}{rgb}{0.7,0.1,0.1} 
+\definecolor{dark-blue}{rgb}{0,0,0.7} 
+\usepackage[linkcolor=dark-red,
+            colorlinks=true,
+            urlcolor=dark-blue,
+            pdfstartview={XYZ null null 1.00},
+            pdfauthor={Gaurav Sood, gsood07@gmail.com; Andy Guess, a.guess@columbia.edu},
+            citecolor=dark-red,
+            bookmarks=false,
+            pdfborder={0 0 0},
+            pdftitle={Quantitative Discipline}]{hyperref}
+
+\usepackage{amsfonts,amssymb,amsbsy,amsmath,amsxtra}
+
+\usepackage{indentfirst}
+\usepackage{setspace} % To set line spacing
+\usepackage{multirow}
+
+\usepackage{verbatim}
+
+\usepackage[multiple]{footmisc}
+\usepackage{fancyvrb}
+
+\usepackage{longtable}
+
+\usepackage[margin=1in]{geometry}
+\usepackage{graphicx}
+
+\raggedright
+\parindent=1.5em % <- or whatever indent you want
+
+\begin{comment}
+
+	tools::texi2dvi("cricket.tex", pdf=TRUE, clean=TRUE)
+
+\end{comment}
+
+\begin{document}
+\title{\vspace{-2.0cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}}
+\author{Gaurav Sood\thanks{\href{mailto:gsood07@gmail.com}{\texttt{gsood07@gmail.com}}.} \and Derek Willis\thanks{\href{mailto:dwillis@gmail.com}{\texttt{dwillis@gmail.com}}.}}
+\maketitle
+\doublespacing
 
@@
 The opportunity to bowl or bat first is decided by a coin toss.
-While this method of deciding on who is advantaged is fair on average, the system isn’t fair in any one game.
+While this method of deciding on who is advantaged is fair on average, the system isn't fair in any one game.
 At first glance, the imbalance seems inevitable.
@@
 One can, however, devise a baseball like system where short innings are interspersed.
-If that violates the nature of the game too much, one can easily create pitches that don’t deteriorate appreciably over the course of a game.
+If that violates the nature of the game too much, one can easily create pitches that don't deteriorate appreciably over the course of a game.
 Or, one can come up with an estimate of the advantage and adjust scores accordingly (something akin to an adjustment issued when matches are shortened due to rain).
@@
 Or it may simply be because teams are bad at predicting it, perhaps because they use bad heuristics.
+
+\begin{figure}[htbp]
+\centering
+\caption{Proportion of Matches Won Minus Matches Lost After Winning the Toss}
+\includegraphics[scale=.75]{../figs/winbyType.pdf}
+\label{fig:summary}
+\end{figure}
+
+No effects across the entire sample may hide some subgroup effects.
+It is often claimed that toss is more crucial in day and night matches, due to dew and lower visibility of the white ball under lights.
+And data show as much.
+
+It may well be the case that toss is more important in tests than one-day matches.
+
+\begin{figure}[htbp]
+\centering
+\caption{Proportion of Matches Won Minus Matches Lost After Winning the Toss}
+\includegraphics[scale=.75]{../figs/winbyDayNight.pdf}
+\label{fig:summary}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Proportion of Matches Won Minus Matches Lost After Winning the Toss}
+\includegraphics[scale=.75]{../figs/winbyDL.pdf}
+\label{fig:summary}
+\end{figure}
+
+\end{document}

### 2015-07-09 scripts/03_cricket.R

@@ -1,5 +1,6 @@
 "
-Unfairly Random? Impact of Winning the Coin Toss on Probability of Winning
+@title: Fairly Random (or Choosing Badly): Impact of Winning the Toss on the Probability of Winning
+@authors: Gaurav Sood and Derek Willis
 
 "
 
@@ -9,23 +10,138 @@ setwd("cricket-stats")
 # Read in data
 cricket <- read.csv("data/final_output.csv")
 
+"
+Take out matches with no toss (~ no match)
+About 3k matches. Don't like the 7% number. Touch too high, imho. But checked data - turns out to be ok.
+
+"
+cricket <- subset(cricket, win_toss!="")
+
+"
+Take out matches where there was no result
+Leaves us w/ 39672 rows
+"
+
+cricket <- subset(cricket, outcome!="No result")
+
+"
+Drawn Matches
+6674 of them ~ 17%
+"
+
+cricket$draw <- 1*(cricket$outcome=="Match drawn")
+
+# table(cricket$win_game[cricket$draw==1])
+
+"
+Winning a toss causes outcome including draws.
+Imp. esp. for first class games
+
+"
+
 # Recode
+# Coding team that wins the toss wins the game
+cricket$tossgame <- 1*(cricket$win_game==cricket$win_toss)
+# The game is drawn
+cricket$tossgame[cricket$draw==1] <- .5
+
+# Results
+ddply(cricket,~type_of_match + day_n_night,summarise,mean=mean(tossgame))
+ddply(cricket,~type_of_match + duckworth_lewis,summarise,mean=mean(tossgame))
 
 # Figs
 library(ggplot2)
 
+"
+By Match Type
+Let us plot 
+"
+
+win_match_type <- ddply(cricket,~type_of_match,summarise, diff=mean(I(tossgame==1) - I(tossgame==0)), count=length(unique(url)))
+win_match_type$diff <- win_match_type$diff*100 # convert to %
+
 # Win by Format
-ggplot(cricket, aes()) + 
+ggplot(win_match_type, aes(type_of_match, diff)) + 
+geom_bar(stat = "identity", position = "identity", fill="#42C4C7") + 
+theme_minimal() + 
+xlab("") +
+scale_y_continuous(breaks=seq(-.5, 7, 1), labels=nolead0s(seq(-.5, 7, 1)), limits=c(-.5, 7), name="") +
+theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted"),
[…]

### 2015-07-09 write_up/cricket.tex
updated graphs

@@
             pdfstartview={XYZ null null 1.00},
-            pdfauthor={Gaurav Sood, gsood07@gmail.com; Andy Guess, a.guess@columbia.edu},
+            pdfauthor={Gaurav Sood, gsood07@gmail.com; Derek Willis, dwillis@gmail.com},
             citecolor=dark-red,
@@
             pdfborder={0 0 0},
-            pdftitle={Quantitative Discipline}]{hyperref}
+            pdftitle={Fairly Random}]{hyperref}
 
@@
 And data show as much.
-
 It may well be the case that toss is more important in tests than one-day matches.

### 2015-07-09 scripts/03_cricket.R

@@ -4,8 +4,11 @@
 
 "
 
+# Load libs
+library(plyr)
+
 # setwd
-setwd("cricket-stats")
+setwd(paste0(githubdir, "/cricket-stats"))
 
 # Read in data
 cricket <- read.csv("data/final_output.csv")
@@ -50,37 +53,39 @@ ddply(cricket,~type_of_match + day_n_night,summarise,mean=mean(tossgame))
 ddply(cricket,~type_of_match + duckworth_lewis,summarise,mean=mean(tossgame))
 
 # Figs
+
+# Fig libs
 library(ggplot2)
+library(grid)
+
+# For figs - let us get type of match is nicer factor order
+cricket$type_of_match <- factor(cricket$type_of_match, levels=c("FC", "TEST", "LISTA", "ODI", "T20", "T20I"))
 
 "
-By Match Type
-Let us plot 
+Win By Match Type
 "
 
 win_match_type <- ddply(cricket,~type_of_match,summarise, diff=mean(I(tossgame==1) - I(tossgame==0)), count=length(unique(url)))
 win_match_type$diff <- win_match_type$diff*100 # convert to %
 
-# Win by Format
 ggplot(win_match_type, aes(type_of_match, diff)) + 
 geom_bar(stat = "identity", position = "identity", fill="#42C4C7") + 
 theme_minimal() + 
 xlab("") +
-scale_y_continuous(breaks=seq(-.5, 7, 1), labels=nolead0s(seq(-.5, 7, 1)), limits=c(-.5, 7), name="") +
+scale_y_continuous(breaks=seq(-1, 7, 1), labels=nolead0s(seq(-1, 7, 1)), limits=c(-1, 7), name="") +
 theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted"),
-        panel.grid.minor.x = element_blank(),
-        panel.grid.major.x = element_line(colour = "#f7f7f7", linetype = "solid"),
-        panel.border       = element_blank(),
-        legend.position  = "none",
-        legend.key       = element_blank(),
-        legend.key.width = unit(1,"cm"),
-        axis.title   = element_text(size=10),
-        axis.text    = element_text(size=8),
-        axis.ticks.y = element_blank(),
-        axis.line.x  = element_line(colour = 'red', size = 3, linetype = 'dashed'),
-        axis.title.x = element_text(vjust=-1),
-        axis.title.y = element_text(vjust= 1),
[…]

### 2015-07-10 write_up/cricket.tex
updated writeup

@@
 The advantage is pointed to by commentators, and by captains of the competing teams in the pre-toss interview.
-And sometimes in the post-match interview.
+And by the losing captain in the post-match interview.
 
@@
 
-But before we move to seriously consider these solutions, one may ask about the evidence.
+But before we move to seriously consider these solutions, we may ask about the evidence.
 
-Data are from nearly five thousand one-day international matches.
+Data are from 43,185 first-class men's cricket matches.
+It is a virtual census of the relevant population.
+We have data from matches in all formats.\footnote{For scripts used to scrape the data and the final data set, go \href{http://github.com/soodoku/get-cricket-data/}{here}.} Of the matches, we do not have data on the toss for 2,807 or roughly 7\% of the matches.
+The primary reason for this is match being abandoned without play.
+We exclude data from these matches.
 
-The team that wins the toss wins the match approximately 49.3\% of the times.
-With 5335 matches, we cannot rule out that the true proportion is 50\%.
+In limited over cricket, a minimum number of overs must be bowled to establish a result.
+In a One Day match, for instance, each side must bat at least 20 overs.
+In 706 matches, or roughly 1.7\% of the remaining matches, not enough overs were bowled to get a result.
+We exclude these matches from our analysis.
+This leaves us with a final tally of 39,672 matches.
+
+In cricket matches, three results are possible -- a draw, a win, or a loss.
+In limited overs cricket, a win or a loss are the most common outcomes.
+In the first part of our analysis, we focus on winning or losing.
+We estimate the difference in probability of winning and losing.
+The team that wins the toss wins the match 2.3\% more often than lose it.
+This effect, however, is highly variable by format, by conditions, and by whether or not a particular formula was used to adjust scores when it rained.
+In particular, it is often claimed that toss is more crucial in day and night matches, due to dew and lower visibility of the white ball under lights.
+And it is often claimed that the toss is more important in tests than one-day matches.
+And data show as much.
+
 Thus, counter to intuition, the effect of winning the toss is, on average, at best minor.
@@
 \end{figure}
-
-No effects across the entire sample may hide some subgroup effects.
-It is often claimed that toss is more crucial in day and night matches, due to dew and lower visibility of the white ball under lights.
-And data show as much.
-It may well be the case that toss is more important in tests than one-day matches.
 

### 2015-07-10 write_up/cricket.tex
fixed more writing

--- 2015-07-09 write_up/cricket.tex
@@
 \begin{document}
-\title{\vspace{-2.0cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}}
-\author{Gaurav Sood\thanks{\href{mailto:gsood07@gmail.com}{\texttt{gsood07@gmail.com}}.} \and Derek Willis\thanks{\href{mailto:dwillis@gmail.com}{\texttt{dwillis@gmail.com}}.}}
+\title{\vspace{-1.5cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}}
+\author{Gaurav Sood\\\small{\href{mailto:gsood07@gmail.com}{\tt{gsood07@gmail.com}}} \and Derek Willis\\\small{\href{mailto:dwillis@gmail.com}{\tt{dwillis@gmail.com}}}}
 \maketitle
@@
 
-In many cricket matches, it is claimed that there is a clear advantage to bowling (batting) first.
-The advantage is pointed to by commentators, and by captains of the competing teams in the pre-toss interview.
-And sometimes in the post-match interview.
+In nearly all cricket matches, it is claimed that there is a clear advantage to bowling (batting) first.
+The advantage is pointed to by commentators, by team captains in the pre-toss interview, and by the captain of the losing team in the post-match interview.
+This isn't to say that there is actually an advantage to winning the toss.
+For it may be impossible to predict well in advance the advantage of bowling or batting first.
+Or it may be that teams squander the potential advantage by using bad heuristics to choose what they do.
 
-The opportunity to bowl or bat first is decided by a coin toss.
-While this method of deciding on who is advantaged is fair on average, the system isn't fair in any one game.
+The opportunity to choose whether to bowl or bat first is decided by a coin toss.
+While this method of granting advatange is fair, on average, the system isn't fair in any one game.
 At first glance, the imbalance seems inevitable.
-After all someone has to bat first.
+After all, someone has to bat first.
 One can, however, devise a baseball like system where short innings are interspersed.
@@
 
-But before we move to seriously consider these solutions, one may ask about the evidence.
+But before we move to seriously consider these solutions, we may ask about the evidence.
 
-Data are from nearly five thousand one-day international matches.
+Data are from 43,185 first-class men's cricket matches.\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{http://github.com/soodoku/get-cricket-data/}{http://github.com/soodoku/get-cricket-data/}.} It is a near census of the relevant population.
+We have data on all types of matches: 20-20, one-day, and five-day.\footnote{In English county cricket, first-class matches last four days.
+Some first class matches last two days.
+Yet others three days.
+Some last just a day.
+And till a particular point, a test match lasted as long as it was needed to finish a game.
+We elide over these differences in our analyses.} Of these matches, we do not have data on the toss for 2,807, or roughly 7\%, of the matches.
+The primary reason we don't have data on these matches is because the match was abandoned without play.
+We exclude data from these matches.
 
-The team that wins the toss wins the match approximately 49.3\% of the times.
-With 5335 matches, we cannot rule out that the true proportion is 50\%.
-Thus, counter to intuition, the effect of winning the toss is, on average, at best minor.
-This may be so because it is impossible to predict well in advance the advantage of bowling or batting first.
-Or it may simply be because teams are bad at predicting it, perhaps because they use bad heuristics.
+In limited overs cricket, a minimum number of overs must be bowled to establish a result.
+In a one-day match, for instance, each side must bat at least 20 overs for a result to be declared.
+In 706 matches, or roughly 1.7\% of the remaining matches, not enough overs were bowled to get a result.
+We exclude these matches from our analysis.
+This leaves us with 39,672 matches.
+We analyse these data.
+
+In cricket matches, three results are possible -- a draw, a win, or a loss.
+We look at draws -- a common outcome in matches in which overs aren't limited -- later on.
+For now, we focus on winning and losing.
+We elide over the small issue of differential number of draws across formats by estimating the difference in probability of winning and losing.
+
+The team that wins the toss wins the match 2.3\% more often than lose it.
+This is a reasonable sized advantage -- though likely much smaller than the number that most commentators carry in their heads -- in a competitive sport.
+This advantage, however, is highly variable by format, by conditions, and by whether or not a particular formula was used to adjust scores when it rained.
+Some of the variability is expected (but as we will see, expectations are often dashed).
+It is often claimed that the toss is more crucial in day and night matches, due to dew and lower visibility of the white ball under lights.
+And it is often claimed that the toss is more important in tests than one-day matches.
+To the results then.
 
@@
 \caption{Proportion of Matches Won Minus Matches Lost After Winning the Toss}
-\includegraphics[scale=.75]{../figs/winbyType.pdf}
-\label{fig:summary}
-\end{figure}
-
-No effects across the entire sample may hide some subgroup effects.
-It is often claimed that toss is more crucial in day and night matches, due to dew and lower visibility of the white ball under lights.
-And data show as much.
-It may well be the case that toss is more important in tests than one-day matches.
-
-\begin{figure}[htbp]
-\centering
-\caption{Proportion of Matches Won Minus Matches Lost After Winning the Toss}
-\includegraphics[scale=.75]{../figs/winbyDayNight.pdf}
+\includegraphics[scale=.85]{../figs/winbyType.pdf}
 \label{fig:summary}
@@
 \caption{Proportion of Matches Won Minus Matches Lost After Winning the Toss}
-\includegraphics[scale=.75]{../figs/winbyDL.pdf}
+\includegraphics[scale=.85]{../figs/winbyDayNight.pdf}
+\label{fig:summary}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\caption{Proportion of Matches Won Minus Matches Lost After Winning the Toss}
+\includegraphics[scale=.85]{../figs/winbyDL.pdf}
 \label{fig:summary}

### 2015-07-10 scripts/03_cricket.R

@@ -155,7 +155,7 @@ theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted")
 	  legend.text        = element_text(size=8),
       plot.margin        = unit(c(0,.5,.5,.5), "cm")) + 
 annotate("text", x = 4.2, y = .15, label = "zero", size=3) 
-ggsave("figs/winbyDL.pdf")
+ggsave("figs/winbyDL.pdf", width=5)
 
 "
 Is there over time learning? If so, toss adv. would increase.

### 2015-07-12 scripts/parse_rankings.py
added rankings parser

@@ -1,0 +1,47 @@
+import csv
+import requests
+from BeautifulSoup import BeautifulSoup
+
+MONTH_MAPPING = {
+    'JANUARY': 1,
+    'FEBRUARY': 2,
+    'MARCH': 3,
+    'APRIL': 4,
+    'MAY': 5,
+    'JUNE': 6,
+    'JULY': 7,
+    'AUGUST': 8,
+    'SEPTEMBER': 9,
+    'OCTOBER': 10,
+    'NOVEMBER': 11,
+    'DECEMBER': 12
+}
+
+FORMAT_RANGES = {
+    'test': range(1952, 2014),
+    'odi': range(1981, 2014)
+}
+
+def parser(format):
+    with open("rankings_%s.csv" % format, "wb") as csvfile:
+        writer = csv.writer(csvfile)
+        writer.writerow(['format', 'year', 'month', 'rank', 'country', 'points'])
+        for year in FORMAT_RANGES[format]:
+            rankings = fetch_rankings(year, format)
+            for ranking in rankings:
+                writer.writerow(ranking)
+
+def fetch_rankings(year, format):
+    results = []
+    url = "http://web.archive.org/web/20130320093711/http://www.icc-cricket.com/match_zone/%s_ranking.php?year=%s" % (format, year)
+    r = requests.get(url)
+    soup = BeautifulSoup(r.text)
+    month_links = soup.findAll('a', attrs={'style':'color:#666666;'})
+    months = [MONTH_MAPPING[m.text] for m in month_links]
+    tables = soup.findAll('table', attrs={'class':'dataBox topMargin'})
+    combo = zip(months, tables)
+    for month, table in combo:
+        for row in table.findAll('tr')[1:-1]:
+            rank, country, points = [td.text for td in row.findAll('td')]
+            results.append([format, year, month, int(rank), country, int(points)])
+    return results

### 2015-07-12 scripts/parse_rankings.py
proper path for rankings csvs

@@ -23,7 +23,7 @@ FORMAT_RANGES = {
 }
 
 def parser(format):
-    with open("rankings_%s.csv" % format, "wb") as csvfile:
+    with open("../data/rankings_%s.csv" % format, "wb") as csvfile:
         writer = csv.writer(csvfile)
         writer.writerow(['format', 'year', 'month', 'rank', 'country', 'points'])
         for year in FORMAT_RANGES[format]:

### 2015-07-12 scripts/parse_rankings.py
rating, not points

@@ -25,7 +25,7 @@ FORMAT_RANGES = {
 def parser(format):
     with open("../data/rankings_%s.csv" % format, "wb") as csvfile:
         writer = csv.writer(csvfile)
-        writer.writerow(['format', 'year', 'month', 'rank', 'country', 'points'])
+        writer.writerow(['format', 'year', 'month', 'rank', 'country', 'rating'])
         for year in FORMAT_RANGES[format]:
             rankings = fetch_rankings(year, format)
             for ranking in rankings:
@@ -42,6 +42,6 @@ def fetch_rankings(year, format):
     combo = zip(months, tables)
     for month, table in combo:
         for row in table.findAll('tr')[1:-1]:
-            rank, country, points = [td.text for td in row.findAll('td')]
-            results.append([format, year, month, int(rank), country, int(points)])
+            rank, country, rating = [td.text for td in row.findAll('td')]
+            results.append([format, year, month, int(rank), country, int(rating)])
     return results

### 2015-07-12 scripts/parse_rankings.py
added parser for ICC rankings API

@@ -45,3 +45,11 @@ def fetch_rankings(year, format):
             rank, country, rating = [td.text for td in row.findAll('td')]
             results.append([format, year, month, int(rank), country, int(rating)])
     return results
+
+def fetch_current_rankings(format):
+    """
+    format is one of: 'test', 'odi' or 't20i'
+    """
+    r = requests.get("http://www.icc-cricket.com/api/getRankings")
+    rankings_json = r.json()
+    return [r['rankings'] for r in rankings_json if r['matchType'] == format.upper()][0]

### 2015-09-01 scripts/04_merge_ranking_data.R
Merge Ranking Data

@@ -1,0 +1,82 @@
+"
+@title: Merge Ranking Data with Cricket Data
+@authors: Gaurav Sood and Derek Willis
+
+"
+
+# setwd
+setwd(paste0(githubdir, "/cricket-stats"))
+
+# Don't want to deal with factors
+options(StringsAsFactors=F)
+
+# Load ranking data
+odi_ranks  <- read.csv("data/rankings_odi.csv")
+test_ranks <- read.csv("data/rankings_test.csv")
+
+"
+Notes: ODI/Test ranks data till 2013
+"
+
+# Load match data
+match 		<- read.csv("data/final_output.csv")
+
+# Match on dates + teams
+# Only have rankings for international ODI and Tests
+# Two teams per match
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+# Match dates - need only month and year
+# Clean a bit
+trim.trailing <- function (x) sub("\\s+$", "", x)
+match$date <- trim.trailing(match$date)
+
+# Month
+# Matches span single, multiple days
+# Some weirdness: table(substr(match$date, 2, 4))
+match$month <- substr(match$date, 2, 4)
+
+# Source of weirdness
+match$date[!(match$month %in% month.abb)] 
+"
+1. Bunch of matches started in 1 year ended in another. Did bad date truncation
+2. No dates
+3. One weird MJ
+"
+
+# Clean up weird cases
+match$month[!(match$month %in% month.abb)] <- substr(match$date[!(match$month %in% month.abb)], 9,11)
+match$month[match$month=='ke '] <- NA
+match$month[match$month==""] <- NA
+ 
+# Year (nearly clean)
+datelen    <- nchar(as.character(match$date))
+match$year <- substr(match$date, datelen-3, datelen)
+match$year[match$year=='unch'] <- NA
+
+# Go for exact match
+# Unique_ID1, Unique_ID2
+match$team1_id <- with(match, paste0(type_of_match, team1, month, year))
+match$team2_id <- with(match, paste0(type_of_match, team2, month, year))
+
+# Month handling for odi and test rank data (just convert to month abb. here as data cleaner)
+odi_ranks$month_abb <- month.abb[odi_ranks$month]
+test_ranks$month_abb <- month.abb[test_ranks$month]
+
+# Uniques for odi and test
[…]

### 2015-09-01 scripts/05_cricket.R

@@ -4,21 +4,24 @@
 
 "
 
-# Load libs
-library(plyr)
-
 # setwd
 setwd(paste0(githubdir, "/cricket-stats"))
 
-# Read in data
-cricket <- read.csv("data/final_output.csv")
+# Source merge script
+source("scripts/04_merge_ranking_data.R")
+
+# Load libs
+library(plyr)
+
+# Read in data/ now sourcing it - see 04_merge_ranking_data
+# cricket <- read.csv("data/final_output.csv")
 
 "
 Take out matches with no toss (~ no match)
 About 3k matches. Don't like the 7% number. Touch too high, imho. But checked data - turns out to be ok.
 
 "
-cricket <- subset(cricket, win_toss!="")
+cricket <- subset(match, win_toss!="")
 
 "
 Take out matches where there was no result
@@ -44,7 +47,7 @@ Imp. esp. for first class games
 
 # Recode
 # Coding team that wins the toss wins the game
-cricket$tossgame <- 1*(cricket$win_game==cricket$win_toss)
+cricket$tossgame <- 1*(as.character(cricket$win_game)==as.character(cricket$win_toss))
 # The game is drawn
 cricket$tossgame[cricket$draw==1] <- .5
 
@@ -162,3 +165,5 @@ Is there over time learning? If so, toss adv. would increase.
 " 
 
 
+ddply(cricket,~type_of_match + day_n_night,summarise,mean=mean(tossgame))
+with(cricket, glm(tossgame ~ diff_ranks))

### 2015-09-01 scripts/04_merge_ranking_data.R
Rank graph + update on merge

@@ -69,14 +69,20 @@ test_ranks$unique <- paste0("TEST", test_ranks$country, test_ranks$month_abb, te
 
 # Bring out data 
 match$team1_rank <- match$team2_rank <- NA
-match$team1_rank <- odi_ranks$rating[match(match$team1_id, odi_ranks$unique)]
-match$team1_rank <- test_ranks$rating[match(match$team1_id, test_ranks$unique)]
-match$team2_rank <- odi_ranks$rating[match(match$team2_id, odi_ranks$unique)]
-match$team2_rank <- test_ranks$rating[match(match$team2_id, test_ranks$unique)]
+odi  <- odi_ranks$rating[match(match$team1_id, odi_ranks$unique)]
+test <- test_ranks$rating[match(match$team1_id, test_ranks$unique)]
+match$team1_rank <- ifelse(is.na(odi), test, odi)
+odi  <- odi_ranks$rating[match(match$team2_id, odi_ranks$unique)]
+test <- test_ranks$rating[match(match$team2_id, test_ranks$unique)]
+match$team2_rank <- ifelse(is.na(odi), test, odi)
 
 # Adhoc data integrity check 
-match$team1_rank[match$type_of_match=='ODI' & month=='Apr']
+match$team1_rank[match$type_of_match=='ODI' & match$month=='Apr']
+table(as.character(match$team1[!is.na(match$team1_rank)]))
+table(as.character(match$team2[!is.na(match$team2_rank)]))
 
 # Diff in ranking
 match$diff_ranks <- abs(match$team1_rank - match$team2_rank)
 
+# Signed diff
+match$signed_diff_ranks <- ifelse(as.character(match$team1)==as.character(match$win_toss), match$diff_ranks, -1*match$diff_ranks)

### 2015-09-01 scripts/05_cricket.R

@@ -51,6 +51,8 @@ cricket$tossgame <- 1*(as.character(cricket$win_game)==as.character(cricket$win_
 # The game is drawn
 cricket$tossgame[cricket$draw==1] <- .5
 
+# Coding ranking diff. sign based on team that wins the toss
+
 # Results
 ddply(cricket,~type_of_match + day_n_night,summarise,mean=mean(tossgame))
 ddply(cricket,~type_of_match + duckworth_lewis,summarise,mean=mean(tossgame))
@@ -160,10 +162,47 @@ theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted")
 annotate("text", x = 4.2, y = .15, label = "zero", size=3) 
 ggsave("figs/winbyDL.pdf", width=5)
 
+
+"
+Win by Diff. in ranks
+
+Probab. of team that wins the toss winning conditional on signed ranking diff. w/ competing team
+
+"
+
+# glm
+with(cricket, glm(tossgame==1 ~ signed_diff_ranks))
+
+rankcricket <- subset(cricket, type_of_match %in% c("ODI", "TEST"))
+win_rank <- ddply(rankcricket,~signed_diff_ranks,summarise, diff=mean(I(tossgame==1) - I(tossgame==0)), count=length(unique(url)))
+win_rank$diff <- win_rank$diff*100
+
+ggplot(win_rank, aes(x=signed_diff_ranks, y=diff)) + 
+geom_smooth(method="auto") +
+theme_minimal() + 
+scale_x_continuous(breaks=seq(-50, 50, 10), labels=nolead0s(seq(-50, 50, 10)), limits=c(-50, 50)) +
+scale_y_continuous(breaks=seq(-35, 15, 5), labels=nolead0s(seq(-35, 15, 5)), limits=c(-35, 15), name="") +
+theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted"),
+      panel.grid.minor.x = element_blank(),
+      panel.grid.major.x = element_line(colour = "#f7f7f7", linetype = "solid"),
+      panel.border       = element_blank(),
+      legend.position  	 = "bottom",
+      legend.text        = element_text(size=10),
+      legend.background  = element_rect(color="#ffffff"),
+      legend.key         = element_rect(color="#ffffff", fill="#ffffff"),
+      legend.key.size    = unit(.1,"cm"),
+      legend.margin      = unit(.2,"cm"),
[…]

### 2015-09-01 scripts/05_cricket.R
Added a line for rankdiff 0

@@ -180,7 +180,8 @@ win_rank$diff <- win_rank$diff*100
 ggplot(win_rank, aes(x=signed_diff_ranks, y=diff)) + 
 geom_smooth(method="auto") +
 theme_minimal() + 
-scale_x_continuous(breaks=seq(-50, 50, 10), labels=nolead0s(seq(-50, 50, 10)), limits=c(-50, 50)) +
+geom_vline(xintercept=0, col="red", size=.1) +
+scale_x_continuous(breaks=seq(-50, 50, 10), labels=nolead0s(seq(-50, 50, 10)), limits=c(-50, 50), name="Ranking Advantage of Team That Won the Toss") +
 scale_y_continuous(breaks=seq(-35, 15, 5), labels=nolead0s(seq(-35, 15, 5)), limits=c(-35, 15), name="") +
 theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted"),
       panel.grid.minor.x = element_blank(),

### 2015-09-01 scripts/05_cricket.R
Different smoother, smaller domain

@@ -178,11 +178,12 @@ win_rank <- ddply(rankcricket,~signed_diff_ranks,summarise, diff=mean(I(tossgame
 win_rank$diff <- win_rank$diff*100
 
 ggplot(win_rank, aes(x=signed_diff_ranks, y=diff)) + 
-geom_smooth(method="auto") +
+geom_hline(yintercept=0, col="#aa0000", linetype="dashed", alpha=.3, size=.1) +
+geom_vline(yintercept=0, col="#aa0000", linetype="dashed", alpha=.5, size=.1) +
+geom_smooth(level = 0.80) +
 theme_minimal() + 
-geom_vline(xintercept=0, col="red", size=.1) +
-scale_x_continuous(breaks=seq(-50, 50, 10), labels=nolead0s(seq(-50, 50, 10)), limits=c(-50, 50), name="Ranking Advantage of Team That Won the Toss") +
-scale_y_continuous(breaks=seq(-35, 15, 5), labels=nolead0s(seq(-35, 15, 5)), limits=c(-35, 15), name="") +
+scale_x_continuous(breaks=seq(-40, 40, 10), labels=nolead0s(seq(-40, 40, 10)), limits=c(-30, 30), name="Ranking Advantage of Team That Won the Toss") +
+scale_y_continuous(breaks=seq(-20, 15, 5), labels=nolead0s(seq(-20, 15, 5)), limits=c(-20, 15), name="") +
 theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted"),
       panel.grid.minor.x = element_blank(),
       panel.grid.major.x = element_line(colour = "#f7f7f7", linetype = "solid"),
@@ -201,7 +202,7 @@ theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted")
 	  strip.text.x       = element_text(size=9),
 	  legend.text        = element_text(size=8),
       plot.margin        = unit(c(0,.5,.5,.5), "cm"))
-ggsave("figs/winbyRank.pdf", width=5)
+ggsave("figs/winbyRank.pdf", width=6)
 
 
 "

### 2015-09-01 write_up/cricket.tex
Added Rank to Cricket

@@
 \centering
-\caption{Proportion of Matches Won Minus Matches Lost After Winning the Toss}
+\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Type of Match}
 \includegraphics[scale=.85]{../figs/winbyType.pdf}
@@
 \centering
-\caption{Proportion of Matches Won Minus Matches Lost After Winning the Toss}
+\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss By Day/Night}
 \includegraphics[scale=.85]{../figs/winbyDayNight.pdf}
@@
 \centering
-\caption{Proportion of Matches Won Minus Matches Lost After Winning the Toss}
+\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Duckworth Lewis}
 \includegraphics[scale=.85]{../figs/winbyDL.pdf}
@@
 
+\begin{figure}[htbp]
+\centering
+\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Difference in Ranks}
+\includegraphics[scale=.85]{../figs/winbyRank.pdf}
+\label{fig:summary}
+\end{figure}
+
 \end{document}

### 2015-09-01 write_up/cricket.tex
Updated write-up

@@
 
+setwd(paste0(githubdir, "/cricket-stats/write_up/"))	
 tools::texi2dvi("cricket.tex", pdf=TRUE, clean=TRUE)	
+setwd(paste0(githubdir, "/cricket-stats"))
 
@@
 The opportunity to choose whether to bowl or bat first is decided by a coin toss.
-While this method of granting advatange is fair, on average, the system isn't fair in any one game.
+While this method of granting advatange is fair on average, the system isn't fair in any one game.
 At first glance, the imbalance seems inevitable.
@@
 Data are from 43,185 first-class men's cricket matches.\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{http://github.com/soodoku/get-cricket-data/}{http://github.com/soodoku/get-cricket-data/}.} It is a near census of the relevant population.
-We have data on all types of matches: 20-20, one-day, and five-day.\footnote{In English county cricket, first-class matches last four days.
-Some first class matches last two days.
+We have data on all types of matches: domestic and international Twenty20s -- T20 and T20I respectively, domestic and international one-dayers -- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches -- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches.
+In English county cricket, first-class matches last four days.
+Some first class matches last just a day.
+Others two days.
 Yet others three days.
-Some last just a day.
-And till a particular point, a test match lasted as long as it was needed to finish a game.
-We elide over these differences in our analyses.} Of these matches, we do not have data on the toss for 2,807, or roughly 7\%, of the matches.
+And till a particular point in history, a test match lasted as long as it was needed to finish a game.
+We elide over such differences.}
+
+Of these matches, we do not have data on the toss for 2,807, or roughly 7\%, of the matches.
 The primary reason we don't have data on these matches is because the match was abandoned without play.
@@
 In 706 matches, or roughly 1.7\% of the remaining matches, not enough overs were bowled to get a result.
-We exclude these matches from our analysis.
+We again exclude these matches from our analysis.
 This leaves us with 39,672 matches.
@@
 For now, we focus on winning and losing.
-We elide over the small issue of differential number of draws across formats by estimating the difference in probability of winning and losing.
+We elide over differential number of draws across formats, estimating the difference in probability of winning and losing.
 
-The team that wins the toss wins the match 2.3\% more often than lose it.
+A caveat about interpretation before we discuss the results.
+As we allude to at the start, it is hard, if not impossible, to distinguish between lack of advantage of winning a toss and not capitalizing on the potential advantage.
+
+The team that wins the toss wins the match 2.3\% more often than it loses it.
 This is a reasonable sized advantage -- though likely much smaller than the number that most commentators carry in their heads -- in a competitive sport.
-This advantage, however, is highly variable by format, by conditions, and by whether or not a particular formula was used to adjust scores when it rained.
-Some of the variability is expected (but as we will see, expectations are often dashed).
-It is often claimed that the toss is more crucial in day and night matches, due to dew and lower visibility of the white ball under lights.
-And it is often claimed that the toss is more important in tests than one-day matches.
-To the results then.
+This advantage, however, is highly variable by format, by conditions, by whether or not a particular formula was used to adjust scores when it rained, and how much better the team that won the toss is vis-\`{a}-vis the competing team.
+Some of the variability is expected, but as we will see, expectations are often dashed.
+
+The conventional wisdom among the lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s respectively.
+And there is good dose of common sense behind the conventional wisdom.
+Pitches invariably deteriorate over multiple days and batting last in a test match is often challenging.
+The pitch deteriorates far less over the course of the day, or in case of Twenty20s -- a few hours.
+And indeed unlimited over matches provide the most consistent advantage -- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{type}).
+But the heftiest advtange is in domestic one-day matches (List A), approximately 4\%.
+The number for ODIs, however, is a bit puzzling -- there is a slight disadvantage to winning the toss.
+That can best be ascribed to choosing badly.
 
@@
 \includegraphics[scale=.85]{../figs/winbyType.pdf}
-\label{fig:summary}
+\label{fig:type}
 \end{figure}
+
+But type of matches cover only major source of variation and theorizing about the advantage granted by the toss.
+It is often claimed that the toss is more crucial in day and night matches.
+Due to dew -- it is thought to make bowling hard, and lower visibility of the white ball under lights -- it allegedly makes catching hard, it is thought that the team that fields second is disadvantaged.
+It turns out that the conventional wisdom is largely vindicated, except in the case of Twenty20 Internationals (see Figure ~\ref{dn}).
+In each, domestic one-day and Twenty20 matches, and one-day internationals, the advantage of winning the toss is at least 3\%, and in case of one-day internationals, nearly 8\%.
 
@@
 \includegraphics[scale=.85]{../figs/winbyDayNight.pdf}
-\label{fig:summary}
+\label{fig:dn}
 \end{figure}
@@
 \includegraphics[scale=.85]{../figs/winbyDL.pdf}
-\label{fig:summary}
+\label{fig:dl}
 \end{figure}
@@
 \includegraphics[scale=.85]{../figs/winbyRank.pdf}
-\label{fig:summary}
+\label{fig:ranks}
 \end{figure}

### 2015-09-01 write_up/cricket.tex
More writing

@@
 The pitch deteriorates far less over the course of the day, or in case of Twenty20s -- a few hours.
-And indeed unlimited over matches provide the most consistent advantage -- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{type}).
+And indeed unlimited over matches provide the most consistent advantage -- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{fig:type}).
 But the heftiest advtange is in domestic one-day matches (List A), approximately 4\%.
@@
 Due to dew -- it is thought to make bowling hard, and lower visibility of the white ball under lights -- it allegedly makes catching hard, it is thought that the team that fields second is disadvantaged.
-It turns out that the conventional wisdom is largely vindicated, except in the case of Twenty20 Internationals (see Figure ~\ref{dn}).
+It turns out that the conventional wisdom is largely vindicated, except in the case of Twenty20 Internationals (see Figure ~\ref{fig:dn}).
 In each, domestic one-day and Twenty20 matches, and one-day internationals, the advantage of winning the toss is at least 3\%, and in case of one-day internationals, nearly 8\%.
@@
 
+In limited over matches sometimes weather intervenes significantly after the match has already begun.
+After certain loss of time, the match is generally curtailed and the scores adjusted using a method invented by Duckworth and Lewis.
+We can use the random nature of who wins the toss to see if winning percentages are strongly conditioned by matches that use Duckworth Lewis and matches that don't.
+We find scant differences except in the case of international Twenty20s (see Figure ~\ref{fig:dl}).
+
 \begin{figure}[htbp]
@@
 \end{figure}
+
+Lastly, winning a toss ought to matter the most when two closely matched teams are playing.
+What good would winning the toss do if you are matched with a team you have no chance beating.
+To study this, we collected ranking data.
+The ranking data are only available for international teams.
+And for reasons to do with convenience, we limited ourselves to collecting data on one-day internationals and tests between the earliest starting point and 2013.
+(These ranking data are updated monthly.) We measure how closely matched the teams are by simply subtracting the ranking points of team that lost the toss from the team that won it.
+The results are expected but new.
+As Figure ~\ref{fig:ranks}, there is a sharp curve around 0.
+When closely matched teams win, the winning ratio is strongly affected by who wins the toss.
 
@@
 
+This isn't a comprehensive inventory of analyses that one could do on the topic.
+But neither is it meant to be.
+There are some simple questions that still unanswered.
+For one, how does the toss advantage vary over time?
+Have teams become smarter?
+How the advantage vary by country -- are some countries better than others?
+Other more nuanced but commonly thought questions also remain.
+For one, students of the game suspect that the advantage of winning a toss in early English season is especially great.
+We propose to investigate these questions in a follow-up piece.
+
 \end{document}

### 2015-09-02 scripts/05_cricket.R
Started on readme + analysis

@@ -208,3 +208,12 @@ ggsave("figs/winbyRank.pdf", width=6)
 "
 Is there over time learning? If so, toss adv. would increase.
 " 
+
+"
+Toss Adv. by Country
+For this - we would want to do Win/Win Toss - Win/Lose Toss to adjust for team probab.
+
+"
+
+
+

### 2015-09-03 scripts/02_parser.py
Removed duplication of scripts, edited readme

@@ -1,142 +1,0 @@
-'''
-
-Parse Downloaded Cricket Data
-
-'''
-
-import os
-
-INPUT_FOLDER = ["espncricinfo-t20","espncricinfo-lista","espncricinfo-fc","espncricinfo-odi","espncricinfo-t20i","espncricinfo-test"]
-FINAL_OUTPUT_FILE = "final_output.csv"
-
-HEADER = "url, team1, team2, win_toss, bat_or_bowl, outcome, win_game, date, day_n_night, ground, rain, duckworth_lewis, match_id, type_of_match"
-
-def removeEndLineCharacter(mystr):
-    if mystr.find("\n"):
-        return mystr[:-1]
-    
-def removeDoubleQuote(myStr):
-    return myStr.replace("\"", "").strip();
-
-def removeDivInText(myStr, divs):
-    for div in divs:
-        myStr = myStr.replace(div,"")
-    return myStr
-
-def refineHTMLToText(myStr):
-    for div in USELESS_TAGS:
-        myStr = myStr.replace(div,"")
-    myStr = myStr.replace("  "," ").strip()
-    return myStr
-
-
-def findWithPattern(mystr, startPattern, endPattern):
-    """
-    Find the string that starts with <startPattern> and ends with <endPattern> in the orginal string <mystr>.
-    Args:
-        + mystr: orginal string.
-        + startPattern: 
-        + endPattern: 
-    Returns:
-        + The found string,
-        + and the remained part of the orginal string.
-    """
-    x = mystr.find(startPattern)
-    if x==-1:
-        return "",mystr
-    mystr = mystr[x + len(startPattern):]
-    y = mystr.find(endPattern)
-    if y==-1:
-        return "",mystr
-    return mystr[:y], mystr[y+len(endPattern):]
-
-def extractDataFrom(data):
-    # Extract 2 teams name
-    team1, tmp = findWithPattern(data,'class="teamLink">','</a>')
-    team2, tmp = findWithPattern(tmp,'class="teamLink">','</a>')
-
-    # Extract win_toss
-    toss_div, tmp = findWithPattern(data,"Toss  - <span class='normal'>","</span>")
-    if toss_div.find(team1)>-1:
-        win_toss = team1
-    elif toss_div.find(team2)>-1:
-        win_toss = team2
-    else:
-        win_toss = "" #!!!
-
-    # Extract bat_or_bowl
-    if toss_div.find("bat")>-1:
-        bat_or_bowl = "bat"
-    elif toss_div.find("field")>-1:
[…]

### 2015-09-03 scripts/04_cricket.R
Modest amendments. Results same.

@@ -8,13 +8,22 @@
 setwd(paste0(githubdir, "/cricket-stats"))
 
 # Source merge script
-source("scripts/04_merge_ranking_data.R")
+source("scripts/03_merge_ranking_data.R")
 
 # Load libs
 library(plyr)
 
 # Read in data/ now sourcing it - see 04_merge_ranking_data
 # cricket <- read.csv("data/final_output.csv")
+
+"
+Fix Data
+# cricket$outcome[cricket$win_game!='']
+"
+
+temp  <- sapply(strsplit(match$outcome, "won", fixed=T), "[", 1) #get the stuff before the word won
+temp2 <- sub("\\s+$", "", temp) # strip trailing space
+match$win_game <- ifelse(match$win_game=="", temp2, match$win_game)
 
 "
 Take out matches with no toss (~ no match)
@@ -28,7 +37,7 @@ Take out matches where there was no result
 Leaves us w/ 39672 rows
 "
 
-cricket <- subset(cricket, outcome!="No result")
+cricket <- subset(cricket, outcome!="No result" & win_game!="")
 
 "
 Drawn Matches
@@ -45,13 +54,36 @@ Imp. esp. for first class games
 
 "
 
+# Win toss then you lose
+cricket$team1_win_toss <- 1*(cricket$team1==cricket$win_toss)
+cricket$team2_win_toss <- 1*(cricket$team2==cricket$win_toss)
+cricket$team1_win_game <- 1*(cricket$team1==cricket$win_game)
+cricket$team2_win_game <- 1*(cricket$team2==cricket$win_game)
+
 # Recode
 # Coding team that wins the toss wins the game
-cricket$tossgame <- 1*(as.character(cricket$win_game)==as.character(cricket$win_toss))
+cricket$tossgame <- 1*(cricket$win_game==cricket$win_toss)
 # The game is drawn
 cricket$tossgame[cricket$draw==1] <- .5
 
+" 
+Data Integrity Check
+
+table(cricket$win_game[!(cricket$team1_win_game | cricket$team2_win_game)])
+
+length(unique(c(cricket$team1, cricket$team2)))
+length(unique(cricket$win_toss))
+
+"
+a <- unique(c(levels(cricket$team1), levels(cricket$team2)))[!(unique(c(levels(cricket$team1), levels(cricket$team2))) %in% levels(cricket$win_toss))]
+cricket$win_toss[cricket$team1 %in% a[1]]
+
+cricket$team1_win_game[cricket$team1_win_toss==1] 
[…]

### 2015-09-04 scripts/04_cricket.R
data integrity check - one resolved issue

@@ -1,5 +1,5 @@
 "
-@title: Fairly Random (or Choosing Badly): Impact of Winning the Toss on the Probability of Winning
+@title:   Fairly Random: Impact of Winning the Toss on the Probability of Winning
 @authors: Gaurav Sood and Derek Willis
 
 "
@@ -17,15 +17,10 @@ library(plyr)
 # cricket <- read.csv("data/final_output.csv")
 
 "
-Fix Data
-# cricket$outcome[cricket$win_game!='']
-"
+Data integrity
+match$outcome[match$win_game==''] # looks good
+table(match$outcome[match$win_toss=='' & match$outcome!=''])
 
-temp  <- sapply(strsplit(match$outcome, "won", fixed=T), "[", 1) #get the stuff before the word won
-temp2 <- sub("\\s+$", "", temp) # strip trailing space
-match$win_game <- ifelse(match$win_game=="", temp2, match$win_game)
-
-"
 Take out matches with no toss (~ no match)
 About 3k matches. Don't like the 7% number. Touch too high, imho. But checked data - turns out to be ok.
 
@@ -71,19 +66,12 @@ Data Integrity Check
 
 table(cricket$win_game[!(cricket$team1_win_game | cricket$team2_win_game)])
 
+# 137 Teams in db that haven't won a toss
 length(unique(c(cricket$team1, cricket$team2)))
 length(unique(cricket$win_toss))
+a <- unique(c(cricket$team1, cricket$team2))[!(unique(c(cricket$team1, cricket$team2)) %in% unique(cricket$win_toss))]
 
 "
-a <- unique(c(levels(cricket$team1), levels(cricket$team2)))[!(unique(c(levels(cricket$team1), levels(cricket$team2))) %in% levels(cricket$win_toss))]
-cricket$win_toss[cricket$team1 %in% a[1]]
-
-cricket$team1_win_game[cricket$team1_win_toss==1] 
-cricket$team2_win_game[cricket$team2_win_toss==1] 
-
-# Coding ranking diff. sign based on team that wins the toss
-team1_wins_toss, team1_wins_game
-
 
 # Results
 ddply(cricket,~type_of_match + day_n_night,summarise,mean=mean(tossgame))

### 2015-09-04 scripts/01_download_data.py
Returning download data to its due place

@@ -1,0 +1,43 @@
+'''
+Download Cricket Data
+Last Edited: 05.27.15
+
+@author: Gaurav Sood
+
+'''
+
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+import requests
+import math
+import csv
+import sys
+import time
+import os
+import simplejson
+import unicodedata
+from BeautifulSoup import BeautifulSoup, SoupStrainer
+
+for match_type in ['odi', 'test', 't20i', 't20', 'list%20a', 'first%20class']:
+    results = []
+    r = requests.get('http://search.espncricinfo.com/ci/content/match/search.html?all=1;page=0;search=' + match_type)
+    soup = BeautifulSoup(r.text)
+    last_match = int(soup.findAll('span', attrs={'class':'PaginationNmbrs'})[-1].text)
+    last_page = int(math.ceil(float(last_match)/float(20)))
+    for i in range(0, last_page):
+        time.sleep(2)
+        results_page = requests.get("http://search.espncricinfo.com/ci/content/match/search.html?search={0};all=1;page={1}".format(match_type, i))
+        soupy = BeautifulSoup(results_page.text)
+        for new_host in soupy.findAll('a', {'class' : 'srchPlyrNmTxt'}):
+            try:
+                new_host = new_host['href']
+            except:
+                continue
+            new_host = unicodedata.normalize('NFKD', new_host).encode('ascii','ignore')
+            #print(type(str.split(new_host)[3]))
+            print str.split(new_host, "/")[4].split('.')[0]
+            results.append(str.split(new_host, "/")[4].split('.')[0])
+
+    with open("matches-{0}.json".format(match_type), "wb") as f:
+        simplejson.dump(results, f)

### 2015-09-04 write_up/cricket.tex
Removed duplication of scripts, edited readme
Adding bib

@@
 The advantage is pointed to by commentators, by team captains in the pre-toss interview, and by the captain of the losing team in the post-match interview.
-This isn't to say that there is actually an advantage to winning the toss.
+This isn't to say that there is actually an advantage to winning the toss, or that in lies in what is claimed, or that teams are able to successfully exploit it.
 For it may be impossible to predict well in advance the advantage of bowling or batting first.
@@
 
+\citep{de1998winning}, \citep{dawson2009bat}, \url{http://www.slate.com/articles/sports/sports_nut/2010/11/flipping_awesome.html}
+
+\section*{Data}
 Data are from 43,185 first-class men's cricket matches.\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{http://github.com/soodoku/get-cricket-data/}{http://github.com/soodoku/get-cricket-data/}.} It is a near census of the relevant population.
@@
 
+\section*{Results}
 A caveat about interpretation before we discuss the results.
@@
 
+\clearpage
+\bibliographystyle{apsr}
+\bibliography{luckybib}
+
 \end{document}

### 2015-09-06 write_up/cricket.tex
Adding bib., url + natbib
Minor fix

@@
 
+\usepackage{natbib}
+\usepackage{url}
 \begin{comment}
@@
 \begin{document}
-\title{\vspace{-1.5cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}}
+\title{\vspace{-1cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}}
 \author{Gaurav Sood\\\small{\href{mailto:gsood07@gmail.com}{\tt{gsood07@gmail.com}}} \and Derek Willis\\\small{\href{mailto:dwillis@gmail.com}{\tt{dwillis@gmail.com}}}}
@@
 
-\citep{de1998winning}, \citep{dawson2009bat}, \url{http://www.slate.com/articles/sports/sports_nut/2010/11/flipping_awesome.html}
+\citep{de1998winning}, \citep{dawson2009bat}, \citep{Weintraub2010}, \citep{Saad2015}
 

### 2015-09-19 write_up/cricket.tex
fixing intro

@@
 tools::texi2dvi("cricket.tex", pdf=TRUE, clean=TRUE)	
-setwd(paste0(githubdir, "/cricket-stats"))
+setwd(paste0(githubdir, "/cricket-stats/"))
 
@@
 In nearly all cricket matches, it is claimed that there is a clear advantage to bowling (batting) first.
-The advantage is pointed to by commentators, by team captains in the pre-toss interview, and by the captain of the losing team in the post-match interview.
-This isn't to say that there is actually an advantage to winning the toss, or that in lies in what is claimed, or that teams are able to successfully exploit it.
-For it may be impossible to predict well in advance the advantage of bowling or batting first.
-Or it may be that teams squander the potential advantage by using bad heuristics to choose what they do.
+The advantage is pointed to by commentators, by the team captains in the pre-toss interview, and by the captain of the losing team in the post-match interview.
+And to wrest this said advantage, one merely needs to pick the side of the coin that will be left facing the sky after it is tossed.
+And while this method of granting advantage is fair on average, the system isn't fair in any one game.
 
-The opportunity to choose whether to bowl or bat first is decided by a coin toss.
-While this method of granting advatange is fair on average, the system isn't fair in any one game.
 At first glance, the imbalance seems inevitable.
@@
 Or, one can come up with an estimate of the advantage and adjust scores accordingly (something akin to an adjustment issued when matches are shortened due to rain).
+
+None of this to say that there is actually an advantage to winning the toss, or, that teams are able to successfully exploit the potential advantage.
+For it may be impossible to predict well in advance the advantage of bowling or batting first.
+Or, it may be that teams squander the potential advantage by using bad heuristics to choose what they do.
 
@@
 This leaves us with 39,672 matches.
-We analyse these data.
+We analyze these data.
 
@@
 And indeed unlimited over matches provide the most consistent advantage -- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{fig:type}).
-But the heftiest advtange is in domestic one-day matches (List A), approximately 4\%.
+But the heftiest advantage is in domestic one-day matches (List A), approximately 4\%.
 The number for ODIs, however, is a bit puzzling -- there is a slight disadvantage to winning the toss.

### 2015-09-19 write_up/cricket.tex
fixed intro + minor correction for readme

@@
 
-But before we move to seriously consider these solutions, we may ask about the evidence.
-
-\citep{de1998winning}, \citep{dawson2009bat}, \citep{Weintraub2010}, \citep{Saad2015}
+To assess the net observed advantage of winning the toss, we leverage data from over 43,000 first-class men's cricket matches --- to our knowledge, largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}.
+Using the data, we estimate the impact of winning the toss on the probability of winning a match.
+Unlike some other studies on the topic, e.g.
+\citet{dawson2009bat, Saad2015}, we do not condition on post coin-toss decisions to avoid post-treatment bias \citep[see][]{acharya2015}.
+We find that winning the toss grants a small but significant advantage, but that advantage varies considerably and systematically.
 

### 2015-09-19 write_up/cricket.tex
added cite to dw etc.

@@
 \section*{Data}
-Data are from 43,185 first-class men's cricket matches.\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{http://github.com/soodoku/get-cricket-data/}{http://github.com/soodoku/get-cricket-data/}.} It is a near census of the relevant population.
-We have data on all types of matches: domestic and international Twenty20s -- T20 and T20I respectively, domestic and international one-dayers -- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches -- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches.
+Data are from 43,185 first-class men's cricket matches.\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.} It is a near census of the relevant population.
+We have data on all types of matches: domestic and international Twenty20s -- T20s and T20Is respectively, domestic and international one-dayers -- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches -- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches.
 In English county cricket, first-class matches last four days.
@@
 In limited over matches sometimes weather intervenes significantly after the match has already begun.
-After certain loss of time, the match is generally curtailed and the scores adjusted using a method invented by Duckworth and Lewis.
-We can use the random nature of who wins the toss to see if winning percentages are strongly conditioned by matches that use Duckworth Lewis and matches that don't.
+After certain loss of time, the match is generally curtailed and the scores adjusted using a method invented by Duckworth and Lewis \citep[see][]{duckworth1998}.
+We can use the random nature of who wins the toss to see if winning percentages are strongly conditioned by matches that use Duckworth-Lewis and matches that don't.
 We find scant differences except in the case of international Twenty20s (see Figure ~\ref{fig:dl}).
@@
 \centering
-\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Duckworth Lewis}
+\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Duckworth-Lewis}
 \includegraphics[scale=.85]{../figs/winbyDL.pdf}

### 2015-09-19 write_up/cricket.tex
better results intro

@@
 \section*{Results}
-A caveat about interpretation before we discuss the results.
-As we allude to at the start, it is hard, if not impossible, to distinguish between lack of advantage of winning a toss and not capitalizing on the potential advantage.
+A caveat about interpretation before we report the results.
+As we discuss in the introduction, we cannot estimate the actual advantage of winning a toss.
+We can only estimate the net observed advantage, which is the extent to which the teams capitalize on the potential advantage.
 

### 2015-12-04 write_up/cricket.tex
revised write-up 1

@@
-\documentclass[11pt]{article}
+\documentclass[12pt]{article}
 \usepackage{color}
@@
             pdfstartview={XYZ null null 1.00},
-            pdfauthor={Gaurav Sood, gsood07@gmail.com; Derek Willis, dwillis@gmail.com},
+            pdfauthor={Gaurav Sood, gsood07@gmail.com},
             citecolor=dark-red,
@@
 \usepackage{amsfonts,amssymb,amsbsy,amsmath,amsxtra}
+
+\usepackage[letterspace=1000]{microtype}
+\usepackage{libertine}
+\usepackage[T1]{fontenc}
 
@@
 \begin{document}
-\title{\vspace{-1cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}}
-\author{Gaurav Sood\\\small{\href{mailto:gsood07@gmail.com}{\tt{gsood07@gmail.com}}} \and Derek Willis\\\small{\href{mailto:dwillis@gmail.com}{\tt{dwillis@gmail.com}}}}
+\title{\vspace{-.5cm}\normalsize{Fairly Random: Impact of Winning the Toss on the Probability of Winning\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}\vspace{.5cm}}}
+\author{\normalsize{Gaurav Sood}\\\href{mailto:gsood07@gmail.com}{\small{gsood07@gmail.com}} \and \normalsize{Derek Willis}\\\href{mailto:dwillis@gmail.com}{\small{dwillis@gmail.com}}}
+\date{\vspace{.5cm}\normalsize{\today}}
 \maketitle
@@
 If that violates the nature of the game too much, one can easily create pitches that don't deteriorate appreciably over the course of a game.
-Or, one can come up with an estimate of the advantage and adjust scores accordingly (something akin to an adjustment issued when matches are shortened due to rain).
+Or, one can come up with an estimate of the advantage, and adjust scores accordingly, akin to an adjustment issued when matches are shortened due to rain.
 
-None of this to say that there is actually an advantage to winning the toss, or, that teams are able to successfully exploit the potential advantage.
+None of this to say that there is actually an advantage in winning the toss, or that teams are able to successfully exploit any such advantage.
 For it may be impossible to predict well in advance the advantage of bowling or batting first.
@@
 \section*{Data}
-Data are from 43,185 first-class men's cricket matches.\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.} It is a near census of the relevant population.
+Data are from 43,185 first-class men's cricket matches.
+It is a near census of the relevant population.
 We have data on all types of matches: domestic and international Twenty20s -- T20s and T20Is respectively, domestic and international one-dayers -- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches -- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches.

### 2015-12-04 write_up/cricket.tex
renamed to add grounds + added grounds to match

@@
 
-In cricket matches, three results are possible -- a draw, a win, or a loss.
-We look at draws -- a common outcome in matches in which overs aren't limited -- later on.
+\section*{Analyses and Results}
+
+We assume that the outcome of a toss is random.
+Conditional on the outcome of a toss being random, the effect of winning a toss can be attributed to the toss itself.
+We begin by assessing the validity of the assumption.
+In particular, we assess whether teams win more tosses in their home country than outside.
+
+In cricket matches, three results are possible: a draw, a win, and a loss.
+We look at draws, a common outcome in matches in which overs aren't limited, later on.
 For now, we focus on winning and losing.
@@
 
-\section*{Results}
 A caveat about interpretation before we report the results.
@@
 The team that wins the toss wins the match 2.3\% more often than it loses it.
-This is a reasonable sized advantage -- though likely much smaller than the number that most commentators carry in their heads -- in a competitive sport.
-This advantage, however, is highly variable by format, by conditions, by whether or not a particular formula was used to adjust scores when it rained, and how much better the team that won the toss is vis-\`{a}-vis the competing team.
+This is a reasonable sized advantage in a competitive sport --- though likely much smaller than the number that most commentators carry in their heads.
+This advantage, however, is highly variable by format, by conditions, by whether or not a particular formula was used to adjust scores when it rains, and how much better the team that won the toss is vis-\`{a}-vis the competing team.
 Some of the variability is expected, but as we will see, expectations are often dashed.

### 2015-12-04 scripts/04_merge_ranking_grounds_data.R

@@ -1,5 +1,5 @@
 "
-@title: Merge Ranking Data with Cricket Data
+@title: Merge Ranking and Grounds Data with Cricket Data
 @authors: Gaurav Sood and Derek Willis
 
 "
@@ -13,6 +13,10 @@ options(StringsAsFactors=F)
 # Load ranking data
 odi_ranks  <- read.csv("data/rankings_odi.csv")
 test_ranks <- read.csv("data/rankings_test.csv")
+
+# Ground
+grounds <- read.csv("data/grounds.csv")
+
 
 "
 Notes: ODI/Test ranks data till 2013
@@ -86,3 +90,10 @@ match$diff_ranks <- abs(match$team1_rank - match$team2_rank)
 
 # Signed diff
 match$signed_diff_ranks <- ifelse(as.character(match$team1)==as.character(match$win_toss), match$diff_ranks, -1*match$diff_ranks)
+
+# Add the grounds data
+
+# Trim leading and trailing spaces for grounds
+match$ground <- gsub("^\\s+", "", trim.trailing(match$ground))
+# Add dat
+match[, c("ground_id", "country", "continent", "latitude", "longitude")] <- grounds[match(match$ground, grounds$ground), c("ground_id", "country", "continent", "latitude", "longitude")]

### 2015-12-04 scripts/05_cricket.R
just add the merge

@@ -8,7 +8,7 @@
 setwd(paste0(githubdir, "/cricket-stats"))
 
 # Source merge script
-source("scripts/03_merge_ranking_data.R")
+source("scripts/04_merge_ranking_grounds_data.R")
 
 # Load libs
 library(plyr)

### 2015-12-05 scripts/05_cricket.R
added test for win toss at home

@@ -44,6 +44,15 @@ cricket$draw <- 1*(cricket$outcome=="Match drawn")
 # table(cricket$win_game[cricket$draw==1])
 
 "
+Country that won toss == Home Country
+International Matches Only --- as they are the easiest
+"
+cricket$country_data  <- cricket$country == cricket$team1 | cricket$country==cricket$team2
+cricket$home_toss_win <- cricket$country == cricket$win_toss
+sum(cricket$home_toss_win)/sum(cricket$country_data)
+binom.test(2433, 4749, p=.5)
+
+"
 Winning a toss causes outcome including draws.
 Imp. esp. for first class games
 

### 2015-12-06 write_up/cricket.tex
home side toss

@@
 We assume that the outcome of a toss is random.
-Conditional on the outcome of a toss being random, the effect of winning a toss can be attributed to the toss itself.
-We begin by assessing the validity of the assumption.
-In particular, we assess whether teams win more tosses in their home country than outside.
+Conditional on the outcome of a toss being random, the effect of winning the toss can be attributed to the toss itself.
+Before we exploit the design, we shed some light on the validity of the assumption.
+In particular, we assess whether the coin toss is somehow rigged, with home sides winning tosses more often.
+For this task, we limited ourselves to only international matches as establishing which of the teams is the home team in local matches is somewhat arduous.
+Of the 4,749 international matches, the home team won the toss in 2,433 matches, or about 51.2\%.
+The probability that we will get as large a proportion by chance after tossing close to 5,000 coins is about 9\%, low but not eyebrow-raisingly so.
 

### 2015-12-06 scripts/05_cricket.R

@@ -12,6 +12,7 @@ source("scripts/04_merge_ranking_grounds_data.R")
 
 # Load libs
 library(plyr)
+library(stringi)
 
 # Read in data/ now sourcing it - see 04_merge_ranking_data
 # cricket <- read.csv("data/final_output.csv")
@@ -35,6 +36,11 @@ Leaves us w/ 39672 rows
 cricket <- subset(cricket, outcome!="No result" & win_game!="")
 
 "
+Team 2 missing in one case
+"
+cricket <- subset(cricket, team2!="")
+
+"
 Drawn Matches
 6674 of them ~ 17%
 "
@@ -47,10 +53,21 @@ cricket$draw <- 1*(cricket$outcome=="Match drawn")
 Country that won toss == Home Country
 International Matches Only --- as they are the easiest
 "
+
 cricket$country_data  <- cricket$country == cricket$team1 | cricket$country==cricket$team2
 cricket$home_toss_win <- cricket$country == cricket$win_toss
 sum(cricket$home_toss_win)/sum(cricket$country_data)
 binom.test(2433, 4749, p=.5)
+
+"
+# Less precise
+# Denominator ...?
+cricket$home_toss_win_t <- stri_detect_fixed(cricket$ground, cricket$win_toss, case_insensitive=TRUE) | 
+                           stri_detect_fixed(cricket$team1, cricket$country, case_insensitive=TRUE) | 
+                           stri_detect_fixed(cricket$team2, cricket$country, case_insensitive=TRUE) |
+                           stri_detect_fixed(cricket$country, cricket$team1, case_insensitive=TRUE) | 
+                           stri_detect_fixed(cricket$country, cricket$team2, case_insensitive=TRUE)
+"
 
 "
 Winning a toss causes outcome including draws.

### 2015-12-06 write_up/cricket.tex
nip/tuck

@@
 Before we exploit the design, we shed some light on the validity of the assumption.
-In particular, we assess whether the coin toss is somehow rigged, with home sides winning tosses more often.
-For this task, we limited ourselves to only international matches as establishing which of the teams is the home team in local matches is somewhat arduous.
+In particular, we assess whether the coin toss is somehow rigged, with the home side enjoying the rub of the green more often.
+For this analysis, we only get to exploit international matches as establishing which of the teams is the home team in local matches is somewhat arduous.
 Of the 4,749 international matches, the home team won the toss in 2,433 matches, or about 51.2\%.
 The probability that we will get as large a proportion by chance after tossing close to 5,000 coins is about 9\%, low but not eyebrow-raisingly so.
+Thus, for now, we choose to trust that the tosses are random.
 
 In cricket matches, three results are possible: a draw, a win, and a loss.
+For now, we focus on winning and losing.
 We look at draws, a common outcome in matches in which overs aren't limited, later on.
-For now, we focus on winning and losing.
 We elide over differential number of draws across formats, estimating the difference in probability of winning and losing.

### 2015-12-07 write_up/cricket.tex
improved intro.

@@
 
-In nearly all cricket matches, it is claimed that there is a clear advantage to bowling (batting) first.
+In nearly all cricket matches, it is claimed that there is a clear advantage to either bowling or batting first.
 The advantage is pointed to by commentators, by the team captains in the pre-toss interview, and by the captain of the losing team in the post-match interview.
-And to wrest this said advantage, one merely needs to pick the side of the coin that will be left facing the sky after it is tossed.
+And to wrest the said advantage, one merely needs to pick the side of the coin left facing the sky after the toss.
 And while this method of granting advantage is fair on average, the system isn't fair in any one game.
@@
 After all, someone has to bat first.
-One can, however, devise a baseball like system where short innings are interspersed.
+One can, however, devise a baseball like system in which short innings are interspersed.
 If that violates the nature of the game too much, one can easily create pitches that don't deteriorate appreciably over the course of a game.
-Or, one can come up with an estimate of the advantage, and adjust scores accordingly, akin to an adjustment issued when matches are shortened due to rain.
+Or, one can come up with an estimate of the advantage, and adjust the scores accordingly, akin to an adjustment issued when matches are shortened due to rain.
 
@@
 
-To assess the net observed advantage of winning the toss, we leverage data from over 43,000 first-class men's cricket matches --- to our knowledge, largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}.
-Using the data, we estimate the impact of winning the toss on the probability of winning a match.
-Unlike some other studies on the topic, e.g.
-\citet{dawson2009bat, Saad2015}, we do not condition on post coin-toss decisions to avoid post-treatment bias \citep[see][]{acharya2015}.
+To assess the net observed advantage of winning the toss, we exploit data from a novel dataset of over 43,000 first-class men's cricket matches --- to our knowledge, the largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}.
+And in analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into.
+To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat, Saad2015} etc., we do not condition on post coin-toss decisions.
 We find that winning the toss grants a small but significant advantage, but that advantage varies considerably and systematically.
@@
 It is a near census of the relevant population.
-We have data on all types of matches: domestic and international Twenty20s -- T20s and T20Is respectively, domestic and international one-dayers -- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches -- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches.
+We have data on all types of matches: domestic and international Twenty20s --- T20s and T20Is respectively, domestic and international one-dayers --- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches --- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches.
 In English county cricket, first-class matches last four days.

### 2015-12-08 write_up/cricket.tex
better org. of data

@@
 
-In cricket matches, three results are possible: a draw, a win, and a loss.
-For now, we focus on winning and losing.
-We look at draws, a common outcome in matches in which overs aren't limited, later on.
-We elide over differential number of draws across formats, estimating the difference in probability of winning and losing.
-
 A caveat about interpretation before we report the results.

### 2015-12-08 scripts/04_merge_ranking_grounds_data.R

@@ -97,3 +97,6 @@ match$signed_diff_ranks <- ifelse(as.character(match$team1)==as.character(match$
 match$ground <- gsub("^\\s+", "", trim.trailing(match$ground))
 # Add dat
 match[, c("ground_id", "country", "continent", "latitude", "longitude")] <- grounds[match(match$ground, grounds$ground), c("ground_id", "country", "continent", "latitude", "longitude")]
+
+# Let us add a unique ID
+match$uniqueid <- 1:nrow(match)

### 2015-12-08 scripts/05_cricket.R

@@ -29,37 +29,49 @@ About 3k matches. Don't like the 7% number. Touch too high, imho. But checked da
 cricket <- subset(match, win_toss!="")
 
 "
-Take out matches where there was no result
-Leaves us w/ 39672 rows
-"
-
-cricket <- subset(cricket, outcome!="No result" & win_game!="")
-
-"
 Team 2 missing in one case
 "
 cricket <- subset(cricket, team2!="")
 
+
+"
+Take out matches where there was no result
+Leaves us w/ 39672 rows
+"
+
+cricket <- subset(cricket, outcome!="No result")
+
+"
+Take out matches where no decision on who is to bowl/bat first is made
+"
+
+cricket <- subset(cricket, bat_or_bowl!="")
+
+
+"
+More recoding
+-----------------
+"
+
 "
 Drawn Matches
-6674 of them ~ 17%
+6673 of them ~ 17%
 "
 
 cricket$draw <- 1*(cricket$outcome=="Match drawn")
 
 # table(cricket$win_game[cricket$draw==1])
 
+# Win toss, win game
+cricket$team1_win_toss <- 1*(cricket$team1==cricket$win_toss)
+cricket$team2_win_toss <- 1*(cricket$team2==cricket$win_toss)
+cricket$team1_win_game <- 1*(cricket$team1==cricket$win_game)
+cricket$team2_win_game <- 1*(cricket$team2==cricket$win_game)
+
 "
 Country that won toss == Home Country
 International Matches Only --- as they are the easiest
-"
 
-cricket$country_data  <- cricket$country == cricket$team1 | cricket$country==cricket$team2
-cricket$home_toss_win <- cricket$country == cricket$win_toss
-sum(cricket$home_toss_win)/sum(cricket$country_data)
-binom.test(2433, 4749, p=.5)
-
-"
 # Less precise
 # Denominator ...?
 cricket$home_toss_win_t <- stri_detect_fixed(cricket$ground, cricket$win_toss, case_insensitive=TRUE) | 
@@ -67,37 +79,75 @@ cricket$home_toss_win_t <- stri_detect_fixed(cricket$ground, cricket$win_toss, c
                            stri_detect_fixed(cricket$team2, cricket$country, case_insensitive=TRUE) |
                            stri_detect_fixed(cricket$country, cricket$team1, case_insensitive=TRUE) | 
[…]

### 2015-12-08 scripts/05_cricket.R
getting close

@@ -13,6 +13,7 @@ source("scripts/04_merge_ranking_grounds_data.R")
 # Load libs
 library(plyr)
 library(stringi)
+library(xtable)
 
 # Read in data/ now sourcing it - see 04_merge_ranking_data
 # cricket <- read.csv("data/final_output.csv")
@@ -109,7 +110,7 @@ crickett$game_win  <- c(cricket$team1_win_game, cricket$team2_win_game)
 # The game is drawn
 crickett$game_win[crickett$draw==1] <- .5
 
-crickett$bat_bowl  <- ifelse(crickett$toss_win, crickett$bat_or_bowl, ifelse(crickett$bat_or_bowl=="bowl", "bat", "bowl"))
+crickett$bat_bowl     <- ifelse(crickett$toss_win, crickett$bat_or_bowl, ifelse(crickett$bat_or_bowl=="bowl", "bat", "bowl"))
 crickett$home_country <- crickett$country == crickett$team
 crickett$countries    <- crickett$country == crickett$team1 | crickett$country == crickett$team2
 
@@ -121,6 +122,8 @@ Analysis
 "
 
 homet <- with(crickett[crickett$countries==1,], xtabs( ~ home_country + toss_win))
+
+ddply(crickett, ~type_of_match + day_n_night, summarise, mean=mean(toss_win))
 
 
 "
@@ -135,24 +138,18 @@ res  <- wint/colSums(wint)
 res
 res[3,2] - res[3,1]
 
-
-
 cricket$home_toss_win <- cricket$country == cricket$win_toss
 sum(cricket$home_toss_win)/sum(cricket$country_data)
 binom.test(2433, 4749, p=.5)
 
+# By Type of Match
+ddply(crickett, ~type_of_match + toss_win, summarise, mean = mean(game_win), n = length(unique(uniqueid)))
+ddply(crickett, ~day_n_night + toss_win, summarise, mean = mean(game_win), n = length(unique(uniqueid)))
+ddply(crickett, ~duckworth_lewis + toss_win, summarise, mean = mean(game_win), n = length(unique(uniqueid)))
+ddply(crickett, ~type_of_match + day_n_night + toss_win, summarise, mean = mean(game_win), n = length(unique(uniqueid)))
+ddply(crickett, ~type_of_match + duckworth_lewis + toss_win, summarise, mean = mean(game_win), n = length(unique(uniqueid)))
 
-
[…]

### 2015-12-13 scripts/05_cricket.R
better script. 90%

@@ -97,22 +97,21 @@ Two rows per match
 match_cols <- c("url", "date", "day_n_night", "ground", "rain", "duckworth_lewis", "match_id", "type_of_match", "month", "year", "diff_ranks", 
       "signed_diff_ranks", "ground_id", "country", "continent", "latitude", "longitude", "uniqueid", "draw")
 
-crickett <- data.frame(matchid=c(cricket$uniqueid, cricket$uniqueid))
-crickett <- merge(cricket, crickett, by.y="matchid", by.x="uniqueid")
+team_cols <- c("team1", "team2", "team1_id", "team2_id", "team2_rank", "team1_rank", "team1_win_toss", "team2_win_toss", "team1_win_game", "team2_win_game")
+rename_cols <- c("team1.name", "team2.name", "team1.id", "team2.id", "team2.rank", "team1.rank", "team1.wintoss", "team2.wintoss", "team1.wingame", "team2.wingame")
+names(cricket)[names(cricket) %in% team_cols] <- rename_cols
 
-# team dat (suboptimal but ok)
-crickett$team      <- c(cricket$team1, cricket$team2)
-crickett$team_id   <- c(cricket$team1_id, cricket$team2_id)
-crickett$team_rank <- c(cricket$team1_rank, cricket$team2_rank)
-crickett$toss_win  <- c(cricket$team1_win_toss, cricket$team2_win_toss)
-crickett$game_win  <- c(cricket$team1_win_game, cricket$team2_win_game)
+crickett <- cricket %>% gather(key, value, starts_with('team')) %>% separate(key, c("var", "col")) %>% arrange(url) %>% spread(col, value)
 
 # The game is drawn
-crickett$game_win[crickett$draw==1] <- .5
+crickett$wingame[crickett$draw==1] <- .5
 
-crickett$bat_bowl     <- ifelse(crickett$toss_win, crickett$bat_or_bowl, ifelse(crickett$bat_or_bowl=="bowl", "bat", "bowl"))
-crickett$home_country <- crickett$country == crickett$team
-crickett$countries    <- crickett$country == crickett$team1 | crickett$country == crickett$team2
+crickett$bat_bowl     <- ifelse(crickett$wintoss, crickett$bat_or_bowl, ifelse(crickett$bat_or_bowl=="bowl", "bat", "bowl"))
+crickett$home_country <- crickett$country == crickett$name
[…]

### 2015-12-13 write_up/cricket.tex
better script. 90%
crick update

@@
 For this analysis, we only get to exploit international matches as establishing which of the teams is the home team in local matches is somewhat arduous.
-Of the 4,749 international matches, the home team won the toss in 2,433 matches, or about 51.2\%.
-The probability that we will get as large a proportion by chance after tossing close to 5,000 coins is about 9\%, low but not eyebrow-raisingly so.
+Of the 5,591 international matches, the home team won the toss in 2,860 matches, or about 51.15\%.
+The probability that we will get as large a proportion by chance after tossing 5,591 coins is about 8.7\%, low but not eyebrow-raisingly so.
 Thus, for now, we choose to trust that the tosses are random.
@@
 We can only estimate the net observed advantage, which is the extent to which the teams capitalize on the potential advantage.
+With that, the results.
 
-The team that wins the toss wins the match 2.3\% more often than it loses it.
+The team that wins the toss wins the match 2.9\% more often than the team that loses the toss.
 This is a reasonable sized advantage in a competitive sport --- though likely much smaller than the number that most commentators carry in their heads.
-This advantage, however, is highly variable by format, by conditions, by whether or not a particular formula was used to adjust scores when it rains, and how much better the team that won the toss is vis-\`{a}-vis the competing team.
-Some of the variability is expected, but as we will see, expectations are often dashed.
+This advantage, however, varies by format, by conditions, by whether or not a particular formula was used to adjust scores when it rains, and how much better the team that won the toss is vis-\`{a}-vis the competing team.
+Some of the variability is expected, but as we will see, much of it is unanticipated.
 
-The conventional wisdom among the lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s respectively.
+The conventional wisdom among lay cricket followers is that toss grants the greatest advantage in multi-day affairs like tests and first class matches, followed by day long affairs, and Twenty20s.
 And there is good dose of common sense behind the conventional wisdom.
-Pitches invariably deteriorate over multiple days and batting last in a test match is often challenging.
-The pitch deteriorates far less over the course of the day, or in case of Twenty20s -- a few hours.
-And indeed unlimited over matches provide the most consistent advantage -- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{fig:type}).
+Pitches invariably deteriorate over multiple days and batting last in a test match is often the most challenging time to bat.
+The pitch deteriorates far less over the course of the day, or in case of Twenty20s, a few hours.
+And indeed unlimited over matches provide the most consistent advantage --- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{fig:type}).
 But the heftiest advantage is in domestic one-day matches (List A), approximately 4\%.

### 2015-12-13 scripts/05_cricket.R

@@ -14,6 +14,7 @@ source("scripts/04_merge_ranking_grounds_data.R")
 library(plyr)
 library(stringi)
 library(xtable)
+library(tidyr)
 
 # Read in data/ now sourcing it - see 04_merge_ranking_data
 # cricket <- read.csv("data/final_output.csv")
@@ -80,9 +81,14 @@ cricket$home_toss_win_t <- stri_detect_fixed(cricket$ground, cricket$win_toss, c
                            stri_detect_fixed(cricket$team2, cricket$country, case_insensitive=TRUE) |
                            stri_detect_fixed(cricket$country, cricket$team1, case_insensitive=TRUE) | 
                            stri_detect_fixed(cricket$country, cricket$team2, case_insensitive=TRUE)
-cricket$home_country_data  <- cricket$country == cricket$team1 | cricket$country==cricket$team2
 
 "
+
+cricket$home_country_data  <- cricket$country == cricket$team1 | cricket$country==cricket$team2
+
+# Home country wins toss
+cricket$home_wins_toss  <- ifelse(cricket$country==cricket$team1, cricket$team1_win_toss, cricket$team2_win_toss)
+# with(cricket[cricket$home_country_data==1,], mean(home_wins_toss))
 
 
 "
@@ -90,28 +96,32 @@ Melt the data
 Two rows per match
 "
 
-# "team1", "team2", "team1_id", "team2_id", "team1_rank", "team2_rank", "team1_win_toss", "team2_win_toss", "team1_win_game", "team2_win_game", 
-# "tossgame", "win_toss", "bat_or_bowl", "outcome", "win_game"
-
 # Match level vars: 
 match_cols <- c("url", "date", "day_n_night", "ground", "rain", "duckworth_lewis", "match_id", "type_of_match", "month", "year", "diff_ranks", 
-      "signed_diff_ranks", "ground_id", "country", "continent", "latitude", "longitude", "uniqueid", "draw")
+      "signed_diff_ranks", "ground_id", "country", "continent", "latitude", "longitude", "uniqueid", "draw", "outcome")
 
+# Team cols, rename for gather/separate to work well
 team_cols <- c("team1", "team2", "team1_id", "team2_id", "team2_rank", "team1_rank", "team1_win_toss", "team2_win_toss", "team1_win_game", "team2_win_game")
[…]

### 2015-12-13 write_up/cricket.tex
a couple more to go

@@
 \centering
-\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Type of Match}
+\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Type of Match}
 \includegraphics[scale=.85]{../figs/winbyType.pdf}
@@
 \centering
-\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss By Day/Night}
+\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Day or Day and Night}
 \includegraphics[scale=.85]{../figs/winbyDayNight.pdf}
@@
 \centering
-\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Duckworth-Lewis}
+\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Whether or not Duckworth-Lewis was invoked}
 \includegraphics[scale=.85]{../figs/winbyDL.pdf}

### 2015-12-13 scripts/05_cricket.R

@@ -121,7 +121,8 @@ crickett$wintoss <- as.numeric(crickett$wintoss)
 crickett$wingame <- as.numeric(crickett$wingame)
 
 # Data Integrity
-# ddply(crickett, ~type_of_match + day_n_night, summarise, mean=mean(wintoss))
+ddply(crickett, ~type_of_match + day_n_night, summarise, mean=mean(wintoss))
+with(crickett, xtabs( ~ type_of_match + wingame))
 
 "
 Analysis
@@ -208,8 +209,9 @@ Win by Day/Night
 No test or first-class
 "
 
-ltdcricket <- subset(cricket, type_of_match!="FC" & type_of_match!="TEST")
-ltd_day_n_night <- ddply(ltdcricket,~type_of_match + day_n_night,summarise, diff=mean(I(tossgame==1) - I(tossgame==0)), count=length(unique(url)))
+ltdcricket <- subset(crickett, type_of_match!="FC" & type_of_match!="TEST")
+
+ltd_day_n_night <- ddply(ltdcricket, ~type_of_match + day_n_night, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
 ltd_day_n_night$diff <- ltd_day_n_night$diff*100
 
 ggplot(ltd_day_n_night, aes(x=type_of_match, y=diff, fill=factor(day_n_night))) + 
@@ -217,27 +219,32 @@ geom_bar(stat="identity", position="dodge") +
 theme_minimal() + 
 xlab("") +
 scale_fill_discrete(name="", labels=c(" Day   ", " Day and Night")) + 
-scale_y_continuous(breaks=seq(-7, 8, 1), labels=nolead0s(seq(-7, 8, 1)), limits=c(-7, 8), name="") +
+scale_y_continuous(breaks=seq(-10, 10, 1), labels= paste0(nolead0s(seq(-10, 10, 1)), "%"), limits=c(-10, 10.5), name="") +
 theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted"),
       panel.grid.minor.x = element_blank(),
       panel.grid.major.x = element_line(colour = "#f7f7f7", linetype = "solid"),
       panel.border       = element_blank(),
-      legend.position  	 = "bottom",
+      legend.position    = "bottom",
       legend.text        = element_text(size=10),
       legend.background  = element_rect(color="#ffffff"),
       legend.key         = element_rect(color="#ffffff", fill="#ffffff"),
[…]

### 2015-12-13 scripts/04_merge_ranking_grounds_data.R
draft v. of graphs done

@@ -88,9 +88,6 @@ table(as.character(match$team2[!is.na(match$team2_rank)]))
 # Diff in ranking
 match$diff_ranks <- abs(match$team1_rank - match$team2_rank)
 
-# Signed diff
-match$signed_diff_ranks <- ifelse(as.character(match$team1)==as.character(match$win_toss), match$diff_ranks, -1*match$diff_ranks)
-
 # Add the grounds data
 
 # Trim leading and trailing spaces for grounds

### 2015-12-13 scripts/05_cricket.R

@@ -98,7 +98,7 @@ Two rows per match
 
 # Match level vars: 
 match_cols <- c("url", "date", "day_n_night", "ground", "rain", "duckworth_lewis", "match_id", "type_of_match", "month", "year", "diff_ranks", 
-      "signed_diff_ranks", "ground_id", "country", "continent", "latitude", "longitude", "uniqueid", "draw", "outcome")
+     "ground_id", "country", "continent", "latitude", "longitude", "uniqueid", "draw", "outcome")
 
 # Team cols, rename for gather/separate to work well
 team_cols <- c("team1", "team2", "team1_id", "team2_id", "team2_rank", "team1_rank", "team1_win_toss", "team2_win_toss", "team1_win_game", "team2_win_game")
@@ -119,6 +119,7 @@ crickett$home_country <- crickett$country == crickett$name
 
 crickett$wintoss <- as.numeric(crickett$wintoss)
 crickett$wingame <- as.numeric(crickett$wingame)
+crickett$signed_diff_ranks <- ifelse(crickett$var=="team1", crickett$diff_ranks, -crickett$diff_ranks)
 
 # Data Integrity
 ddply(crickett, ~type_of_match + day_n_night, summarise, mean=mean(wintoss))
@@ -250,7 +251,7 @@ ggsave("figs/winbyDayNight.pdf", width=6)
 Win by DL
 "
 
-ltd_dl <- ddply(ltdcricket,~type_of_match + duckworth_lewis,summarise, diff=mean(I(tossgame==1) - I(tossgame==0)), count=length(unique(url)))
+ltd_dl <- ddply(ltdcricket, ~type_of_match + duckworth_lewis, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
 ltd_dl$diff <- ltd_dl$diff*100
 
 ggplot(ltd_dl, aes(x=type_of_match, y=diff, fill=factor(duckworth_lewis))) + 
@@ -263,21 +264,27 @@ theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted")
       panel.grid.minor.x = element_blank(),
       panel.grid.major.x = element_line(colour = "#f7f7f7", linetype = "solid"),
       panel.border       = element_blank(),
-      legend.position  	 = "bottom",
+      legend.position    = "bottom",
       legend.text        = element_text(size=10),
       legend.background  = element_rect(color="#ffffff"),
[…]

### 2015-12-13 write_up/cricket.tex
better graphs

@@
 \begin{figure}[htbp]
+\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Type of Match.}
+\begin{minipage}{0.85\textwidth}
 \centering
-\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Type of Match}
 \includegraphics[scale=.85]{../figs/winbyType.pdf}
+{\footnotesize Note: \emph{n} refers to the number of matches.\par}
+\end{minipage}
 \label{fig:type}

### 2015-12-13 scripts/04_merge_ranking_grounds_data.R

@@ -89,7 +89,6 @@ table(as.character(match$team2[!is.na(match$team2_rank)]))
 match$diff_ranks <- abs(match$team1_rank - match$team2_rank)
 
 # Add the grounds data
-
 # Trim leading and trailing spaces for grounds
 match$ground <- gsub("^\\s+", "", trim.trailing(match$ground))
 # Add dat

### 2015-12-13 scripts/05_cricket.R

@@ -5,7 +5,7 @@
 "
 
 # setwd
-setwd(paste0(githubdir, "/cricket-stats"))
+setwd(paste0(githubdir, "cricket-stats"))
 
 # Source merge script
 source("scripts/04_merge_ranking_grounds_data.R")
@@ -69,6 +69,8 @@ cricket$team1_win_toss <- 1*(cricket$team1==cricket$win_toss)
 cricket$team2_win_toss <- 1*(cricket$team2==cricket$win_toss)
 cricket$team1_win_game <- 1*(cricket$team1==cricket$win_game)
 cricket$team2_win_game <- 1*(cricket$team2==cricket$win_game)
+cricket$team1_win_game[cricket$draw==1] <- .5
+cricket$team2_win_game[cricket$draw==1] <- .5
 
 "
 Country that won toss == Home Country
@@ -89,7 +91,6 @@ cricket$home_country_data  <- cricket$country == cricket$team1 | cricket$country
 # Home country wins toss
 cricket$home_wins_toss  <- ifelse(cricket$country==cricket$team1, cricket$team1_win_toss, cricket$team2_win_toss)
 # with(cricket[cricket$home_country_data==1,], mean(home_wins_toss))
-
 
 "
 Melt the data
@@ -112,7 +113,6 @@ crickett <- cricket %>% gather(key, value, starts_with('team')) %>% separate(key
 Recode, Fix Variable Type
 "
 # The game is drawn
-crickett$wingame[crickett$draw==1] <- .5
 
 crickett$bat_bowl     <- ifelse(crickett$wintoss, crickett$bat_or_bowl, ifelse(crickett$bat_or_bowl=="bowl", "bat", "bowl"))
 crickett$home_country <- crickett$country == crickett$name
@@ -177,8 +177,10 @@ cricket$type_of_match <- factor(cricket$type_of_match, levels=c("FC", "TEST", "L
 Win By Match Type
 "
 
-win_match_type <- ddply(crickett, ~type_of_match, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]))
+win_match_type <- ddply(crickett, ~type_of_match, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
 win_match_type$diff <- win_match_type$diff*100
[…]

### 2015-12-13 scripts/05_cricket.R
added by country

@@ -298,7 +298,6 @@ ggsave("figs/winbyDL.pdf", width=5)
 
 "
 Win by Diff. in ranks
-
 Probab. of team that wins the toss winning conditional on signed ranking diff. w/ competing team
 
 "
@@ -339,8 +338,8 @@ ggsave("figs/winbyRank.pdf", width=7)
 Is there over time learning? If so, toss adv. would increase. 
 Or it could be that teams develop better strategies to offset toss advantage. 
 If you are going to come up short half the times, you develop strategies to counter that.
+But lots of moving parts for over time stuff. For instance, format is a concern. Certain formats not around earlier.
 " 
-
 
 
 "
@@ -349,6 +348,40 @@ For this - we would want to do Win/Win Toss - Win/Lose Toss to adjust for team p
 
 "
 
+small_set <- subset(crickett, name %in% c("India", "Australia", "West Indies", "England", "New Zealand", "Pakistan", "Sri Lanka"))
 
+by_country <- ddply(small_set, ~ name, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
+by_country$diff <- by_country$diff*100
+
+ggplot(by_country, aes(x=name, y=diff)) + 
+geom_bar(stat = "identity", position = "identity", fill="#42c4c7") + 
+theme_minimal() + 
+xlab("") +
+scale_y_continuous(breaks=seq(-3, 7, 1), labels= paste0(nolead0s(seq(-3, 7, 1)), "%"), limits=c(-3, 7), name="") +
+theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted"),
+      panel.grid.minor.x = element_blank(),
+      panel.grid.major.x = element_line(colour = "#f7f7f7", linetype = "solid"),
+      panel.border       = element_blank(),
+      legend.position    = "bottom",
+      legend.text        = element_text(size=10),
+      legend.background  = element_rect(color="#ffffff"),
+      legend.key         = element_rect(color="#ffffff", fill="#ffffff"),
+      legend.key.size    = unit(.1,"cm"),
+      legend.margin      = unit(.2,"cm"),
+      title              = element_text(size=8),
+      axis.title         = element_text(size=8),
+      axis.text          = element_text(size=8),
+      axis.ticks.y       = element_blank(),
[…]

### 2015-12-14 write_up/cricket.tex
added by country
draft cricket

@@
 This advantage, however, varies by format, by conditions, by whether or not a particular formula was used to adjust scores when it rains, and how much better the team that won the toss is vis-\`{a}-vis the competing team.
-Some of the variability is expected, but as we will see, much of it is unanticipated.
+Much of the variability follows expected patterns.
 
@@
 The pitch deteriorates far less over the course of the day, or in case of Twenty20s, a few hours.
-And indeed unlimited over matches provide the most consistent advantage --- the average advantage over FC and test matches is north of 2.2\% (see Figure~\ref{fig:type}).
-But the heftiest advantage is in domestic one-day matches (List A), approximately 4\%.
-The number for ODIs, however, is a bit puzzling -- there is a slight disadvantage to winning the toss.
+And indeed unlimited over matches provide the most consistent advantage --- the average advantage over FC and test matches is north of 2.7\% (see Figure~\ref{fig:type}).
+Looked in relative terms, the advantage of winning the toss is also close to the greatest in multi-day affairs.
+Only about 60\% of test matches end in a  clear decision, the rest end in a draw.
+Thus, the advantage is closer to 4.5\%.
+
+The heftiest advantage, however, is in domestic one-day matches (List A), approximately 4.8\%.
+The number for ODIs, however, is a bit puzzling --- there is an advantage to winning the toss but it is really slight (.83\%).
 That can best be ascribed to choosing badly.
+In T20s and T20Is, the advantage is 1.3\% and 3.8\% respectively.
 
@@
 It is often claimed that the toss is more crucial in day and night matches.
-Due to dew -- it is thought to make bowling hard, and lower visibility of the white ball under lights -- it allegedly makes catching hard, it is thought that the team that fields second is disadvantaged.
-It turns out that the conventional wisdom is largely vindicated, except in the case of Twenty20 Internationals (see Figure ~\ref{fig:dn}).
-In each, domestic one-day and Twenty20 matches, and one-day internationals, the advantage of winning the toss is at least 3\%, and in case of one-day internationals, nearly 8\%.
+Due to dew --- it is thought to make bowling hard, and the visibility of the white ball is thought to be lower under lights, which makes catching hard --- it is thought that the team that fields second is disadvantaged.
+It turns out that the conventional wisdom is largely vindicated, except perhaps for Twenty20 Internationals (but we have very little on Day and Night T20 Internationals to say anything with great certainty) (see Figure ~\ref{fig:dn}).
+In each, domestic one-day and Twenty20 matches, and one-day internationals, the advantage of winning the toss in Day and Night matches is heftier, and in the case of ODIs and domestic T20s, markedly so --- the difference in advantage of winning the toss between Day and Night and Day versions is over 9\% for both.
+Splitting by whether or not the match spans Day and Night also reveals that in day ODIs, the team that wins the toss loses more often than the team that wins it.
+We can  attribute that to choosing badly.
 
@@
 We can use the random nature of who wins the toss to see if winning percentages are strongly conditioned by matches that use Duckworth-Lewis and matches that don't.
-We find scant differences except in the case of international Twenty20s (see Figure ~\ref{fig:dl}).
+Again the conventional wisdom suggests that the team that wins the toss is at an advantage in such matches.
+And again, the conventional wisdom is vindicated.
+In List A, ODI and domestic Twenty20 matches, the advantage of winning the toss is greater when Duckworth Lewis is used to adjust the scores than when it isn't --- the difference in advantage ranges from about .75\% to over 2\% (see Figure ~\ref{fig:dl}).
 
@@
 
-Lastly, winning a toss ought to matter the most when two closely matched teams are playing.
+Winning a toss ought to matter the most when two closely matched teams are playing.
 What good would winning the toss do if you are matched with a team you have no chance beating.
@@
 As Figure ~\ref{fig:ranks}, there is a sharp curve around 0.
-When closely matched teams win, the winning ratio is strongly affected by who wins the toss.
+When closely matched teams win, the team that wins the toss has a consistent advantage over the team that loses the toss.
 
@@
 
+How the advantage vary by country?
+Are some countries better than others?
+We investigated the question by tallying the advantage by team.
+Sri Lanka and India do especially well (see ~\ref{fig:country}), coming in at 4.53\% and 3.35\% respectively.
+Australia and West Indies hover around 1.3--1.5\% and England at 2.33\%.
+New Zealand, however, does worse when it wins the toss than when it loses it --- difference is about 1.21\%.
+
+\begin{figure}[htbp]
+\centering
+\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Country}
+\includegraphics[scale=.85]{../figs/winbyCountry.pdf}
+\label{fig:country}
+\end{figure}
+
 This isn't a comprehensive inventory of analyses that one could do on the topic.
@@
 Have teams become smarter?
-How the advantage vary by country -- are some countries better than others?
 Other more nuanced but commonly thought questions also remain.

### 2015-12-14 scripts/05_cricket.R

@@ -153,6 +153,9 @@ res[3,2] - res[3,1]
 # By Type of Match
 ddply(crickett, ~type_of_match + wintoss, summarise, mean = mean(wingame==1), n = length(unique(uniqueid)),  se=2*100*sqrt(mean*(1-mean)/n))
 
+# To interpret, need baserate
+ddply(crickett, ~type_of_match + wintoss, summarise, mean = mean(wingame==1), n = length(unique(uniqueid)),  se=2*100*sqrt(mean*(1-mean)/n))
+
 # By Day/N/Night
 ddply(crickett, ~day_n_night +   wintoss, summarise,   mean = mean(wingame==1), n = length(unique(uniqueid)), se=2*100*sqrt(mean*(1-mean)/n))
 

### 2015-12-15 write_up/cricket.tex
split ranking by odi, tests, new write up for ranking

@@
 
-Winning a toss ought to matter the most when two closely matched teams are playing.
-What good would winning the toss do if you are matched with a team you have no chance beating.
-To study this, we collected ranking data.
-The ranking data are only available for international teams.
-And for reasons to do with convenience, we limited ourselves to collecting data on one-day internationals and tests between the earliest starting point and 2013.
-(These ranking data are updated monthly.) We measure how closely matched the teams are by simply subtracting the ranking points of team that lost the toss from the team that won it.
-The results are expected but new.
+Winning the toss ought to matter the most when the difference between the quality of the teams that are playing is the least.
+Similarly, it is unlikely that winning the toss would change who wins the game when two ill-matched teams are playing.
+To study this, we collected data on team quality.
+The ICC publishes team ratings for international test and one-day teams each month.\footnote{For details about how the ICC produces these ratings, see \href{http://icc-live.s3.amazonaws.com/cms/media/about_docs/536b1a48c16e5-Reliance\%20ICC\%20ODI\%20Team\%20Rankings\%20FAQs\%202014.pdf}{ICC Rating FAQs}.} Ratings of the ODI teams have been published since 1981, and of test teams, since 1952.
+Of the entire dataset that spans 1981-today and 1952-today, we only have we have data till 2013.\footnote{The format in which the ICC publishes the ratings changed in 2013.
+And scraping the latter data posed additional hurdles.
+We decided that the additional effort wasn't worth the small amount of additional data.}
+
+Team ratings range from 0 to 143 in our data.
+So, for instance, Bangladesh had a rating of 0 in tests for most of 2002 and 2003.
+And Australia in 2007 twice held a rating of 143.
+We measure how closely matched the teams by differencing the ranking points of one team from the other.
+We interact this with whether or not the team won the toss.
+
+The results are expected, but new.
 As Figure ~\ref{fig:ranks}, there is a sharp curve around 0.

### 2015-12-15 scripts/04_merge_ranking_grounds_data.R

@@ -68,6 +68,7 @@ odi_ranks$month_abb <- month.abb[odi_ranks$month]
 test_ranks$month_abb <- month.abb[test_ranks$month]
 
 # Uniques for odi and test
+# It is not ranking but rating data (higher the better)
 odi_ranks$unique  <- paste0("ODI", odi_ranks$country,  odi_ranks$month_abb,  odi_ranks$year)
 test_ranks$unique <- paste0("TEST", test_ranks$country, test_ranks$month_abb, test_ranks$year)
 

### 2015-12-15 scripts/05_cricket.R

@@ -302,11 +302,14 @@ ggsave("figs/winbyDL.pdf", width=5)
 "
 Win by Diff. in ranks
 Probab. of team that wins the toss winning conditional on signed ranking diff. w/ competing team
+We have to separate by ODI and Tests also
 
 "
-rankcricket <- subset(crickett, type_of_match %in% c("ODI", "TEST") & !is.na(signed_diff_ranks))
 
-ggplot(rankcricket, aes(x=signed_diff_ranks, y=wingame*100, colour=factor(wintoss))) + 
+rankcricket_odi <- subset(crickett, type_of_match %in% c("ODI") & !is.na(signed_diff_ranks))
+rankcricket_test <- subset(crickett, type_of_match %in% c("TEST") & !is.na(signed_diff_ranks))
+
+ggplot(rankcricket_odi, aes(x=signed_diff_ranks, y=wingame*100, colour=factor(wintoss))) + 
 geom_hline(yintercept=50, col="#333333", linetype="dashed", alpha=.3, size=.1) +
 geom_vline(xintercept=0, col="#333333", linetype="dashed", alpha=.3, size=.1) +
 geom_smooth(se = FALSE, method = "loess", formula = y ~ x, span=.7, size = .5) +
@@ -334,8 +337,37 @@ theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted")
       plot.margin        = unit(c(0,.5,.5,.5), "cm")) + 
 annotate("text", y=52, x=43, label="Lose Toss", size=3.5, colour="#42c4c7") +
 annotate("text", y=70, x=43, label="Win Toss", size=3.5, colour="#FF9999")
-ggsave("figs/winbyRank.pdf", width=7)
+ggsave("figs/winbyRank_odi.pdf", width=7)
 
+ggplot(rankcricket_test, aes(x=signed_diff_ranks, y=wingame*100, colour=factor(wintoss))) + 
+geom_hline(yintercept=50, col="#333333", linetype="dashed", alpha=.3, size=.1) +
+geom_vline(xintercept=0, col="#333333", linetype="dashed", alpha=.3, size=.1) +
+geom_smooth(se = FALSE, method = "loess", formula = y ~ x, span=.7, size = .5) +
+theme_minimal() + 
+scale_x_continuous(breaks=seq(-40, 45, 5), labels=nolead0s(seq(-40, 45, 5)), limits=c(-40, 45), name="Difference in Ranking Points") +
[…]

### 2015-12-15 scripts/05_cricket.R
win by month in eng

@@ -376,6 +376,47 @@ If you are going to come up short half the times, you develop strategies to coun
 But lots of moving parts for over time stuff. For instance, format is a concern. Certain formats not around earlier.
 " 
 
+"
+Early English Season
+"
+eng_season <- subset(crickett, country=="England")
+
+by_month <- ddply(eng_season, ~ month, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
+by_month <- subset(by_month, month!='Mar') # only 5 entries
+by_month$month <- factor(by_month$month, month.abb, ordered=T)
+by_month$diff <- by_month$diff*100
+by_month <- by_month[order(by_month$month),]
+
+ggplot(by_month, aes(x=month, y=diff)) + 
+geom_bar(stat = "identity", position = "identity", fill="#42c4c7") + 
+theme_minimal() + 
+xlab("") +
+scale_y_continuous(breaks=seq(-5, 5, 1), labels= paste0(nolead0s(seq(-5, 5, 1)), "%"), limits=c(-5, 5), name="") +
+theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted"),
+      panel.grid.minor.x = element_blank(),
+      panel.grid.major.x = element_line(colour = "#f7f7f7", linetype = "solid"),
+      panel.border       = element_blank(),
+      legend.position    = "bottom",
+      legend.text        = element_text(size=10),
+      legend.background  = element_rect(color="#ffffff"),
+      legend.key         = element_rect(color="#ffffff", fill="#ffffff"),
+      legend.key.size    = unit(.1,"cm"),
+      legend.margin      = unit(.2,"cm"),
+      title              = element_text(size=8),
+      axis.title         = element_text(size=8),
+      axis.text          = element_text(size=8),
+      axis.ticks.y       = element_blank(),
+      axis.ticks.x       = element_line(colour = '#f1f1f1'),
+      strip.text.x       = element_text(size=9),
+      legend.text        = element_text(size=8),
+      plot.margin        = unit(c(0,.5,.5,.5), "cm")) + 
+annotate("text", 
+   x = seq(1, 6, 1), 
+   y = ifelse(by_month$diff > 0, by_month$diff + .35, by_month$diff - .35), 
+   label = paste0(round(by_month$diff,2), "% \n (n =", format(by_month$count, big.mark=",", scientific=FALSE), ")"), 
[…]

### 2015-12-15 scripts/05_cricket.R
updated

@@ -306,15 +306,14 @@ We have to separate by ODI and Tests also
 
 "
 
-rankcricket_odi <- subset(crickett, type_of_match %in% c("ODI") & !is.na(signed_diff_ranks))
-rankcricket_test <- subset(crickett, type_of_match %in% c("TEST") & !is.na(signed_diff_ranks))
+rankcricket <- subset(crickett, type_of_match %in% c("ODI", "TEST") & !is.na(signed_diff_ranks))
 
-ggplot(rankcricket_odi, aes(x=signed_diff_ranks, y=wingame*100, colour=factor(wintoss))) + 
+ggplot(rankcricket, aes(x=signed_diff_ranks, y=wingame*100, colour=factor(wintoss))) + 
 geom_hline(yintercept=50, col="#333333", linetype="dashed", alpha=.3, size=.1) +
 geom_vline(xintercept=0, col="#333333", linetype="dashed", alpha=.3, size=.1) +
 geom_smooth(se = FALSE, method = "loess", formula = y ~ x, span=.7, size = .5) +
 theme_minimal() + 
-scale_x_continuous(breaks=seq(-40, 45, 5), labels=nolead0s(seq(-40, 45, 5)), limits=c(-40, 45), name="Difference in Ranking Points") +
+scale_x_continuous(breaks=seq(-40, 47, 10), labels=nolead0s(seq(-40, 47, 10)), limits=c(-40, 47), name="Difference in Ranking Points") +
 scale_y_continuous(breaks=seq(0, 100, 10), labels=paste0(nolead0s(seq(0, 100, 10)), "%"), limits=c(0, 100), name="Percentage Won") +
 scale_colour_manual(values = c("#42c4c7","#FF9999")) + 
 theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted"),
@@ -336,38 +335,10 @@ theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted")
       legend.text        = element_text(size=8),
       plot.margin        = unit(c(0,.5,.5,.5), "cm")) + 
 annotate("text", y=52, x=43, label="Lose Toss", size=3.5, colour="#42c4c7") +
-annotate("text", y=70, x=43, label="Win Toss", size=3.5, colour="#FF9999")
-ggsave("figs/winbyRank_odi.pdf", width=7)
+annotate("text", y=70, x=43, label="Win Toss", size=3.5, colour="#FF9999") + 
+facet_grid(. ~ type_of_match)
 
[…]

### 2015-12-16 scripts/04_merge_ranking_grounds_data.R
update on rank stuff --- not happy yet

@@ -10,13 +10,14 @@ setwd(paste0(githubdir, "/cricket-stats"))
 # Don't want to deal with factors
 options(StringsAsFactors=F)
 
-# Load ranking data
+# Load and merge ranking data
 odi_ranks  <- read.csv("data/rankings_odi.csv")
 test_ranks <- read.csv("data/rankings_test.csv")
+ranks <- rbind(odi_ranks, test_ranks)
+ranks$format <- ifelse(ranks$format=="odi", "ODI", "TEST")
 
 # Ground
 grounds <- read.csv("data/grounds.csv")
-
 
 "
 Notes: ODI/Test ranks data till 2013
@@ -64,27 +65,23 @@ match$team1_id <- with(match, paste0(type_of_match, team1, month, year))
 match$team2_id <- with(match, paste0(type_of_match, team2, month, year))
 
 # Month handling for odi and test rank data (just convert to month abb. here as data cleaner)
-odi_ranks$month_abb <- month.abb[odi_ranks$month]
-test_ranks$month_abb <- month.abb[test_ranks$month]
+ranks$month_abb  <- month.abb[ranks$month]
 
 # Uniques for odi and test
 # It is not ranking but rating data (higher the better)
-odi_ranks$unique  <- paste0("ODI", odi_ranks$country,  odi_ranks$month_abb,  odi_ranks$year)
-test_ranks$unique <- paste0("TEST", test_ranks$country, test_ranks$month_abb, test_ranks$year)
+ranks$unique  <- paste0(ranks$format, ranks$country,  ranks$month_abb,  ranks$year)
 
 # Bring out data 
 match$team1_rank <- match$team2_rank <- NA
-odi  <- odi_ranks$rating[match(match$team1_id, odi_ranks$unique)]
-test <- test_ranks$rating[match(match$team1_id, test_ranks$unique)]
-match$team1_rank <- ifelse(is.na(odi), test, odi)
-odi  <- odi_ranks$rating[match(match$team2_id, odi_ranks$unique)]
-test <- test_ranks$rating[match(match$team2_id, test_ranks$unique)]
-match$team2_rank <- ifelse(is.na(odi), test, odi)
+match$team1_rank  <- ranks$rating[match(match$team1_id, ranks$unique)]
+match$team2_rank  <- ranks$rating[match(match$team2_id, ranks$unique)]
 
 # Adhoc data integrity check 
[…]

### 2015-12-16 scripts/05_cricket.R

@@ -119,11 +119,19 @@ crickett$home_country <- crickett$country == crickett$name
 
 crickett$wintoss <- as.numeric(crickett$wintoss)
 crickett$wingame <- as.numeric(crickett$wingame)
-crickett$signed_diff_ranks <- ifelse(crickett$var=="team1", crickett$diff_ranks, -crickett$diff_ranks)
+crickett$diff_ranks <- as.numeric(crickett$diff_ranks)
 
-# Data Integrity
+crickett$signed_diff_ranks <- ifelse(crickett$var=="team1", crickett$diff_ranks, -1*crickett$diff_ranks)
+
+"
+Ad Hoc Data Integrity Checks
+"
+
 ddply(crickett, ~type_of_match + day_n_night, summarise, mean=mean(wintoss))
 with(crickett, xtabs( ~ type_of_match + wingame))
+
+crickett$name[!is.na(crickett$signed_diff_ranks) & crickett$signed_diff_ranks < -110]
+crickett$name[!is.na(crickett$signed_diff_ranks) & crickett$signed_diff_ranks > 110]
 
 "
 Analysis
@@ -168,10 +176,32 @@ ddply(crickett, ~type_of_match + day_n_night, summarise, mean = mean(wingame==1)
 # By Type of Match, D/L
 ddply(crickett, ~type_of_match + duckworth_lewis, summarise, mean = mean(wingame==1), n = length(unique(uniqueid)),  se=2*100*sqrt(mean*(1-mean)/n))
 
-# Fig libs
+# Figs
+# ~~~~~~~~~~
+
+# Fig Libs
 library(ggplot2)
 library(grid)
 library(goji)
+
+theme_base <- theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linetype = "dotted"),
+      panel.grid.minor.x = element_blank(),
+      panel.grid.major.x = element_line(colour = "#f7f7f7", linetype = "solid"),
+      panel.border       = element_blank(),
+      legend.position    = "bottom",
+      legend.text        = element_text(size=10),
+      legend.background  = element_rect(color="#ffffff"),
+      legend.key         = element_rect(color="#ffffff", fill="#ffffff"),
+      legend.key.size    = unit(.1,"cm"),
+      legend.margin      = unit(.2,"cm"),
+      title              = element_text(size=8),
+      axis.title         = element_text(size=8),
+      axis.text          = element_text(size=8),
+      axis.ticks.y       = element_blank(),
[…]

### 2015-12-16 write_up/cricket.tex
updated
update on rank stuff --- not happy yet
more writing

@@
 The advantage is pointed to by commentators, by the team captains in the pre-toss interview, and by the captain of the losing team in the post-match interview.
-And to wrest the said advantage, one merely needs to pick the side of the coin left facing the sky after the toss.
+And to wrest the said advantage, the captain merely needs to pick the side of the coin that will be left facing the sky after the toss.
 And while this method of granting advantage is fair on average, the system isn't fair in any one game.
 
-At first glance, the imbalance seems inevitable.
+At first glance, this imbalance seems inevitable.
 After all, someone has to bat first.
-One can, however, devise a baseball like system in which short innings are interspersed.
-If that violates the nature of the game too much, one can easily create pitches that don't deteriorate appreciably over the course of a game.
-Or, one can come up with an estimate of the advantage, and adjust the scores accordingly, akin to an adjustment issued when matches are shortened due to rain.
+One can, however, devise a baseball like system, with a series of short innings woven together.
+If that violates the nature of the game too much, one can easily create pitches that don't deteriorate appreciably over the course of a match.
+Or, one can come up with an estimate of the advantage, and adjust the scores accordingly, akin to an adjustment that is issued when the matches are shortened due to rain.
 
@@
 To assess the net observed advantage of winning the toss, we exploit data from a novel dataset of over 43,000 first-class men's cricket matches --- to our knowledge, the largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}.
-And in analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into.
-To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat, Saad2015} etc., we do not condition on post coin-toss decisions.
+In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into.
+To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on post coin-toss decisions.
 We find that winning the toss grants a small but significant advantage, but that advantage varies considerably and systematically.
@@
 Conditional on the outcome of a toss being random, the effect of winning the toss can be attributed to the toss itself.
-Before we exploit the design, we shed some light on the validity of the assumption.
+Any decision made after the outcome of the toss is known, however, is `post-treatment.' In particular, the decision to bat or bowl first is made after accounting for the relative strengths and weaknesses vis-\`{a}-vis the competing team at that particular instance, and thus not independent of team attributes.
+Hence, conditioning on decision to bowl or bat first can induce bias in estimates of advantage of winning a toss.
+Thus, unlike \citet{dawson2009bat}, \citet{Saad2015}, we solely rely on the assumption that the outcome of a toss is random.
+
+But before we exploit the design, we shed some light on the validity of the assumption.
 In particular, we assess whether the coin toss is somehow rigged, with the home side enjoying the rub of the green more often.
@@
 This is a reasonable sized advantage in a competitive sport --- though likely much smaller than the number that most commentators carry in their heads.
+% Also interpret it by comparing to probability of winning when playing with the worst placed team (not defined from last team)
 This advantage, however, varies by format, by conditions, by whether or not a particular formula was used to adjust scores when it rains, and how much better the team that won the toss is vis-\`{a}-vis the competing team.
@@
 \caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Difference in Ranks}
-\includegraphics[scale=.85]{../figs/winbyRank.pdf}
+\includegraphics[width=1\textwidth]{../figs/winbyRank.pdf}
 \label{fig:ranks}

### 2015-12-17 write_up/cricket.tex
restate point on ranking

@@
 Team ratings range from 0 to 143 in our data.
-So, for instance, Bangladesh had a rating of 0 in tests for most of 2002 and 2003.
+For instance, Bangladesh had a rating of 0 in tests for most of 2002 and 2003.
 And Australia in 2007 twice held a rating of 143.
+Both seem reasonable.
 We measure how closely matched the teams by differencing the ranking points of one team from the other.
-We interact this with whether or not the team won the toss.
+Commercial considerations mean that a majority of the games are played among highly ranked and closely matched teams.
+So, we limit our analysis to matches where teams ratings differ by no more than 30 points.
 

### 2015-12-17 write_up/cricket.tex
added english season data

@@
 
+This covers some of the more straightforward questions.
+Other more nuanced but commonly thought questions remain.
+For one, students of the game suspect that the advantage of winning a toss in early English season is especially great.
+We next check that using data on International matches from England.
+
+\begin{figure}[htbp]
+\centering
+\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Month in England}
+\includegraphics[scale=.85]{../figs/winbyMonthEngland.pdf}
+\label{fig:season}
+\end{figure}
+
 This isn't a comprehensive inventory of analyses that one could do on the topic.
 But neither is it meant to be.
-There are some simple questions that still unanswered.
+Some simple questions still remain unanswered.
 For one, how does the toss advantage vary over time?
 Have teams become smarter?
-Other more nuanced but commonly thought questions also remain.
-For one, students of the game suspect that the advantage of winning a toss in early English season is especially great.
 We propose to investigate these questions in a follow-up piece.

### 2016-01-24 scripts/01a_download_match_data.py
updated output

@@ -19,7 +19,7 @@ import simplejson
 import unicodedata
 from BeautifulSoup import BeautifulSoup, SoupStrainer
 
-for match_type in ['list%20a', 'first%20class']: # 'odi', 'test', 't20i', 't20', 
+for match_type in ['list%20a', 'first%20class', 'odi', 'test', 't20i', 't20']: 
     results = []
     r = requests.get('http://search.espncricinfo.com/ci/content/match/search.html?all=1;page=0;search=' + match_type)
     soup = BeautifulSoup(r.text)

### 2016-01-24 scripts/02_parse_cric.py

@@ -8,13 +8,13 @@ import csv
 import simplejson
 from espncricinfo.match import Match
 
-headers = ["team1", "team1_id", "team2", "team2_id", "win_toss", "bat_or_bowl", "outcome", "win_game", "date", "day_n_night", "ground", "rain", "duckworth_lewis", "match_id", "type_of_match", "home_team_id", "umpire_1_id", "umpire_1_name", "umpire_1_country", "umpire_2_id", "umpire_2_name", "umpire_2_country", "tv_umpire_id", "tv_umpire_name", "tv_umpire_country", "referree_id", "referee_name", "referee_country"]
+headers = ["team1", "team1_id", "team2", "team2_id", "win_toss", "bat_or_bowl", "outcome", "win_game", "date", "day_n_night", "ground", "rain", "duckworth_lewis", "match_id", "type_of_match", "match_type_id", "home_team_id", "umpire_1_id", "umpire_1_name", "umpire_1_country", "umpire_2_id", "umpire_2_name", "umpire_2_country", "tv_umpire_id", "tv_umpire_name", "tv_umpire_country", "referree_id", "referee_name", "referee_country", "url"]
 
 results = simplejson.loads(open('matches-first-class.json').read())
-results = list(set(results))[9193:-1]
+results = list(set(results))
 
 ##################################START PROCESSING DATA#########################################
-with open("first_class_output2.csv", "wb") as csvfile:
+with open("first_class_output.csv", "wb") as csvfile:
     writer = csv.writer(csvfile)
     writer.writerow(headers)
     for result in results:
@@ -37,6 +37,10 @@ with open("first_class_output2.csv", "wb") as csvfile:
                 m.team_1()['team_name']
             except KeyError:
                 continue
+            if m.match_json()['international_class_card'] != "":
+                match_type_id = m.match_json()['international_class_id']
+            else:
+                match_type_id = m.match_json()['general_class_id']
             try:
                 if len(m.officials()) > 0:
                     umpires = [o for o in m.officials() if o['player_type_name'] == 'umpire']
@@ -72,7 +76,7 @@ with open("first_class_output2.csv", "wb") as csvfile:
                 mr_id = None
                 mr_name = None
                 mr_country = None
[…]

### 2016-01-25 scripts/02_parse_cric.py
updated final output

@@ -10,17 +10,17 @@ from espncricinfo.match import Match
 
 headers = ["team1", "team1_id", "team2", "team2_id", "win_toss", "bat_or_bowl", "outcome", "win_game", "date", "day_n_night", "ground", "rain", "duckworth_lewis", "match_id", "type_of_match", "match_type_id", "home_team_id", "umpire_1_id", "umpire_1_name", "umpire_1_country", "umpire_2_id", "umpire_2_name", "umpire_2_country", "tv_umpire_id", "tv_umpire_name", "tv_umpire_country", "referree_id", "referee_name", "referee_country", "url"]
 
-results = simplejson.loads(open('matches-first-class.json').read())
-results = list(set(results))
+matches = simplejson.loads(open('all_matches.json').read())
+matches = list(set(matches)) # dedupe
 
 ##################################START PROCESSING DATA#########################################
-with open("first_class_output.csv", "wb") as csvfile:
+with open("final_output.csv", "wb") as csvfile:
     writer = csv.writer(csvfile)
     writer.writerow(headers)
-    for result in results:
+    for match in matches:
         time.sleep(1)
         try:
-            m = Match(int(result))
+            m = Match(int(match))
             if m.match_json()['match_status'] == 'forthcoming':
                 continue
             if m.result() == '':
@@ -76,7 +76,7 @@ with open("first_class_output.csv", "wb") as csvfile:
                 mr_id = None
                 mr_name = None
                 mr_country = None
-            writer.writerow([m.team_1()['team_name'], m.team_1_id(), m.team_2()['team_name'], m.team_2_id(), m.toss_winner(), m.toss_decision(), m.result(), m.match_winner(), m.date(), m.lighting(), m.ground_name(), None, duckworth_lewis, result, m.match_class(), match_type_id, m.match_json()['home_team_id'], ump_1['object_id'], ump_1['known_as'], ump_1['team_abbreviation'], ump_2['object_id'], ump_2['known_as'], ump_2['team_abbreviation'], tvu_id, tvu_name, tvu_country, mr_id, mr_name, mr_country, m.match_url])
[…]

### 2016-01-26 scripts/05_cricket.R
dot plots

@@ -66,10 +66,10 @@ cricket$draw <- 1*grepl("Match drawn", cricket$outcome)
 # table(cricket$win_game[cricket$draw==1])
 
 # Win toss, win game
-cricket$team1_win_toss <- 1*(cricket$team1==cricket$win_toss)
-cricket$team2_win_toss <- 1*(cricket$team2==cricket$win_toss)
-cricket$team1_win_game <- 1*(cricket$team1==cricket$win_game)
-cricket$team2_win_game <- 1*(cricket$team2==cricket$win_game)
+cricket$team1_win_toss <- 1*(cricket$team1_id==cricket$win_toss)
+cricket$team2_win_toss <- 1*(cricket$team2_id==cricket$win_toss)
+cricket$team1_win_game <- 1*(cricket$team1_id==cricket$win_game)
+cricket$team2_win_game <- 1*(cricket$team2_id==cricket$win_game)
 cricket$team1_win_game[cricket$draw==1] <- .5
 cricket$team2_win_game[cricket$draw==1] <- .5
 
@@ -90,8 +90,11 @@ cricket$home_toss_win_t <- stri_detect_fixed(cricket$ground, cricket$win_toss, c
 cricket$home_country_data  <- cricket$country == cricket$team1 | cricket$country==cricket$team2
 
 # Home country wins toss
-cricket$home_wins_toss  <- ifelse(cricket$country==cricket$team1, cricket$team1_win_toss, cricket$team2_win_toss)
+cricket$home_wins_toss  <- ifelse(cricket$home_team_id==cricket$team1_id, cricket$team1_win_toss, cricket$team2_win_toss)
 # with(cricket[cricket$home_country_data==1,], mean(home_wins_toss))
+
+# Fix Day/Night
+cricket$day_n_night <- ifelse(cricket$day_n_night=="night match", "day/night match", cricket$day_n_night)
 
 "
 Melt the data
@@ -205,7 +208,7 @@ theme_base <- theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linety
       plot.margin        = unit(c(0,.5,.5,.5), "cm"))
 
 # For figs - let us get type of match is nicer factor order
-crickett$type_of_match <- factor(crickett$type_of_match, levels=c("FC", "TEST", "LISTA", "ODI", "T20", "T20I"))
[…]

### 2016-01-26 write_up/cricket.tex
viva dotplots

@@
 \begin{document}
-\title{\vspace{-.5cm}\normalsize{Fairly Random: Impact of Winning the Toss on the Probability of Winning\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}\vspace{.5cm}}}
+\title{\vspace{-.5cm}\normalsize{Fairly Random: The Impact of Winning the Toss on the Probability of Winning\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}\vspace{.5cm}}}
 \author{\normalsize{Gaurav Sood}\\\href{mailto:gsood07@gmail.com}{\small{gsood07@gmail.com}} \and \normalsize{Derek Willis}\\\href{mailto:dwillis@gmail.com}{\small{dwillis@gmail.com}}}

### 2016-01-26 scripts/05_cricket.R

@@ -266,16 +266,16 @@ Win by DL
 ltd_dl <- ddply(ltdcricket, ~type_of_match + duckworth_lewis, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
 ltd_dl$diff <- ltd_dl$diff*100
 
-ggplot(ltd_dl, aes(x=type_of_match, y=diff, fill=factor(duckworth_lewis))) + 
-geom_bar(stat="identity", position="dodge") +
+ggplot(ltd_dl, aes(y=type_of_match, x=diff, color=factor(duckworth_lewis))) + 
+geom_point() + 
 theme_minimal() + 
-xlab("") +
+ylab("") +
 scale_fill_discrete(name="", labels=c(" No D/L   ", " Duckworth Lewis")) + 
-scale_y_continuous(breaks=seq(-1, 7, 1), labels=nolead0s(seq(-1, 7, 1)), limits=c(-1, 7), name="") +
+scale_x_continuous(breaks=seq(-1, 7, 1), labels= paste0(nolead0s(seq(-1, 7, 1)), "%"), limits=c(-1, 7), name="") +
 theme_base + 
 annotate("text", 
-   x = seq(.75,4.25,.5), 
-   y = ifelse(ltd_dl$diff > 0, ltd_dl$diff+ .25, ltd_dl$diff-.25), 
+   y = seq(.75,4.25,.5), 
+   x = ifelse(ltd_dl$diff > 0, ltd_dl$diff+ .25, ltd_dl$diff-.25), 
    label = paste0(round(ltd_dl$diff,2), "% \n (n =", format(ltd_dl$count, big.mark=",", scientific=FALSE), ")"), 
    colour = "#444444", 
    size = 2.5) + 
@@ -324,20 +324,20 @@ Early English Season
 eng_season <- subset(crickett, country=="England")
 
 by_month <- ddply(eng_season, ~ month, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
-by_month <- subset(by_month, month!='Mar') # only 5 matches
-by_month$month <- factor(by_month$month, month.abb, ordered=T)
+by_month <- subset(by_month, month!=3) # only 5 matches
+by_month$month <- month.abb[by_month$month]
 by_month$diff <- by_month$diff*100
 by_month <- by_month[order(by_month$month),]
 
[…]

### 2016-01-26 scripts/05_cricket.R
plots being finalized, with s.e. now

@@ -210,21 +210,48 @@ theme_base <- theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linety
 # For figs - let us get type of match is nicer factor order
 crickett$type_of_match2 <- factor(crickett$type_of_match, levels=c("FC", "TEST", "LISTA", "ODI", "T20", "T20I"))
 
+# Bootstrap s.e.
+
+# set.seed
+set.seed(97689)
+boot.se <- function(dat, n_boots = 1000) {
+
+   n_uniques <- length(unique(dat$url))
+   samps     <- replicate(n_boots, sample(1:n_uniques, n_uniques, replace=T))
+   all_diffs <- NA
+
+   for (i in 1:ncol(samps)){
+      small_dat <- dat[match(dat$url[samps[,i]], dat$url, nomatch = 0),]
+      diff      <- with(small_dat, mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]))
+      all_diffs[i] <- diff
+   }
+   sd(all_diffs)
+   #res <- all_diffs[order(all_diffs)]
+   #c(res[c(.025, .975)*n_boots])
+}
+
 "
 Win By Match Type
 "
-# crickett <- subset(crickett, !is.na(type_of_match))
 
-win_match_type <- ddply(crickett, ~type_of_match, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
+crickett$basic_type_of_match <- car::recode(crickett$type_of_match, "c('TEST', 'FC')='FC/TEST';c('T20','T20I')='T20/T20I';c('LISTA', 'ODI')='LISTA/ODI'")
+
+win_match_type <- ddply(crickett, ~basic_type_of_match, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
+se <- ddply(crickett,  ~basic_type_of_match, function(x) c(se = boot.se(x)))
+win_match_type$se   <- se$se[match(win_match_type$basic_type_of_match, se$basic_type_of_match)]
 win_match_type$diff <- win_match_type$diff*100
-win_match_type$type_of_match <- factor(win_match_type$type_of_match, levels=c("T20I", "T20", "ODI", "LISTA",  "TEST", "FC"))
-win_match_type <- win_match_type[order(win_match_type$type_of_match),]
[…]

### 2016-01-26 scripts/05_cricket.R
fancy pants version

@@ -268,45 +268,58 @@ No test or first-class
 
 ltdcricket <- subset(crickett, type_of_match!="FC" & type_of_match!="TEST")
 
-ltd_day_n_night <- ddply(ltdcricket, ~type_of_match + day_n_night, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
-ltd_day_n_night$diff <- ltd_day_n_night$diff*100
+ltd_day_n_night <- ddply(ltdcricket, ~basic_type_of_match + day_n_night, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
+se <- ddply(ltdcricket,  ~basic_type_of_match + day_n_night, function(x) c(se = boot.se(x)))
+ltd_day_n_night$se    <- se$se[match(ltd_day_n_night$basic_type_of_match, se$basic_type_of_match)]
+ltd_day_n_night$diff  <-  ltd_day_n_night$diff*100
+ltd_day_n_night$lci   <-  ltd_day_n_night$diff - 2*ltd_day_n_night$se*100
+ltd_day_n_night$hci   <-  ltd_day_n_night$diff + 2*ltd_day_n_night$se*100
+ltd_day_n_night$id <- c(1,1.1,2,2.1)
 
-ggplot(ltd_day_n_night, aes(y=type_of_match, x=diff, color=day_n_night)) + 
+ggplot(ltd_day_n_night, aes(y=id, x=diff, xmin = lci, xmax = hci, color=day_n_night)) + 
 geom_point() + 
+geom_errorbarh(height = 0) +
+geom_vline(xintercept = 0, color="grey", linetype="dashed") + 
 theme_minimal() + 
 ylab("") +
-scale_fill_discrete(name="", labels=c(" Day   ", " Day and Night")) + 
-scale_x_continuous(breaks=seq(-10, 11, 2), labels= paste0(nolead0s(seq(-10, 11, 2)), "%"), limits=c(-10, 11), name="") +
+scale_y_continuous(breaks=c(1,2), labels= c("LISTA/ODI", "T20/T20I"), limits=c(.5, 2.5), name="") +
+scale_color_discrete(name="", labels=c(" Day   ", " Day and Night")) + 
[…]

### 2016-01-26 scripts/05_cricket.R
super cadillac

@@ -270,7 +270,7 @@ ltdcricket <- subset(crickett, type_of_match!="FC" & type_of_match!="TEST")
 
 ltd_day_n_night <- ddply(ltdcricket, ~basic_type_of_match + day_n_night, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
 se <- ddply(ltdcricket,  ~basic_type_of_match + day_n_night, function(x) c(se = boot.se(x)))
-ltd_day_n_night$se    <- se$se[match(ltd_day_n_night$basic_type_of_match, se$basic_type_of_match)]
+ltd_day_n_night$se    <-  se$se[match(ltd_day_n_night$basic_type_of_match, se$basic_type_of_match)]
 ltd_day_n_night$diff  <-  ltd_day_n_night$diff*100
 ltd_day_n_night$lci   <-  ltd_day_n_night$diff - 2*ltd_day_n_night$se*100
 ltd_day_n_night$hci   <-  ltd_day_n_night$diff + 2*ltd_day_n_night$se*100
@@ -282,7 +282,7 @@ geom_errorbarh(height = 0) +
 geom_vline(xintercept = 0, color="grey", linetype="dashed") + 
 theme_minimal() + 
 ylab("") +
-scale_y_continuous(breaks=c(1,2), labels= c("LISTA/ODI", "T20/T20I"), limits=c(.5, 2.5), name="") +
+scale_y_continuous(breaks=c(1.05,2.05), labels= c("LISTA/ODI", "T20/T20I"), limits=c(.5, 2.5), name="") +
 scale_color_discrete(name="", labels=c(" Day   ", " Day and Night")) + 
 scale_x_continuous(breaks=seq(-4, 10, 2), labels= paste0(nolead0s(seq(-4, 10, 2)), "%"), limits=c(-3, 10), name="") +
 theme_base + 
@@ -298,9 +298,9 @@ ggsave("figs/winbyDayNight.pdf", width=5.5)
 Win by DL
 "
 
-ltd_dl       <- ddply(ltdcricket, ~basic_type_of_match + duckworth_lewis, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
-se           <- ddply(ltdcricket,  ~basic_type_of_match + duckworth_lewis, function(x) c(se = boot.se(x)))
[…]

### 2016-01-28 scripts/05_cricket.R
plots are done. s.e. remains

@@ -1,6 +1,10 @@
 "
-@title:   Fairly Random: Impact of Winning the Toss on the Probability of Winning
-@authors: Gaurav Sood and Derek Willis
+@description:
+1. Analyses
+2. Graphs 
+
+@authors: 
+Gaurav Sood and Derek Willis
 
 "
 
@@ -16,85 +20,34 @@ library(stringi)
 library(xtable)
 library(tidyr)
 
-# Read in data/ now sourcing it - see 04_merge_ranking_data
-# cricket <- read.csv("data/final_output.csv")
+"
+Notes About the Data:
+
+Not scraped 'matches' where the match was abandoned without a toss
+In a previous dataset, about 3k matches were abandoned so. Don't like the 7% number. Touch too high, imho. But checked data - turns out to be ok.
+sum(match$win_toss=='')
+
+Any matches with missing team name info.?
+sum(match$team2=='')
+"
 
 "
-Data integrity
-match$outcome[match$win_game==''] # looks good
-table(match$outcome[match$win_toss=='' & match$outcome!=''])
+Take out matches where there was no result. See for e.g.,
+http://www.espncricinfo.com/matches/engine/match/329869.html
 
-Take out matches with no toss (~ no match)
-About 3k matches. Don't like the 7% number. Touch too high, imho. But checked data - turns out to be ok.
+Takes out 780 matches
+"
+temp <- grepl("No result", match$outcome)
+cricket <- subset(match, !temp)
 
 "
-cricket <- subset(match, win_toss!="")
-
-"
-Team 2 missing in one case
-"
-cricket <- subset(cricket, team2!="")
-
-
-"
-Take out matches where there was no result
-Leaves us w/ 39672 rows
-"
-temp <- grepl("No result", cricket$outcome)
-cricket <- subset(cricket, !temp)
-
-
-"
-Take out matches where no decision on who is to bowl/bat first is made
+Take out matches where decision on who bowled/batted first is unknown. See this, for e.g.:
+http://www.espncricinfo.com/matches/engine/match/537589.html 
+Takes out 2384 matches
 "
 
 cricket <- subset(cricket, bat_or_bowl!="")
 
-
-"
-More recoding
[…]

### 2016-03-03 scripts/04_merge_ranking_grounds_data.R
recode done

@@ -87,6 +87,9 @@ match$di_type_of_match <- car::recode(match$type_of_match, "c('LISTA', 'T20', 'F
 # Distinguish Men's ODI, Test, T20I from rest as rankings only for men's 
 match$men_type_of_match <- ifelse(match$women | match$youth | match$unofficial, paste0("WYU", match$type_of_match), match$type_of_match)
 
+# Fix Day/Night
+match$day_n_night <- ifelse(match$day_n_night=="night match", "day/night match", match$day_n_night)
+
 # Go for exact match
 # Unique_ID1, Unique_ID2
 match$team1_spid <- with(match, tolower(paste0(men_type_of_match, team1, month, year)))
@@ -128,6 +131,9 @@ match$uniqueid <- 1:nrow(match)
 # Drawn Matches
 match$draw <- 1*grepl("Match drawn", match$outcome)
 
+# Data integrity check: 
+table(match$draw, match$basic_type_of_match)
+
 # Win toss, win game
 match$team1_win_toss <- 1*(match$team1_id==match$win_toss)
 match$team2_win_toss <- 1*(match$team2_id==match$win_toss)
@@ -137,11 +143,24 @@ match$team1_win_game[match$draw==1] <- .5
 match$team2_win_game[match$draw==1] <- .5
 
 # Home country
-match$home_country_data  <- match$country == match$team1 | match$country==match$team2
+match$team1_home_country  <- match$country == match$team1
+match$team2_home_country  <- match$country == match$team2
+match$team1_home_country[!match$team1_home_country & !match$team2_home_country] <- NA
+match$team2_home_country[!match$team1_home_country & !match$team2_home_country] <- NA
 
 # Home country wins toss
-match$home_wins_toss  <- ifelse(match$home_team_id==match$team1_id, match$team1_win_toss, match$team2_win_toss)
-# with(cricket[match$home_country_data==1,], mean(home_wins_toss))
+match$home_wins_toss  <- 0
+match$home_wins_toss[!is.na(match$team1_home_country) & match$team1_home_country & match$team1_win_toss] <- 1
+match$home_wins_toss[!is.na(match$team1_home_country) & match$team2_home_country & match$team2_win_toss] <- 1
+match$home_wins_toss[is.na(match$team1_home_country)] <- NA
 
-# Fix Day/Night
-match$day_n_night <- ifelse(match$day_n_night=="night match", "day/night match", match$day_n_night)
+# Umpiring
[…]

### 2016-03-03 scripts/04_merge_ranking_grounds_data.R
triple venti coding of home umpires

@@ -164,3 +164,6 @@ match$team2_home_umpire2 <- match$team2 == match$umpire_2_country
 
 match$team1_home_tv_umpire <- match$team1 == match$tv_umpire_country
 match$team2_home_tv_umpire <- match$team2 == match$tv_umpire_country
+
+match$team1_home_umpire <- rowSums(cbind(match$team1_home_umpire1, match$team1_home_umpire2, match$team1_home_tv_umpire))
+match$team2_home_umpire <- rowSums(cbind(match$team2_home_umpire1, match$team2_home_umpire2, match$team2_home_tv_umpire))

### 2016-03-03 scripts/04_merge_ranking_grounds_data.R
one more change

@@ -142,6 +142,9 @@ match$team2_win_game <- 1*(match$team2_id==match$win_game)
 match$team1_win_game[match$draw==1] <- .5
 match$team2_win_game[match$draw==1] <- .5
 
+# Win toss and win game
+match$win_toss_win_game <- ((match$team1_win_toss & match$team1_win_game==1) | (match$team2_win_toss & match$team2_win_game==1))
+
 # Home country
 match$team1_home_country  <- match$country == match$team1
 match$team2_home_country  <- match$country == match$team2

### 2016-03-03 scripts/04_merge_ranking_grounds_data.R
fixed home country issue

@@ -149,7 +149,7 @@ match$win_toss_win_game <- ((match$team1_win_toss & match$team1_win_game==1) | (
 match$team1_home_country  <- match$country == match$team1
 match$team2_home_country  <- match$country == match$team2
 match$team1_home_country[!match$team1_home_country & !match$team2_home_country] <- NA
-match$team2_home_country[!match$team1_home_country & !match$team2_home_country] <- NA
+match$team2_home_country[is.na(match$team1_home_country)] <- NA
 
 # Home country wins toss
 match$home_wins_toss  <- 0

### 2016-03-03 scripts/04_merge_ranking_grounds_data.R
scripts + figs updated

@@ -159,14 +159,14 @@ match$home_wins_toss[is.na(match$team1_home_country)] <- NA
 
 # Umpiring
 # It was tested out with 1 umpire beginning in 1992 and then made standard with 2 in 2002: http://www.espncricinfo.com/magazine/content/story/511175.html
-match$team1_home_umpire1 <- match$team1 == match$umpire_1_country
-match$team2_home_umpire1 <- match$team2 == match$umpire_1_country
+match$team1_umpire1 <- match$team1 == match$umpire_1_country
+match$team2_umpire1 <- match$team2 == match$umpire_1_country
 
-match$team1_home_umpire2 <- match$team1 == match$umpire_2_country
-match$team2_home_umpire2 <- match$team2 == match$umpire_2_country
+match$team1_umpire2 <- match$team1 == match$umpire_2_country
+match$team2_umpire2 <- match$team2 == match$umpire_2_country
 
-match$team1_home_tv_umpire <- match$team1 == match$tv_umpire_country
-match$team2_home_tv_umpire <- match$team2 == match$tv_umpire_country
+match$team1_tv_umpire <- match$team1 == match$tv_umpire_country
+match$team2_tv_umpire <- match$team2 == match$tv_umpire_country
 
-match$team1_home_umpire <- rowSums(cbind(match$team1_home_umpire1, match$team1_home_umpire2, match$team1_home_tv_umpire))
-match$team2_home_umpire <- rowSums(cbind(match$team2_home_umpire1, match$team2_home_umpire2, match$team2_home_tv_umpire))
+match$team1_umpire <- rowSums(cbind(match$team1_umpire1, match$team1_umpire2, match$team1_tv_umpire))
+match$team2_umpire <- rowSums(cbind(match$team2_umpire1, match$team2_umpire2, match$team2_tv_umpire))

### 2016-03-03 scripts/05_cricket.R

@@ -48,6 +48,8 @@ Takes out 2384 matches
 
 cricket <- subset(cricket, bat_or_bowl!="")
 
+# Prop. who win toss, win game. (not including draws)
+table(cricket$win_toss_win_game, cricket$basic_type_of_match)
 
 "
 Melt the data
@@ -55,16 +57,25 @@ Two rows per match
 "
 
 # Match level vars: 
-match_cols <- c("url", "date", "day_n_night", "ground", "rain", "duckworth_lewis", "match_id", "type_of_match", "basic_type_of_match", "type_of_match2", 
-                 "di_type_of_match", "month", "year", "diff_ranks", "ground_id", "country", "continent", "latitude", "longitude", "uniqueid", "draw", "outcome")
+match_cols <- c("url", "match_id", "uniqueid", "date", "month", "day", "year", "rain", 
+                 "type_of_match", "basic_type_of_match", "type_of_match2", "di_type_of_match", "men_type_of_match", "day_n_night", "youth", "women", "youth.women", "unofficial", "duckworth_lewis", 
+                 "win_toss", "win_game", "home_wins_toss", "draw", "outcome", "diff_ranks", "bat_or_bowl", 
+                 "ground", "ground_id", "country", "continent", "latitude", "longitude")
 
 # Team cols, rename for gather/separate to work well
-team_cols <- c("team1", "team2", "team1_id", "team2_id", "team2_rank", "team1_rank", "team1_win_toss", "team2_win_toss", "team1_win_game", "team2_win_game")
-rename_cols <- c("team1.name", "team2.name", "team1.id", "team2.id", "team2.rank", "team1.rank", "team1.wintoss", "team2.wintoss", "team1.wingame", "team2.wingame")
+team_cols <- c("team1", "team2", "team1_id", "team2_id", "team2_rank", "team1_rank", "team1_win_toss", "team2_win_toss", "team1_win_game", "team2_win_game", 
+               "team1_home_country", "team2_home_country", 
+               "team1_umpire1", "team2_umpire1", "team1_umpire2", "team2_umpire2", "team1_tv_umpire", "team2_tv_umpire", "team1_umpire", "team2_umpire")
[…]

### 2016-03-03 scripts/04_merge_ranking_grounds_data.R
umpire stuff in the scripts

@@ -160,13 +160,20 @@ match$home_wins_toss[is.na(match$team1_home_country)] <- NA
 # Umpiring
 # It was tested out with 1 umpire beginning in 1992 and then made standard with 2 in 2002: http://www.espncricinfo.com/magazine/content/story/511175.html
 match$team1_umpire1 <- match$team1 == match$umpire_1_country
+match$team1_umpire1[match$umpire_1_country==""] <- NA
 match$team2_umpire1 <- match$team2 == match$umpire_1_country
+match$team2_umpire1[match$umpire_1_country==""] <- NA
 
 match$team1_umpire2 <- match$team1 == match$umpire_2_country
+match$team1_umpire2[match$umpire_2_country==""] <- NA
 match$team2_umpire2 <- match$team2 == match$umpire_2_country
+match$team2_umpire2[match$umpire_2_country==""] <- NA
+
 
 match$team1_tv_umpire <- match$team1 == match$tv_umpire_country
+match$team1_tv_umpire[match$tv_umpire_country==""] <- NA
 match$team2_tv_umpire <- match$team2 == match$tv_umpire_country
+match$team2_tv_umpire[match$tv_umpire_country==""] <- NA
 
-match$team1_umpire <- rowSums(cbind(match$team1_umpire1, match$team1_umpire2, match$team1_tv_umpire))
-match$team2_umpire <- rowSums(cbind(match$team2_umpire1, match$team2_umpire2, match$team2_tv_umpire))
+match$team1_umpire <- rowSums(cbind(match$team1_umpire1, match$team1_umpire2, match$team1_tv_umpire), na.rm=T)
+match$team2_umpire <- rowSums(cbind(match$team2_umpire1, match$team2_umpire2, match$team2_tv_umpire), na.rm=T)

### 2016-03-03 scripts/05_cricket.R

@@ -398,7 +398,15 @@ annotate("text",
 ggsave("figs/winbyCountry.pdf", width=6)
 
 
-# Split by Home Umpires on Winning Toss
+# Split by Home Umpires on Winning Toss and Winning Match
 # It was tested out with 1 umpire beginning in 1992 and then made standard with 2 in 2002: 
 # http://www.espncricinfo.com/magazine/content/story/511175.html
+
+# Only international matches
+with(crickett[!is.na(crickett$umpire1) & !is.na(crickett$umpire2) & crickett$di_type_of_match=="International",], xtabs( ~ umpire + wintoss))
+with(crickett[!is.na(crickett$umpire1) & !is.na(crickett$umpire2) & crickett$di_type_of_match=="International",], xtabs( ~ umpire + wingame))
+
+# By Type of Match
+ddply(crickett[!is.na(crickett$umpire1) & !is.na(crickett$umpire2) & crickett$di_type_of_match=="International",], ~ umpire, summarise, win_toss = mean(wintoss==1), lose_toss = mean(wintoss==0), n = length(unique(uniqueid)),  se_win=2*sqrt(win_toss*(1-win_toss)/n))
+ddply(crickett[!is.na(crickett$umpire1) & !is.na(crickett$umpire2) & crickett$di_type_of_match=="International",], ~ umpire, summarise, win_game = mean(wingame==1), lose_game = mean(wintoss==0), n = length(unique(uniqueid)),  se_win=2*sqrt(win_game*(1-win_game)/n))
 

### 2016-03-04 write_up/cricket.tex
close

@@
 For it may be impossible to predict well in advance the advantage of bowling or batting first.
-Or, it may be that teams squander the potential advantage by using bad heuristics to choose what they do.
+(If pre-match assessments of the pitch by media commentators are anything to go by, error in assessment of conditions is likely large.) Or, it may be that teams squander the potential advantage by using bad heuristics to choose what they do.
+For instance, teams may weigh outcomes from recent matches `too' heavily; e.g., a team that has, of late, won chasing may choose to chase even though pitch conditions favor batting first.
 
-To assess the net observed advantage of winning the toss, we exploit data from a novel dataset of over 43,000 first-class men's cricket matches --- to our knowledge, the largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}.
+To assess the net observed advantage of winning the toss, we exploit data from a novel dataset of over 43,000 first-class men's cricket matches--- to our knowledge, the largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}.
 In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into.
@@
 \section*{Data}
-Data are from 43,185 first-class men's cricket matches.
-It is a near census of the relevant population.
-We have data on all types of matches: domestic and international Twenty20s --- T20s and T20Is respectively, domestic and international one-dayers --- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches --- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches.
+Data are from 44,224 first-class cricket matches.
+It is a near census of the relevant population.\footnote{Data excludes scheduled matches that were abandoned without the toss being conducted.} We have data on all types of matches: domestic and international Twenty20s --- T20s and T20Is respectively, domestic and international one-dayers --- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches --- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches.
 In English county cricket, first-class matches last four days.
@@
 
-Of these matches, we do not have data on the toss for 2,807, or roughly 7\%, of the matches.
-The primary reason we don't have data on these matches is because the match was abandoned without play.
-We exclude data from these matches.
+Of the 44,224 matches, 1,019 matches were abandoned without play.
+We exclude these matches.
+In another 1,376 matches, we do not have information on whether the team chose to bat or bowl after the toss.
+Informal inspection suggests that data are missing because no match was played.
+We excludes these matches as well.
 
@@
 In a one-day match, for instance, each side must bat at least 20 overs for a result to be declared.
-In 706 matches, or roughly 1.7\% of the remaining matches, not enough overs were bowled to get a result.
-We again exclude these matches from our analysis.
-This leaves us with 39,672 matches.
+In 769 matches, or roughly 1.7\% of the total matches, not enough overs were bowled to get a result.
+We exclude these matches from our analysis.
+This leaves us with data from 41,060 matches.
 We analyze these data.
@@
 Any decision made after the outcome of the toss is known, however, is `post-treatment.' In particular, the decision to bat or bowl first is made after accounting for the relative strengths and weaknesses vis-\`{a}-vis the competing team at that particular instance, and thus not independent of team attributes.
-Hence, conditioning on decision to bowl or bat first can induce bias in estimates of advantage of winning a toss.
+Hence, conditioning on decision to bowl or bat first can bias estimates of advantage of winning a toss.
 Thus, unlike \citet{dawson2009bat}, \citet{Saad2015}, we solely rely on the assumption that the outcome of a toss is random.
@@
 For this analysis, we only get to exploit international matches as establishing which of the teams is the home team in local matches is somewhat arduous.
-Of the 5,591 international matches, the home team won the toss in 2,860 matches, or about 51.15\%.
-The probability that we will get as large a proportion by chance after tossing 5,591 coins is about 8.7\%, low but not eyebrow-raisingly so.
-Thus, for now, we choose to trust that the tosses are random.
+Of the 5,684 international matches for which we can match the country of the ground to the country of one of the teams, the home team won the toss in 2,892 matches, or about 50.87\%.
+The chance of getting as many wins by fluke after tossing 5,684 coins is about 18.91\%.
+The chance is low, but not eyebrow raisingly so.
 
-A caveat about interpretation before we report the results.
+But, rather than consider all home matches, we may instead want to only consider matches that are officiated only by home umpires --- the norm till 1992.\footnote{For more information on move to neutral umpires, see \href{http://www.espncricinfo.com/magazine/content/story/511175.html}{Neutral Umpires} by S.
+Rajesh on ESPNcricinfo.} In matches featuring umpires from only the home country, the team with `home umpire' advantage won toss nearly 51.9\% of the times.
+And the chance of getting a greater percentage of wins than 51.9\% in 2,965 is a shade less than 4\%.
+Thus, there is some reason to worry that the tosses are rigged.
+Any such rigging would bias estimates of advantage of winning the coin toss to the extent that it is correlated with ability.
+More plainly, if stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards.
+And vice versa, if otherwise.
+We, however, do not have good reasons to think that there is a correlation.
+So for now, we proceed as if the tosses are random.
+
+Another caveat about interpretation before we present the results.
 As we discuss in the introduction, we cannot estimate the actual advantage of winning a toss.
@@
 
-The team that wins the toss wins the match 2.9\% more often than the team that loses the toss.
+The team that wins the toss wins the match 2.8\% more often than the team that loses the toss.
 This is a reasonable sized advantage in a competitive sport --- though likely much smaller than the number that most commentators carry in their heads.
-% Also interpret it by comparing to probability of winning when playing with the worst placed team (not defined from last team)
 This advantage, however, varies by format, by conditions, by whether or not a particular formula was used to adjust scores when it rains, and how much better the team that won the toss is vis-\`{a}-vis the competing team.

### 2016-03-04 scripts/05_cricket.R

@@ -32,21 +32,29 @@ sum(match$team2=='')
 "
 
 "
-Take out matches where there was no result. See for e.g.,
-http://www.espncricinfo.com/matches/engine/match/329869.html
-
-Takes out 780 matches
-"
-temp <- grepl("No result", match$outcome)
-cricket <- subset(match, !temp)
-
-"
 Take out matches where decision on who bowled/batted first is unknown. See this, for e.g.:
 http://www.espncricinfo.com/matches/engine/match/537589.html 
-Takes out 2384 matches
+
+Takes out abandoned matches
+# abandoned
+table(match$bat_or_bowl, grepl('abandon', match$outcome))
+# 1019
+
+
+Takes out 3164 matches
 "
 
-cricket <- subset(cricket, bat_or_bowl!="")
+cricket <- subset(match, bat_or_bowl!="")
+
+"
+Takes out matches where there was no result. See for e.g.,
+http://www.espncricinfo.com/matches/engine/match/329869.html
+"
+
+temp <- grepl('No result', cricket$outcome)
+cricket <- subset(cricket, !temp)
+
+# ---------------- 
 
 # Prop. who win toss, win game. (not including draws)
 table(cricket$win_toss_win_game, cricket$basic_type_of_match)
@@ -104,7 +112,7 @@ crickett$name[!is.na(crickett$signed_diff_ranks) & crickett$signed_diff_ranks > 
 "
 Analysis
 
-1. Do teams win more tosses at home?
+1a. Do teams win more tosses at home?
    Evidence from International Matches
 "
 # Proportion of tosses won in home country
@@ -112,7 +120,25 @@ Analysis
 
 homet <- with(crickett[!is.na(crickett$home_country),], xtabs( ~ home_country + wintoss))
 homet/rowSums(homet)
-binom.test(2892, 5774, p=.5)
+binom.test(2892, 5684, p=.5)
+
+"
+1b. Do team win more matches when matches officiated by home umpires?
+
+"
+# Split by Home Umpires on Winning Toss and Winning Match
+# It was tested out with 1 umpire beginning in 1992 and then made standard with 2 in 2002: 
+# http://www.espncricinfo.com/magazine/content/story/511175.html
+
+# Only international matches
[…]

### 2016-03-06 write_up/cricket.tex
cricket draft

@@
 \begin{document}
-\title{\vspace{-.5cm}\normalsize{Fairly Random: The Impact of Winning the Toss on the Probability of Winning\footnote{You can find the scripts used to scrape the data, the final data set, and scripts used for analysis at: \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}\vspace{.5cm}}}
+\title{\vspace{-.5cm}\normalsize{Fairly Random: The Impact of Winning the Toss on the Probability of Winning\footnote{The paper benefitted from comments by Andrew Gelman, Donald Green, and Daniel Stone.
+Scripts used to scrape the data, the final data set, and scripts used for analysis can be found at: \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.
+}\vspace{.5cm}}}
 \author{\normalsize{Gaurav Sood}\\\href{mailto:gsood07@gmail.com}{\small{gsood07@gmail.com}} \and \normalsize{Derek Willis}\\\href{mailto:dwillis@gmail.com}{\small{dwillis@gmail.com}}}
@@
 
-But, rather than consider all home matches, we may instead want to only consider matches that are officiated only by home umpires --- the norm till 1992.\footnote{For more information on move to neutral umpires, see \href{http://www.espncricinfo.com/magazine/content/story/511175.html}{Neutral Umpires} by S.
+However, rather than consider all home matches, we may instead want to only consider matches that are officiated only by home umpires---the norm till 1992.\footnote{For more information on move to neutral umpires, see \href{http://www.espncricinfo.com/magazine/content/story/511175.html}{Neutral Umpires} by S.
 Rajesh on ESPNcricinfo.} In matches featuring umpires from only the home country, the team with `home umpire' advantage won toss nearly 51.9\% of the times.
@@
 The pitch deteriorates far less over the course of the day, or in case of Twenty20s, a few hours.
-And indeed unlimited over matches provide the most consistent advantage --- the average advantage over FC and test matches is north of 2.7\% (see Figure~\ref{fig:type}).
+And indeed unlimited over matches provide the greatest advantage--- the average advantage over FC and test matches is north of 2.6\% (see Figure~\ref{fig:type}).
 Looked in relative terms, the advantage of winning the toss is also close to the greatest in multi-day affairs.
@@
 Thus, the advantage is closer to 4.5\%.
-
-The heftiest advantage, however, is in domestic one-day matches (List A), approximately 4.8\%.
-The number for ODIs, however, is a bit puzzling --- there is an advantage to winning the toss but it is really slight (.83\%).
-That can best be ascribed to choosing badly.
-In T20s and T20Is, the advantage is 1.3\% and 3.8\% respectively.
+The heftiest raw advantage, however, is in one-day matches (List A and ODIs), approximately 3.3\%.
+In T20s and T20Is, the advantage is considerably smaller, just about 1.27\%.
+And unlike the estimate of advantage for multi-day and one-day affairs, we cannot statistically reject that the possibility that there is no advantage.
 
@@
 \includegraphics[scale=.85]{../figs/winbyType.pdf}
-{\footnotesize Note: \emph{n} refers to the number of matches.\par}
+{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
 \end{minipage}
@@
 
-But type of matches cover only major source of variation and theorizing about the advantage granted by the toss.
+Type of matches are but one source of variation and theorizing about the advantage granted by the toss.
 It is often claimed that the toss is more crucial in day and night matches.
-Due to dew --- it is thought to make bowling hard, and the visibility of the white ball is thought to be lower under lights, which makes catching hard --- it is thought that the team that fields second is disadvantaged.
-It turns out that the conventional wisdom is largely vindicated, except perhaps for Twenty20 Internationals (but we have very little on Day and Night T20 Internationals to say anything with great certainty) (see Figure ~\ref{fig:dn}).
-In each, domestic one-day and Twenty20 matches, and one-day internationals, the advantage of winning the toss in Day and Night matches is heftier, and in the case of ODIs and domestic T20s, markedly so --- the difference in advantage of winning the toss between Day and Night and Day versions is over 9\% for both.
-Splitting by whether or not the match spans Day and Night also reveals that in day ODIs, the team that wins the toss loses more often than the team that wins it.
-We can  attribute that to choosing badly.
+Due to dew---it is thought to make bowling hard, and the visibility of the white ball is thought to be lower under lights, which makes catching hard---the team that fields second is thought to be at a disadvantage.
+The conventional wisdom is largely vindicated for one day affairs (see Figure ~\ref{fig:dn}).
+In one-day matches, the advantage of winning the toss in a day and night match is 5.92\%, whereas the advantage of winning the toss in a one-dayer played during the day is less than half---2.89\%.
+In Twenty20s---domestic and international---, however, we cannot distinguish between the advantage granted by the toss in day and night matches and day time affairs.
 
@@
 \includegraphics[scale=.85]{../figs/winbyDayNight.pdf}
+{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
 \label{fig:dn}
@@
 
-In limited over matches sometimes weather intervenes significantly after the match has already begun.
-After certain loss of time, the match is generally curtailed and the scores adjusted using a method invented by Duckworth and Lewis \citep[see][]{duckworth1998}.
-We can use the random nature of who wins the toss to see if winning percentages are strongly conditioned by matches that use Duckworth-Lewis and matches that don't.
-Again the conventional wisdom suggests that the team that wins the toss is at an advantage in such matches.
-And again, the conventional wisdom is vindicated.
-In List A, ODI and domestic Twenty20 matches, the advantage of winning the toss is greater when Duckworth Lewis is used to adjust the scores than when it isn't --- the difference in advantage ranges from about .75\% to over 2\% (see Figure ~\ref{fig:dl}).
+Weather has a large impact on the playing conditions in cricket.
+For instance, cool overcast weather is thought to aid swing bowling, especially on certain pitches.
+More generally, the advantage of winning the toss likely varies by weather.
+However, we do not have data on weather.
+But, we can proxy it with seasons.
+In particular, students of the game suspect that the advantage of winning the toss in early English season is especially great.
+We next assess whether that is so.
+
+There is some evidence of a seasonal pattern, with advantage of winning the toss somewhat greater in spring and early summer (May and June) than in mid and late summer and early fall (July to September) ~\ref{fig:season}.
+However, the thing that catches attention is the large disadvantage of winning the toss in April.
+We don't have a good explanation for the pattern, except for teams choosing badly.
+
+\begin{figure}[htbp]
+\centering
+\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Month in England}
+\includegraphics[scale=.85]{../figs/winbyMonthEngland.pdf}
+{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
+\label{fig:season}
+\end{figure}
+
+Aside from affecting the playing conditions, weather affects cricket matches in other, more forceful ways---interrupting, and sometimes ending matches.
+When a limited over match that is already underway is interrupted by bad weather, and more than a certain amount of time is lost due to the interruption, the match is curtailed and the total that the team batting second must achieve to win is adjusted using a method invented by Duckworth and Lewis \citep[see][]{duckworth1998}.\footnote{Before the Duckworth and Lewis method was adopted, weather affected matches sometimes continued on the next day; an extra day was deliberately left in the schedule for dealing with such eventualities.
+In cases where the spare day proved inadequate, the match was declared a draw.}
+
+We can use the random nature of who wins the toss to see if winning percentages of the teams that win the toss are strongly conditioned by whether or not Duckworth-Lewis is used.
+If the advantage of winning the toss in matches using Duckworth-Lewis is different from matches that don't use it, it suggests that the Duckworth-Lewis method is biased.
+(For a precise estimate of the bias, ideally, we would want to compare matches using Duckworth-Lewis method with matches held in similar conditions.)
+
+In both one-day and Twenty20 matches, the advantage of winning the toss in a match where the target is adjusted using Duckworth-Lewis, is considerably greater (see Figure ~\ref{fig:dl}).
+In one-day matches, the advantage is 5.35\% in matched adjudicated by Duckworth-Lewis and 3.17\% in matches that don't use it.
+Statistically, the chances that the two numbers are the same is less than 10\%.
+(And if you make the plausible assumption that winning a toss only improves the chances of winning, the chance that the two numbers are the same is half that.) In Twenty 20s, the advantage of winning the toss shrinks from 3.9\% in matches using Duckworth-Lewis to 1.17\% in matches without it.
+Once again, the chance that the two numbers are the same is about 10\%.
 
@@
 \includegraphics[scale=.85]{../figs/winbyDL.pdf}
+{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
 \label{fig:dl}
@@
 To study this, we collected data on team quality.
-The ICC publishes team ratings for international test and one-day teams each month.\footnote{For details about how the ICC produces these ratings, see \href{http://icc-live.s3.amazonaws.com/cms/media/about_docs/536b1a48c16e5-Reliance\%20ICC\%20ODI\%20Team\%20Rankings\%20FAQs\%202014.pdf}{ICC Rating FAQs}.} Ratings of the ODI teams have been published since 1981, and of test teams, since 1952.
-Of the entire dataset that spans 1981-today and 1952-today, we only have we have data till 2013.\footnote{The format in which the ICC publishes the ratings changed in 2013.
+The ICC publishes team ratings for international test and one-day teams each month.\footnote{For details about how the ICC produces these ratings, see \href{http://icc-live.s3.amazonaws.com/cms/media/about_docs/536b1a48c16e5-Reliance\%20ICC\%20ODI\%20Team\%20Rankings\%20FAQs\%202014.pdf}{ICC Rating FAQs}.} Ratings of the men's ODI teams have been published since 1981, and of the test teams, since 1952.
+Of the entire ranking dataset, that spans 1981--today and 1952--today for ODI and test teams respectively, we only have we have data till 2013.\footnote{The format in which the ICC publishes the ratings changed in 2013.
 And scraping the latter data posed additional hurdles.
@@
 And Australia in 2007 twice held a rating of 143.
-Both seem reasonable.
 We measure how closely matched the teams by differencing the ranking points of one team from the other.
 Commercial considerations mean that a majority of the games are played among highly ranked and closely matched teams.
-So, we limit our analysis to matches where teams ratings differ by no more than 30 points.
+Thus, the precision of our estimates is greatest for matches between closely matched teams.
 
 The results are expected, but new.
-As Figure ~\ref{fig:ranks}, there is a sharp curve around 0.
-When closely matched teams win, the team that wins the toss has a consistent advantage over the team that loses the toss.
+As Figure ~\ref{fig:ranks} --- which plots percentage of matches won or drawn by the team that won the toss--- illustrates, there is a sharp curve around 0.
+When closely matched teams win, winning the toss has a large impact on the probability of winning.
 
@@
 \includegraphics[width=1\textwidth]{../figs/winbyRank.pdf}
+{\footnotesize Note: Smoothed relationship between difference in ranks and winning probability by whether or not the team won the toss.\par}
 \label{fig:ranks}
@@
 
-How the advantage vary by country?
+Lastly, we investigate how advantage varies by country in international matches?
 Are some countries better than others?
 We investigated the question by tallying the advantage by team.
-Sri Lanka and India do especially well (see ~\ref{fig:country}), coming in at 4.53\% and 3.35\% respectively.
-Australia and West Indies hover around 1.3--1.5\% and England at 2.33\%.
-New Zealand, however, does worse when it wins the toss than when it loses it --- difference is about 1.21\%.
+All of our estimates are imprecise enough that we cannot discount the possibility that the true effect is zero.
+The most puzzling result is from New Zealand, which seems to do more poorly in matches where it wins tosses than where it loses it.
+On the other hand, Sri Lanka and India do especially well (see ~\ref{fig:country}), coming in at 3.85\% and 3.09\% respectively.
+Australia, Pakistan, and West Indies hover around 1--1.5\% and England is at 2.29\%.
 
@@
 \includegraphics[scale=.85]{../figs/winbyCountry.pdf}
+{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
 \label{fig:country}
@@
 This covers some of the more straightforward questions.
-Other more nuanced but commonly thought questions remain.
-For one, students of the game suspect that the advantage of winning a toss in early English season is especially great.
-We next check that using data on International matches from England.
-
-\begin{figure}[htbp]
-\centering
-\caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Month in England}
-\includegraphics[scale=.85]{../figs/winbyMonthEngland.pdf}
-\label{fig:season}
-\end{figure}
-
-This isn't a comprehensive inventory of analyses that one could do on the topic.
-But neither is it meant to be.
-Some simple questions still remain unanswered.
+However, some simple questions still remain unanswered.
 For one, how does the toss advantage vary over time?
-Have teams become smarter?
+For another, how does winning the toss affect the margin of victory?
 We propose to investigate these questions in a follow-up piece.

### 2016-03-06 scripts/05_cricket.R

@@ -132,6 +132,7 @@ binom.test(2892, 5684, p=.5)
 
 # Only international matches
 with(crickett[!is.na(crickett$umpire1) & !is.na(crickett$umpire2) & crickett$di_type_of_match=="International",], xtabs( ~ umpire + wintoss))
+# with(crickett[!is.na(crickett$umpire1) & !is.na(crickett$umpire2) & crickett$di_type_of_match=="International",], xtabs( ~ umpire + wintoss + country))
 with(crickett[!is.na(crickett$umpire1) & !is.na(crickett$umpire2) & crickett$di_type_of_match=="International",], xtabs( ~ umpire + wingame))
 
 # By Type of Match
@@ -196,7 +197,6 @@ theme_base <- theme(panel.grid.major.y = element_line(colour = "#e3e3e3", linety
       strip.text.x       = element_text(size=9),
       legend.text        = element_text(size=8),
       plot.margin        = unit(c(0,.5,.5,.5), "cm"))
-
 
 # Bootstrap s.e.
 
@@ -334,14 +334,16 @@ crickett$name[!is.na(crickett$signed_diff_ranks) & crickett$signed_diff_ranks < 
 crickett$name[!is.na(crickett$signed_diff_ranks) & crickett$signed_diff_ranks > 110]
 "
 
-rankcricket <- subset(crickett, !is.na(signed_diff_ranks))
+rankcricket <- subset(crickett, !is.na(signed_diff_ranks) & wintoss==1)
 rankcricket$wingamer <- as.numeric(rankcricket$wingame)*100
 
-ggplot(rankcricket, aes(x=signed_diff_ranks, y=wingamer, colour=factor(wintoss))) + 
-geom_smooth(method="loess", span=.80, se=F, size=.4) + 
+library(lme4)
+summary(with(rankcricket, lm(wingamer ~ zero1(signed_diff_ranks))))
+
+ggplot(rankcricket, aes(x=signed_diff_ranks, y=wingamer)) + 
+geom_smooth(size=.4, col="#2b8cbe") + 
 geom_vline(xintercept=0, col="#333333", linetype="dashed", alpha=.3, size=.1) +
-scale_x_continuous(breaks=seq(-30, 30, 10), labels=nolead0s(seq(-30, 30, 10)), limits=c(-30, 30), name="Difference in Ranking Points") +
-scale_colour_manual(values = c("#2b8cbe", "#31a354"), labels=c("Lose Toss", "Win Toss")) +
[…]

### 2016-03-06 write_up/cricket.tex
figure formatting

@@
 \caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Type of Match.}
-\begin{minipage}{0.85\textwidth}
 \centering
 \includegraphics[scale=.85]{../figs/winbyType.pdf}
-{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
-\end{minipage}
+{\footnotesize \\ Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
 \label{fig:type}
@@
 \caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Day or Day and Night}
-\includegraphics[scale=.85]{../figs/winbyDayNight.pdf}
-{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
+\includegraphics[scale=.95]{../figs/winbyDayNight.pdf}
+{\footnotesize \\ Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
 \label{fig:dn}
@@
 \caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Month in England}
-\includegraphics[scale=.85]{../figs/winbyMonthEngland.pdf}
-{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
+\includegraphics[scale=1]{../figs/winbyMonthEngland.pdf}
+{\footnotesize \\ Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
 \label{fig:season}
@@
 \caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Whether or not Duckworth-Lewis was invoked}
-\includegraphics[scale=.85]{../figs/winbyDL.pdf}
-{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
+\includegraphics[scale=1]{../figs/winbyDL.pdf}
+{\footnotesize \\ Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
 \label{fig:dl}
@@
 \caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Country}
-\includegraphics[scale=.85]{../figs/winbyCountry.pdf}
-{\footnotesize Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
+\includegraphics[scale=1]{../figs/winbyCountry.pdf}
+{\footnotesize \\ Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
 \label{fig:country}

### 2016-03-07 write_up/cricket.tex
margin of victory

@@
 Data are from 44,224 first-class cricket matches.
-It is a near census of the relevant population.\footnote{Data excludes scheduled matches that were abandoned without the toss being conducted.} We have data on all types of matches: domestic and international Twenty20s --- T20s and T20Is respectively, domestic and international one-dayers --- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches --- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches.
+It is a near census of the relevant population.\footnote{Data excludes scheduled matches that were abandoned without the toss being conducted.} We have data on all types of matches: domestic and international Twenty20s---T20s and T20Is respectively, domestic and international one-dayers---List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches --- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches.
 In English county cricket, first-class matches last four days.

### 2016-03-07 scripts/04_merge_ranking_grounds_data.R

@@ -177,3 +177,12 @@ match$team2_tv_umpire[match$tv_umpire_country==""] <- NA
 
 match$team1_umpire <- rowSums(cbind(match$team1_umpire1, match$team1_umpire2, match$team1_tv_umpire), na.rm=T)
 match$team2_umpire <- rowSums(cbind(match$team2_umpire1, match$team2_umpire2, match$team2_tv_umpire), na.rm=T)
+
+# Margin of victory
+# number of runs, balls, wickets, innings
+split_by <- sapply(strsplit(match$outcome, " by "), "[", 2)
+
+match$wickets  <- as.numeric(sapply(strsplit(split_by, " wicket"), "[", 1))
+match$runs     <- as.numeric(sapply(strsplit(split_by, " runs"), "[", 1))
+match$balls    <- as.numeric(sapply(strsplit(sapply(strsplit(split_by, "with "), "[", 2), " ball"), "[", 1))
+match$innings  <- grepl("inning", split_by)

### 2016-03-07 scripts/05_cricket.R

@@ -65,9 +65,9 @@ Two rows per match
 "
 
 # Match level vars: 
-match_cols <- c("url", "match_id", "uniqueid", "date", "month", "day", "year", "rain", 
+match_cols <- c("url", "match_id", "uniqueid", "date", "month", "day", "year", "rain",  "diff_ranks", "bat_or_bowl", 
                  "type_of_match", "basic_type_of_match", "type_of_match2", "di_type_of_match", "men_type_of_match", "day_n_night", "youth", "women", "youth.women", "unofficial", "duckworth_lewis", 
-                 "win_toss", "win_game", "home_wins_toss", "draw", "outcome", "diff_ranks", "bat_or_bowl", 
+                 "win_toss", "win_game", "home_wins_toss", "draw", "outcome", "wickets", "runs", "balls", "innings", 
                  "ground", "ground_id", "country", "continent", "latitude", "longitude")
 
 # Team cols, rename for gather/separate to work well
@@ -424,3 +424,12 @@ Or it could be that teams develop better strategies to offset toss advantage.
 If you are going to come up short half the times, you develop strategies to counter that.
 But lots of moving parts for over time stuff. For instance, format is a concern. Certain formats not around earlier.
 " 
+small_set <- subset(crickett, name %in% c("India", "Australia", "West Indies", "England", "New Zealand", "Pakistan", "Sri Lanka"))
+ddply(crickett, ~ year + name, summarise, diff = mean(wingame[wintoss==1]) - mean(wingame[wintoss==0]), count=length(unique(url)))
+
+# Margin of victory
+ddply(crickett[crickett$wingame==1,], ~ wintoss + basic_type_of_match, summarise, wickets = mean(wickets, na.rm=T))
+ddply(crickett[crickett$wingame==1,], ~ wintoss + basic_type_of_match, summarise, runs = mean(runs, na.rm=T))
+ddply(crickett[crickett$basic_type_of_match!="FC/TEST" & crickett$wingame==1,], ~ wintoss + basic_type_of_match, summarise, balls = mean(balls, na.rm=T))
+ddply(crickett[crickett$basic_type_of_match=="FC/TEST" & crickett$wingame==1,], ~ wintoss, summarise, innings = mean(innings, na.rm=T))
+

### 2016-03-07 write_up/cricket.tex
updated ms. w/ margin + split by international/domestic

@@
 The heftiest raw advantage, however, is in one-day matches (List A and ODIs), approximately 3.3\%.
-In T20s and T20Is, the advantage is considerably smaller, just about 1.27\%.
-And unlike the estimate of advantage for multi-day and one-day affairs, we cannot statistically reject that the possibility that there is no advantage.
+In T20s and T20Is, the advantage is considerably smaller, just about 1.27\%.\footnote{Splitting data by whether the match was domestic or international yields some additional insights.
+Like \citet{de1998winning}, who based on analysis of data from 427 international one-day matches conclude that `winning the toss at the outset of a match provides no competitive advantage' in one-day international matches, we find that in ODIs teams that win the toss win games at about the same rate as those that lose the toss.
+For first-class and test matches, the advantage to winning the toss is roughly the same.
+Meanwhile in Twenty20s, the advantage of winning the toss is greater in international than domestic matches.} And unlike the estimate of advantage for multi-day and one-day affairs, we cannot statistically reject that the possibility that there is no advantage.
 
@@
 Similarly, it is unlikely that winning the toss would change who wins the game when two ill-matched teams are playing.
-To study this, we collected data on team quality.
+To study the issue, we collected data on team quality.
 The ICC publishes team ratings for international test and one-day teams each month.\footnote{For details about how the ICC produces these ratings, see \href{http://icc-live.s3.amazonaws.com/cms/media/about_docs/536b1a48c16e5-Reliance\%20ICC\%20ODI\%20Team\%20Rankings\%20FAQs\%202014.pdf}{ICC Rating FAQs}.} Ratings of the men's ODI teams have been published since 1981, and of the test teams, since 1952.
@@
 
-Lastly, we investigate how advantage varies by country in international matches?
-Are some countries better than others?
-We investigated the question by tallying the advantage by team.
-All of our estimates are imprecise enough that we cannot discount the possibility that the true effect is zero.
-The most puzzling result is from New Zealand, which seems to do more poorly in matches where it wins tosses than where it loses it.
-On the other hand, Sri Lanka and India do especially well (see ~\ref{fig:country}), coming in at 3.85\% and 3.09\% respectively.
-Australia, Pakistan, and West Indies hover around 1--1.5\% and England is at 2.29\%.
+Lastly, we investigate how the advantage varies by country winning the toss in international matches.
+Are some countries better than others at capitalizing on a toss win?
+We investigated the question by tallying the advantage by team that wins the toss.
+As ~\ref{fig:country} illustrates, all of our estimates are imprecise enough that we cannot say with confidence that any of the teams capitalizes on winning the toss.
+Neither can we discount the possibility that the actual differences across the teams are zero.
+The most puzzling result is from New Zealand.
+Like \citet{Saad2015}, data suggest that New Zealand does more poorly in matches where it wins tosses than where it loses it.
+On the other hand, Sri Lanka and India appear to do especially well, winning 3.85\% and 3.09\% additional matches, respectively, when they win the toss.
+Australia, Pakistan, and West Indies hover around 1--1.5\%, and England is at 2.29\%.
 
@@
 
-This covers some of the more straightforward questions.
-However, some simple questions still remain unanswered.
-For one, how does the toss advantage vary over time?
-For another, how does winning the toss affect the margin of victory?
-We propose to investigate these questions in a follow-up piece.
+Till now, we have focused on assessing the impact of winning the toss on the probability of winning, and how the impact is conditioned by playing conditions, by the type of match, and by the teams involved.
+Winning the toss, however, likely not only affects the probability of winning, but also the margin of victory.
+But before we assess the impact of winning the toss on the margin of victory, a short primer.
+
+In limited over matches, when the team chasing the total falls short, the margin of victory is given in difference in runs.
+When the team is able to successfully chase the total, the margin of victory is the given by two numbers: number of balls remaining, and the number of wickets in hand.
+In unlimited overs matches, the metrics for margin of victory differ in two small ways.
+We don't tally the number of balls remaining when the winning team achieves the target (principally we could).
+Instead, we note whether or not the winning team had to bat twice---whether or not the team won by an `innings' and additional runs.
+
+Teams that win the toss and the match in first-class and test matches win with more wickets in hand (Mean $= 6.91$) than winning teams that lose the toss (Mean $= 6.64$).
+In one-dayers and Twenty20s, the teams that win the toss and the match have about the same number of wickets in hand as the teams that lose the toss but win the match (One-Day: Mean$_{\text{Lose Toss}} = 5.57$, Mean$_{\text{Win Toss}} = 5.65$; Twenty20: Mean$_{\text{Lose Toss}} = 6.37$; Mean$_{\text{Win Toss}} = 6.21$).
+
+In first-class and test matches teams that win the toss and the match also win by few more runs on average (Mean $= 136.48$; Median $=124$) than teams that lose the toss but win the match (Mean $= 133.71$; Median $=122$).
+Similarly, in Twenty20s, the team that wins the toss wins by a few more runs (Mean $= 37.60$; Median $=28$) than team that loses it (Mean $= 34.50$; Median $=26$).
+In one-dayers, however, the margin of victory is largely indistinguishable across cases where the winning team wins the toss and where it loses it (Mean$_{\text{Lose Toss}} = 63.95$, Median$_{\text{Lose Toss}} = 51$; Mean$_{\text{Win Toss}} = 63.15$, Median$_{\text{Win Toss}} = 51$).
+
+Similar patterns hold for balls remaining---teams that win the toss generally win with a few more balls remaining than teams that lose the toss.
+In one-day matches, Mean$_{\text{Lose Toss}} = 49.87$, Median$_{\text{Lose Toss}} = 28$, Mean$_{\text{Win Toss}} = 53.27$, and Median$_{\text{Win Toss}} = 30$.
+And in Twenty20s, Mean$_{\text{Lose Toss}} = 18.69$, Median$_{\text{Lose Toss}} = 12$, Mean$_{\text{Win Toss}} = 16.41$, and Median$_{\text{Win Toss}} = 10$.
+
+And lastly, on number of innings, the teams that win the toss win by an innings about as often as teams that lose the toss (Mean$_{\text{Lose Toss}} = 22.26\%$, Mean$_{\text{Win Toss}} = 22.43\%$).
+
+\section*{Discussion}
+
+The data suggest that winning the toss has a sizable impact on the probability of winning, especially in closely contested games.
+The data also suggest that the advantage varies considerably and systematically in expected ways, with advantage being greater in  day and night matches, matches in which Duckworth-Lewis is used to adjust scores, and where the match was played between closely matched teams.
 

### 2016-03-07 scripts/05_cricket.R

@@ -419,6 +419,14 @@ annotate("text",
 ggsave("figs/winbyCountry.pdf", width=6)
 
 "
+Domestic Verus International
+Presumably diff. smaller in international
+"
+ddply(crickett, ~di_type_of_match + wintoss, summarise, mean=mean(wingame))
+ddply(crickett, ~basic_type_of_match + di_type_of_match + wintoss, summarise, mean=mean(wingame))
+
+
+"
 Is there over time learning? If so, toss adv. would increase. 
 Or it could be that teams develop better strategies to offset toss advantage. 
 If you are going to come up short half the times, you develop strategies to counter that.
@@ -432,4 +440,3 @@ ddply(crickett[crickett$wingame==1,], ~ wintoss + basic_type_of_match, summarise
 ddply(crickett[crickett$wingame==1,], ~ wintoss + basic_type_of_match, summarise, runs = mean(runs, na.rm=T))
 ddply(crickett[crickett$basic_type_of_match!="FC/TEST" & crickett$wingame==1,], ~ wintoss + basic_type_of_match, summarise, balls = mean(balls, na.rm=T))
 ddply(crickett[crickett$basic_type_of_match=="FC/TEST" & crickett$wingame==1,], ~ wintoss, summarise, innings = mean(innings, na.rm=T))
-

### 2016-03-07 write_up/cricket.tex
added an abstract + abstract v. of conditioning factors

@@
 
+\begin{center}\textbf{Abstract}\\\end{center}
+In a competitive sport, every little thing matters.
+Yet many sports leave some large levers out of the reach of the teams, in the hands of fate.
+In cricket, world's second most popular sport by some measures \footnote{See \citet{Economist2011} in The Economist.}, one such lever---the toss---has been subject to much recent attention.
+Using a large novel dataset of 44,224 cricket matches, we estimate the impact of winning the toss on the probability of winning.
+The data suggest that winning the toss increases the chance of winning by a small ($\sim$ 2.8\%) but significant margin.
+The advantage varies heftily and systematically, by how closely matched the competing teams are, and by playing conditions---tautologically, winning the toss in conditions where the toss grants a greater advantage, for e.g., in day and night matches, has a larger impact on the probability of winning.
+\clearpage
+
 In nearly all cricket matches, it is claimed that there is a clear advantage to either bowling or batting first.
@@
 We find that winning the toss grants a small but significant advantage, but that advantage varies considerably and systematically.
+We next assess whether the advantage of winning the toss varies, broadly speaking, by how closely matched the teams are, and by how large an advantage winning the toss grants---the advantage of winning the toss is greater in certain playing conditions than others.
+We find that the advantage of winning the toss varies widely and systematically, in expected ways.
 
@@
 The data suggest that winning the toss has a sizable impact on the probability of winning, especially in closely contested games.
-The data also suggest that the advantage varies considerably and systematically in expected ways, with advantage being greater in  day and night matches, matches in which Duckworth-Lewis is used to adjust scores, and where the match was played between closely matched teams.
+The data also suggest that the advantage varies considerably and systematically ---in expected ways---, with advantage greater in day and night matches, matches in which Duckworth-Lewis is used to adjust scores, and where the match is played between closely matched teams.
+In showing so, the data lend credence to, and quantify, the suspicion that many of the cricket fans have long had---that tosses matter.
+Besides that, the data also help quantify the bias in Duckworth-Lewis method.
+More generally, the analysis we do here could be replicated elsewhere to assess bias in competing methods, and used to prove that a particular method is better or worse than the Duckworth-Lewis.
 

### 2016-03-07 write_up/cricket.tex
more regular format for author affil.

@@
 }\vspace{.5cm}}}
-\author{\normalsize{Gaurav Sood}\\\href{mailto:gsood07@gmail.com}{\small{gsood07@gmail.com}} \and \normalsize{Derek Willis}\\\href{mailto:dwillis@gmail.com}{\small{dwillis@gmail.com}}}
+\author{\normalsize{Gaurav Sood}\footnote{Gaurav is an independent researcher.
+He can be reached at: \href{mailto:gsood07@gmail.com}{\small{gsood07@gmail.com}}} \and \normalsize{Derek Willis}\footnote{Derek is a  news applications developer at ProPublica.
+He can be reached at: \href{mailto:dwillis@gmail.com}{\small{dwillis@gmail.com}}}}
 \date{\vspace{.5cm}\normalsize{\today}}

### 2016-03-07 write_up/cricket.tex
nip/tuck

@@
 In a competitive sport, every little thing matters.
-Yet many sports leave some large levers out of the reach of the teams, in the hands of fate.
-In cricket, world's second most popular sport by some measures \footnote{See \citet{Economist2011} in The Economist.}, one such lever---the toss---has been subject to much recent attention.
+Yet, many sports leave some large levers out of the reach of the teams, and in the hands of fate.
+In cricket, world's second most popular sport by some measures, \footnote{See \citet{Economist2011} in The Economist.} one such lever---the toss---has been subject to much recent attention.
 Using a large novel dataset of 44,224 cricket matches, we estimate the impact of winning the toss on the probability of winning.

### 2016-05-27 write_up/arxiv/arxiv_cricket.tex
arxiv submission

--- 2016-03-07 write_up/cricket.tex
@@
 
-\usepackage{natbib}
+\usepackage[authoryear]{natbib}
 \usepackage{url}
@@
 setwd(paste0(githubdir, "/cricket-stats/write_up/"))	
-tools::texi2dvi("cricket.tex", pdf=TRUE, clean=TRUE)	
+tools::texi2dvi("cricket.tex", pdf=TRUE)	
 setwd(paste0(githubdir, "/cricket-stats/"))
@@
 \title{\vspace{-.5cm}\normalsize{Fairly Random: The Impact of Winning the Toss on the Probability of Winning\footnote{The paper benefitted from comments by Andrew Gelman, Donald Green, and Daniel Stone.
-Scripts used to scrape the data, the final data set, and scripts used for analysis can be found at: \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.
+Scripts used to scrape the data, the final data set, and scripts used for analysis can be found at: \href{https://github.com/dwillis/toss-up}{https://github.com/dwillis/toss-up}.
 }\vspace{.5cm}}}
@@
 \maketitle
-\doublespacing
 
@@
 \centering
-\includegraphics[scale=.85]{../figs/winbyType.pdf}
+\includegraphics[scale=1]{winbyType.pdf}
 {\footnotesize \\ Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
@@
 \caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Day or Day and Night}
-\includegraphics[scale=.95]{../figs/winbyDayNight.pdf}
+\includegraphics[scale=.95]{winbyDayNight.pdf}
 {\footnotesize \\ Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
@@
 \caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Month in England}
-\includegraphics[scale=1]{../figs/winbyMonthEngland.pdf}
+\includegraphics[scale=1]{winbyMonthEngland.pdf}
 {\footnotesize \\ Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
@@
 \caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Whether or not Duckworth-Lewis was invoked}
-\includegraphics[scale=1]{../figs/winbyDL.pdf}
+\includegraphics[scale=1]{winbyDL.pdf}
 {\footnotesize \\ Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}
@@
 \caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Difference in Ranks}
-\includegraphics[width=1\textwidth]{../figs/winbyRank.pdf}
+\includegraphics[width=1\textwidth]{winbyRank.pdf}
 {\footnotesize Note: Smoothed relationship between difference in ranks and winning probability by whether or not the team won the toss.\par}
@@
 \caption{Difference in Winning Percentages of Teams that Won the Toss and Teams that Lost the Toss by Country}
-\includegraphics[scale=1]{../figs/winbyCountry.pdf}
+\includegraphics[scale=1]{winbyCountry.pdf}
 {\footnotesize \\ Note: Means and 95\% confidence intervals. \emph{n} refers to the number of matches.\par}

### 2020-09-18 scripts/04_prep.R
"added lal scripts + export of tables"

@@ -1,0 +1,318 @@
+# %%
+###################################################
+rm(list = ls())
+library(LalRUtils)
+LalRUtils::libreq(tidyverse, stargazer2, lfe, rio,
+  foreach, magrittr, data.table, janitor) #,  lib2)
+theme_set(lal_plot_theme())
+options(repr.plot.width=12, repr.plot.height=9)
+
+# %% [markdown]
+# # Data Cleaning
+# run this whole section once, then start with [Empirics](#Empirics)
+
+# %%
+# setwd(root)
+root = '/home/alal/Dropbox/1_Research/cricket/'
+datadir = file.path(root, "repos/cricket-stats/data")
+
+# %% [markdown]
+# ## Ingest
+
+# %%
+odi_ranks  <- fread(file.path(datadir,"rankings_odi.csv"))
+test_ranks <- fread(file.path(datadir,"rankings_test.csv"))
+ranks <- rbind(odi_ranks, test_ranks)
+ranks[, format := ifelse(format=="odi", "ODI", "TEST")]
+
+# Ground
+grounds <- fread(file.path(datadir, "grounds.csv"))
+
+# Load match data
+match 		<- fread(file.path(datadir, "final_output.csv"))
+# %% drop matches with no toss info - mainly abandoned matches, with some weird county matches thrown in
+match[bat_or_bowl %in% c(0, NA), .N, by = outcome] [order(-N)]
+match = match[!(bat_or_bowl %in% c(0, NA))]
+
+# %% date processing
+# Match dates - need only month and year
+# Clean a bit
+match[, date2 := str_trim(date)]
+setorder(match, -date)
+# Month
+# Matches span single, multiple days
+# 18th and 19th century dates in yyyy-mm-dd format
+# No information about century for other dates
+# Let us standardize the format:
+temp_date <- match$date2
+fix_date <- function (x) {
+  s <- as.data.frame(do.call(rbind, strsplit(x, "-")))
+  apply(s[,c(2,3,1)], 1, paste, collapse="/")
+}
+
+match$date2 <- ifelse(grepl("-", temp_date), fix_date(temp_date), temp_date)
+temp <- as.data.frame(do.call(rbind, strsplit(match$date2, "/")))
+
+temp %>% head
+
+# %%
+match$month <- as.numeric(temp[,1])
+match$day   <- as.numeric(temp[,2])
+match$year  <- ifelse(as.numeric(temp[,3]) < 1700,
[…]

### 2020-09-18 scripts/05_empirics.R

@@ -1,0 +1,124 @@
+rm(list = ls())
+library(LalRUtils)
+LalRUtils::libreq(tidyverse, stargazer2, lfe, rio,
+  foreach, magrittr, data.table) #,  lib2)
+theme_set(lal_plot_theme())
+options(repr.plot.width=12, repr.plot.height=9)
+LalRUtils::libreq(tidyverse, stargazer2, lfe, rio, janitor, tictoc, broom,
+  foreach, magrittr, data.table, ggstatsplot)
+
+# %%
+
+root = '/home/alal/Dropbox/1_Research/cricket/'
+datadir = file.path(root, "repos/cricket-stats/data")
+
+##        #######  ##    ##  ######
+##       ##     ## ###   ## ##    ##
+##       ##     ## ####  ## ##
+##       ##     ## ## ## ## ##   ####
+##       ##     ## ##  #### ##    ##
+##       ##     ## ##   ### ##    ##
+########  #######  ##    ##  ######
+
+
+# %% long data
+crickett = fread(file.path(root, "tmp/matches_long.csv"))
+setorder(crickett, -year, -month, -date, mid)
+crickett %>% glimpse
+# check that match id is unique
+crickett[, .N, mid][order(-N)][1]
+# %% # ## Subsample (for now)
+crickett$type_of_match %>% tabyl
+regsamp = crickett[type_of_match  %in%
+  c("ODI", "T20", "T20I", "Twenty20", "TEST", "FC", "LISTA")]
+regsamp %>% dim
+
+# %% throw out matches with no results / cancelled / walkover
+regsamp[outcome %like% 'abandoned' | outcome %like% 'cancelled', .N]
+regsamp = regsamp[!(outcome %like% 'abandoned' | outcome %like% 'cancelled')]
+
+regsamp[outcome %like% 'walkover' , .N]
+regsamp = regsamp[!(outcome %like% 'walkover')]
+
+# %% bat first checks
+regsamp$bat_first %>% tabyl
+# win toss checks
+regsamp$wintoss %>% tabyl
+
+# %% final batting order checks
+regsamp[, bat_first_min := min(bat_first), by = mid]
+regsamp$bat_first_min %>% tabyl
+# these have no toss info
+regsamp[bat_first_min == 1,
+  .(mid, outcome, wintoss, wingame, bat_first, bat_first_min)]
+# drop them
+regsamp = regsamp[!(bat_first_min == 1)]
+
+# %% check coding of draws
+regsamp %>% glimpse
[…]

### 2020-11-06 scripts/99_eda.R
"results for 06-11 meeting"

@@ -1,0 +1,51 @@
+# %% ####################################################
+rm(list = ls())
+library(LalRUtils)
+LalRUtils::libreq(tidyverse, data.table, stargazer2, patchwork,
+                  lfe, rio, foreach, magrittr, janitor)
+theme_set(lal_plot_theme())
+options(repr.plot.width=12, repr.plot.height=9)
+set.seed(42)
+# %% ####################################################
+root = "/home/alal/Dropbox/1_Research/cricket"
+
+root = '/home/alal/Dropbox/1_Research/cricket/'
+datadir = file.path(root, "repos/cricket-stats/data")
+crickett = fread(file.path(root, "tmp/matches_long.csv"))
+
+# %%
+
+
+# %%
+df  = import(file.path(root, "tmp/overs_odi/overs/overs_1000891.json")) %>% setDT
+df1 = import(file.path(root, "tmp/overs_odi/overs/overs_1000893.json")) %>% setDT
+
+
+
+# %%
+cum_run_curves = function(df){
+  subt = paste(unique(df$teamShortName) , collapse = ' vs ')
+  df[, remaining_wickets := 11 - totalWicket]
+  # max run tallies
+  totruns = df[, max(totalRuns), by = 'innings']
+  # figure
+  f = ggplot(df, aes(over, totalRuns, colour = as.factor(innings))) +
+    geom_point(aes(alpha = remaining_wickets)) +
+    geom_hline(yintercept = totruns[1, 2][[1]], colour = 'red', alpha = 0.5, linetype = 'dashed') +
+    geom_hline(yintercept = totruns[2, 2][[1]], colour = 'blue' , alpha = 0.5, linetype = 'dashed') +
+    geom_vline(xintercept = 50, alpha = 0.5, linetype = 'dotted') +
+    scale_colour_brewer(palette = "Set1") +
+    labs(subtitle = subt, caption = 'alpha scaled by number of wickets left',
+      colour = 'Innings') + scale_alpha(guide = 'none')
+  return(f)
+}
+
+p1 = cum_run_curves(df1) + ggtitle('Failed Chase')
+p2 = cum_run_curves(df) + ggtitle('Successful Chase')
+
+# %%
+p = p1 / p2
+ggsave('/home/alal/Dropbox/1_Research/cricket/tmp/cum_run_curve.pdf', p, device = cairo_pdf)
+ggsave('/home/alal/Dropbox/1_Research/cricket/tmp/cum_run_curve.png', p)
+
+# %%

### 2020-11-06 scripts/06_results.R
"R script for results for easier read"

@@ -1,0 +1,349 @@
+## ---- echo=FALSE, include=FALSE-----------------------------------------------
+# %% ####################################################
+rm(list = ls())
+library(LalRUtils)
+LalRUtils::libreq(tidyverse, data.table, stargazer2,
+  lfe, ivmodel, fastDummies, patchwork,
+  rio, foreach, magrittr, janitor, knitr, rmarkdown)
+theme_set(lal_plot_theme())
+options(repr.plot.width=12, repr.plot.height=9)
+set.seed(42)
+# %% ####################################################
+opts_chunk$set(echo=TRUE, autodep=TRUE, cache.comments=FALSE, cache = TRUE,
+               message=FALSE, warning=FALSE)
+outf = 'html'
+
+
+## -----------------------------------------------------------------------------
+# %% long data
+root = '/home/alal/Dropbox/1_Research/cricket/'
+datadir = file.path(root, "repos/cricket-stats/data")
+crickett = fread(file.path(root, "tmp/matches_long.csv"))
+setorder(crickett, -year, -month, -date, mid)
+
+
+## ---- echo=F, include = F-----------------------------------------------------
+# check that match id is unique
+crickett[, .N, mid][order(-N)][1]
+# dedupe
+# %% # ## Subsample (for now)
+crickett$type_of_match %>% tabyl
+regsamp = crickett[type_of_match  %in%
+  c("ODI", "T20", "T20I", "Twenty20", "TEST", "FC", "LISTA", "Women's T20")]
+# %% throw out matches with no results / cancelled / walkover
+regsamp[outcome %like% 'abandoned' | outcome %like% 'cancelled', .N]
+regsamp = regsamp[!(outcome %like% 'abandoned' | outcome %like% 'cancelled')]
+regsamp[outcome %like% 'walkover' , .N]
+regsamp = regsamp[!(outcome %like% 'walkover')]
+# fix bat first and toss indicators
+########     ###    ########  #######  ########  ########  ######## ########
+##     ##   ## ##      ##    ##     ## ##     ## ##     ## ##       ##     ##
+##     ##  ##   ##     ##    ##     ## ##     ## ##     ## ##       ##     ##
[…]

### 2021-01-08 scripts/04_prep.R
"jan 7 2021 draft code"

@@ -7,19 +7,13 @@ LalRUtils::libreq(tidyverse, stargazer2, lfe, rio,
 theme_set(lal_plot_theme())
 options(repr.plot.width=12, repr.plot.height=9)
 
-# %% [markdown]
-# # Data Cleaning
-# run this whole section once, then start with [Empirics](#Empirics)
 
 # %%
 # setwd(root)
 root = '/home/alal/Dropbox/1_Research/cricket/'
 datadir = file.path(root, "repos/cricket-stats/data")
 
-# %% [markdown]
-# ## Ingest
-
-# %%
+# %% input data
 odi_ranks  <- fread(file.path(datadir,"rankings_odi.csv"))
 test_ranks <- fread(file.path(datadir,"rankings_test.csv"))
 ranks <- rbind(odi_ranks, test_ranks)
@@ -35,6 +29,9 @@ match[bat_or_bowl %in% c(0, NA), .N, by = outcome] [order(-N)]
 match = match[!(bat_or_bowl %in% c(0, NA))]
 
 # %% date processing
+
+match %>% glimpse
+
 # Match dates - need only month and year
 # Clean a bit
 match[, date2 := str_trim(date)]
@@ -64,14 +61,13 @@ match$year  <- ifelse(as.numeric(temp[,3]) < 1700,
 match[order(-year), .(month, day, year)]
 
 # %%
-# manual fix for bad years
+# manual fix for bad year variables
 match[, year := ifelse(year > 2020, year - 100,year)]
 match[, .(month, day, year, date, date2)][order(-year)][1:10]
 
-# %% [markdown]
-# ## Match Types
 
-# %%
+
+# %%# ## Match Types
 # International can be split by men, women, youth
 # Domestic matches apparently cannot be as we don't have that info.
 # See: https://github.com/dwillis/cricket-stats/issues/15
@@ -114,24 +110,13 @@ match[, day_n_night := ifelse(day_n_night=="night match", "day/night match", day
 match[, team1_spid  := tolower(paste0(men_type_of_match, team1, month, year))]
 match[, team2_spid  := tolower(paste0(men_type_of_match, team2, month, year))]
 
-# %% [markdown]
-# ## Rankings
+match[, .(team1_spid, team2_spid)] %>% head
 
 # %%
+# %% ## Rankings
 # Uniques for odi and test
 # It is not ranking but rating data (higher the better)
[…]

### 2021-01-08 scripts/05_cricket.R

@@ -1,15 +1,15 @@
 "
 @description:
 1. Analyses
-2. Graphs 
+2. Graphs
 
-@authors: 
+@authors:
 Gaurav Sood and Derek Willis
 
 "
 
 # setwd
-setwd(paste0(githubdir, "cricket-stats"))
+# setwd(paste0(githubdir, "cricket-stats"))
 
 # Source merge script
 source("scripts/04_merge_ranking_grounds_data.R")
@@ -20,6 +20,7 @@ library(stringi)
 library(xtable)
 library(tidyr)
 
+# %%
 "
 Notes About the Data:
 
@@ -33,7 +34,7 @@ sum(match$team2=='')
 
 "
 Take out matches where decision on who bowled/batted first is unknown. See this, for e.g.:
-http://www.espncricinfo.com/matches/engine/match/537589.html 
+http://www.espncricinfo.com/matches/engine/match/537589.html
 
 Takes out abandoned matches
 # abandoned
@@ -54,7 +55,7 @@ http://www.espncricinfo.com/matches/engine/match/329869.html
 temp <- grepl('No result', cricket$outcome)
 cricket <- subset(cricket, !temp)
 
-# ---------------- 
+# ----------------
 
 # Prop. who win toss, win game. (not including draws)
 table(cricket$win_toss_win_game, cricket$basic_type_of_match)
@@ -64,18 +65,18 @@ Melt the data
 Two rows per match
 "
 
-# Match level vars: 
-match_cols <- c("url", "match_id", "uniqueid", "date", "month", "day", "year", "rain",  "diff_ranks", "bat_or_bowl", 
-                 "type_of_match", "basic_type_of_match", "type_of_match2", "di_type_of_match", "men_type_of_match", "day_n_night", "youth", "women", "youth.women", "unofficial", "duckworth_lewis", 
-                 "win_toss", "win_game", "home_wins_toss", "draw", "outcome", "wickets", "runs", "balls", "innings", 
+# Match level vars:
+match_cols <- c("url", "match_id", "uniqueid", "date", "month", "day", "year", "rain",  "diff_ranks", "bat_or_bowl",
[…]

### 2021-01-08 scripts/06_results.R

@@ -1,152 +1,32 @@
-## ---- echo=FALSE, include=FALSE-----------------------------------------------
 # %% ####################################################
 rm(list = ls())
 library(LalRUtils)
-LalRUtils::libreq(tidyverse, data.table, stargazer2,
-  lfe, ivmodel, fastDummies, patchwork,
-  rio, foreach, magrittr, janitor, knitr, rmarkdown)
+LalRUtils::libreq(tidyverse, data.table, stargazer2, broom, fixest,
+  lfe, ivmodel, fastDummies, patchwork, purrr, kableExtra, xtable,
+  rio, magrittr, janitor, knitr, ggrepel)
 theme_set(lal_plot_theme())
 options(repr.plot.width=12, repr.plot.height=9)
 set.seed(42)
-# %% ####################################################
-opts_chunk$set(echo=TRUE, autodep=TRUE, cache.comments=FALSE, cache = TRUE,
-               message=FALSE, warning=FALSE)
-outf = 'html'
-
-
+outf = 'latex'
 ## -----------------------------------------------------------------------------
 # %% long data
 root = '/home/alal/Dropbox/1_Research/cricket/'
 datadir = file.path(root, "repos/cricket-stats/data")
-crickett = fread(file.path(root, "tmp/matches_long.csv"))
-setorder(crickett, -year, -month, -date, mid)
+outdir = file.path(root, "output")
 
-
-## ---- echo=F, include = F-----------------------------------------------------
-# check that match id is unique
-crickett[, .N, mid][order(-N)][1]
-# dedupe
-# %% # ## Subsample (for now)
-crickett$type_of_match %>% tabyl
-regsamp = crickett[type_of_match  %in%
-  c("ODI", "T20", "T20I", "Twenty20", "TEST", "FC", "LISTA", "Women's T20")]
-# %% throw out matches with no results / cancelled / walkover
-regsamp[outcome %like% 'abandoned' | outcome %like% 'cancelled', .N]
-regsamp = regsamp[!(outcome %like% 'abandoned' | outcome %like% 'cancelled')]
-regsamp[outcome %like% 'walkover' , .N]
-regsamp = regsamp[!(outcome %like% 'walkover')]
-# fix bat first and toss indicators
-########     ###    ########  #######  ########  ########  ######## ########
[…]

### 2021-01-08 scripts/initialParser.py

@@ -1,0 +1,39 @@
+import json
+import re
+import csv
+import os
+from tqdm import tqdm
+
+
+def structure(filename, data_writer):
+    with open(filename, "r") as content:
+        text_form = json.load(content)
+        for i in range(len(text_form)):
+            current = text_form[i]
+            rain_info = re.findall("Match State(?:.(?!<))*Rain.*?<", current['postText'], re.IGNORECASE)
+            if len(rain_info) == 0:
+                continue
+            match_state = rain_info[len(rain_info) - 1]
+            formatted_version = match_state[:len(match_state) - 1]
+            if formatted_version.lower().endswith('rain stopped'):
+                continue
+            day = re.findall(r'\d+', formatted_version)[0]
+            row = [filename, day, str(current['innings']), str(current['over']), formatted_version]
+            data_writer.writerow(row)
+
+
+def main():
+    outfile = open('rain_data.csv', 'w')
+    data_writer = csv.writer(outfile, delimiter=',', quotechar='"', quoting=csv.QUOTE_MINIMAL)
+    data_writer.writerow(['filename', 'day', 'innings', 'overs', 'match_state'])
+    directory = "comms"
+    for filename in tqdm(os.listdir(directory)):
+        if filename.endswith(".json"):
+            structure(os.path.join(directory, filename), data_writer)
+
+
+if __name__ == '__main__':
+    main()
+
+
+

### 2021-03-07 scripts/01_prep.R
"2021 draft codebase; purged old output and code" "

@@ -10,8 +10,7 @@ options(repr.plot.width=12, repr.plot.height=9)
 
 # %%
 # setwd(root)
-root = '/home/alal/Dropbox/1_Research/cricket/'
-datadir = file.path(root, "repos/cricket-stats/data")
+datadir = file.path("../data/")
 
 # %% input data
 odi_ranks  <- fread(file.path(datadir,"rankings_odi.csv"))
@@ -316,3 +315,85 @@ crickett$signed_diff_ranks <- ifelse(crickett$var=="team1",
 
 # %%
 fwrite(crickett, file.path(root, "tmp/matches_long.csv"))
+
+# %%
+# %%
+########  ########  ######## ########
+##     ## ##     ## ##       ##     ##
+##     ## ##     ## ##       ##     ##
+########  ########  ######   ########
+##        ##   ##   ##       ##
+##        ##    ##  ##       ##
+##        ##     ## ######## ##
+
+crickett = fread(file.path(datadir, "/matches_long.csv"))
+setorder(crickett, -year, -month, -date, mid)
+# %%
+crickett %>% glimpse
+crickett$type_of_match %>% tabyl %>% mutate(n/2)
+
+# %%
+# check that match id is unique
+crickett[, .N, mid][order(-N)][1]
+# dedupe
+crickett$type_of_match %>% tabyl
+# # ## Subsample of well understood match types
+regsamp = crickett[type_of_match  %in%
+  c("ODI", "T20", "T20I", "Twenty20", "TEST", "FC")]
+# throw out matches with no results / cancelled / walkover
+regsamp[outcome %like% 'abandoned' | outcome %like% 'cancelled', .N]
+regsamp = regsamp[!(outcome %like% 'abandoned' | outcome %like% 'cancelled')]
+regsamp[outcome %like% 'walkover' , .N]
+regsamp = regsamp[!(outcome %like% 'walkover')]
+# fix bat first and toss indicators
+regsamp$bat_first %>% tabyl
+# win toss checks
+regsamp$wintoss %>% tabyl
+# final batting order checks
+regsamp[, bat_first_min := min(bat_first), by = mid]
+regsamp$bat_first_min %>% tabyl
+# these have no toss info
+regsamp[bat_first_min == 1,
+  .(mid, outcome, wintoss, wingame, bat_first, bat_first_min)]
+# drop them
+regsamp = regsamp[!(bat_first_min == 1)]
+
+# check coding of draws
+regsamp[, .N, wingame]
+
[…]

### 2021-03-07 scripts/02_results.R

@@ -9,10 +9,18 @@ options(repr.plot.width=12, repr.plot.height=9)
 set.seed(42)
 outf = 'latex'
 ## -----------------------------------------------------------------------------
-# %% long data
-root = '/home/alal/Dropbox/1_Research/cricket/'
-datadir = file.path(root, "repos/cricket-stats/data")
-outdir = file.path(root, "output")
+# %% paths
+outdir = file.path("../output")
+regsamp = readRDS(file.path("../data/regression_sample.rds"))
+
+# %%
+######## ##     ## ##    ##  ######  ######## ####  #######  ##    ##  ######
+##       ##     ## ###   ## ##    ##    ##     ##  ##     ## ###   ## ##    ##
+##       ##     ## ####  ## ##          ##     ##  ##     ## ####  ## ##
+######   ##     ## ## ## ## ##          ##     ##  ##     ## ## ## ##  ######
+##       ##     ## ##  #### ##          ##     ##  ##     ## ##  ####       ##
+##       ##     ## ##   ### ##    ##    ##     ##  ##     ## ##   ### ##    ##
+##        #######  ##    ##  ######     ##    ####  #######  ##    ##  ######
 
 # %% intro subgroup stats + figs
 reg_by_group2 = function(df, groupvar, fml){
@@ -62,90 +70,20 @@ ann_coefplotter = function(df, title, vn = 1.5, hn = 0){
 
 
 # %%
-crickett = fread(file.path(root, "tmp/matches_long.csv"))
-setorder(crickett, -year, -month, -date, mid)
-# %%
-crickett %>% glimpse
-crickett$type_of_match %>% tabyl %>% mutate(n/2)
-
-# %%
-########  ########  ######## ########
-##     ## ##     ## ##       ##     ##
-##     ## ##     ## ##       ##     ##
-########  ########  ######   ########
-##        ##   ##   ##       ##
-##        ##    ##  ##       ##
-##        ##     ## ######## ##
-
-# %%
-# check that match id is unique
-crickett[, .N, mid][order(-N)][1]
-# dedupe
-crickett$type_of_match %>% tabyl
-# # ## Subsample of well understood match types
-regsamp = crickett[type_of_match  %in%
-  c("ODI", "T20", "T20I", "Twenty20", "TEST", "FC")]
[…]

### 2021-03-07 scripts/03_iv.R

@@ -1,4 +1,19 @@
+# %% ####################################################
+rm(list = ls())
+library(LalRUtils)
+LalRUtils::libreq(tidyverse, data.table, stargazer2, broom, fixest,
+  lfe, ivmodel, fastDummies, patchwork, purrr, kableExtra, xtable,
+  rio, magrittr, janitor, knitr, ggrepel)
+theme_set(lal_plot_theme())
+options(repr.plot.width=12, repr.plot.height=9)
+set.seed(42)
+outf = 'latex'
+# %% long data
+outdir = file.path("../output")
+regsamp = readRDS("../data/regression_sample.rds")
 # %%
+fstatrow = function(m) round(m$stage1$iv1fstat[[1]][['F']], 2)
+marker = function (name, yesno) return(c(name, map_chr(yesno, function(x) ifelse(x, "$\\checkmark$", ""))))
 ########     ###    ########
 ##     ##   ## ##      ##
 ##     ##  ##   ##     ##
@@ -6,9 +21,6 @@
 ##     ## #########    ##
 ##     ## ##     ##    ##
 ########  ##     ##    ##
-
-# %%
-
 # %%
 ols          = felm(wingame ~ bat_first | 0 | 0 | mid , regsamp)
 # matches where toss-winner chose to bat first

### 2021-03-08 scripts/01_prep.R
"added draws benchmarking - tables tbd"

@@ -325,7 +325,6 @@ fwrite(crickett, file.path(root, "tmp/matches_long.csv"))
 ##        ##   ##   ##       ##
 ##        ##    ##  ##       ##
 ##        ##     ## ######## ##
-
 crickett = fread(file.path(datadir, "/matches_long.csv"))
 setorder(crickett, -year, -month, -date, mid)
 # %%
@@ -337,6 +336,8 @@ crickett$type_of_match %>% tabyl %>% mutate(n/2)
 crickett[, .N, mid][order(-N)][1]
 # dedupe
 crickett$type_of_match %>% tabyl
+
+# %%
 # # ## Subsample of well understood match types
 regsamp = crickett[type_of_match  %in%
   c("ODI", "T20", "T20I", "Twenty20", "TEST", "FC")]

### 2021-03-08 scripts/02_results.R

@@ -2,8 +2,8 @@
 rm(list = ls())
 library(LalRUtils)
 LalRUtils::libreq(tidyverse, data.table, stargazer2, broom, fixest,
-  lfe, ivmodel, fastDummies, patchwork, purrr, kableExtra, xtable,
-  rio, magrittr, janitor, knitr, ggrepel, mediation)
+  lfe, fastDummies, patchwork, purrr, kableExtra, xtable,
+  rio, magrittr, janitor, knitr, ggrepel)
 theme_set(lal_plot_theme())
 options(repr.plot.width=12, repr.plot.height=9)
 set.seed(42)
@@ -77,8 +77,7 @@ ann_coefplotter = function(df, title, vn = 1.5, hn = 0){
      ## ##     ## ##     ## ##     ## ######### ##   ##      ##
 ##    ## ##     ## ##     ## ##     ## ##     ## ##    ##     ##
 ######   #######  ##     ## ##     ## ##     ## ##     ##    ##
-
-regsamp[, tabyl(type_of_match)]
+regsamp[, tabyl(type_of_match2)]
 # %%
 regsamp[, .N, bat_first]
 regsamp[, .N, wingame]
@@ -131,6 +130,23 @@ regsamp[type_of_match == "FC"] %>%
   tabyl(wintoss, wingame) %>%
   adorn_percentages("row") %>%
   adorn_pct_formatting()
+
+# %%
+########  ########     ###    ##      ##  ######
+##     ## ##     ##   ## ##   ##  ##  ## ##    ##
+##     ## ##     ##  ##   ##  ##  ##  ## ##
+##     ## ########  ##     ## ##  ##  ##  ######
+##     ## ##   ##   ######### ##  ##  ##       ##
+##     ## ##    ##  ##     ## ##  ##  ## ##    ##
+########  ##     ## ##     ##  ###  ###   ######
+
+regsamp[type_of_match2 %in% c("FC", "TEST") & wintoss == 1 & wingame %in% c(0.0, 0.5)] %>% tabyl(wingame)
+regsamp[type_of_match2 %in% c("FC", "TEST") & wintoss == 0 & wingame %in% c(0.0, 0.5)] %>% tabyl(wingame)
+
+regsamp[type_of_match2 %in% c("FC", "TEST") & wintoss == 1 & wingame %in% c(0.5, 1.0)] %>% tabyl(wingame)
[…]

### 2021-05-02 scripts/02_results.R
"added heterogeneity plots"

@@ -13,6 +13,8 @@ outf = 'latex'
 outdir = file.path("../output")
 regsamp = readRDS(file.path("../data/regression_sample.rds"))
 
+regsamp[, last_outcome := lag(wingame), by = name]
+regsamp[, last_toss    := lag(wintoss), by = name]
 # %%
 ######## ##     ## ##    ##  ######  ######## ####  #######  ##    ##  ######
 ##       ##     ## ###   ## ##    ##    ##     ##  ##     ## ###   ## ##    ##
@@ -103,12 +105,10 @@ regsamp[bat_or_bowl == 2]  %>%
   adorn_pct_formatting()
 
 
+# %% sanity checks
+.381 * .565 + .408 * .435
+
 # %%
-.381 * .565 + .408 * .435
-
-.381 * .565 + .408 * .435
-
-
 # %%
 
 regsamp[type_of_match == "ODI"] %>%
@@ -184,8 +184,6 @@ ggsave(file.path(outdir, 'matchcounts.pdf'), p0, width = 5, height = 7.5, device
 
 # %% balance checks
 setorder(regsamp, name, date)
-regsamp[, last_outcome := lag(wingame), by = name]
-regsamp[, last_toss    := lag(wintoss), by = name]
 balsamp2 = regsamp[!is.na(rank) & rank != 0];
 balsamp2[, `:=`(lowrank = min(rank), nm = .N) , by = mid]
 balsamp2 = balsamp2[nm == 2]
@@ -451,10 +449,12 @@ tests$time_unit %>% tabyl
 rank_exists = regsamp[type_of_match2 %in% c("OD", "TEST") & !is.na(rank)]
 rank_exists[, diff_rank := abs(rank[1] - rank[2]), by = mid]
 rank_exists[, big_gap   := ifelse(diff_rank > 15, 1, 0), by = mid]
+rank_exists[, diff_std := scale(diff_rank)]
 
 # %%
+# %%
 (p_rank_test = reg_by_group2(rank_exists[!is.na(big_gap) & type_of_match == "TEST"], big_gap,
-        formula_lfe('wingame', 'wintoss', C = "mid")) %>% rbindlist %>%
+        formula_fixest('wingame', 'wintoss')) %>% rbindlist %>%
     .[term == "wintoss"]  %>%
     inner_join(counter(rank_exists[!is.na(big_gap) & type_of_match == "TEST"], big_gap), by = "group") %>%
     mutate(group = as.factor(group)) %>%
[…]

### 2021-05-04 scripts/01_prep.R
see changelog

@@ -1,6 +1,7 @@
 # %%
 ###################################################
 rm(list = ls())
+# devtools::install_github("apoorvalal/LalRUtils")
 library(LalRUtils)
 LalRUtils::libreq(tidyverse, stargazer2, lfe, rio,
   foreach, magrittr, data.table, janitor) #,  lib2)

### 2021-05-04 scripts/02_results.R

@@ -1,5 +1,9 @@
 # %% ####################################################
 rm(list = ls())
+
+# devtools::install_github("apoorvalal/LalRUtils")
+# devtools::install_github("apoorvalal/stargazer2")
+
 library(LalRUtils)
 LalRUtils::libreq(tidyverse, data.table, stargazer2, broom, fixest,
   lfe, fastDummies, patchwork, purrr, kableExtra, xtable,
@@ -435,7 +439,7 @@ ggsave(file.path(outdir, 'reduced_form_by_matchtype.pdf'), p_rf_het,
 
 # %%
 p_rf_het = (p_overtime / p_test_time  / p_fc_time / ( p_odi_time | p_t20_time )) +
-  plot_annotation('Heterogeneous Treatment Effects of the Toss on Win Probability by Format over time')
+  plot_annotation('Heterogeneous Treatment Effects of the Toss on Win Probability by Format Over Time')
 
 ggsave(file.path(outdir, 'reduced_form_by_format_overtime.pdf'), p_rf_het,
   height = 15, width = 12, device = cairo_pdf)
@@ -524,8 +528,8 @@ sortedeffects_od <- spe(fm = fml, data = het_te_wide[type_of_match == "ODI"],
 toc()
 # %%
 pdf(file.path(outdir, 'het_rank_te_odi.pdf'))
-plot(x = sortedeffects_od, ylim = c(-.05, 0.05), ylab = "Change in Probability from winning the toss",
-  main = "ATE and Sorted Effect of Toss on Win Probability \n One-day Matches with ILO rankings",
+plot(x = sortedeffects_od, ylim = c(-.05, 0.05), ylab = "Change in Probability from Winning the Toss",
+  main = "ATE and Sorted Effect of Toss on Win Probability \n One-day Matches with ELO rankings",
   sub = "Logit ; het-TE by rank, last outcome and toss, and team dummies")
 dev.off()
 # %%
@@ -534,8 +538,8 @@ sortedeffects_test <- spe(fm = fml, data = het_te_wide[type_of_match == "TEST"],
         b = 100, bc = TRUE)
 # %%
 pdf(file.path(outdir, 'het_rank_te_test.pdf'))
-plot(x = sortedeffects_test, ylim = c(-.05, 0.1), ylab = "Change in Probability from winning the toss",
-  main = "ATE and Sorted Effect of Toss on Win Probability \n Test Matches with ILO rankings",
[…]

### 2021-05-04 scripts/archive/05_cricket.R
nuke archive

@@ -1,443 +1,0 @@
-"
-@description:
-1. Analyses
-2. Graphs
-
-@authors:
-Gaurav Sood and Derek Willis
-
-"
-
-# setwd
-# setwd(paste0(githubdir, "cricket-stats"))
-
-# Source merge script
-source("scripts/04_merge_ranking_grounds_data.R")
-
-# Load libs
-library(plyr)
-library(stringi)
-library(xtable)
-library(tidyr)
-
-# %%
-"
-Notes About the Data:
-
-Not scraped 'matches' where the match was abandoned without a toss
-In a previous dataset, about 3k matches were abandoned so. Don't like the 7% number. Touch too high, imho. But checked data - turns out to be ok.
-sum(match$win_toss=='')
-
-Any matches with missing team name info.?
-sum(match$team2=='')
-"
-
-"
-Take out matches where decision on who bowled/batted first is unknown. See this, for e.g.:
-http://www.espncricinfo.com/matches/engine/match/537589.html
-
-Takes out abandoned matches
-# abandoned
-table(match$bat_or_bowl, grepl('abandon', match$outcome))
-# 1019
-
-
-Takes out 3164 matches
-"
-
-cricket <- subset(match, bat_or_bowl!="")
-
-"
-Takes out matches where there was no result. See for e.g.,
-http://www.espncricinfo.com/matches/engine/match/329869.html
-"
-
-temp <- grepl('No result', cricket$outcome)
-cricket <- subset(cricket, !temp)
-
-# ----------------
-
-# Prop. who win toss, win game. (not including draws)
-table(cricket$win_toss_win_game, cricket$basic_type_of_match)
-
-"
-Melt the data
-Two rows per match
-"
-
-# Match level vars:
-match_cols <- c("url", "match_id", "uniqueid", "date", "month", "day", "year", "rain",  "diff_ranks", "bat_or_bowl",
-                 "type_of_match", "basic_type_of_match", "type_of_match2", "di_type_of_match", "men_type_of_match", "day_n_night", "youth", "women", "youth.women", "unofficial", "duckworth_lewis",
-                 "win_toss", "win_game", "home_wins_toss", "draw", "outcome", "wickets", "runs", "balls", "innings",
[…]

### 2021-05-04 scripts/archive/eng_county_champ.py

@@ -1,58 +1,0 @@
-'''
-Parse Downloaded Cricket Data
-
-'''
-
-import time
-import csv
-import simplejson
-from espncricinfo.match import Match
-
-headers = ["team1", "team1_id", "team2", "team2_id", "win_toss", "bat_or_bowl", "outcome", "win_game", "start_date", "end_date", "country", "ground", "match_id", "type_of_match", "match_type_id", "home_team_id", "first_innings_runs", "first_innings_batting_team", "url"]
-
-matches = simplejson.loads(open('/Users/derekwillis/code/cricket-stats/data/json/matches-first-class.json').read())
-matches = sorted(list(set(matches))) # dedupe
-
-##################################START PROCESSING DATA#########################################
-with open("english_county_championship_innings.csv", "wb") as csvfile:
-    writer = csv.writer(csvfile)
-    writer.writerow(headers)
-    for match in matches:
-        if int(match) < 539738:
-            continue
-        time.sleep(1)
-        try:
-            m = Match(int(match))
-            if m.match_json()['country_name'] != 'England':
-                continue
-            if m.match_json()['match_status'] == 'forthcoming':
-                continue
-            if m.result() == '':
-                continue
-            if m.result() == 'Match abandoned without a ball bowled':
-                continue
-            if len(m.innings()) == 0:
-                continue
-            print match
-            try:
-                m.team_2()['team_name']
-            except KeyError:
-                continue
-            try:
-                m.team_1()['team_name']
-            except KeyError:
-                continue
-            if m.team_1_id == m.innings()[0]['batting_team_id']:
-                first_innings_batting_team = m.team_1()['team_name']
-            else:
-                first_innings_batting_team = m.team_2()['team_name']
-            if m.match_json()['international_class_card'] != "":
-                match_type_id = m.match_json()['international_class_id']
-            else:
-                match_type_id = m.match_json()['general_class_id']
-            writer.writerow([m.team_1()['team_name'], m.team_1_id(), m.team_2()['team_name'], m.team_2_id(), m.toss_winner(), m.toss_decision(), m.result(), m.match_json()['winner_team_id'], m.date(), m.match_json()['end_date_raw'], m.lighting(), m.match_json()['country_name'], m.ground_name(), match, m.match_class(), match_type_id, m.match_json()['home_team_id'], m.innings()[0]['runs'], first_innings_batting_team, m.match_url])
-        except simplejson.scanner.JSONDecodeError:
-            continue
-
[…]

### 2021-05-04 scripts/archive/initialParser.py

@@ -1,39 +1,0 @@
-import json
-import re
-import csv
-import os
-from tqdm import tqdm
-
-
-def structure(filename, data_writer):
-    with open(filename, "r") as content:
-        text_form = json.load(content)
-        for i in range(len(text_form)):
-            current = text_form[i]
-            rain_info = re.findall("Match State(?:.(?!<))*Rain.*?<", current['postText'], re.IGNORECASE)
-            if len(rain_info) == 0:
-                continue
-            match_state = rain_info[len(rain_info) - 1]
-            formatted_version = match_state[:len(match_state) - 1]
-            if formatted_version.lower().endswith('rain stopped'):
-                continue
-            day = re.findall(r'\d+', formatted_version)[0]
-            row = [filename, day, str(current['innings']), str(current['over']), formatted_version]
-            data_writer.writerow(row)
-
-
-def main():
-    outfile = open('rain_data.csv', 'w')
-    data_writer = csv.writer(outfile, delimiter=',', quotechar='"', quoting=csv.QUOTE_MINIMAL)
-    data_writer.writerow(['filename', 'day', 'innings', 'overs', 'match_state'])
-    directory = "comms"
-    for filename in tqdm(os.listdir(directory)):
-        if filename.endswith(".json"):
-            structure(os.path.join(directory, filename), data_writer)
-
-
-if __name__ == '__main__':
-    main()
-
-
-

### 2021-05-04 scripts/99_eda.R
nuke eda and requirements and improve readme

@@ -1,51 +1,0 @@
-# %% ####################################################
-rm(list = ls())
-library(LalRUtils)
-LalRUtils::libreq(tidyverse, data.table, stargazer2, patchwork,
-                  lfe, rio, foreach, magrittr, janitor)
-theme_set(lal_plot_theme())
-options(repr.plot.width=12, repr.plot.height=9)
-set.seed(42)
-# %% ####################################################
-root = "/home/alal/Dropbox/1_Research/cricket"
-
-root = '/home/alal/Dropbox/1_Research/cricket/'
-datadir = file.path(root, "repos/cricket-stats/data")
-crickett = fread(file.path(root, "tmp/matches_long.csv"))
-
-# %%
-
-
-# %%
-df  = import(file.path(root, "tmp/overs_odi/overs/overs_1000891.json")) %>% setDT
-df1 = import(file.path(root, "tmp/overs_odi/overs/overs_1000893.json")) %>% setDT
-
-
-
-# %%
-cum_run_curves = function(df){
-  subt = paste(unique(df$teamShortName) , collapse = ' vs ')
-  df[, remaining_wickets := 11 - totalWicket]
-  # max run tallies
-  totruns = df[, max(totalRuns), by = 'innings']
-  # figure
-  f = ggplot(df, aes(over, totalRuns, colour = as.factor(innings))) +
-    geom_point(aes(alpha = remaining_wickets)) +
-    geom_hline(yintercept = totruns[1, 2][[1]], colour = 'red', alpha = 0.5, linetype = 'dashed') +
-    geom_hline(yintercept = totruns[2, 2][[1]], colour = 'blue' , alpha = 0.5, linetype = 'dashed') +
-    geom_vline(xintercept = 50, alpha = 0.5, linetype = 'dotted') +
-    scale_colour_brewer(palette = "Set1") +
-    labs(subtitle = subt, caption = 'alpha scaled by number of wickets left',
-      colour = 'Innings') + scale_alpha(guide = 'none')
-  return(f)
-}
-
-p1 = cum_run_curves(df1) + ggtitle('Failed Chase')
-p2 = cum_run_curves(df) + ggtitle('Successful Chase')
-
-# %%
-p = p1 / p2
-ggsave('/home/alal/Dropbox/1_Research/cricket/tmp/cum_run_curve.pdf', p, device = cairo_pdf)
-ggsave('/home/alal/Dropbox/1_Research/cricket/tmp/cum_run_curve.png', p)
-
-# %%

### 2021-05-04 scripts/02_results.R
update

@@ -8,6 +8,7 @@ library(LalRUtils)
 LalRUtils::libreq(tidyverse, data.table, stargazer2, broom, fixest,
   lfe, fastDummies, patchwork, purrr, kableExtra, xtable,
   rio, magrittr, janitor, knitr, ggrepel)
+source("00_func.R")
 theme_set(lal_plot_theme())
 options(repr.plot.width=12, repr.plot.height=9)
 set.seed(42)
@@ -19,60 +20,6 @@ regsamp = readRDS(file.path("../data/regression_sample.rds"))
 
 regsamp[, last_outcome := lag(wingame), by = name]
 regsamp[, last_toss    := lag(wintoss), by = name]
-# %%
-######## ##     ## ##    ##  ######  ######## ####  #######  ##    ##  ######
-##       ##     ## ###   ## ##    ##    ##     ##  ##     ## ###   ## ##    ##
-##       ##     ## ####  ## ##          ##     ##  ##     ## ####  ## ##
-######   ##     ## ## ## ## ##          ##     ##  ##     ## ## ## ##  ######
-##       ##     ## ##  #### ##          ##     ##  ##     ## ##  ####       ##
-##       ##     ## ##   ### ##    ##    ##     ##  ##     ## ##   ### ##    ##
-##        #######  ##    ##  ######     ##    ####  #######  ##    ##  ######
-
-# %% intro subgroup stats + figs
-reg_by_group2 = function(df, groupvar, fml){
-  require(dplyr) ; require(purrr) ; require(broom)
-  group_reg_tables = df %>%
-    group_by({{groupvar}}) %>%
-    group_map( ~ feols(fml, .x, cluster = ~mid) %>%
-      tidy %>% mutate(group = .y[[1]]), .keep = T)
-  return(group_reg_tables)
-}
-
-counter = function(df, groupvar){
-  df %>% tabyl({{groupvar}}) %>% mutate(n = n/2) %>% dplyr::select(-percent) %>%
-  rename(group = {{groupvar}})
-}
-
-coefplotter = function(df, title){
-  df %>%
-    mutate(term2 = paste0(group, '_', term)) %>%
-    ggplot(. , aes(y = group, x = estimate) ) +
-      geom_point(colour = '#6495ed') +
-      geom_pointrange(aes(
-        xmin = estimate - 1.96 * `std.error`,
-        xmax = estimate + 1.96 * `std.error`,
[…]

### 2021-05-04 scripts/03_iv.R

@@ -4,12 +4,13 @@ library(LalRUtils)
 LalRUtils::libreq(tidyverse, data.table, stargazer2, broom, fixest,
   lfe, ivmodel, fastDummies, patchwork, purrr, kableExtra, xtable,
   rio, magrittr, janitor, knitr, ggrepel)
+source("00_func.R")
 theme_set(lal_plot_theme())
 options(repr.plot.width=12, repr.plot.height=9)
 set.seed(42)
 outf = 'latex'
 # %% long data
-outdir = file.path("../output")
+outdir = file.path("../output/iv/")
 regsamp = readRDS("../data/regression_sample.rds")
 # %%
 fstatrow = function(m) round(m$stage1$iv1fstat[[1]][['F']], 2)

### 2021-05-06 ms/ms.tex
.tex

\documentclass[11pt,  letterpaper]{article}

\usepackage{mathpazo}
\usepackage{fontspec}
\setmainfont{TeX Gyre Pagella}

\input{boilerplate}

\title{\textbf{Asymmetries in Cricket \\ \vspace{5pt} \large{The Effects of Winning the Toss on Winning the Match}}\thanks{This paper incorporates some of the text and 
results from Sood \& Willis's (2019) working paper: ``Fairly Random: The Impact of Winning the Toss on the Probability of Winning.'' The authors thank Matt Blackwell, 
Andrew Gelman, Donald Green, Jens Hainmueller, and Daniel Stone for their comments and suggestions. We are grateful to Sajan Mehotra for outstanding research assistance. 
% The first three authors contributed roughly equally to the research design, data collection, and implementation of the whole project, and are listed alphabetically. Acharya
% contributed to the research design and interpretation of results.
}}

\author{
Apoorva Lal\thanks{Dept.~of Political Science, Stanford University;  \textsf{apoorval@stanford.edu}.} \;\;
Derek Willis\thanks{News applications developer at ProPublica;  \textsf{dwillis@gmail.com}.}  \;\;
Gaurav Sood\thanks{Independent researcher;  \textsf{gsood07@gmail.com}.} \;\;
Avidit Acharya\thanks{Dept.~of Political Science and Graduate School of Business, Stanford University; \textsf{avidit@stanford.edu}.}
}
\date{\today}

%\usepackage[
%  backend=biber,
%  style=authoryear,
%  citestyle=authoryear,
%  ]{biblatex}
%\addbibresource{biblio.bib}
%\renewbibmacro{in:}{}

\begin{document}

\maketitle

\begin{abstract}

The rules in cricket treat the competing teams asymmetrically, even before the start of physical play. In the one-day version of the game, for instance, which team bats first is determined by the captain that wins the toss. This raises questions about the effects of the asymmetry created by luck. Does the team that wins the toss have an advantage over the other team?

We find that, on average, the effect of winning the toss on winning the match is essentially nil in international limited-overs daytime matches but substantial (3.3 percentage points) in day/night matches. In test cricket, the advantage is even larger (4.9 percentage points). The advantage has exploded in the last decade (to nearly 15 percentage points) after a steady decline. More surprisingly, the effect of the toss does not appear to vary between competitive and noncompetitive pairings of teams. And as for the decision to bat or field first, we find the win-loss differential is significantly larger for the teams that elect to field than those that elect to bat, in which case it is negligible. Lastly, and perhaps most surprisingly, in rain-delayed matches in which the chasing team's total was adjusted by the Duckworth-Lewis formula, we find that teams that elected to bat first did just as well than those that elected to field first.

\smallskip

\textbf{Key words:} cricket, fairness, randomization

%\textbf{JEL codes:} C72
\end{abstract}

% #### ##    ## ######## ########   #######
%  ##  ###   ##    ##    ##     ## ##     ##
%  ##  ####  ##    ##    ##     ## ##     ##
%  ##  ## ## ##    ##    ########  ##     ##
%  ##  ##  ####    ##    ##   ##   ##     ##
%  ##  ##   ###    ##    ##    ##  ##     ##
% #### ##    ##    ##    ##     ##  #######

%
\section{Introduction}
In a competitive sport, every little thing matters. Yet, many sports leave some large levers out of the reach of the teams and in the
hands of fate. In cricket, one such lever---the toss---has been the subject of much attention. In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview. To wrest the advantage, the captain merely needs to pick the side of the coin that will be left facing the sky after the toss and choose whichever one---batting second or batting second---gives the greater advantage.

In this paper, we measure how well teams have been able to capitalize on any advantage conferred by winning the toss. Since the only consequence of the toss is letting the toss-winning side decide whether to bat or field first, we also attempt to measure the advantage of electing to bat first. To answer these questions, we collected data on over 35,000 cricket matches across four formats (First Class, ODIs, T20s, and tests) played over the course of more than one hundred fifty years. (To our knowledge, this is the most comprehensive professional cricket match dataset to date.)

Winning the toss increases the chances of winning the match by 1.8 percentage points: toss winners win 39.2\% of the time and lose 37.4\% of the time, with the remaining share being draws or ties. The effect of winning the toss is the largest in tests, where it is 4.9 percentage points and is smallest in ODIs and T20s, where it is essentially zero. That being said, in ODIs and T20s that are day/night matches, the effect of winning the toss is 3.3 percentage points, whereas, in the day-time matches, the effect is a precise zero.

The effect of toss in the ODIs has remained relatively stable since the 1980's when the one-day format first became popular. In tests, however, the importance of the toss has varied considerably. The effect of winning the toss in test cricket was a little over five percentage points before 1990. It then dropped nearly to 0 from 1990 to 2010. In the last decade, however, it has exploded, averaging 14.5 percentage points. Counter to conventional wisdom, the effect of winning the toss is relatively similar across continents, though the toss appears less important on European and African wickets than elsewhere. The effect on Aussie and Kiwi wickets (3 percentage points) is not substantially different from its effect on South Asian wickets (2.8 percentage points) but is higher than its effect on European and African wickets (1.1 and 0 percentage points respectively).

In sports like soccer, the conventional wisdom is that the toss is inconsequential \citep{csato2020comparison}. In cricket, it is believed that the toss has little significance in lopsided matches that have clear favorites and underdogs. For example, it is thought to be less relevant in an India-Ireland match than in an India-Australia match. For this reason, we look at the effect of the toss separately in subsamples where teams are similarly ranked according to ICC ELO ratings and where their ratings differ by more substantially. We find that the effect of the toss doesn't vary by how well matched the teams are. Since ELO ratings are available only for matches played in recent decades, these findings (along with our finding that influence of the toss has declined over time) suggest that the toss has little influence in modern cricket.

An important caveat about the null findings is in order, however. Our empirical approach cannot tell whether the effects are nil because the toss is unimportant or because the toss is important, but that (modern) teams have not been able to capitalize on the advantage. The toss may matter quite a bit in principle, but we would only be able to observe its effect if a team is able to use the information on the conditions of play and how it pairs strategically against the opposing team to be able to make the optimal decision of whether to bat or field first. If teams simply do not have this knowledge and are often choosing poorly, then we would not see any ``toss effects.'} So, when we refer to ``toss effects'' in this paper, what we are measuring is not whether there is an advantage to winning the toss in principle, but whether any such advantage is exploited in practice.

\footnote{The median rating is 17, so a cutoff of 15 roughly splits the sample of matches in which we have ratings for both teams in two halves.}

%Another way to benchmark the effect of the toss is to compute it in
%terms of ``run- or wicket-advantage.'' To do this, we consider ODI
%and T20 matches, and estimate [.Apoorva: describe the methodology
%that we discussed here]. We find that the effect of winning the toss
%and electing to bat first is an [X] run advantage. In other words,
%the effect of the toss in these instances is nullified only by
%allowing the chasing team to start their innings with [X] runs
%already on the board. Similarly, the effect of winning the toss and
%electing to field first is a [Y] run advantage, so that the effect is
%nullified only if we make the chasing team start with [-Y] runs on
%the board. \todo{We had an idea here that I vaguely recall of how to
%address the complication that if a team loses all ten wickets, then
%we don't observe its 50 over total.}

With this, we turn to the batting first decision. When pitch and weather conditions are hard to read, and it is not obvious that fielding first has a clear advantage, many cricket commentators suggest that batting first is the appropriate choice: get runs on the board, and put the opposing team under psychological pressure by having them chase the target, worrying about ``required run rates,'' etc. We see that the effect of winning the toss on batting first is 13 percentage points. If one team bats, the other team must field, so this translates to saying that among toss winners, 56.5\% of the teams elect to bat, and the remaining 43.5\% choose to field. Thus, compared to the case where teams are randomly assigned to bat, i.e., the case in which the toss has no effect on the batting decision, toss-winning teams bat an excess of 6.5\% of matches.

However, we find that of the toss winners that elected to bat, only 38.1\% won the match compared to 37.9\% that lost--- a small difference of 0.2 percentage points. On the other hand, of the toss winners that elected to field, 40.8\% won the match compared to 36.9\% that lost--- a larger difference of 3.9 percentage points. This means that 94\% of the 1.8 percentage point effect of winning the toss can be attributed to instances in which the toss winning team elected to field, and only 6\% of the effect can be attributed to the toss winning team electing to bat. This suggests that teams are better able to capitalize on the toss by electing to field when it is advantageous to do so. 
%
%Combining this with the fact that we saw the reduced form effect of the toss on the outcome of the match to be 1.8 percentage points, we can uncover the average effect of batting first on the so-called ``complier'' teams: those that got to bat (because they won the toss) that would not have batted if batting first were randomly assigned \citep{angrist1996identification}. Since winning the toss results in a 0.9\% excess of match wins (from the benchmark case in which the toss has no effect), the effect of batting first in this complier population is 0.9/6.5 = 13.9 percentage points. This is the instrumented effect of winning the toss-- or, the rate at which the excess decisions to bat were converted into excess wins.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sings of all of our estimates gives the instrumented effect of electing to field first. For example, the instrumented effect of electing to field first is -13.9 percentage points.} 
%
%This conversion rate is different in sign and considerably larger than the naive difference in win
%probabilities between teams that bat first and those that don't, which is $-1.6$ percentage points. Since the rules of cricket imply that the toss cannot affect the outcome of the match other than through the decision of the toss-winning team to bat or field, this discrepancy is likely due to the complier population being very different from the never-taker or always-taker populations that form the bulk of our team-match data. The IV estimates are thus
%telling us that some teams in some matches are able to take advantage of the decision to bat first when the batting choice really matters.

We also look at how the toss advantage breaks down between teams that elected to bat versus those that elected to field within the various subgroups mentioned above and others. Perhaps the most striking result from this analysis for many cricket followers is that there does not appear to be the case that toss winners that elect to bat first do better than those that elected to field first in rain delayed matches where the chasing total is adjusted according to the \citet{duckworth1998} formula.

Our results are pertinent to debates on how cricket should be governed. One argument is that since the toss is random, no team is advantaged by the format prior to the toss; and in fact the luck of the toss along with the strategic choice to bat first or second are integral parts of the game--- not very different from the luck of weather affecting the teams differently. 
%or the strategic choices that captains make as to how to set the field, or the order in which they deploy their batsmen.
An alternative argument, however, is that the toss is problematic because it advantages one team over the other even before the first ball is bowled; and care should be taken to mitigate this asymmetry--- for example, by making cricket pitches more durable so that they don't deteriorate so much so that batting on the final day of a test match is miserable. Otherwise, the toss-winning team could exploit this advantage by batting first, reducing the chance that it has to endure this misery. Moreover, the fact that no side is advantaged \emph{ex-ante} if the toss is random is not enough to justify not correcting any asymmetry created by the toss. In fact, the rules of cricket already try to correct the asymmetry between teams that bat first and those that bat second in one particular instance: In rain-delayed one-day matches, the Duckworth-Lewis formula is used to set a new target for the chasing team.\footnote{The argument that no side was advantaged \emph{ex-ante} could have been used to justify not doing any correction in such rain-delayed matches, and simply declaring that the team that batted first won the match because the chasing team was unable to reach its target. But the designers of one-day cricket were clearly unwilling to allow the bad luck of weather to have this much influence on the game. If addressing the misfortunes of weather are important, it may also be important to address the misfortune of losing the toss.}

For those who take this latter viewpoint, a straightforward policy prescription that comes out of our analysis is that, at least for ODIs, day-night matches should be avoided. In general, when new formats are introduced (such as the recent introduction of T20s) care should be taken to ensure the toss has little effect. The designers of cricket should always consider ways of leveling the playing field for the competing teams starting from the point at which the game really begins--- the moment the first ball is bowled.

We do not address these optimal design questions in this paper, but our results are relevant to these issues depending on what the designers of cricket seek to optimize. %However, at the end of the paper, we propose (and discuss the possible consequences of) a subtle change to the rules governing the toss: Rather than tossing a coin to confer the choice to bat first or second upon the captain of the toss-winning team, umpires may consider simply tossing a coin to determine which team must bat first.

% ########     ###    ########    ###
% ##     ##   ## ##      ##      ## ##
% ##     ##  ##   ##     ##     ##   ##
% ##     ## ##     ##    ##    ##     ##
% ##     ## #########    ##    #########
% ##     ## ##     ##    ##    ##     ##
% ########  ##     ##    ##    ##     ##

\section{Data}

\subsection{Creating the sample}

We obtained data on cricket matches from the ESPN Cricinfo website, which provides coverage of international and many domestic matches,
and maintains statistics on a wide variety of matches across formats.\footnote{Information on all available matches was downloaded in June 2020 using a purpose-built parsing library available at \href{https://github.com/dwillis/python-espncricinfo}{https://github.com/dwillis/python-espncricinfo}. The data will be available as a separate repository available at \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}

We focus on four types of formats: T20, ODI, Test, and First Class matches for both men's and women's teams.\footnote{We do not include the List A matches that are not ODIs or T20s, which are often considered a less professional category (with worse data recording) than the formats that we include.} The data are a near census of the relevant population, excluding only scheduled matches that were abandoned without the toss being conducted, and a small number of limited-overs matches in which the minimum number of overs that must be bowled to establish a result was not met.\footnote{In ODIs, for example, each side must bat at least 20 overs for a result to be declared. We have no reason to suspect that the small matches that were excluded because they did not meet this threshold could be biased enough to affect our conclusions. We drop 426 matches recorded as ``no result,'' 3 matches that are recorded as ``abandoned'' or ``canceled,'' and another three that don't record who bat first. The former constitutes the largest exclusion we make and appears to be comprised of matches canceled for idiosyncratic reasons, ranging from the \href{https://www.espncricinfo.com/series/shell-tri-series-1991-92-61203/australia-women-vs-england-women-final-66975/full-scorecard}{a team being unable to bat because of the weather} to cancellations because of \href{https://www.espncricinfo.com/series/australia-tour-of-pakistan-1982-83-61392/pakistan-vs-australia-3rd-odi-64196/full-scorecard}{unwanted audience participation}. This results in us dropping approximately 0.07\% of matches.}

Our final sample comprises of 35,357 cricket matches across four formats played over the last 150 years. We report the distribution of matches over time and by type in Figure \ref{fig:timeseries_decomp}. As the figure shows, the number of matches has exploded over the last two decades, largely due to the introduction of the enormously popular Twenty-20 (T20) format of the game both in both international and domestic tournaments. T20 matches constitute over a quarter of all matches played in the last decade.

% subsubsection data_collection (end)

\subsection{Is the toss random?}

Since our analysis will be based on the assumption that the coin toss is random, we start by examining this assumption.\footnote{Prior work in the sports analytics of cricket by \citet{shafqat2015analysis} and \citet{dawson2009bat} conditions on post-treatment variables, whereas we use only the randomization of the toss to identify toss effects.} Is there any evidence that the coin toss is somehow rigged--- for example, with the home side enjoying the rub of the green more often? In order to study this, we compare the rates at which home teams, teams that won the toss in their last match, and underdogs (defined as the lower-ranked team whenever rankings are available for both teams) win the toss relative to their opponents.

To examine whether the toss is rigged, we compute the odds of seeing as many successes in the same number of draws of a sequence of independent Bernoulli random variables with parameter 0.5. For example, home teams have won 50.81\% of tosses in the 6377 matches for which we have data. The likelihood of tossing a fair coin 6377 times and getting at least this percentage of heads is 20.1\%. This chances appears low, but not eyebrow-raisingly so. We find that toss wins are even more balanced for teams that won and lost the last toss, and for teams that are underdogs and favorites. Table \ref{table: balance} reports these balance results.

That said, we do see one potential source of concern. We find that teams that share their nationality with at least two of the three umpires of the match win 51.3\% tosses in the 4044 international matches for which we have this data. The likelihood that tossing a fair coin 4044 times and and getting at least this share of heads is only 10\%. This number is certainly low enough to raise some eyebrows. That said, the practice of allowing home-team umpires in international matches ended in 1992, and so these issues are less relevant in recent decades which comprise the bulk of our sample.

In addition, a rigged toss in favor of the home team would bias estimates of the advantage of winning the coin toss to the extent that it is
correlated with ability. If stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards. And vice versa, if otherwise. We do not have good reasons to think that there is a correlation. So for now, we proceed as if the toss is random.

% robust standard errors and report these estimates in
% table~\ref{table:baltab} alongside the number of matches used to test
% for balance. Wefail to reject the null that the toss is random.

%N = 6377
%Wins = 3240 home team wins
% prob succes = 50.81
% p value = 20.14\%.

%Wins = 17189
%N = 34283
%50.14
%p value = 0.6114

%Underdog = lower ranked team in any dyad with nonmissing ranking info (so int'l matches)
%Underdog won = 2001
%N = 4054
%Win prob = 49.36\%
%p value = 0.4231

% \input{output/baltab.tex}

% ########  #######   ######   ######
%    ##    ##     ## ##    ## ##    ##
%    ##    ##     ## ##       ##
%    ##    ##     ##  ######   ######
%    ##    ##     ##       ##       ##
%    ##    ##     ## ##    ## ##    ##
%    ##     #######   ######   ######

\section{Do teams exploit the advantage of winning the toss?}

To estimate the advantage of winning the toss, we run regressions on a team $\times$ match level dataset. Let $i$ index teams, and $j$ index matches. $Y_{ij}$ is a dummy for team $i$ winning match $j$, and takes a value of $0.5$ for draws. $Z_{ij}$ is a dummy for team $i$ winning the toss in match $j$. For the main specifications of interest, we estimate regression equations of the form
\begin{equation}\label{eqn:reduced_form}
Y_{ij} = \alpha_0 + \tau Z_{ij} + \gamma_i + \delta_t +
\psi_k + \varepsilon_{ij}
\end{equation}
which includes fixed effects for teams $\gamma_i$, time $\delta_t$, and match-type $\psi_k$, so that all variables can be interpreted as residuals from partialing out team, match-type, and time fixed effects. Since unobservable factors are correlated across opposing teams in the same match, we cluster the standard errors at the match level.

We report results in Table \ref{table:reduced_form}. We progressively introduce team, match-type, and time period fixed-effects in columns 2, 3, and 4 to make more fine-grained comparisons, and find that results are remarkably stable across specifications. The first column, for example gives
the baseline estimate of the toss effect being 1.8 percentage points in all matches. As we mentioned in the introduction, this effect represents toss winners winning 39.2\% of the time and losing 37.4\% of the time, with the remaining matches being draws. The estimated effect remains at 1.7 percentage points when we progressively add the fixed effects.

We then proceed to examine the effect of the toss withing various subgroups. We report the estimates graphically in Figures \ref{fig:rf_het_TE} - \ref{fig:omni_het_test} with point estimates and sample sizes as annotations. The results are summarized as follows.

\textbf{Format.} Toss effects are most pronounced in test matches followed by first class matches, and are statistically indistinguishable from zero in ODIs and T20s. We estimate the effect of the toss on winning a test match to be 4.7 percentage points, allowing for the fact that many tests end in draws. Given that only 65\% of test matches end with one team winning, the effect of the toss conditional on the match producing a winner is higher: 7.2 percentage points. The effect of the toss in first class matches is 2.3 percentage points. These results support the conventional wisdom among lay cricket followers that the toss grants the greatest advantage in multi-day affairs. Pitches invariably deteriorate over the course of these matches, and the fourth innings in test match is often the most challenging time to bat. The pitch deteriorates far less over the course of a day, or in the case of T20s, a few hours. Our null results for limited over matches also extend \citet{de1998winning}, who, based on analysis of data from 427 international one-day matches conclude that ``winning the toss at the outset of a match provides no competitive advantage.''

\textbf{International vs.~Domestic matches.} Toss effects are virtually the same for international and domestic matches (point estimates are 1.7 percentage points in the former, and 1.9 percentage points in the latter). This is a bit surprising given differential use of cricket formats differs in international and domestic matches (for example, there are no domestic tests), and our findings of heterogeneous effects across formats.

\textbf{Effects over time.} Toss effects overall appear to have declined over time from a high of 9.9 percentage points in the second half of the 19th century, to 1.4 percentage points in the last decade. We also look at these trends within format. T20s have not been played long enough for us to draw any conclusions about how toss effects have changed over time. But within the other formats, we find that toss effects follow a similar pattern, except that in test cricket there is a peculiar uptick in in the last decade: the toss effect jumps from 0 in the decade before to 14.5 percentage points now.  We investigate potential sources of this variation below  but there are some factors that could be important drivers for which we do not have data. For example, cricketers may have become more fit, and able to play for longer periods without tiring as much as they once did. Pitches now may be more resilient to wear than they once were. Teams may have acquired better knowledge on how to play more effectively whether they bat first or second. Etc.

\textbf{Relative strength.} Common sense suggests that the toss should be less important in lopsided matches where there is one dominant team that is expected to win, whether it bats first or second. It should be more important in competitive matches where the opposing teams are more even in cricketing strength. We define competitive matches as those in which the gap in ELO ratings of the two teams is less than 15 points. Since the median ELO rating difference is 17 points, this threshold roughly splits our sample of matches for which we have ELO ratings for both teams in two halves. Surprisingly, we find the toss effect to be nearly the same, and indistinguishable from zero within both competitive and noncompetitive matches even when we look at tests and international ODIs separately (these are the two formats for which ELO ratings apply). Part of the reason for this is that the ELO ratings are available only for recent years, and we saw above that toss effects have declined over time. We also perform an omnibus test of treatment effect heterogeneity by estimating sorted partial effects \citep{chernozhukov2018sorted,chen2019sortedeffects} with respect to ranking and result history, and find minimal evidence for effect heterogeneity.

\textbf{Continents.} We find that toss effects vary only somewhat across continents, hovering around 2.5 percentage points for matches played in Asia, Oceania, and the Americas (though the estimates for the Americas are noisier in part due to the smaller number of matches played there). In Europe (mainly England) the toss effect is only 1.1 percentage points (though we do not rule out a null effect), and in Africa it is essentially zero. Overall, despite there being much talk about the different nature of wickets across the world, with wickets in certain regions being drier, or stickier, or subject to more wear, these differences do not seem to translate into substantial toss effect differences at least when we look coarsely at averages.

\textbf{Day/Night matches.} It is often claimed that the toss is more crucial in day-and-night matches. We find some evidence for this. In the limited overs formats where day/night matches take place (ODIs and T20s), the advantage of winning the toss in a day/night match is 3.3 percentage points, whereas the
advantage of winning the toss in a limited overs match played during the day is zero.

\textbf{Seasonality.} Weather is also thought to have a large impact on the playing conditions in cricket. Cool
overcast weather, for example, is thought to aid swing bowling, especially in English pitches. More generally,
the advantage of winning the toss likely varies by the weather. Although do not have data on
the weather, we can proxy it with seasons. In particular, students of the game suspect that
the advantage of winning the toss in the early English season is especially significant.
There is some evidence of a seasonal pattern for matches played in England, with the advantage of winning the toss
growing over the summer, though the trend is weak and broken in the month of July.

\textbf{Rain delays.} Finally, we look at the differential effects of winning the toss in ODI matches that were rain-delayed, causing the number of overs bowled in the second innings to be adjusted down. We examine the subset of matches where rain-delays result in revised targets computed using the \citet{duckworth1998} method. This method has been the subject of much controversy, with many cricket fans suggesting that it is rigged against the chasing team. To our surprise, we find no evidence that toss winners are able to exploit the possibility of second inning rain delays to their advantage. If anything, we see that the toss-winner has a \emph{disadvantage} of 2.4 percentage points in matches subject to the Duckworth-Lewis adjustment, though this point estimate is statistically indistinguishable from zero \footnote{Toss-winners won 160 out of 333 ODIs with DL adjustment and 3085 out of 6187 ODIs without DL adjustment}. This, however, does not mean that the Duckworth-Lewis method is not rigged against the chasing team; that issue will be discussed further in the next section.

% \begin{table}
% \centering
% \caption{Reduced Form Effect of Toss on win probability by match type}
% \input{output/reduced_form_het.tex}
% \label{table:reduced_form_het}
% \end{table}

% subsubsection heterogeneous_treatment_effects (end)

% \newpage

% ########     ###    ########
% ##     ##   ## ##      ##
% ##     ##  ##   ##     ##
% ########  ##     ##    ##
% ##     ## #########    ##
% ##     ## ##     ##    ##
% ########  ##     ##    ##

\section{Batting versus Fielding First}

\subsection{The Choice to Bat}

The first column of Table \ref{table:iv_table} shows that 49.2\% of teams that batted first won the match, while 50.8\% of teams that fielded first won. The second column shows that among teams that won the toss and batted first, the ``win-loss differential'' is small: these teams won only 0.2 percentage points more frequently than they lost, winning 38.1\% of matches and losing 37.9\% of matches, as reported in the introduction. This win-loss differential is smaller than it is for teams that won the toss and elected to field. The third column of the table shows that these teams won 3.9 percentage points more frequently than they lost: they won 40.8\% of their matches and lost 36.9\% of them. Combining this information with the fact that 56.5\% of teams that won the toss elected to bat (and only 43.5\% elected to field, as implied by the fourth column of the table) this means that 94\% ($= .435 \times .039/.018$) of the 1.8 percentage point advantage to winning the toss that we reported in the previous section can be attributed to the success of toss-winning teams that elected to field.

These findings are not surprising given the conventional wisdom that captains who are not sure whether they should bat or field first often choose to bat. If a team wins the toss and elects to field, they often have a good reason to do so. It is interesting to ask what these reasons could be. Although we cannot answer this question fully, comparing  the win-loss differentials within subgroups may offer some clues.

\textbf{Format.} In test matches 72.8\% of teams that win the toss elect to bat, whereas in T20s and ODIs this share is only about 52.9\%. In first class matches it is 56.9\%. It appears that teams err heavily on the side of batting first in longer match formats.

\textbf{International vs. Domestic matches.} We see no big differences in propensity to elect to bat first across international versus domestic matches.

\textbf{Effects over time.} Historically, toss-winning teams tilted heavily in favor of batting first, but this declines over the course of the 20th century, and now they show no preference (on average) to  bat first. Even for test matches, toss winning teams tend to be tilting less toward batting first than they once did. Pre-1950, more than 90\% of teams that won the toss elected to bat, whereas in the last two decades that number is closer to two-thirds. Perhaps players have gotten more fit and are willing to trade-off expecting to have to bat on the final days of the match in favor of taking advantage of playing conditions that are favorable to fielding first. As for ODIs, captains curiously erred heavily on the side of fielding first in the early decades of the format, but in the last thirty years, they have tilted more towards batting first, with just over half of toss-winning teams electing to bat.

\textbf{Relative strength.} We see no notable differences in the toss-winning captain's choice to bat first across matches that are lopsided and those in which the teams are more evenly matched. This is especially true in tests, but also appears to be true in ODIs, though a slightly larger share of toss-winning teams elect to bat in matches that are lopsided (54.6\%) in comparison to ones that are less lopsided (51.9\%).

\textbf{Continents.} Toss-winning captains show a preference to bat first in matches played in Oceania and Europe, more so than in other pitches. For example, in Europe (mainly England) $64.9\%$ of toss winning teams elect to bat, and in Oceania this number is $59.9\%$. By contrast, on Asian wickets it is only $50.1\%$.

\textbf{Day/Night matches.} The ball being harder to see at night afflicts both the night-time batting team, as well as the fielding team, making it prone to dropping catches. But the difficulty for batsmen is probably greater than it is for fielders. Indeed, within the shorter formats, the variation between day matches and day/night matches is greatest when it comes to the choice to bat first. In day/night matches, $59\%$ of toss-winning teams elect to bat first, whereas in day matches just about only half of them do.

\textbf{Seasonality.} The preference to bat first in England follows a broadly similar trend to the toss effect over the course of the year-- rising steadily over the summer, and drops in September. At its peak in August, top-winning captains choose to bat first $71.9\%$ of the time. At its lowest in April, they choose to bat $57.4\%$ of the time.

\textbf{Rain delays.} Remarkably, there is an overall preference to field first in matches that ended up using the Duckworth-Lewis adjustment: only $43.5\%$ of toss winning teams elected to bat in these matches. If captains are at all able to predict rain (and other) delays, they seem to show a preference to bat a fewer number of overs and chase the adjusted target.

\subsection{The Difficulty in Estimating the Advantage to Electing to Bat/Field First}

Despite having reported the `reduced form' and `first stage' effects of the toss, we cannot not interpret the ratio of these effects as the causal effect of batting first. If we used the toss as an instrument for batting first, say, we could at best hope to estimate the local average treatment effect (LATE) of electing to bat on winning the match \citep{angrist1996identification}. This is because in our setting  treatment effects likely vary across groups: some teams in some circumstances will be better able to capitalize on the ability to bat or field first.

However, even the conditions under which the LATE is known to be identified are probably not satisfied in our setting. To see why, first note that our setting has two-sided noncompliance: there are teams that win the toss that bat first, and those that lose the toss that bat first as well. Then, recall that under two-sided noncompliance, the LATE is identified under a monotonicity assumption that rules out defiers. However, defiers are likely to be present in our setting: they correspond to dyads of opposing teams that walk on to the pitch, both intending to field first if they win the toss. In such a setting, the team that wins the toss will chose to field first and the team loses the toss will be compelled to bat first. If these defiers are present, which they likely are, the monotonicity assumption does not hold, and so we cannot assert that the IV estimates are giving us the LATE.

We make these points a bit more concrete. Recall that we saw that 56.5\% of teams that won the toss elect to bat, which corresponds to an excess of 6.5 percentage points over the benchmark case in which the toss has no effect on the batting/fielding decision. In addition, the reduced form estimated effect of winning the toss on winning the match was 1.8 percentage points, implying that toss winners win an excess of 0.9 percent of matches over parity. The instrumented effect of batting first would then be estimated as 0.9/6.5 = 13.9 percentage points.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sing of this number gives the estimated instrumented effect of electing to field first.}  Can we interpret this as the average rate at which the complier teams that won the toss and elected to bat generated the excess wins? Not really, because these excess wins could have been generated by the defiers making good decisions to field. Since we saw that teams that win the toss and elect to field first have a higher win-loss differential than those that elect to bat first, it would be irresponsible to rule out this possibility. Consequently, it is unlikely the LATE can be identified in our setting.

% \section{Changing the Rules of the Toss}

% We end with a small but potentially important policy proposal that cricket boards might consider. The proposal is to eliminate the choice that the toss confers upon the toss-winning captain. How can this be done? The umpire assigns either of the two opposing captains to take heads, and flips the coin. If the coin lands heads, that captain must bat; otherwise, that captain must field. The captains do not get to make any choices.

% What does this take away from cricket? What problem does it fix? \todo{Guys, we should have a discussion about this!}

%
%This conversion rate is different in sign and considerably larger than the naive difference in win
%probabilities between teams that bat first and those that don't, which is $-1.6$ percentage points. Since the rules of cricket imply that the toss cannot affect the outcome of the match other than through the decision of the toss-winning team to bat or field, this discrepancy is likely due to the complier population being very different from the never-taker or always-taker populations that form the bulk of our team-match data. The IV estimates are thus
%telling us that some teams in some matches are able to take advantage of the decision to bat first when the batting choice really matters.
%
%Batting first is a strategic decision made by the captain of the team
%that won the toss in any given game. As such, a naive comparison of
%win rates between teams that chose to bat first versus those who
%didn't is unlikely to uncover the causal effect of batting first since
%the treatment (batting) is endogenous. To get around this problem,
%we use the randomly assigned ability to chose to bat first
%afforded to one of the two captains by virtue of winning the toss as
%an instrument for batting first. Note that because the decision to field first is the prefectly colinear inverse of the decision to bat, we can flip the sign of our point estimates to interpret them as the instrumented effects of choosing to field. 
%
%Since after winning the toss, a captain has the choice to bat field, instrumenting for batting first by the coin toss amounts to
%analysing an encouragement experiment with one-sided non-compliance
%. In this context, this corresponds to the outcome
%where the captain that won the toss still chooses to field first. The
%toss readily constitutes a valid instrument--- it
%gives the captain the choice to bat first, and almost certainly
%satisfies the exclusion restriction, as the toss is unlikely to have an independent effect
%on the outcome of the match besides its effect on the order that teams bat.\footnote{The only other possibility is that winning or losing the toss may have a dire psychological effect that directly affects their performance. We believe this to be unlikely.}
%Since we are in the ``just-identified'' case with a single endogenous
%regressor (the choice to bat first) and a single instrument
%(winning the toss), we can recover the causal instrumented effect of batting first as the ratio of the
%reduced-form effect of the toss on win probability divided by the first stage effect of the
%toss on the choice to bat.
%
%These assumptions are sufficient if we assume that the instrumented effect is constant in
%the population. However, we believe this to be an implausible assumption and
%therefore operate in the heterogeneous-effects framework proposed by
%\citet{angrist1996identification}. This necessitates an additional
%assumption--- that of {monotonicity}, which requires that the toss
%only moves batting choices in one direction: teams that would have
%otherwise chosen to bat first don't switch to fielding first upon
%winning the toss.\footnote{Formally, if $D_1$
%is the binary choice to bat upon winning the toss, while $D_0$ is the
%assignment when losing the toss, then monotonicity  assumption is stated as
%$\Prob{D_1 \geq D_0} = 1$.} Despite being unverifiable, this assumption is
%is likely to be valid. By using the instrumental variables (IV) estimator, we are therefore
%estimating the Local Average Treatment Effect (LATE) for
%{compliers} -- teams that are able to harness the
%available information about whether or not it is advantageous to bat
%first, and chooses to do so when it is in fact there is such an
%advantage.\footnote{An alternative formulation of the IV framework
%proposed by
%\citet{heckman2005structural} permits a
%characterisation of treatment selection based on a latent variable
%$U_i$, which yields a monotonicity assumption of the form $D =
%\Indic{U_i \leq p(\vee{X}_i, Z_i)}$, where $U_i$ is an unobserved
%propensity to take treatment (in our setting, choose to bat first
%given available information), and $p(\vee{X}_i, Z_i)$ is the propensity score.
%$U_i$ is normalised to be distributed on the unit interval and encodes
%information about an individual $i$'s treatment choices
%$\SetB{D_i(z)}_{z \in \Sett{Z}}$. This framework allows for
%extrapolation of the LATE into other target parameters.}
%
%We report the naive OLS estimate of the estimated difference in win
%rates in column 1 of Table \ref{table:iv_table}, which is
%counter-intuitively small and negative. In column 2 we estimate this
%specification separately across matches in which the toss-winner
%chooses to bat first, and in column 3 we estimate it in matches in which the toss-winner
%chooses to field first. We find that matches in which the
%toss-winner chooses to field first exhibit a substantial first-batter
%disadvantage, which suggests that captains are doing better than
%choosing randomly between batting and fielding first. We report the
%first stage effect of winning the toss on the choice to bat first in
%column 4, and find that winning the toss increases the probability
%of batting first by 13 percentage points, which is a sizeable effect.
%Columns 5-7 report the IV estimates. We find that in
%contrast to the naive OLS specification, the IV estimate gives a
%sizeable first-batter advantage of between 13.4 and 13.9 percentage
%points across specifications with team- and match-type fixed effects.
%
%Like with the toss, we then proceed to analyse heterogeneity in the instrumented effect. The results are depicted in Figures
%\ref{fig:iv_het_TE1} - \ref{fig:iv_het_TE3}.  
%

%
%
%\subsection{Characterising Compliers} % (fold)
%\label{sub:characterising_compliers}
%
%A complier in our setting is a team that is able to harness the
%available information about whether or not it is advantageous to bat
%first, and chooses to do so when it is in fact there is such an
%advantage. Since the monotonicity assumption rules out defiers,
%we can decompose the sample into {always takers}, {never
%takers}, and {compliers}
%\citep{angrist1996identification,Abadie2003-hj}. The fractions of each of these are given by 
%\begin{align*}
%\pi_{\text{compliers}} & = \Prob{D_1 > D_0 }= \Exp{D | Z = 1} - \Exp{D | Z = 0} \\
%\pi_{\text{always-takers}} & = \Prob{D_1 = D_0 = 1 } = \Exp{D | Z = 0}  \\
%\pi_{\text{never-takers}} & = \Prob{D_1 = D_0 = 0 } = 1 - \Exp{D | Z = 1}
%\end{align*}
%where $D_1$ is a dummy for batting upon winning the toss, $D_0$ is a dummy for batting first upon losing the toss, and $Z \in \{0,1\}$ is the outcome of the toss, with 1 coded as a win.  Applying the above formulae to our sample yields 13\% compliers,
%43.5\% always takers and 43.5\% never takers respectively. This is a small fraction of compilers, suggesting that teams are generally not very adept at capitalizing on the toss. However, it could also be that since a complier is a team-match, match conditions in which the toss would matter may not be that prevalent. We cannot tell. 
%
%To better understand the population of compliers, we apply \citeauthor{Abadie2003-hj}'s (2003) kappa to compute the mean of covariates for compliers. The kappa can be applied to any
%characteristic or outcome to get its mean for compliers by removing
%the means for never and always takers using the kappa weights. %\footnote{Abadie's formulation is
%%considerably more general and applies to any measurable function $g(Y,
%%D, \vee{X})$ with finite variance.
%%$$
%%\Exp{g(Y, D, \vee{X}) | D_1 > D_0} = \frac{\Exp{\kappa \cdot g(Y, D,
%%\vee{X})}}{\Prob{D_1 > D_0}} = \frac{\Exp{\kappa \cdot g(Y, D,
%%\vee{X})}}{\Exp{\kappa}}
%%$$. In our application, $g()$ simply returns $x_i$.
%% }. 
%
%Kappa weights are defined by
%$$
%\kappa = 1 - \frac{D(1-Z)}{1 - \Prob{Z = 1| X}} - \frac{(1-D)
%Z}{\Prob{Z=1|X}}
%$$ where $X$ is a characteristic. We average a characteristic $X$ using kappa weights computed using the
%above expression where $\Prob{Z = 1 | X}$  is estimated using a
%logistic regression \cite[p
%181-183]{angristMostlyHarmlessEconometrics2008}. Applying this procedure to some team-specific
%characteristics yields the averages reported in Tables 4 and 5. We find
%that compliers are marginally more likely to be underdogs (defined as
%the lower rated team in the match, according to ELO ratings), away teams, and teams playing in
%Asia. Among the teams that made the right decision to bat first after winning the toss, 78\% of them were playing day/night matches-- a remarkably high number. This suggests that it is relatively easier to make a good choice of whether to bat or field first in day/night matches. 
%
%Another result that is noteworthy, is that over a quarter of team-match compliers are Australian teams-- by far the highest frequency of any international team. England, Pakistan, South Africa, and India are all essentially tied for second place with a frequency of around 15\% each. What is surprising is how poorly West Indian, Sri Lankan and Kiwi teams do in choosing to bat when it matters. These are strong teams with quite a bit of cricketing experience in comparison to teams like Bangladesh, Kenya and Zimbabwe. And yet they do not seem to enter the complier population at higher rates.\footnote{For this interpretation, the match conditions are unlikely to be driving these differences. What we mean by this is the following. Suppose that Aussie teams are better at capitalizing on choosing to bat or field because Aussie wickets are particularly more likely to provide the playing conditions in which choosing to bat or field matters; and Aussie teams play on Aussie wickets more often than other teams. However, we see that matches played in Asian wickets constitute more than half of the complier population.} 
%
%%estimating the following
%%system of equations in order to uncover the batting advantage:
%%\begin{align}
%%Y_{ij} & = \alpha_0 + \beta D_{ij} + \varepsilon_{ij} \label{eqn:struct_eqn} \\
%%D_{ij} & = \gamma_0 + \delta Z_{ij} + \eta_{ij} \label{eqn:first_stage}
%%\end{align}
%%where $Y_{ij}$ is an indicator for the outcome of the match, $D_{ij}$ is the decision to bat first, $Z_{ij}$ is an indicator for winning the toss, $\alpha_0$ and $\gamma_0$ are constants, and $\varepsilon_{ij}$ and $\eta_{ij}$ are noise terms. $\delta$ is the first stage coefficient, and $\beta$ is the second stage coefficient of interest.
%
%
%% subsection characterising_compliers (end)
%
%% \section{Duckworth-Lewis} % (fold)
%% \label{sec:duckworth_lewis}
%
%% \subsection{Document effects of Duckworth Lewis} % (fold)
%% \label{sub:document_effects_of_duckworth_lewis}
%
%% \missingfigure{Event study Test Cricket and ODIs around 1999}
%
%% \subsubsection{Improving on DL} % (fold)
%% \label{ssub:improving_on_dl}
%
%% subsubsection improving_on_dl (end)
%
%% subsection document_effects_of_duckworth_lewis (end)
%
%% section improving_duckworth_lewis (end)
%

\section{Summing Up}

When the toss is unimportant, the game is fair: neither the batting side nor fielding side has an advantage derived simply from the luck of the toss prior to the first ball being bowled.  The toss appears to be unimportant in some settings, but important in others, like day/night matches. Although day/night matches are enjoyable, if we care about eliminating the influence of the toss, they should be avoided. Cricket pitches should be made durable, and with an eye to not disadvantaging teams that bat first or second. As for the much maligned Duckworth-Lewis formula, it looks as though there is no need to change it for now. We have no evidence that it is helping or harming teams that bat second.

\bibliography{biblio.bib}

\clearpage

\begin{figure}[tbh]
  \centering
  \includegraphics[width=0.6\textwidth,keepaspectratio]{output/matchcounts.pdf}
  \caption{Number of matches by format over time}
  \label{fig:timeseries_decomp}
\end{figure}

\begin{table}[b!]\centering
\begin{tabular}{llll}
 & Toss Wins & Matches & $p$ value\\ \hline
Home team & 3240 & 6377 & 0.201\\
Last toss winners & 17189 & 34283 & 0.611\\
Underdog & 2001 & 4054 & 0.423\\
Home Umpire &  2075	& 4044	& 0.098 \\
\end{tabular}
\caption{Balance \label{table: balance}}
\end{table}

% \clearpage

\begin{table}
\centering
\caption{Reduced Form Effect of Toss on win probability}
\input{output/reduced_form_table.tex}
\label{table:reduced_form}
\end{table}

\begin{figure}[b]
  \centering
  \figinc{output/reduced_form_by_matchtype.pdf}
  \caption{Effect of winning the toss on win probability decomposed across different
  match types, over time, and by continent.}
  \label{fig:rf_het_TE}
\end{figure}

% \clearpage

\begin{figure}[b]
  \centering
  \figinc{output/reduced_form_by_format_overtime.pdf}
  \caption{Effect of winning the toss on win probability decomposed by format over time}
  \label{fig:rf_het_TE2}
\end{figure}

\begin{figure}[b]
  \centering
  \figinc{output/reduced_form_by_rank_dl_season.pdf}
  \caption{Effect of winning the toss on win probability decomposed by competitiveness, use of DL, and Seasonality}
  \label{fig:rf_het_TE3}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{output/het_rank_te_odi.pdf}
    \caption{Treatment effect heterogeneity by ranking, toss and win history, and team dummies. Minimal evidence of effect heterogeneity}
    \label{fig:omni_het_odi}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{output/het_rank_te_test.pdf}
    \caption{Treatment effect heterogeneity for test matches by ranking, toss and win history, and team dummies. Minimal evidence of effect heterogeneity}
    \label{fig:omni_het_test}
\end{figure}

% \clearpage

\begin{table}
\scriptsize
\centering
\caption{Batting first and win probability}
\input{output/iv_table.tex}
\label{table:iv_table}
\end{table}

\begin{figure}
 \centering
 \figinc{output/first_stage_by_matchtype.pdf}
 \caption{Toss effects on choice to bat decomposed across different match
 types, timing, and continent.}
 \label{fig:fs_het_TE1}
\end{figure}

\begin{figure}
 \centering
 \includegraphics[width=\textwidth,keepaspectratio]{output/first_stage_by_format_overtime.pdf}
 \caption{Toss effects on choice to bat decomposed across different formats
 over time}
 \label{fig:fs_het_TE2}
\end{figure}

\begin{figure}
 \centering
 \includegraphics[width=0.9\textwidth,keepaspectratio]{output/first_stage_by_rank_dl_season.pdf}
 \caption{Toss effects on choice to bat decomposed by ranking differences, rain-interruptions, and season (in England)}
 \label{fig:fs_het_TE3}
\end{figure}

%\begin{figure}
%  \centering
%  \includegraphics[width=0.9\textwidth,keepaspectratio]{output/first_stage_season.pdf}
%  \caption{First stage (probability of choosing to bat first after winning the toss) by month in England}
%  \label{fig:first_stage_seasonality}
%\end{figure}
%
%
%\clearpage
%
%
%\begin{table}[!htb]
%    \caption*{}
%    \begin{minipage}{.5\linewidth}
%      \caption{Team Characteristics}
%      \centering
%        \begin{tabular}{l|r}
%        % \label{table:kappa_res1}
%        \hline
%        & $\Ol{X}$\\
%        \hline
%        underdog & 0.52\\
%        \hline
%        day/night & 0.78\\
%        \hline
%        home & 0.42\\ \hline \hline \\
%        \textbf{continent} & \\
%        \hline
%        africa & 0.11\\
%        \hline
%        asia & 0.62\\
%        \hline
%        europe & 0.03\\
%        \hline
%        oceania & 0.24\\
%        \hline
%        \end{tabular}
%    \end{minipage}%
%    \begin{minipage}{.5\linewidth}
%      \centering
%        \caption{Team}
%        \begin{tabular}{l|r}
%        % \label{table:kappa_res2}
%        \hline
%        team & $\Ol{X}$\\
%        \hline
%        australia & 0.27\\
%        \hline
%        bangladesh & 0.01\\
%        \hline
%        england & 0.15\\
%        \hline
%        india & 0.15\\
%        \hline
%        kenya & 0.02\\
%        \hline
%        new zealand & 0.00\\
%        \hline
%        pakistan & 0.15\\
%        \hline
%        south africa & 0.14\\
%        \hline
%        sri lanka & 0.03\\
%        \hline
%        west indies & 0.04\\
%        \hline
%        zimbabwe & 0.04\\
%        \hline
%        \end{tabular}
%    \end{minipage}
%\end{table}
%
%

%\renewcommand{\mkbibnamefamily}[1]{\textsc{#1}}
%\printbibliography
%
%\appendix

\end{document}

### 2023-01-26 ms/ms.tex
fix abstract

@@
 
-The rules in cricket treat the competing teams asymmetrically, even before the start of physical play.
-In the one-day version of the game, for instance, which team bats first is determined by the captain that wins the toss.
-This raises questions about the effects of the asymmetry created by luck.
+In cricket, which team bats first is determined by a toss.
 Does the team that wins the toss have an advantage over the other team?
-
 We find that, on average, the effect of winning the toss on winning the match is essentially nil in international limited-overs daytime matches but substantial (3.3 percentage points) in day/night matches.
-In test cricket, the advantage is even larger (4.9 percentage points).
+In Test cricket, the advantage is even larger (4.9 percentage points).
 The advantage has exploded in the last decade (to nearly 15 percentage points) after a steady decline.
-More surprisingly, the effect of the toss does not appear to vary between competitive and noncompetitive pairings of teams.
-And as for the decision to bat or field first, we find the win-loss differential is significantly larger for the teams that elect to field than those that elect to bat, in which case it is negligible.
-Lastly, and perhaps most surprisingly, in rain-delayed matches in which the chasing team's total was adjusted by the Duckworth-Lewis formula, we find that teams that elected to bat first did just as well than those that elected to field first.
+Surprisingly, the effect of the toss does not appear to vary between competitive and non-competitive pairings of teams.
+And as for the decision to bat or field first, we find the win-loss differential is significantly larger for teams that elect to field; the advantage is negligible among those that elect to bat.
+Lastly, and perhaps most surprisingly, in rain-delayed matches in which the chasing team's total is adjusted by the Duckworth-Lewis formula, we find that teams that elected to bat first did just as well as those that elected to field first.
 

### 2023-01-27 ms/ms.tex
revised ms

@@
 
-\title{\textbf{Asymmetries in Cricket \\ \vspace{5pt} \large{The Effects of Winning the Toss on Winning the Match}}\thanks{This paper incorporates some of the text and 
-results from Sood \& Willis's (2019) working paper: ``Fairly Random: The Impact of Winning the Toss on the Probability of Winning.'' The authors thank Matt Blackwell, 
-Andrew Gelman, Donald Green, Jens Hainmueller, and Daniel Stone for their comments and suggestions. We are grateful to Sajan Mehotra for outstanding research assistance. 
-% The first three authors contributed roughly equally to the research design, data collection, and implementation of the whole project, and are listed alphabetically. Acharya
-% contributed to the research design and interpretation of results.
-}}
+\title{Asymmetries in Cricket: The Effects of Winning the Toss on Winning the Match\thanks{This paper incorporates some of the text and results from Sood \& Willis's (2019) working paper: ``Fairly Random: The Impact of Winning the Toss on the Probability of Winning.'' The authors thank Matt Blackwell, Andrew Gelman, Donald Green, Jens Hainmueller, and Daniel Stone for their comments.}}
 
@@
 
-%
 \section{Introduction}
-In a competitive sport, every little thing matters. Yet, many sports leave some large levers out of the reach of the teams and in the
-hands of fate.
+
+In competitive sports, every little thing matters.
+Nevertheless, many sports leave some key levers out of teams' control and in the hands of fate.
 In cricket, one such lever---the toss---has been the subject of much attention.
 In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview.
-To wrest the advantage, the captain merely needs to pick the side of the coin that will be left facing the sky after the toss and choose whichever one---batting second or batting second---gives the greater advantage.
+To wrest the advantage, the captain merely needs to pick the side of the coin that will be left facing the sky after the toss and choose whichever one---batting first or batting second---gives the greater advantage.
 
 In this paper, we measure how well teams have been able to capitalize on any advantage conferred by winning the toss.
-Since the only consequence of the toss is letting the toss-winning side decide whether to bat or field first, we also attempt to measure the advantage of electing to bat first.
-To answer these questions, we collected data on over 35,000 cricket matches across four formats (First Class, ODIs, T20s, and tests) played over the course of more than one hundred fifty years.
-(To our knowledge, this is the most comprehensive professional cricket match dataset to date.)
+To answer the question, we collected data on over 35,000 cricket matches across four formats (First Class, ODIs, T20s, and Tests) played over the course of more than one hundred fifty years.
 
-Winning the toss increases the chances of winning the match by 1.8 percentage points: toss winners win 39.2\% of the time and lose 37.4\% of the time, with the remaining share being draws or ties.
-The effect of winning the toss is the largest in tests, where it is 4.9 percentage points and is smallest in ODIs and T20s, where it is essentially zero.
-That being said, in ODIs and T20s that are day/night matches, the effect of winning the toss is 3.3 percentage points, whereas, in the day-time matches, the effect is a precise zero.
+We find that winning the toss increases the chances of winning the match by 1.8 percentage points: toss winners prevail 39.2\% of the time and lose 37.4\% of the time; the rest of the matches are drawn or tied.
+The effect of winning the toss is the largest in Tests, where it is 4.9 percentage points and the smallest in ODIs and T20s, where it is essentially zero on average.
+That said, in day/night ODIs and T20s, the effect of winning the toss is 3.3 percentage points; in day-time matches, the effect is a precise zero.
 
-The effect of toss in the ODIs has remained relatively stable since the 1980's when the one-day format first became popular.
-In tests, however, the importance of the toss has varied considerably.
-The effect of winning the toss in test cricket was a little over five percentage points before 1990.
-It then dropped nearly to 0 from 1990 to 2010.
+The effect of the toss in ODIs has remained relatively stable since the 1980's when the one-day format first became popular.
+In Tests, however, the importance of the toss has varied considerably.
+The effect of winning the toss in Test cricket was a little over five percentage points before 1990.
+It then dropped to almost 0 from 1990 to 2010.
 In the last decade, however, it has exploded, averaging 14.5 percentage points.
-Counter to conventional wisdom, the effect of winning the toss is relatively similar across continents, though the toss appears less important on European and African wickets than elsewhere.
-The effect on Aussie and Kiwi wickets (3 percentage points) is not substantially different from its effect on South Asian wickets (2.8 percentage points) but is higher than its effect on European and African wickets (1.1 and 0 percentage points respectively).
 
-In sports like soccer, the conventional wisdom is that the toss is inconsequential \citep{csato2020comparison}.
-In cricket, it is believed that the toss has little significance in lopsided matches that have clear favorites and underdogs.
-For example, it is thought to be less relevant in an India-Ireland match than in an India-Australia match.
-For this reason, we look at the effect of the toss separately in subsamples where teams are similarly ranked according to ICC ELO ratings and where their ratings differ by more substantially.
-We find that the effect of the toss doesn't vary by how well matched the teams are.
-Since ELO ratings are available only for matches played in recent decades, these findings (along with our finding that influence of the toss has declined over time) suggest that the toss has little influence in modern cricket.
-
-An important caveat about the null findings is in order, however.
-Our empirical approach cannot tell whether the effects are nil because the toss is unimportant or because the toss is important, but that (modern) teams have not been able to capitalize on the advantage.
-The toss may matter quite a bit in principle, but we would only be able to observe its effect if a team is able to use the information on the conditions of play and how it pairs strategically against the opposing team to be able to make the optimal decision of whether to bat or field first.
-If teams simply do not have this knowledge and are often choosing poorly, then we would not see any ``toss effects.'} So, when we refer to ``toss effects'' in this paper, what we are measuring is not whether there is an advantage to winning the toss in principle, but whether any such advantage is exploited in practice.
-
-\footnote{The median rating is 17, so a cutoff of 15 roughly splits the sample of matches in which we have ratings for both teams in two halves.}
-
-%Another way to benchmark the effect of the toss is to compute it in
-%terms of ``run- or wicket-advantage.'' To do this, we consider ODI
-%and T20 matches, and estimate [.Apoorva: describe the methodology
-%that we discussed here]. We find that the effect of winning the toss
-%and electing to bat first is an [X] run advantage. In other words,
-%the effect of the toss in these instances is nullified only by
-%allowing the chasing team to start their innings with [X] runs
-%already on the board. Similarly, the effect of winning the toss and
-%electing to field first is a [Y] run advantage, so that the effect is
-%nullified only if we make the chasing team start with [-Y] runs on
-%the board. \todo{We had an idea here that I vaguely recall of how to
-%address the complication that if a team loses all ten wickets, then
-%we don't observe its 50 over total.}
+Contrary to conventional wisdom, the effect of winning the toss does not vary dramatically across continents, though the toss appears less critical on European and African wickets than elsewhere.
+The effect on Aussie and Kiwi wickets (3 percentage points) is not substantially different from its effect on South Asian wickets (2.8 percentage points) but is higher than on European and African wickets (1.1 and 0 percentage points respectively).
 
@@
 This suggests that teams are better able to capitalize on the toss by electing to field when it is advantageous to do so.
-%
-%Combining this with the fact that we saw the reduced form effect of the toss on the outcome of the match to be 1.8 percentage points, we can uncover the average effect of batting first on the so-called ``complier'' teams: those that got to bat (because they won the toss) that would not have batted if batting first were randomly assigned \citep{angrist1996identification}.
-Since winning the toss results in a 0.9\% excess of match wins (from the benchmark case in which the toss has no effect), the effect of batting first in this complier population is 0.9/6.5 = 13.9 percentage points.
-This is the instrumented effect of winning the toss-- or, the rate at which the excess decisions to bat were converted into excess wins.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sings of all of our estimates gives the instrumented effect of electing to field first.
-For example, the instrumented effect of electing to field first is -13.9 percentage points.}
-%
-%This conversion rate is different in sign and considerably larger than the naive difference in win
-%probabilities between teams that bat first and those that don't, which is $-1.6$ percentage points.
-Since the rules of cricket imply that the toss cannot affect the outcome of the match other than through the decision of the toss-winning team to bat or field, this discrepancy is likely due to the complier population being very different from the never-taker or always-taker populations that form the bulk of our team-match data.
-The IV estimates are thus
-%telling us that some teams in some matches are able to take advantage of the decision to bat first when the batting choice really matters.
 
 We also look at how the toss advantage breaks down between teams that elected to bat versus those that elected to field within the various subgroups mentioned above and others.
-Perhaps the most striking result from this analysis for many cricket followers is that there does not appear to be the case that toss winners that elect to bat first do better than those that elected to field first in rain delayed matches where the chasing total is adjusted according to the \citet{duckworth1998} formula.
+Perhaps the most striking result from this analysis for many cricket followers is that there does not appear to be the case that toss winners that elect to bat first do better than those that elect to field first in rain delayed matches where the chasing total is adjusted according to the \citet{duckworth1998} formula.
 
-Our results are pertinent to debates on how cricket should be governed.
+Lastly, a caveat about our findings.
+Our empirical approach estimates the extent to which teams have been able to capitalize on the advantage granted by winning the toss.
+The toss may matter a lot in principle, but teams may decide poorly.
+So, when we refer to ``toss effects'' in this paper, we do not refer to whether there is an advantage to winning the toss in principle, whether any such advantage is exploited in practice.
+
+Our results are pertinent to debates about how cricket should be governed.
 One argument is that since the toss is random, no team is advantaged by the format prior to the toss; and in fact the luck of the toss along with the strategic choice to bat first or second are integral parts of the game--- not very different from the luck of weather affecting the teams differently.
-%or the strategic choices that captains make as to how to set the field, or the order in which they deploy their batsmen.
+
 An alternative argument, however, is that the toss is problematic because it advantages one team over the other even before the first ball is bowled; and care should be taken to mitigate this asymmetry--- for example, by making cricket pitches more durable so that they don't deteriorate so much so that batting on the final day of a test match is miserable.
@@
 We do not address these optimal design questions in this paper, but our results are relevant to these issues depending on what the designers of cricket seek to optimize.
-%However, at the end of the paper, we propose (and discuss the possible consequences of) a subtle change to the rules governing the toss: Rather than tossing a coin to confer the choice to bat first or second upon the captain of the toss-winning team, umpires may consider simply tossing a coin to determine which team must bat first.
+However, at the end of the paper, we propose (and discuss the possible consequences of) a subtle change to the rules governing the toss: Rather than tossing a coin to confer the choice to bat first or second upon the captain of the toss-winning team, umpires may consider simply tossing a coin to determine which team must bat first.
 
@@
 
-We obtained data on cricket matches from the ESPN Cricinfo website, which provides coverage of international and many domestic matches,
-and maintains statistics on a wide variety of matches across formats.\footnote{Information on all available matches was downloaded in June 2020 using a purpose-built parsing library available at \href{https://github.com/dwillis/python-espncricinfo}{https://github.com/dwillis/python-espncricinfo}.
-The data will be available as a separate repository available at \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}
+We obtained data on cricket matches from the ESPNCricinfo website, which covers international and domestic matches, and maintains statistics on a wide variety of matches across formats.\footnote{Information on the matches was downloaded in June 2020 using a purpose-built library available at \href{https://github.com/dwillis/python-espncricinfo}{https://github.com/dwillis/python-espncricinfo}.
+The data is available at \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}
 
-We focus on four types of formats: T20, ODI, Test, and First Class matches for both men's and women's teams.\footnote{We do not include the List A matches that are not ODIs or T20s, which are often considered a less professional category (with worse data recording) than the formats that we include.} The data are a near census of the relevant population, excluding only scheduled matches that were abandoned without the toss being conducted, and a small number of limited-overs matches in which the minimum number of overs that must be bowled to establish a result was not met.\footnote{In ODIs, for example, each side must bat at least 20 overs for a result to be declared.
+We focus on four formats: T20, ODI, Test, and First Class for both men's and women's teams.\footnote{We exclude List A matches that are not ODIs or T20s.
+These matches are considered a less professional category (with worse data recording) than the formats that we include.} The data are a near census of the relevant population, only excluding scheduled matches that were abandoned without the toss being conducted, and a small number of limited-overs matches in which the minimum number of overs that must be bowled to establish a result were not met.\footnote{In ODIs, for example, each side must bat at least 20 overs for a result to be declared.
 We have no reason to suspect that the small matches that were excluded because they did not meet this threshold could be biased enough to affect our conclusions.
-We drop 426 matches recorded as ``no result,'' 3 matches that are recorded as ``abandoned'' or ``canceled,'' and another three that don't record who bat first.
-The former constitutes the largest exclusion we make and appears to be comprised of matches canceled for idiosyncratic reasons, ranging from the \href{https://www.espncricinfo.com/series/shell-tri-series-1991-92-61203/australia-women-vs-england-women-final-66975/full-scorecard}{a team being unable to bat because of the weather} to cancellations because of \href{https://www.espncricinfo.com/series/australia-tour-of-pakistan-1982-83-61392/pakistan-vs-australia-3rd-odi-64196/full-scorecard}{unwanted audience participation}.
-This results in us dropping approximately 0.07\% of matches.}
+We drop 426 matches recorded as ``no result,'' 3 matches that are recorded as ``abandoned'' or ``canceled,'' and another three that don't record who batted first.
+The "no result" matches constitute the largest exclusion we make.
+They appear to be comprised of matches canceled for idiosyncratic reasons, ranging from the \href{https://www.espncricinfo.com/series/shell-tri-series-1991-92-61203/australia-women-vs-england-women-final-66975/full-scorecard}{a team being unable to bat because of the weather} to cancellations because of \href{https://www.espncricinfo.com/series/australia-tour-of-pakistan-1982-83-61392/pakistan-vs-australia-3rd-odi-64196/full-scorecard}{unwanted audience participation}.
+In all, we drop approximately 0.07\% of matches.}
 
-Our final sample comprises of 35,357 cricket matches across four formats played over the last 150 years.
+Our final sample comprises 35,357 cricket matches across four formats played over the last 150 years.
 We report the distribution of matches over time and by type in Figure \ref{fig:timeseries_decomp}.
-As the figure shows, the number of matches has exploded over the last two decades, largely due to the introduction of the enormously popular Twenty-20 (T20) format of the game both in both international and domestic tournaments.
+As the figure shows, the number of matches has exploded over the last two decades, largely due to the introduction of the enormously popular Twenty-20 (T20) format of the game both in international and domestic tournaments.
 T20 matches constitute over a quarter of all matches played in the last decade.
@@
 
-Since our analysis will be based on the assumption that the coin toss is random, we start by examining this assumption.\footnote{Prior work in the sports analytics of cricket by \citet{shafqat2015analysis} and \citet{dawson2009bat} conditions on post-treatment variables, whereas we use only the randomization of the toss to identify toss effects.} Is there any evidence that the coin toss is somehow rigged--- for example, with the home side enjoying the rub of the green more often?
+Since our analysis rests on the assumption that the coin toss is random, we start by examining this assumption.\footnote{Prior work by \citet{shafqat2015analysis} and \citet{dawson2009bat} conditions on post-treatment variables, whereas we use only the randomization of the toss to identify toss effects.} Is there any evidence that the coin toss is rigged---for example, does the home side enjoy the rub of the green more often?
 In order to study this, we compare the rates at which home teams, teams that won the toss in their last match, and underdogs (defined as the lower-ranked team whenever rankings are available for both teams) win the toss relative to their opponents.
@@
 To examine whether the toss is rigged, we compute the odds of seeing as many successes in the same number of draws of a sequence of independent Bernoulli random variables with parameter 0.5.
-For example, home teams have won 50.81\% of tosses in the 6377 matches for which we have data.
-The likelihood of tossing a fair coin 6377 times and getting at least this percentage of heads is 20.1\%.
-This chances appears low, but not eyebrow-raisingly so.
+For example, home teams have won 50.81\% of tosses in the 6,377 matches for which we have data.
+The likelihood of tossing a fair coin 6,377 times and getting at least this percentage of heads is 20.1\%.
+This chance appears low, but not eyebrow-raisingly so.
 We find that toss wins are even more balanced for teams that won and lost the last toss, and for teams that are underdogs and favorites.
@@
 That said, we do see one potential source of concern.
-We find that teams that share their nationality with at least two of the three umpires of the match win 51.3\% tosses in the 4044 international matches for which we have this data.
-The likelihood that tossing a fair coin 4044 times and and getting at least this share of heads is only 10\%.
+We find that teams that share their nationality with at least two of the three umpires of the match win 51.3\% tosses in the 4,044 international matches for which we have this data.
+The likelihood of tossing a fair coin 4,044 times and getting at least this share of heads is only 10\%.
 This number is certainly low enough to raise some eyebrows.
 That said, the practice of allowing home-team umpires in international matches ended in 1992, and so these issues are less relevant in recent decades which comprise the bulk of our sample.
-
-In addition, a rigged toss in favor of the home team would bias estimates of the advantage of winning the coin toss to the extent that it is
-correlated with ability.
+In addition, a rigged toss in favor of the home team would bias estimates of the advantage of winning the coin toss to the extent that it is correlated with ability.
 If stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards.
 And vice versa, if otherwise.
-We do not have good reasons to think that there is a correlation.
-So for now, we proceed as if the toss is random.
-
-% robust standard errors and report these estimates in
-% table~\ref{table:baltab} alongside the number of matches used to test
-% for balance. Wefail to reject the null that the toss is random.
-
-%N = 6377
-%Wins = 3240 home team wins
-% prob succes = 50.81
-% p value = 20.14\%.
-
-%Wins = 17189
-%N = 34283
-%50.14
-%p value = 0.6114
-
-%Underdog = lower ranked team in any dyad with nonmissing ranking info (so int'l matches)
-%Underdog won = 2001
-%N = 4054
-%Win prob = 49.36\%
-%p value = 0.4231
-
-% \input{output/baltab.tex}
+Given home teams win more matches and are seen as 'stronger' at home, it is possible that the results are slightly biased upwards.
+We investigate this point later in the paper, looking at effects post 1992.
+For now, we proceed as if the toss was random.
 
@@
 For the main specifications of interest, we estimate regression equations of the form
+
 \begin{equation}\label{eqn:reduced_form}
-Y_{ij} = \alpha_0 + \tau Z_{ij} + \gamma_i + \delta_t +
-\psi_k + \varepsilon_{ij}
+Y_{ij} = \alpha_0 + \tau Z_{ij} + \gamma_i + \delta_t + \psi_k + \varepsilon_{ij}
 \end{equation}
-which includes fixed effects for teams $\gamma_i$, time $\delta_t$, and match-type $\psi_k$, so that all variables can be interpreted as residuals from partialing out team, match-type, and time fixed effects.
-Since unobservable factors are correlated across opposing teams in the same match, we cluster the standard errors at the match level.
+
+which include fixed effects for teams $\gamma_i$, time $\delta_t$, and match-type $\psi_k$. We cluster the standard errors at the match level.
 
 We report results in Table \ref{table:reduced_form}.
-We progressively introduce team, match-type, and time period fixed-effects in columns 2, 3, and 4 to make more fine-grained comparisons, and find that results are remarkably stable across specifications.
-The first column, for example gives
-the baseline estimate of the toss effect being 1.8 percentage points in all matches.
-As we mentioned in the introduction, this effect represents toss winners winning 39.2\% of the time and losing 37.4\% of the time, with the remaining matches being draws.
-The estimated effect remains at 1.7 percentage points when we progressively add the fixed effects.
+We progressively introduce team, match-type, and time period fixed effects in columns 2, 3, and 4, and find that results are remarkably stable.
+The first column gives the baseline estimate of the toss effect.
+The 1.8 percentage points difference reflects the fact that toss winners win 39.2\% of the time and lose 37.4\% of the time; the remaining matches are drawn.
+The estimated effect remains at 1.7 percentage points when we progressively add various fixed effects.
 
-We then proceed to examine the effect of the toss withing various subgroups.
-We report the estimates graphically in Figures \ref{fig:rf_het_TE} - \ref{fig:omni_het_test} with point estimates and sample sizes as annotations.
-The results are summarized as follows.
+We then proceed to examine the effect of the toss within various subgroups.
+We report the estimates graphically in Figures \ref{fig:rf_het_TE} - 4 with point estimates and sample sizes as annotations.
+We summarize the results below:
 
-\textbf{Format.} Toss effects are most pronounced in test matches followed by first class matches, and are statistically indistinguishable from zero in ODIs and T20s.
-We estimate the effect of the toss on winning a test match to be 4.7 percentage points, allowing for the fact that many tests end in draws.
-Given that only 65\% of test matches end with one team winning, the effect of the toss conditional on the match producing a winner is higher: 7.2 percentage points.
+\textbf{Format.} Toss effects are most pronounced in Test matches followed by first class matches, and are statistically indistinguishable from zero in ODIs and T20s.
+We estimate the effect of the toss on winning a Test match to be 4.7 percentage points.
 The effect of the toss in first class matches is 2.3 percentage points.
 These results support the conventional wisdom among lay cricket followers that the toss grants the greatest advantage in multi-day affairs.
-Pitches invariably deteriorate over the course of these matches, and the fourth innings in test match is often the most challenging time to bat.
+Pitches invariably deteriorate over the course of these matches, and the fourth innings in a Test match is often the most challenging time to bat.
 The pitch deteriorates far less over the course of a day, or in the case of T20s, a few hours.
@@
 \textbf{International vs.~Domestic matches.} Toss effects are virtually the same for international and domestic matches (point estimates are 1.7 percentage points in the former, and 1.9 percentage points in the latter).
-This is a bit surprising given differential use of cricket formats differs in international and domestic matches (for example, there are no domestic tests), and our findings of heterogeneous effects across formats.
 
-\textbf{Effects over time.} Toss effects overall appear to have declined over time from a high of 9.9 percentage points in the second half of the 19th century, to 1.4 percentage points in the last decade.
-We also look at these trends within format.
+\textbf{Effects over time.} Toss effects appear to have declined over time from a high of 9.9 percentage points in the second half of the 19th century to 1.4 percentage points in the last decade.
+This trend may be a result of a changing composition of matches so we also looked at these trends within formats.
 T20s have not been played long enough for us to draw any conclusions about how toss effects have changed over time.
-But within the other formats, we find that toss effects follow a similar pattern, except that in test cricket there is a peculiar uptick in in the last decade: the toss effect jumps from 0 in the decade before to 14.5 percentage points now.
-We investigate potential sources of this variation below  but there are some factors that could be important drivers for which we do not have data.
-For example, cricketers may have become more fit, and able to play for longer periods without tiring as much as they once did.
-Pitches now may be more resilient to wear than they once were.
-Teams may have acquired better knowledge on how to play more effectively whether they bat first or second.
-Etc.
+But within the other formats, we find that toss effects follow a similar pattern, except that in Test cricket there is a peculiar uptick in the last decade: the toss effect jumps from 0 in the decade before to 14.5 percentage points now.
+We investigate potential sources of the ecological decline below but there are some factors that could be important drivers but for which we do not have the data.
+For example, cricketers may have become fitter, and able to play for longer periods without tiring as much as they once did, reducing the impact of the order in which they ball or bat.
+Pitches may have become more resilient to wear than they once were.
+Teams may have become better informed about how to play more effectively whether they bat first or second, etc.
 
-\textbf{Relative strength.} Common sense suggests that the toss should be less important in lopsided matches where there is one dominant team that is expected to win, whether it bats first or second.
-It should be more important in competitive matches where the opposing teams are more even in cricketing strength.
-We define competitive matches as those in which the gap in ELO ratings of the two teams is less than 15 points.
-Since the median ELO rating difference is 17 points, this threshold roughly splits our sample of matches for which we have ELO ratings for both teams in two halves.
-Surprisingly, we find the toss effect to be nearly the same, and indistinguishable from zero within both competitive and noncompetitive matches even when we look at tests and international ODIs separately (these are the two formats for which ELO ratings apply).
-Part of the reason for this is that the ELO ratings are available only for recent years, and we saw above that toss effects have declined over time.
-We also perform an omnibus test of treatment effect heterogeneity by estimating sorted partial effects \citep{chernozhukov2018sorted,chen2019sortedeffects} with respect to ranking and result history, and find minimal evidence for effect heterogeneity.
+\textbf{Relative strength.} We expect the toss to be less important in matches where one team is clearly stronger than the other.
+We expect the stronger team to win, whether it bats first or second.
+To test this hypothesis, we use team ranking data published by cricket's official governing body, the International Cricket Council (ICC).
+The median match is played between teams where the ICC rating difference is 17 points.
+We use that as the approximate splitting point, splitting the matches into teams where the gap in the ICC ratings of the two teams is less than 15 points and where it is more.
+In this case, we believe the standard errors are too large for us to arrive at any substantively meaningful conclusions.
+Statistically, we believe effects larger than 8\% are unlikely.
 
-\textbf{Continents.} We find that toss effects vary only somewhat across continents, hovering around 2.5 percentage points for matches played in Asia, Oceania, and the Americas (though the estimates for the Americas are noisier in part due to the smaller number of matches played there).
-In Europe (mainly England) the toss effect is only 1.1 percentage points (though we do not rule out a null effect), and in Africa it is essentially zero.
-Overall, despite there being much talk about the different nature of wickets across the world, with wickets in certain regions being drier, or stickier, or subject to more wear, these differences do not seem to translate into substantial toss effect differences at least when we look coarsely at averages.
+\textbf{Continents.} Toss effects vary only somewhat across continents, hovering around 2.5 percentage points for matches played in Asia, Oceania, and the Americas (though the estimates for the Americas are noisier in part due to the smaller number of matches played there).
+In Europe (effectively England) the toss effect is only 1.1 percentage points (though we cannot rule out a null effect), and in Africa it is essentially zero.
+Overall, despite there being much talk about the different kinds of wickets across the world, with wickets in certain regions being drier, or stickier, or subject to more wear, these differences do not seem to translate into substantial toss effect differences at least when we look at coarse geographic averages.
 
@@
 In the limited overs formats where day/night matches take place (ODIs and T20s), the advantage of winning the toss in a day/night match is 3.3 percentage points, whereas the
+
 advantage of winning the toss in a limited overs match played during the day is zero.
 
-\textbf{Seasonality.} Weather is also thought to have a large impact on the playing conditions in cricket. Cool
-overcast weather, for example, is thought to aid swing bowling, especially in English pitches. More generally,
-the advantage of winning the toss likely varies by the weather. Although do not have data on
-the weather, we can proxy it with seasons. In particular, students of the game suspect that
-the advantage of winning the toss in the early English season is especially significant.
-There is some evidence of a seasonal pattern for matches played in England, with the advantage of winning the toss
-growing over the summer, though the trend is weak and broken in the month of July.
+\textbf{Seasonality.} Weather is also thought to have a large impact on the playing conditions in cricket.
+Cool overcast weather, for example, is thought to aid swing bowling, especially on English pitches.
+More generally, the advantage of winning the toss likely varies by weather.
+Although we do not have data on the weather, we can proxy it with seasons.
+In particular, students of the game suspect that the advantage of winning the toss in the early English season is especially significant.
+There is some evidence of a seasonal pattern for matches played in England, with the advantage of winning the toss growing over the summer, though the trend is weak and broken in the month of July.
 
@@
 This method has been the subject of much controversy, with many cricket fans suggesting that it is rigged against the chasing team.
-To our surprise, we find no evidence that toss winners are able to exploit the possibility of second inning rain delays to their advantage.
-If anything, we see that the toss-winner has a \emph{disadvantage} of 2.4 percentage points in matches subject to the Duckworth-Lewis adjustment, though this point estimate is statistically indistinguishable from zero \footnote{Toss-winners won 160 out of 333 ODIs with DL adjustment and 3085 out of 6187 ODIs without DL adjustment}.
-This, however, does not mean that the Duckworth-Lewis method is not rigged against the chasing team; that issue will be discussed further in the next section.
-
-% \begin{table}
-% \centering
-% \caption{Reduced Form Effect of Toss on win probability by match type}
-% \input{output/reduced_form_het.tex}
-% \label{table:reduced_form_het}
-% \end{table}
+To our surprise, we find no evidence that toss winners are able to exploit the possibility of second-inning rain delays to their advantage.
+If anything, we see that the toss-winner has a \emph{disadvantage} of 2.4 percentage points in matches subject to the Duckworth-Lewis adjustment, though this point estimate is statistically indistinguishable from zero.\footnote{Toss-winners won 160 out of 333 ODIs with the Duckworth-Lewis adjustment (i.e.~48\%) and 3085 out of 6187 ODIs without the Duckworth-Lewis adjustment (49.9\%).} This, however, does not mean that the Duckworth-Lewis method is not rigged against the chasing team; that issue will be discussed further in the next section.
 
 % subsubsection heterogeneous_treatment_effects (end)
-
-% \newpage
 
@@
 
-\subsection{The Choice to Bat}
-
+56.5\% of teams that won the toss elected to bat.
+This is not surprising given the conventional wisdom is that when you are not sure about whether to bat or field first, bat first.
 The first column of Table \ref{table:iv_table} shows that 49.2\% of teams that batted first won the match, while 50.8\% of teams that fielded first won.
@@
 The third column of the table shows that these teams won 3.9 percentage points more frequently than they lost: they won 40.8\% of their matches and lost 36.9\% of them.
-Combining this information with the fact that 56.5\% of teams that won the toss elected to bat (and only 43.5\% elected to field, as implied by the fourth column of the table) this means that 94\% ($= .435 \times .039/.018$) of the 1.8 percentage point advantage to winning the toss that we reported in the previous section can be attributed to the success of toss-winning teams that elected to field.
+Combining this information with the fact that 56.5\% of teams that won the toss elected to bat (and only 43.5\% elected to field, as implied by the fourth column of the table), we learn that 94\% ($= .435 \times .039/.018$) of the 1.8 percentage point advantage to winning the toss that we reported in the previous section can be attributed to the success of toss-winning teams that elected to field.
+To further understand what explains the preference to bat or bowl first, we see how winning the toss affects the choice in various subsets:
 
-These findings are not surprising given the conventional wisdom that captains who are not sure whether they should bat or field first often choose to bat.
-If a team wins the toss and elects to field, they often have a good reason to do so.
-It is interesting to ask what these reasons could be.
-Although we cannot answer this question fully, comparing  the win-loss differentials within subgroups may offer some clues.
+\textbf{Format.} Teams appear to have a strong preference for batting first in longer match formats.
+In Test matches 72.8\% of teams that win the toss elected to bat, whereas in T20s and ODIs this share is only about 52.9\%.
+In first-class matches it is 56.9\%.
 
-\textbf{Format.} In test matches 72.8\% of teams that win the toss elect to bat, whereas in T20s and ODIs this share is only about 52.9\%.
-In first class matches it is 56.9\%.
-It appears that teams err heavily on the side of batting first in longer match formats.
+\textbf{International vs. Domestic matches.} We see no big differences in the propensity to elect to bat first across international versus domestic matches.
 
-\textbf{International vs. Domestic matches.} We see no big differences in propensity to elect to bat first across international versus domestic matches.
-
-\textbf{Effects over time.} Historically, toss-winning teams tilted heavily in favor of batting first, but this declines over the course of the 20th century, and now they show no preference (on average) to  bat first.
-Even for test matches, toss winning teams tend to be tilting less toward batting first than they once did.
-Pre-1950, more than 90\% of teams that won the toss elected to bat, whereas in the last two decades that number is closer to two-thirds.
-Perhaps players have gotten more fit and are willing to trade-off expecting to have to bat on the final days of the match in favor of taking advantage of playing conditions that are favorable to fielding first.
-As for ODIs, captains curiously erred heavily on the side of fielding first in the early decades of the format, but in the last thirty years, they have tilted more towards batting first, with just over half of toss-winning teams electing to bat.
+\textbf{Effects over time.} Historically, toss-winning teams tilted heavily in favor of batting first, but this declined over the course of the 20th century, and now they show no preference (on average) to bat first.
+Even for test matches, toss-winning teams tend to be tilting less toward batting first than they once did.
+Pre-1950, more than 90\% of the teams that won the toss elected to bat, whereas in the last two decades that number is closer to two-thirds.
+Perhaps players have gotten fitter and are willing to trade off expecting to have to bat on the final day of the match in favor of taking advantage of playing conditions that are favorable to fielding first.
+As for ODIs, teams interestingly heavily preferred fielding first in the early decades of the format, but in the last thirty years, they have tilted more towards batting first, with just over half of the toss-winning teams electing to bat.
 
 \textbf{Relative strength.} We see no notable differences in the toss-winning captain's choice to bat first across matches that are lopsided and those in which the teams are more evenly matched.
-This is especially true in tests, but also appears to be true in ODIs, though a slightly larger share of toss-winning teams elect to bat in matches that are lopsided (54.6\%) in comparison to ones that are less lopsided (51.9\%).
+This is especially true in Tests but also appears to be true in ODIs, though a slightly larger share of toss-winning teams elect to bat in matches that are lopsided (54.6\%) in comparison to ones that are less lopsided (51.9\%).
 
-\textbf{Continents.} Toss-winning captains show a preference to bat first in matches played in Oceania and Europe, more so than in other pitches.
-For example, in Europe (mainly England) $64.9\%$ of toss winning teams elect to bat, and in Oceania this number is $59.9\%$.
-By contrast, on Asian wickets it is only $50.1\%$.
+\textbf{Continents.} Toss-winning captains show a preference to bat first in matches played in Oceania and Europe, more so than on other pitches.
+For example, in Europe (mainly England) $64.9\%$ of toss-winning teams elect to bat, and in Oceania, this number is $59.9\%$.
+By contrast, on Asian wickets, it is only $50.1\%$.
 
-\textbf{Day/Night matches.} The ball being harder to see at night afflicts both the night-time batting team, as well as the fielding team, making it prone to dropping catches.
+\textbf{Day/Night matches.} The ball being harder to see at night afflicts both batting and fielding.
 But the difficulty for batsmen is probably greater than it is for fielders.
 Indeed, within the shorter formats, the variation between day matches and day/night matches is greatest when it comes to the choice to bat first.
-In day/night matches, $59\%$ of toss-winning teams elect to bat first, whereas in day matches just about only half of them do.
+In day/night matches, $59\%$ of toss-winning teams elect to bat first, whereas in day matches just about half of them do.
 
-\textbf{Seasonality.} The preference to bat first in England follows a broadly similar trend to the toss effect over the course of the year-- rising steadily over the summer, and drops in September.
+\textbf{Seasonality.} The preference to bat first in England follows a broadly similar trend to the toss effect over the course of the year-- rising steadily over the summer, and dropping in September.
 At its peak in August, top-winning captains choose to bat first $71.9\%$ of the time.
-At its lowest in April, they choose to bat $57.4\%$ of the time.
+At its lowest in April, the  teams that win the toss choose to bat $57.4\%$ of the time.
 
-\textbf{Rain delays.} Remarkably, there is an overall preference to field first in matches that ended up using the Duckworth-Lewis adjustment: only $43.5\%$ of toss winning teams elected to bat in these matches.
-If captains are at all able to predict rain (and other) delays, they seem to show a preference to bat a fewer number of overs and chase the adjusted target.
+\textbf{Rain delays.} Remarkably, there is an overall preference to field first in matches that ended up using the Duckworth-Lewis adjustment: only $43.5\%$ of toss-winning teams elected to bat in these matches.
+If captains have the ability to predict rain (and other) delays, they seem to prefer to bat fewer overs and chase the adjusted target.
 
@@
 
-Despite having reported the `reduced form' and `first stage' effects of the toss, we cannot not interpret the ratio of these effects as the causal effect of batting first.
+Despite having reported the `reduced form' and `first stage' effects of the toss, we cannot interpret the ratio of these effects as the causal effect of batting first.
 If we used the toss as an instrument for batting first, say, we could at best hope to estimate the local average treatment effect (LATE) of electing to bat on winning the match \citep{angrist1996identification}.
@@
 
-However, even the conditions under which the LATE is known to be identified are probably not satisfied in our setting.
+However, even the conditions under which the LATE is known to be identified are not satisfied in our setting.
 To see why, first note that our setting has two-sided noncompliance: there are teams that win the toss that bat first, and those that lose the toss that bat first as well.
 Then, recall that under two-sided noncompliance, the LATE is identified under a monotonicity assumption that rules out defiers.
-However, defiers are likely to be present in our setting: they correspond to dyads of opposing teams that walk on to the pitch, both intending to field first if they win the toss.
-In such a setting, the team that wins the toss will chose to field first and the team loses the toss will be compelled to bat first.
-If these defiers are present, which they likely are, the monotonicity assumption does not hold, and so we cannot assert that the IV estimates are giving us the LATE.
+However, defiers are likely to be present in our setting: they correspond to dyads of opposing teams that walk on to the pitch intending to field first if they win the toss.
+In such a setting, the team that wins the toss will choose to field first and the team that loses the toss will be compelled to bat first.
+If these defiers are present, which they likely are, the monotonicity assumption does not hold, so we cannot assert that the IV estimates are giving us the LATE.
 
-We make these points a bit more concrete.
-Recall that we saw that 56.5\% of teams that won the toss elect to bat, which corresponds to an excess of 6.5 percentage points over the benchmark case in which the toss has no effect on the batting/fielding decision.
+We make these points more concrete.
+Recall that we saw that 56.5\% of teams that won the toss elect to bat, which is 6.5 percentage points more than what you would expect if the toss had no effect on the batting/fielding decision.
 In addition, the reduced form estimated effect of winning the toss on winning the match was 1.8 percentage points, implying that toss winners win an excess of 0.9 percent of matches over parity.
 The instrumented effect of batting first would then be estimated as 0.9/6.5 = 13.9 percentage points.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sing of this number gives the estimated instrumented effect of electing to field first.} Can we interpret this as the average rate at which the complier teams that won the toss and elected to bat generated the excess wins?
-Not really, because these excess wins could have been generated by the defiers making good decisions to field.
+No, because these excess wins could have been generated by the defiers making good decisions to field.
 Since we saw that teams that win the toss and elect to field first have a higher win-loss differential than those that elect to bat first, it would be irresponsible to rule out this possibility.
@@
 
-% \section{Changing the Rules of the Toss}
+\section{Conclusion}
 
-% We end with a small but potentially important policy proposal that cricket boards might consider.
-The proposal is to eliminate the choice that the toss confers upon the toss-winning captain.
-How can this be done?
-The umpire assigns either of the two opposing captains to take heads, and flips the coin.
-If the coin lands heads, that captain must bat; otherwise, that captain must field.
-The captains do not get to make any choices.
+When the toss is unimportant, the game is fair: neither the batting side nor the fielding side has an advantage derived simply from the luck of the toss prior to the first ball being bowled.
+The toss appears to be unimportant in some settings, but important in others, like day/night matches.
 
-% What does this take away from cricket? What problem does it fix? \todo{Guys, we should have a discussion about this!}
+We end with a small but potentially important policy proposal that cricket boards might consider.
+We propose auctioning off the ability to choose whether to bat or bowl first.
+Ask teams to use their knowledge and experience to price the value of getting to pick whether to bowl or bat first.
+We propose giving teams the time to inspect the playing surface and then asking the teams to place bids for how many runs should be added to the other team's total.
+The team with the higher bid wins the right to choose whether to bowl or bat.
+(If the bids are tied, which suggests that teams price the ability to choose the same, an actual coin toss can be used to decide which team chooses and the runs would be added to the other team's score.) Such ways of resolving who has the right to make the choice to bat or field first could make games more competitive, and thus more enjoyable to many viewers.
+Cricket could benefit from our willingness think outside the box.
 
-%
-%This conversion rate is different in sign and considerably larger than the naive difference in win
-%probabilities between teams that bat first and those that don't, which is $-1.6$ percentage points.
-Since the rules of cricket imply that the toss cannot affect the outcome of the match other than through the decision of the toss-winning team to bat or field, this discrepancy is likely due to the complier population being very different from the never-taker or always-taker populations that form the bulk of our team-match data.
-The IV estimates are thus
-%telling us that some teams in some matches are able to take advantage of the decision to bat first when the batting choice really matters.
-%
-%Batting first is a strategic decision made by the captain of the team
-%that won the toss in any given game. As such, a naive comparison of
-%win rates between teams that chose to bat first versus those who
-%didn't is unlikely to uncover the causal effect of batting first since
-%the treatment (batting) is endogenous. To get around this problem,
-%we use the randomly assigned ability to chose to bat first
-%afforded to one of the two captains by virtue of winning the toss as
-%an instrument for batting first.
-Note that because the decision to field first is the prefectly colinear inverse of the decision to bat, we can flip the sign of our point estimates to interpret them as the instrumented effects of choosing to field.
-%
-%Since after winning the toss, a captain has the choice to bat field, instrumenting for batting first by the coin toss amounts to
-%analysing an encouragement experiment with one-sided non-compliance
-%. In this context, this corresponds to the outcome
-%where the captain that won the toss still chooses to field first. The
-%toss readily constitutes a valid instrument--- it
-%gives the captain the choice to bat first, and almost certainly
-%satisfies the exclusion restriction, as the toss is unlikely to have an independent effect
-%on the outcome of the match besides its effect on the order that teams bat.\footnote{The only other possibility is that winning or losing the toss may have a dire psychological effect that directly affects their performance.
-We believe this to be unlikely.}
-%Since we are in the ``just-identified'' case with a single endogenous
-%regressor (the choice to bat first) and a single instrument
-%(winning the toss), we can recover the causal instrumented effect of batting first as the ratio of the
-%reduced-form effect of the toss on win probability divided by the first stage effect of the
-%toss on the choice to bat.
-%
-%These assumptions are sufficient if we assume that the instrumented effect is constant in
-%the population. However, we believe this to be an implausible assumption and
-%therefore operate in the heterogeneous-effects framework proposed by
-%\citet{angrist1996identification}. This necessitates an additional
-%assumption--- that of {monotonicity}, which requires that the toss
-%only moves batting choices in one direction: teams that would have
-%otherwise chosen to bat first don't switch to fielding first upon
-%winning the toss.\footnote{Formally, if $D_1$
-%is the binary choice to bat upon winning the toss, while $D_0$ is the
-%assignment when losing the toss, then monotonicity  assumption is stated as
-%$\Prob{D_1 \geq D_0} = 1$.} Despite being unverifiable, this assumption is
-%is likely to be valid. By using the instrumental variables (IV) estimator, we are therefore
-%estimating the Local Average Treatment Effect (LATE) for
-%{compliers} -- teams that are able to harness the
-%available information about whether or not it is advantageous to bat
-%first, and chooses to do so when it is in fact there is such an
-%advantage.\footnote{An alternative formulation of the IV framework
-%proposed by
-%\citet{heckman2005structural} permits a
-%characterisation of treatment selection based on a latent variable
-%$U_i$, which yields a monotonicity assumption of the form $D =
-%\Indic{U_i \leq p(\vee{X}_i, Z_i)}$, where $U_i$ is an unobserved
-%propensity to take treatment (in our setting, choose to bat first
-%given available information), and $p(\vee{X}_i, Z_i)$ is the propensity score.
-%$U_i$ is normalised to be distributed on the unit interval and encodes
-%information about an individual $i$'s treatment choices
-%$\SetB{D_i(z)}_{z \in \Sett{Z}}$. This framework allows for
-%extrapolation of the LATE into other target parameters.}
-%
-%We report the naive OLS estimate of the estimated difference in win
-%rates in column 1 of Table \ref{table:iv_table}, which is
-%counter-intuitively small and negative. In column 2 we estimate this
-%specification separately across matches in which the toss-winner
-%chooses to bat first, and in column 3 we estimate it in matches in which the toss-winner
-%chooses to field first. We find that matches in which the
-%toss-winner chooses to field first exhibit a substantial first-batter
-%disadvantage, which suggests that captains are doing better than
-%choosing randomly between batting and fielding first. We report the
-%first stage effect of winning the toss on the choice to bat first in
-%column 4, and find that winning the toss increases the probability
-%of batting first by 13 percentage points, which is a sizeable effect.
-%Columns 5-7 report the IV estimates. We find that in
-%contrast to the naive OLS specification, the IV estimate gives a
-%sizeable first-batter advantage of between 13.4 and 13.9 percentage
-%points across specifications with team- and match-type fixed effects.
-%
-%Like with the toss, we then proceed to analyse heterogeneity in the instrumented effect. The results are depicted in Figures
-%\ref{fig:iv_het_TE1} - \ref{fig:iv_het_TE3}.  
-%
-
-%
-%
-%\subsection{Characterising Compliers} % (fold)
-%\label{sub:characterising_compliers}
-%
-%A complier in our setting is a team that is able to harness the
-%available information about whether or not it is advantageous to bat
-%first, and chooses to do so when it is in fact there is such an
-%advantage. Since the monotonicity assumption rules out defiers,
-%we can decompose the sample into {always takers}, {never
-%takers}, and {compliers}
-%\citep{angrist1996identification,Abadie2003-hj}. The fractions of each of these are given by 
-%\begin{align*}
-%\pi_{\text{compliers}} & = \Prob{D_1 > D_0 }= \Exp{D | Z = 1} - \Exp{D | Z = 0} \\
-%\pi_{\text{always-takers}} & = \Prob{D_1 = D_0 = 1 } = \Exp{D | Z = 0}  \\
-%\pi_{\text{never-takers}} & = \Prob{D_1 = D_0 = 0 } = 1 - \Exp{D | Z = 1}
-%\end{align*}
-%where $D_1$ is a dummy for batting upon winning the toss, $D_0$ is a dummy for batting first upon losing the toss, and $Z \in \{0,1\}$ is the outcome of the toss, with 1 coded as a win.
-Applying the above formulae to our sample yields 13\% compliers,
-%43.5\% always takers and 43.5\% never takers respectively.
-This is a small fraction of compilers, suggesting that teams are generally not very adept at capitalizing on the toss.
-However, it could also be that since a complier is a team-match, match conditions in which the toss would matter may not be that prevalent.
-We cannot tell.
-%
-%To better understand the population of compliers, we apply \citeauthor{Abadie2003-hj}'s (2003) kappa to compute the mean of covariates for compliers. The kappa can be applied to any
-%characteristic or outcome to get its mean for compliers by removing
-%the means for never and always takers using the kappa weights. %\footnote{Abadie's formulation is
-%%considerably more general and applies to any measurable function $g(Y,
-%%D, \vee{X})$ with finite variance.
-%%$$
-%%\Exp{g(Y, D, \vee{X}) | D_1 > D_0} = \frac{\Exp{\kappa \cdot g(Y, D,
-%%\vee{X})}}{\Prob{D_1 > D_0}} = \frac{\Exp{\kappa \cdot g(Y, D,
-%%\vee{X})}}{\Exp{\kappa}}
-%%$$. In our application, $g()$ simply returns $x_i$.
-%% }. 
-%
-%Kappa weights are defined by
-%$$
-%\kappa = 1 - \frac{D(1-Z)}{1 - \Prob{Z = 1| X}} - \frac{(1-D)
-%Z}{\Prob{Z=1|X}}
-%$$ where $X$ is a characteristic. We average a characteristic $X$ using kappa weights computed using the
-%above expression where $\Prob{Z = 1 | X}$  is estimated using a
-%logistic regression \cite[p
-%181-183]{angristMostlyHarmlessEconometrics2008}. Applying this procedure to some team-specific
-%characteristics yields the averages reported in Tables 4 and 5. We find
-%that compliers are marginally more likely to be underdogs (defined as
-%the lower rated team in the match, according to ELO ratings), away teams, and teams playing in
-%Asia.
-Among the teams that made the right decision to bat first after winning the toss, 78\% of them were playing day/night matches-- a remarkably high number.
-This suggests that it is relatively easier to make a good choice of whether to bat or field first in day/night matches.
-%
-%Another result that is noteworthy, is that over a quarter of team-match compliers are Australian teams-- by far the highest frequency of any international team.
-England, Pakistan, South Africa, and India are all essentially tied for second place with a frequency of around 15\% each.
-What is surprising is how poorly West Indian, Sri Lankan and Kiwi teams do in choosing to bat when it matters.
-These are strong teams with quite a bit of cricketing experience in comparison to teams like Bangladesh, Kenya and Zimbabwe.
-And yet they do not seem to enter the complier population at higher rates.\footnote{For this interpretation, the match conditions are unlikely to be driving these differences.
-What we mean by this is the following.
-Suppose that Aussie teams are better at capitalizing on choosing to bat or field because Aussie wickets are particularly more likely to provide the playing conditions in which choosing to bat or field matters; and Aussie teams play on Aussie wickets more often than other teams.
-However, we see that matches played in Asian wickets constitute more than half of the complier population.}
-%
-%%estimating the following
-%%system of equations in order to uncover the batting advantage:
-%%\begin{align}
-%%Y_{ij} & = \alpha_0 + \beta D_{ij} + \varepsilon_{ij} \label{eqn:struct_eqn} \\
-%%D_{ij} & = \gamma_0 + \delta Z_{ij} + \eta_{ij} \label{eqn:first_stage}
-%%\end{align}
-%%where $Y_{ij}$ is an indicator for the outcome of the match, $D_{ij}$ is the decision to bat first, $Z_{ij}$ is an indicator for winning the toss, $\alpha_0$ and $\gamma_0$ are constants, and $\varepsilon_{ij}$ and $\eta_{ij}$ are noise terms.
-$\delta$ is the first stage coefficient, and $\beta$ is the second stage coefficient of interest.
-%
-%
-%% subsection characterising_compliers (end)
-%
-%% \section{Duckworth-Lewis} % (fold)
-%% \label{sec:duckworth_lewis}
-%
-%% \subsection{Document effects of Duckworth Lewis} % (fold)
-%% \label{sub:document_effects_of_duckworth_lewis}
-%
-%% \missingfigure{Event study Test Cricket and ODIs around 1999}
-%
-%% \subsubsection{Improving on DL} % (fold)
-%% \label{ssub:improving_on_dl}
-%
-%% subsubsection improving_on_dl (end)
-%
-%% subsection document_effects_of_duckworth_lewis (end)
-%
-%% section improving_duckworth_lewis (end)
-%
-
-\section{Summing Up}
-
-When the toss is unimportant, the game is fair: neither the batting side nor fielding side has an advantage derived simply from the luck of the toss prior to the first ball being bowled.
-The toss appears to be unimportant in some settings, but important in others, like day/night matches.
-Although day/night matches are enjoyable, if we care about eliminating the influence of the toss, they should be avoided.
-Cricket pitches should be made durable, and with an eye to not disadvantaging teams that bat first or second.
-As for the much maligned Duckworth-Lewis formula, it looks as though there is no need to change it for now.
-We have no evidence that it is helping or harming teams that bat second.
+\newpage
 
@@
   \figinc{output/reduced_form_by_matchtype.pdf}
-  \caption{Effect of winning the toss on win probability decomposed across different
-  match types, over time, and by continent.}
+  \caption{Effect of winning the toss on win probability decomposed across different match types, over time, and by continent.}
   \label{fig:rf_het_TE}
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{output/het_rank_te_odi.pdf}
-    \caption{Treatment effect heterogeneity by ranking, toss and win history, and team dummies. Minimal evidence of effect heterogeneity}
-    \label{fig:omni_het_odi}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\textwidth]{output/het_rank_te_test.pdf}
-    \caption{Treatment effect heterogeneity for test matches by ranking, toss and win history, and team dummies. Minimal evidence of effect heterogeneity}
-    \label{fig:omni_het_test}
-\end{figure}
-
-% \clearpage
+\clearpage
 
@@
 
-%\begin{figure}
-%  \centering
-%  \includegraphics[width=0.9\textwidth,keepaspectratio]{output/first_stage_season.pdf}
-%  \caption{First stage (probability of choosing to bat first after winning the toss) by month in England}
-%  \label{fig:first_stage_seasonality}
-%\end{figure}
-%
-%
-%\clearpage
-%
-%
-%\begin{table}[!htb]
-%    \caption*{}
-%    \begin{minipage}{.5\linewidth}
-%      \caption{Team Characteristics}
-%      \centering
-%        \begin{tabular}{l|r}
-%        % \label{table:kappa_res1}
-%        \hline
-%        & $\Ol{X}$\\
-%        \hline
-%        underdog & 0.52\\
-%        \hline
-%        day/night & 0.78\\
-%        \hline
-%        home & 0.42\\ \hline \hline \\
-%        \textbf{continent} & \\
-%        \hline
-%        africa & 0.11\\
-%        \hline
-%        asia & 0.62\\
-%        \hline
-%        europe & 0.03\\
-%        \hline
-%        oceania & 0.24\\
-%        \hline
-%        \end{tabular}
-%    \end{minipage}%
-%    \begin{minipage}{.5\linewidth}
-%      \centering
-%        \caption{Team}
-%        \begin{tabular}{l|r}
-%        % \label{table:kappa_res2}
-%        \hline
-%        team & $\Ol{X}$\\
-%        \hline
-%        australia & 0.27\\
-%        \hline
-%        bangladesh & 0.01\\
-%        \hline
-%        england & 0.15\\
-%        \hline
-%        india & 0.15\\
-%        \hline
-%        kenya & 0.02\\
-%        \hline
-%        new zealand & 0.00\\
-%        \hline
-%        pakistan & 0.15\\
-%        \hline
-%        south africa & 0.14\\
-%        \hline
-%        sri lanka & 0.03\\
-%        \hline
-%        west indies & 0.04\\
-%        \hline
-%        zimbabwe & 0.04\\
-%        \hline
-%        \end{tabular}
-%    \end{minipage}
-%\end{table}
-%
-%
-
-%\renewcommand{\mkbibnamefamily}[1]{\textsc{#1}}
-%\printbibliography
-%
-%\appendix
-
 \end{document}

### 2023-09-04 ms/ms.tex
replicate (broad sense) + revise ms.

@@
-\documentclass[11pt,  letterpaper]{article}
+\documentclass[12pt, letterpaper]{article}
+\usepackage[titletoc,title]{appendix}
+\usepackage{color}
+\usepackage{booktabs}
+\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+\definecolor{dark-red}{rgb}{0.75,0.10,0.10}
+\definecolor{bluish}{rgb}{0.05,0.05,0.85}
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{naturalnames}{hyperref}
+\usepackage[margin=1in]{geometry}
+\usepackage[linkcolor=blue,
+			colorlinks=true,
+			urlcolor=blue,
+			pdfstartview={XYZ null null 1.00},
+			pdfpagemode=UseNone,
+			citecolor={bluish},
+			pdftitle={social proof}]{hyperref}
 
-\usepackage{mathpazo}
-\usepackage{fontspec}
-\setmainfont{TeX Gyre Pagella}
+%\newcites{SI}{SI References}
+\usepackage{natbib}
 
-\input{boilerplate}
+\usepackage{float}
+\usepackage{placeins}
 
-\title{Asymmetries in Cricket: The Effects of Winning the Toss on Winning the Match\thanks{This paper incorporates some of the text and results from Sood \& Willis's (2019) working paper: ``Fairly Random: The Impact of Winning the Toss on the Probability of Winning.'' The authors thank Matt Blackwell, Andrew Gelman, Donald Green, Jens Hainmueller, and Daniel Stone for their comments.}}
+\usepackage{geometry}  % see geometry.pdf on how to lay out the page. There's lots.
+\geometry{letterpaper} % This is 8.5x11 paper. Options are a4paper or a5paper or other...
+\usepackage{graphicx}  % Handles inclusion of major graphics formats and allows use of
+\usepackage{amsfonts,amssymb,amsbsy}
+\usepackage{amsxtra}
+\usepackage{verbatim}
+%\setcitestyle{round,semicolon,aysep={},yysep={;}}
+\usepackage{setspace} % Permits line spacing control. Options are:
+%\doublespacing
+%\onehalfspace
+%\usepackage{sectsty}    % Permits control of section header styles
+\usepackage{pdflscape}
+\usepackage{fancyhdr}   % Permits header customization. See header section below.
+\usepackage{url}        % Correctly formats URLs with the \url{} tag
+\usepackage{fullpage}   %1-inch margins
+\usepackage{multirow}
+\usepackage{verbatim}
+\usepackage{rotating}
+\setlength{\parindent}{3em}
+
+%\usepackage[T1]{fontenc}
+%\usepackage[bitstream-charter]{mathdesign}
+
+\usepackage{chngcntr}
+\usepackage{longtable}
+\usepackage{adjustbox}
+\usepackage{dcolumn}
+
+\usepackage[nameinlink, capitalize, noabbrev]{cleveref}
+
+\def\citeapos#1{\citeauthor{#1}'s (\citeyear{#1})}
+
+\makeatother
+
+\usepackage{footmisc}
+\setlength{\footnotesep}{\baselineskip}
+\makeatother
+\renewcommand{\footnotelayout}{\normalsize \doublespacing}
+
+% Colors
+\usepackage{color}
+
+\newcommand{\bch}{\color{blue}\em  }   % begin change
+\newcommand{\ying} {\color{orange}\em  }   % begin change
+\newcommand{\bgcd} {\color{purple}\em }
+\newcommand{\ech}{\color{black}\rm  }    % end change
+
+% Caption
+\usepackage{caption}
+%\captionsetup[subtable]{font=small,skip=0pt}
+\usepackage{subcaption}
+
+% tt font issues
+% \renewcommand*{\ttdefault}{qcr}
+\renewcommand{\ttdefault}{pcr}
+
+\usepackage{tocloft}
+
+\usepackage{lscape}
+\renewcommand{\textfraction}{0}
+\renewcommand{\topfraction}{0.95}
+\renewcommand{\bottomfraction}{0.95}
+\renewcommand{\floatpagefraction}{0.40}
+\setcounter{totalnumber}{5}
+\makeatletter
+\providecommand\phantomcaption{\caption@refstepcounter\@captype}
+\makeatother
+
+\usepackage{footmisc}
+\renewcommand{\footnotelayout}{\setstretch{2}}
+
+\title{Asymmetry in Cricket: The Effect of Winning the Toss on Winning the Match\thanks{Data and scripts needed to replicate the results are available at: https://github.com/outside-edge/asymmetry/}}
 
 \author{
-Apoorva Lal\thanks{Dept.~of Political Science, Stanford University;  \textsf{apoorval@stanford.edu}.} \;\;
-Derek Willis\thanks{News applications developer at ProPublica;  \textsf{dwillis@gmail.com}.}  \;\;
-Gaurav Sood\thanks{Independent researcher;  \textsf{gsood07@gmail.com}.} \;\;
-Avidit Acharya\thanks{Dept.~of Political Science and Graduate School of Business, Stanford University; \textsf{avidit@stanford.edu}.}
+Apoorva Lal\thanks{Data Scientist, Netflix;  \href{mailto:lal.apoorva@gmail.com}{lal.apoorva@gmail.com }.} \;\;
+Derek Willis\thanks{Lecturer in Data and Computational Journalism, University of Maryland;  \href{mailto:dwillis@gmail.com}{dwillis@gmail.com}.}  \;\;
+Gaurav Sood\thanks{Independent researcher;  \href{mailto:gsood07@gmail.com}{gsood07@gmail.com}; Corresponding author. Phone: 650 269 7946. Address: 156th Place NE, Redmond, WA, 98052.}\;\;
+Avidit Acharya\thanks{Department of Political Science and Graduate School of Business, Stanford University; \href{mailto:avidit@stanford.edu}{avidit@stanford.edu}.}
 }
 \date{\today}
-
-%\usepackage[
-%  backend=biber,
-%  style=authoryear,
-%  citestyle=authoryear,
-%  ]{biblatex}
-%\addbibresource{biblio.bib}
-%\renewbibmacro{in:}{}
 
@@
 \maketitle
+\doublespacing
 
@@
 In cricket, which team bats first is determined by a toss.
-Does the team that wins the toss have an advantage over the other team?
+Does the team that wins the toss do better?
 We find that, on average, the effect of winning the toss on winning the match is essentially nil in international limited-overs daytime matches but substantial (3.3 percentage points) in day/night matches.
-In Test cricket, the advantage is even larger (4.9 percentage points).
-The advantage has exploded in the last decade (to nearly 15 percentage points) after a steady decline.
-Surprisingly, the effect of the toss does not appear to vary between competitive and non-competitive pairings of teams.
-And as for the decision to bat or field first, we find the win-loss differential is significantly larger for teams that elect to field; the advantage is negligible among those that elect to bat.
-Lastly, and perhaps most surprisingly, in rain-delayed matches in which the chasing team's total is adjusted by the Duckworth-Lewis formula, we find that teams that elected to bat first did just as well as those that elected to field first.
+In Test cricket, the advantage is yet larger (4.9 percentage points).
+Surprisingly, the effect of winning the toss does not vary substantially across matches between evenly matched and mismatched teams.
+Lastly, and perhaps most surprisingly, in rain-delayed matches in which the chasing team's total is adjusted by the Duckworth-Lewis formula, we find that the effect of winning the toss is statistically indistinguishable from zero.
 
@@
 
-%\textbf{JEL codes:} C72
 \end{abstract}
-
-% #### ##    ## ######## ########   #######
-%  ##  ###   ##    ##    ##     ## ##     ##
-%  ##  ####  ##    ##    ##     ## ##     ##
-%  ##  ## ## ##    ##    ########  ##     ##
-%  ##  ##  ####    ##    ##   ##   ##     ##
-%  ##  ##   ###    ##    ##    ##  ##     ##
-% #### ##    ##    ##    ##     ##  #######
 
@@
 
-In competitive sports, every little thing matters.
-Nevertheless, many sports leave some key levers out of teams' control and in the hands of fate.
-In cricket, one such lever---the toss---has been the subject of much attention.
-In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview.
-To wrest the advantage, the captain merely needs to pick the side of the coin that will be left facing the sky after the toss and choose whichever one---batting first or batting second---gives the greater advantage.
+In competitive sports, teams fight for every little advantage.
+Many sports, however, leave some important things to chance.
+In cricket, a coin toss is used to decide which team bats first.
+The choice is widely seen as consequential.
+In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}.} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview.
 
-In this paper, we measure how well teams have been able to capitalize on any advantage conferred by winning the toss.
-To answer the question, we collected data on over 35,000 cricket matches across four formats (First Class, ODIs, T20s, and Tests) played over the course of more than one hundred fifty years.
+In this paper, we measure the advantage conferred by winning the toss.
+To answer the question, we collected data on over 35,000 cricket matches played over 150 years---the largest ever dataset assembled for the question, and more than 50 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}.
+In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into.
+To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on post coin-toss decisions.
 
 We find that winning the toss increases the chances of winning the match by 1.8 percentage points: toss winners prevail 39.2\% of the time and lose 37.4\% of the time; the rest of the matches are drawn or tied.
-The effect of winning the toss is the largest in Tests, where it is 4.9 percentage points and the smallest in ODIs and T20s, where it is essentially zero on average.
-That said, in day/night ODIs and T20s, the effect of winning the toss is 3.3 percentage points; in day-time matches, the effect is a precise zero.
-
-The effect of the toss in ODIs has remained relatively stable since the 1980's when the one-day format first became popular.
-In Tests, however, the importance of the toss has varied considerably.
-The effect of winning the toss in Test cricket was a little over five percentage points before 1990.
-It then dropped to almost 0 from 1990 to 2010.
-In the last decade, however, it has exploded, averaging 14.5 percentage points.
+The effect of winning the toss is the largest in Tests, where it is 4.9 percentage points, and the smallest in ODIs and T20s, where it is essentially zero on average.
+That said, in day/night ODIs and T20s, the effect of winning the toss is 3.3 percentage points.
 
@@
 
-With this, we turn to the batting first decision.
-When pitch and weather conditions are hard to read, and it is not obvious that fielding first has a clear advantage, many cricket commentators suggest that batting first is the appropriate choice: get runs on the board, and put the opposing team under psychological pressure by having them chase the target, worrying about ``required run rates,'' etc.
-We see that the effect of winning the toss on batting first is 13 percentage points.
-If one team bats, the other team must field, so this translates to saying that among toss winners, 56.5\% of the teams elect to bat, and the remaining 43.5\% choose to field.
-Thus, compared to the case where teams are randomly assigned to bat, i.e., the case in which the toss has no effect on the batting decision, toss-winning teams bat an excess of 6.5\% of matches.
-
-However, we find that of the toss winners that elected to bat, only 38.1\% won the match compared to 37.9\% that lost--- a small difference of 0.2 percentage points.
-On the other hand, of the toss winners that elected to field, 40.8\% won the match compared to 36.9\% that lost--- a larger difference of 3.9 percentage points.
-This means that 94\% of the 1.8 percentage point effect of winning the toss can be attributed to instances in which the toss winning team elected to field, and only 6\% of the effect can be attributed to the toss winning team electing to bat.
-This suggests that teams are better able to capitalize on the toss by electing to field when it is advantageous to do so.
-
-We also look at how the toss advantage breaks down between teams that elected to bat versus those that elected to field within the various subgroups mentioned above and others.
-Perhaps the most striking result from this analysis for many cricket followers is that there does not appear to be the case that toss winners that elect to bat first do better than those that elect to field first in rain delayed matches where the chasing total is adjusted according to the \citet{duckworth1998} formula.
+With this, we turn to what teams choose after winning the toss.
+When pitch and weather conditions are hard to read, many cricket commentators suggest that batting first is the appropriate default choice: get runs on the board and put the opposing team under psychological pressure by having them chase the target.
+That advice is borne out in the data.
+Toss winners elect to bat somewhat more often (56.5\% of the time) than they elect to field.
+However, of the toss winners that elect to bat, only 38.1\% win the match compared to 37.9\% that lose.
+On the other hand, of the toss winners that opt to field, 40.8\% win the match compared to 36.9\% that lose.
+This means that 94\% of the 1.8 percentage point effect of winning the toss can be attributed to instances in which the toss-winning team elected to field, and just 6\% of the effect can be attributed to the toss-winning team electing to bat.
 
 Lastly, a caveat about our findings.
-Our empirical approach estimates the extent to which teams have been able to capitalize on the advantage granted by winning the toss.
+Our empirical approach estimates the extent to which teams have capitalized on the advantage granted by winning the toss and not the theoretical advantage granted by the toss.
 The toss may matter a lot in principle, but teams may decide poorly.
-So, when we refer to ``toss effects'' in this paper, we do not refer to whether there is an advantage to winning the toss in principle, whether any such advantage is exploited in practice.
 
-Our results are pertinent to debates about how cricket should be governed.
-One argument is that since the toss is random, no team is advantaged by the format prior to the toss; and in fact the luck of the toss along with the strategic choice to bat first or second are integral parts of the game--- not very different from the luck of weather affecting the teams differently.
-
-An alternative argument, however, is that the toss is problematic because it advantages one team over the other even before the first ball is bowled; and care should be taken to mitigate this asymmetry--- for example, by making cricket pitches more durable so that they don't deteriorate so much so that batting on the final day of a test match is miserable.
-Otherwise, the toss-winning team could exploit this advantage by batting first, reducing the chance that it has to endure this misery.
-Moreover, the fact that no side is advantaged \emph{ex-ante} if the toss is random is not enough to justify not correcting any asymmetry created by the toss.
-In fact, the rules of cricket already try to correct the asymmetry between teams that bat first and those that bat second in one particular instance: In rain-delayed one-day matches, the Duckworth-Lewis formula is used to set a new target for the chasing team.\footnote{The argument that no side was advantaged \emph{ex-ante} could have been used to justify not doing any correction in such rain-delayed matches, and simply declaring that the team that batted first won the match because the chasing team was unable to reach its target.
-But the designers of one-day cricket were clearly unwilling to allow the bad luck of weather to have this much influence on the game.
-If addressing the misfortunes of weather are important, it may also be important to address the misfortune of losing the toss.}
-
-For those who take this latter viewpoint, a straightforward policy prescription that comes out of our analysis is that, at least for ODIs, day-night matches should be avoided.
-In general, when new formats are introduced (such as the recent introduction of T20s) care should be taken to ensure the toss has little effect.
-The designers of cricket should always consider ways of leveling the playing field for the competing teams starting from the point at which the game really begins--- the moment the first ball is bowled.
-
-We do not address these optimal design questions in this paper, but our results are relevant to these issues depending on what the designers of cricket seek to optimize.
-However, at the end of the paper, we propose (and discuss the possible consequences of) a subtle change to the rules governing the toss: Rather than tossing a coin to confer the choice to bat first or second upon the captain of the toss-winning team, umpires may consider simply tossing a coin to determine which team must bat first.
-
-% ########     ###    ########    ###
-% ##     ##   ## ##      ##      ## ##
-% ##     ##  ##   ##     ##     ##   ##
-% ##     ## ##     ##    ##    ##     ##
-% ##     ## #########    ##    #########
-% ##     ## ##     ##    ##    ##     ##
-% ########  ##     ##    ##    ##     ##
+The rest of the paper is organized as follows.
+We start by describing the data.
+We then analyze if the data are consistent with the toss being random.
+Next, we present the results.
+We end with discussing the policy implications.
 
@@
 
-\subsection{Creating the sample}
+We collected data on cricket matches from ESPNCricinfo using a purpose-built library \citep{Willis_python-espncricinfo_A_Python_2022} in June 2020.
+From the data we collected, we discarded List A matches that were not ODIs or T20s because they are considered a less professional category (with worse data recording).
+This leaves us with data on all the men's and women's cricket matches in four formats: T20, ODI, Test, and First Class.
+From this census of the relevant population, we discard matches that were abandoned without the toss being conducted, and a small number of limited-overs (ODI and T20) matches in which the minimum number of overs that must be bowled to establish a result were not bowled.
+(In ODIs, for example, each side must bat at least 20 overs for a result to be declared.) In all, we dropped 426 matches with ``no result,'' 3 matches that were ``abandoned'' or ``canceled,'' and another three where we have no record for who batted first.
+The data also suggest little reason to be concerned about biased attrition.
+The ``no result'' matches, which constitute the overwhelming majority of matches we discard, comprise of matches that were canceled for idiosyncratic reasons, ranging from the \href{https://www.espncricinfo.com/series/shell-tri-series-1991-92-61203/australia-women-vs-england-women-final-66975/full-scorecard}{a team being unable to bat because of the weather} to cancellations because of \href{https://www.espncricinfo.com/series/australia-tour-of-pakistan-1982-83-61392/pakistan-vs-australia-3rd-odi-64196/full-scorecard}{unwanted audience participation}.
 
-We obtained data on cricket matches from the ESPNCricinfo website, which covers international and domestic matches, and maintains statistics on a wide variety of matches across formats.\footnote{Information on the matches was downloaded in June 2020 using a purpose-built library available at \href{https://github.com/dwillis/python-espncricinfo}{https://github.com/dwillis/python-espncricinfo}.
-The data is available at \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}
+Our final sample comprises 35,357 cricket matches across four formats played over 150 years.
+We report the distribution of matches over time and by type in Figure \ref{fig:timeseries_decomp}.
+As the figure shows, the number of matches has exploded over the last two decades.
+This is largely due to the introduction of the enormously popular T20 format of the game; T20 matches constitute over a quarter of all matches played in the last decade.
 
-We focus on four formats: T20, ODI, Test, and First Class for both men's and women's teams.\footnote{We exclude List A matches that are not ODIs or T20s.
-These matches are considered a less professional category (with worse data recording) than the formats that we include.} The data are a near census of the relevant population, only excluding scheduled matches that were abandoned without the toss being conducted, and a small number of limited-overs matches in which the minimum number of overs that must be bowled to establish a result were not met.\footnote{In ODIs, for example, each side must bat at least 20 overs for a result to be declared.
-We have no reason to suspect that the small matches that were excluded because they did not meet this threshold could be biased enough to affect our conclusions.
-We drop 426 matches recorded as ``no result,'' 3 matches that are recorded as ``abandoned'' or ``canceled,'' and another three that don't record who batted first.
-The "no result" matches constitute the largest exclusion we make.
-They appear to be comprised of matches canceled for idiosyncratic reasons, ranging from the \href{https://www.espncricinfo.com/series/shell-tri-series-1991-92-61203/australia-women-vs-england-women-final-66975/full-scorecard}{a team being unable to bat because of the weather} to cancellations because of \href{https://www.espncricinfo.com/series/australia-tour-of-pakistan-1982-83-61392/pakistan-vs-australia-3rd-odi-64196/full-scorecard}{unwanted audience participation}.
-In all, we drop approximately 0.07\% of matches.}
+\section{Is The Toss Random?}
 
-Our final sample comprises 35,357 cricket matches across four formats played over the last 150 years.
-We report the distribution of matches over time and by type in Figure \ref{fig:timeseries_decomp}.
-As the figure shows, the number of matches has exploded over the last two decades, largely due to the introduction of the enormously popular Twenty-20 (T20) format of the game both in international and domestic tournaments.
-T20 matches constitute over a quarter of all matches played in the last decade.
-
-% subsubsection data_collection (end)
-
-\subsection{Is the toss random?}
-
-Since our analysis rests on the assumption that the coin toss is random, we start by examining this assumption.\footnote{Prior work by \citet{shafqat2015analysis} and \citet{dawson2009bat} conditions on post-treatment variables, whereas we use only the randomization of the toss to identify toss effects.} Is there any evidence that the coin toss is rigged---for example, does the home side enjoy the rub of the green more often?
-In order to study this, we compare the rates at which home teams, teams that won the toss in their last match, and underdogs (defined as the lower-ranked team whenever rankings are available for both teams) win the toss relative to their opponents.
-
-To examine whether the toss is rigged, we compute the odds of seeing as many successes in the same number of draws of a sequence of independent Bernoulli random variables with parameter 0.5.
-For example, home teams have won 50.81\% of tosses in the 6,377 matches for which we have data.
-The likelihood of tossing a fair coin 6,377 times and getting at least this percentage of heads is 20.1\%.
-This chance appears low, but not eyebrow-raisingly so.
-We find that toss wins are even more balanced for teams that won and lost the last toss, and for teams that are underdogs and favorites.
+Since our analysis rests on the assumption that the coin toss is random, we start by examining this assumption.
+Is there any evidence that the coin toss is rigged?
+For example, does the home side enjoy the rub of the green more often?
+To study this, we compare the rates at which home teams, teams that won the toss in their last match, and underdogs (defined as the lower-ranked team whenever rankings are available for both teams) win the toss relative to their opponents.
+We compute the odds of seeing as many successes in the same number of Bernoulli trials with a probability of 0.5.
+For example, home teams have won 50.8\% of tosses in the 6,377 matches for which we have data.
+The likelihood of tossing a fair coin 6,377 times and getting as large a percentage of heads is 20.1\%.
+We find that toss wins are even more balanced for teams that won (lost) the last toss, and for teams that are underdogs and favorites.
 Table \ref{table: balance} reports these balance results.
@@
 That said, we do see one potential source of concern.
-We find that teams that share their nationality with at least two of the three umpires of the match win 51.3\% tosses in the 4,044 international matches for which we have this data.
-The likelihood of tossing a fair coin 4,044 times and getting at least this share of heads is only 10\%.
+We find that teams that share their nationality with at least two of the three umpires of the match win 51.3\% of the tosses in the 4,044 international matches for which we have this data.
+The likelihood of tossing a fair coin 4,044 times and getting this large a share of heads is only 10\%.
 This number is certainly low enough to raise some eyebrows.
@@
 And vice versa, if otherwise.
-Given home teams win more matches and are seen as 'stronger' at home, it is possible that the results are slightly biased upwards.
-We investigate this point later in the paper, looking at effects post 1992.
+Given that home teams win more matches and are seen as stronger, it is possible that the results are biased upwards.
 For now, we proceed as if the toss was random.
 
-% ########  #######   ######   ######
-%    ##    ##     ## ##    ## ##    ##
-%    ##    ##     ## ##       ##
-%    ##    ##     ##  ######   ######
-%    ##    ##     ##       ##       ##
-%    ##    ##     ## ##    ## ##    ##
-%    ##     #######   ######   ######
+\section{The Effect of Winning the Toss on the Probability of Winning}
 
-\section{Do teams exploit the advantage of winning the toss?}
-
-To estimate the advantage of winning the toss, we run regressions on a team $\times$ match level dataset.
+To estimate the advantage of winning the toss, we run regressions on a team $\times$ match-level dataset.
 Let $i$ index teams, and $j$ index matches.
@@
 We report results in Table \ref{table:reduced_form}.
-We progressively introduce team, match-type, and time period fixed effects in columns 2, 3, and 4, and find that results are remarkably stable.
 The first column gives the baseline estimate of the toss effect.
-The 1.8 percentage points difference reflects the fact that toss winners win 39.2\% of the time and lose 37.4\% of the time; the remaining matches are drawn.
-The estimated effect remains at 1.7 percentage points when we progressively add various fixed effects.
+We progressively add team, match-type, and time period fixed effects in columns 2, 3, and 4, and find that the results are remarkably stable.
+The baseline estimate of 1.8 percentage points reflects the fact that toss winners win 39.2\% of the time and lose 37.4\% of the time; the remaining matches are drawn.
+The estimated effect shuffles to 1.7 percentage points when we add various fixed effects.
 
-We then proceed to examine the effect of the toss within various subgroups.
-We report the estimates graphically in Figures \ref{fig:rf_het_TE} - 4 with point estimates and sample sizes as annotations.
+We next examine the effect of the toss across various subgroups.
+We report the estimates graphically in Figures \ref{fig:rf_het_TE} - \ref{fig:rf_het_TE3} with point estimates and sample sizes as annotations.
 We summarize the results below:
-
-\textbf{Format.} Toss effects are most pronounced in Test matches followed by first class matches, and are statistically indistinguishable from zero in ODIs and T20s.
-We estimate the effect of the toss on winning a Test match to be 4.7 percentage points.
-The effect of the toss in first class matches is 2.3 percentage points.
+\begin{itemize}
+\item \textbf{Format.} Toss effects are most pronounced in Test matches followed by first-class matches, and are statistically indistinguishable from zero in ODIs and T20s.
+The effect of the toss on winning a Test match is 4.7 percentage points and 2.3 percentage points in first-class matches.
 These results support the conventional wisdom among lay cricket followers that the toss grants the greatest advantage in multi-day affairs.
-Pitches invariably deteriorate over the course of these matches, and the fourth innings in a Test match is often the most challenging time to bat.
-The pitch deteriorates far less over the course of a day, or in the case of T20s, a few hours.
+Pitches invariably deteriorate over the course of these matches, and the fourth innings of a Test match is often the most challenging time to bat.
+The pitch deteriorates far less over the course of a day or in the case of T20s, a few hours.
 Our null results for limited over matches also extend \citet{de1998winning}, who, based on analysis of data from 427 international one-day matches conclude that ``winning the toss at the outset of a match provides no competitive advantage.''
 
-\textbf{International vs.~Domestic matches.} Toss effects are virtually the same for international and domestic matches (point estimates are 1.7 percentage points in the former, and 1.9 percentage points in the latter).
+\item \textbf{International vs.~Domestic matches.} Toss effects are virtually the same for international and domestic matches (point estimates are 1.7 percentage points in the former and 1.9 percentage points in the latter).
 
-\textbf{Effects over time.} Toss effects appear to have declined over time from a high of 9.9 percentage points in the second half of the 19th century to 1.4 percentage points in the last decade.
-This trend may be a result of a changing composition of matches so we also looked at these trends within formats.
+\item \textbf{Effects over time.} Toss effects appear to have declined over time from a high of 9.9 percentage points in the second half of the 19th century to 1.4 percentage points in the last decade.
+This trend may be a result of a changing composition of matches so we also looked at trends within formats.
 T20s have not been played long enough for us to draw any conclusions about how toss effects have changed over time.
-But within the other formats, we find that toss effects follow a similar pattern, except that in Test cricket there is a peculiar uptick in the last decade: the toss effect jumps from 0 in the decade before to 14.5 percentage points now.
-We investigate potential sources of the ecological decline below but there are some factors that could be important drivers but for which we do not have the data.
-For example, cricketers may have become fitter, and able to play for longer periods without tiring as much as they once did, reducing the impact of the order in which they ball or bat.
+But in the other formats, except for Test cricket, where there has been an uptick in the last decade, we find a similar pattern as above.
+The potential reasons for the decline are numerous.
+For example, cricketers may have become fitter, and more able to play for long periods without tiring as much as they once did, reducing the impact of the order in which they batted.
 Pitches may have become more resilient to wear than they once were.
@@
 
-\textbf{Relative strength.} We expect the toss to be less important in matches where one team is clearly stronger than the other.
+\item \textbf{Relative strength.} We expect the toss to be less important in matches where one team is clearly stronger than the other.
 We expect the stronger team to win, whether it bats first or second.
 To test this hypothesis, we use team ranking data published by cricket's official governing body, the International Cricket Council (ICC).
-The median match is played between teams where the ICC rating difference is 17 points.
-We use that as the approximate splitting point, splitting the matches into teams where the gap in the ICC ratings of the two teams is less than 15 points and where it is more.
-In this case, we believe the standard errors are too large for us to arrive at any substantively meaningful conclusions.
-Statistically, we believe effects larger than 8\% are unlikely.
+The median difference between teams is 17 points.
+We use that as an approximate splitting point, splitting the matches into teams where the gap in the ICC ratings of the two teams is less than 15 points and where it is more.
+In this case, we believe the standard errors are too large for us to draw substantively meaningful conclusions.
 
-\textbf{Continents.} Toss effects vary only somewhat across continents, hovering around 2.5 percentage points for matches played in Asia, Oceania, and the Americas (though the estimates for the Americas are noisier in part due to the smaller number of matches played there).
-In Europe (effectively England) the toss effect is only 1.1 percentage points (though we cannot rule out a null effect), and in Africa it is essentially zero.
-Overall, despite there being much talk about the different kinds of wickets across the world, with wickets in certain regions being drier, or stickier, or subject to more wear, these differences do not seem to translate into substantial toss effect differences at least when we look at coarse geographic averages.
+\item \textbf{Continents.} Toss effects vary somewhat across continents, hovering around 2.5 percentage points for matches played in Asia, Oceania, and the Americas (though the estimates for the Americas are noisier in part due to the smaller number of matches played there).
+In Europe (effectively England) the toss effect is only 1.1 percentage points (though we cannot rule out a null effect), and in Africa, it is essentially zero.
 
-\textbf{Day/Night matches.} It is often claimed that the toss is more crucial in day-and-night matches.
-We find some evidence for this.
-In the limited overs formats where day/night matches take place (ODIs and T20s), the advantage of winning the toss in a day/night match is 3.3 percentage points, whereas the
+\item \textbf{Day/Night matches.} It is often claimed that the toss is more crucial in day-and-night matches.
+Part of the claim rests on the theory that batting is easier when there is dew, which generally comes in the evening.
+Another popular theory is that the ball used for day/night limited overs cricket is harder to see at night.
+We do have the data to adjudicate between the theories but have data that shows that the toss is more consequential in day/night limited over games.
+In the limited overs formats where day/night matches take place (ODIs and T20s), the advantage of winning the toss is 3.3 percentage points, compared to 0 for limited overs matches during the day.
 
-advantage of winning the toss in a limited overs match played during the day is zero.
-
-\textbf{Seasonality.} Weather is also thought to have a large impact on the playing conditions in cricket.
+\item \textbf{Seasonality.} Weather is thought to have a large impact on the playing conditions in cricket.
 Cool overcast weather, for example, is thought to aid swing bowling, especially on English pitches.
@@
 
-\textbf{Rain delays.} Finally, we look at the differential effects of winning the toss in ODI matches that were rain-delayed, causing the number of overs bowled in the second innings to be adjusted down.
-We examine the subset of matches where rain-delays result in revised targets computed using the \citet{duckworth1998} method.
+\item \textbf{Rain delays.} Finally, we look at the differential effects of winning the toss in ODI matches that were rain-delayed, causing the number of overs bowled in the second innings to be adjusted down.
+We examine the subset of matches where rain delays result in revised targets computed using the \citet{duckworth1998} method.
 This method has been the subject of much controversy, with many cricket fans suggesting that it is rigged against the chasing team.
-To our surprise, we find no evidence that toss winners are able to exploit the possibility of second-inning rain delays to their advantage.
-If anything, we see that the toss-winner has a \emph{disadvantage} of 2.4 percentage points in matches subject to the Duckworth-Lewis adjustment, though this point estimate is statistically indistinguishable from zero.\footnote{Toss-winners won 160 out of 333 ODIs with the Duckworth-Lewis adjustment (i.e.~48\%) and 3085 out of 6187 ODIs without the Duckworth-Lewis adjustment (49.9\%).} This, however, does not mean that the Duckworth-Lewis method is not rigged against the chasing team; that issue will be discussed further in the next section.
+To our surprise, we find no evidence that toss winners are able to exploit the prediction of rain delays to their advantage.
+If anything, we see that the toss-winners have a 2.4 percentage points \emph{disadvantage} in matches subject to the Duckworth-Lewis adjustment, though this point estimate is statistically indistinguishable from zero.\footnote{Toss-winners won 160 out of 333 ODIs with the Duckworth-Lewis adjustment (i.e.~48\%) and 3,085 out of 6,187 ODIs without the Duckworth-Lewis adjustment (49.9\%).}
 
-% subsubsection heterogeneous_treatment_effects (end)
+\end{itemize}
 
-% ########     ###    ########
-% ##     ##   ## ##      ##
-% ##     ##  ##   ##     ##
-% ########  ##     ##    ##
-% ##     ## #########    ##
-% ##     ## ##     ##    ##
-% ########  ##     ##    ##
-
-\section{Batting versus Fielding First}
+\section{Batting Vs. Fielding First}
 
 56.5\% of teams that won the toss elected to bat.
-This is not surprising given the conventional wisdom is that when you are not sure about whether to bat or field first, bat first.
+This is not surprising given the conventional wisdom is that when you are not sure about whether it is advantageous to bat or field first, bat first.
 The first column of Table \ref{table:iv_table} shows that 49.2\% of teams that batted first won the match, while 50.8\% of teams that fielded first won.
-The second column shows that among teams that won the toss and batted first, the ``win-loss differential'' is small: these teams won only 0.2 percentage points more frequently than they lost, winning 38.1\% of matches and losing 37.9\% of matches, as reported in the introduction.
+The second column shows that among teams that won the toss and batted first the ``win-loss differential'' is small: these teams won only 0.2 percentage points more frequently than they lost, winning 38.1\% of matches and losing 37.9\% of matches, as reported in the introduction.
 This win-loss differential is smaller than it is for teams that won the toss and elected to field.
@@
 
-\textbf{Format.} Teams appear to have a strong preference for batting first in longer match formats.
-In Test matches 72.8\% of teams that win the toss elected to bat, whereas in T20s and ODIs this share is only about 52.9\%.
+\begin{itemize}
+\item \textbf{Format.} Teams prefer batting first in longer match formats.
+In Test matches 72.8\% of teams that win the toss bat first, whereas in T20s and ODIs this share is only about 52.9\%.
 In first-class matches it is 56.9\%.
 
-\textbf{International vs. Domestic matches.} We see no big differences in the propensity to elect to bat first across international versus domestic matches.
+\item \textbf{International vs. Domestic matches.} We see no big differences in the propensity to bat first across international versus domestic matches.
 
-\textbf{Effects over time.} Historically, toss-winning teams tilted heavily in favor of batting first, but this declined over the course of the 20th century, and now they show no preference (on average) to bat first.
-Even for test matches, toss-winning teams tend to be tilting less toward batting first than they once did.
-Pre-1950, more than 90\% of the teams that won the toss elected to bat, whereas in the last two decades that number is closer to two-thirds.
+\item \textbf{Effects over time.} Historically, toss-winning teams tilted heavily in favor of batting first, but this declined over the course of the 20th century, and now they show no preference (on average) to bat first.
+Even in test matches, toss-winning teams tend to be tilting less toward batting first than they once did.
+Pre-1950, more than 90\% of the teams that won the toss elected to bat, whereas in the last two decades that number has been closer to two-thirds.
 Perhaps players have gotten fitter and are willing to trade off expecting to have to bat on the final day of the match in favor of taking advantage of playing conditions that are favorable to fielding first.
@@
 
-\textbf{Relative strength.} We see no notable differences in the toss-winning captain's choice to bat first across matches that are lopsided and those in which the teams are more evenly matched.
+\item \textbf{Relative strength.} We see no notable differences in the toss-winning captain's choice to bat first across matches that are lopsided and those in which the teams are more evenly matched.
 This is especially true in Tests but also appears to be true in ODIs, though a slightly larger share of toss-winning teams elect to bat in matches that are lopsided (54.6\%) in comparison to ones that are less lopsided (51.9\%).
 
-\textbf{Continents.} Toss-winning captains show a preference to bat first in matches played in Oceania and Europe, more so than on other pitches.
+\item \textbf{Continents.} Toss-winning captains show a preference to bat first in matches played in Oceania and Europe, more so than on other pitches.
 For example, in Europe (mainly England) $64.9\%$ of toss-winning teams elect to bat, and in Oceania, this number is $59.9\%$.
@@
 
-\textbf{Day/Night matches.} The ball being harder to see at night afflicts both batting and fielding.
+\item \textbf{Day/Night matches.} The ball being harder to see at night afflicts both batting and fielding.
 But the difficulty for batsmen is probably greater than it is for fielders.
@@
 
-\textbf{Seasonality.} The preference to bat first in England follows a broadly similar trend to the toss effect over the course of the year-- rising steadily over the summer, and dropping in September.
+\item \textbf{Seasonality.} The preference to bat first in England follows a broadly similar trend to the toss effect over the course of the year-- rising steadily over the summer and dropping in September.
 At its peak in August, top-winning captains choose to bat first $71.9\%$ of the time.
@@
 
-\textbf{Rain delays.} Remarkably, there is an overall preference to field first in matches that ended up using the Duckworth-Lewis adjustment: only $43.5\%$ of toss-winning teams elected to bat in these matches.
+\item \textbf{Rain delays.} Remarkably, there is an overall preference to field first in matches that ended up using the Duckworth-Lewis adjustment: only $43.5\%$ of toss-winning teams elected to bat in these matches.
 If captains have the ability to predict rain (and other) delays, they seem to prefer to bat fewer overs and chase the adjusted target.
+
+\end{itemize}
 
@@
 
-Despite having reported the `reduced form' and `first stage' effects of the toss, we cannot interpret the ratio of these effects as the causal effect of batting first.
+Despite reporting the `reduced form' and `first stage' effects of the toss, we cannot interpret the ratio of these effects as the causal effect of batting first.
 If we used the toss as an instrument for batting first, say, we could at best hope to estimate the local average treatment effect (LATE) of electing to bat on winning the match \citep{angrist1996identification}.
-This is because in our setting treatment effects likely vary across groups: some teams in some circumstances will be better able to capitalize on the ability to bat or field first.
+This is because in our setting treatment effects likely vary across groups: some teams in some circumstances will be better able to capitalize on batting first.
 
-However, even the conditions under which the LATE is known to be identified are not satisfied in our setting.
-To see why, first note that our setting has two-sided noncompliance: there are teams that win the toss that bat first, and those that lose the toss that bat first as well.
+However, even the conditions under which the LATE is identified are not satisfied in our setting.
+To see why, first note that our setting has two-sided noncompliance: there are teams that win the toss that bat first and those that lose the toss that bat first as well.
 Then, recall that under two-sided noncompliance, the LATE is identified under a monotonicity assumption that rules out defiers.
-However, defiers are likely to be present in our setting: they correspond to dyads of opposing teams that walk on to the pitch intending to field first if they win the toss.
+However, defiers are likely to be present in our setting: they correspond to dyads of opposing teams that walk onto the pitch intending to field first if they win the toss.
 In such a setting, the team that wins the toss will choose to field first and the team that loses the toss will be compelled to bat first.
@@
 In addition, the reduced form estimated effect of winning the toss on winning the match was 1.8 percentage points, implying that toss winners win an excess of 0.9 percent of matches over parity.
-The instrumented effect of batting first would then be estimated as 0.9/6.5 = 13.9 percentage points.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sing of this number gives the estimated instrumented effect of electing to field first.} Can we interpret this as the average rate at which the complier teams that won the toss and elected to bat generated the excess wins?
+The instrumented effect of batting first would then be estimated as 0.9/6.5 = 13.9 percentage points.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sign of this number gives the estimated instrumented effect of electing to field first.} Can we interpret this as the average rate at which the complier teams that won the toss and elected to bat generated the excess wins?
 No, because these excess wins could have been generated by the defiers making good decisions to field.
@@
 
-\section{Conclusion}
+\section{Policy Proposals}
 
-When the toss is unimportant, the game is fair: neither the batting side nor the fielding side has an advantage derived simply from the luck of the toss prior to the first ball being bowled.
-The toss appears to be unimportant in some settings, but important in others, like day/night matches.
+The data suggest that in certain kinds of cricket matches, e.g., Day/Night matches and Tests, the team that wins the toss wins the match significantly more often.
+However, from one perspective, this is not a problem.
+Given that the toss is random, no team has an advantage in expectation.
+But this view has at least two problems.
+First, for the advantage to average out, we need a series of games.
+This isn't always the case, especially in big tournaments.
+Second, in each match, the toss grants an advantage to one of the teams before the match starts.
+This makes matches less interesting when the decision of who bats first is pivotal to the outcome.
+Furthermore, elsewhere, the rules of cricket already try to correct the asymmetry between teams that bat first and those that bat second.
+In rain-delayed one-day matches, the Duckworth-Lewis formula is used to set a new target for the chasing team.
+The argument that no side was advantaged \emph{ex-ante} could have been used to justify not doing any correction in such rain-delayed matches, and simply declaring that the team that batted first won the match because the chasing team was unable to reach its target.
+But the designers of one-day cricket were clearly unwilling to allow the bad luck of weather to have this much influence on the game.
+If addressing the misfortunes of weather is important, it may also be important to address the misfortune of losing the toss.
 
-We end with a small but potentially important policy proposal that cricket boards might consider.
-We propose auctioning off the ability to choose whether to bat or bowl first.
+For those who take this latter view, there are a variety of policy levers.
+For instance, to address the large advantage in day/night ODIs and T20s, one way institute a policy of not holding day/night matches.
+Or, as some English counties have done, the right to choose whether to bat or bowl first can be given to the visiting team.
+Or, we may want to build wickets that do not deteriorate appreciably.
+Or, institute a baseball-like innings system where short innings are interspersed so that no one team benefits from the conditions.
+All of these solutions likely have deleterious side effects.
+For instance, making pitches less prone to deterioration means that certain kinds of bowling (spin) will have less of an impact robbing the game of an interesting aspect.
+A better solution may be to auction off the ability to choose whether to bat or bowl first \citep{sood2021}.
 Ask teams to use their knowledge and experience to price the value of getting to pick whether to bowl or bat first.
-We propose giving teams the time to inspect the playing surface and then asking the teams to place bids for how many runs should be added to the other team's total.
-The team with the higher bid wins the right to choose whether to bowl or bat.
+We propose giving teams time to inspect the playing surface and then asking the teams to place bids for how many runs should be added to the other team's total.
+The team with the highest bid wins the right to choose whether to bowl or bat.
 (If the bids are tied, which suggests that teams price the ability to choose the same, an actual coin toss can be used to decide which team chooses and the runs would be added to the other team's score.) Such ways of resolving who has the right to make the choice to bat or field first could make games more competitive, and thus more enjoyable to many viewers.
-Cricket could benefit from our willingness think outside the box.
 
+\section*{Acknowledgements}
+This paper incorporates some of the text and results from Sood \& Willis's (2016) working paper \href{https://arxiv.org/abs/1605.08753}{Fairly Random: The Impact of Winning the Toss on the Probability of Winning} \citep{sood2016fairly}.
+The authors thank Matt Blackwell, Andrew Gelman, Donald Green, Jens Hainmueller, and Daniel Stone for their comments.
 \newpage
+\bibliographystyle{agsm}
+\bibliography{biblio.bib}
 
-\bibliography{biblio.bib}
+\clearpage
+
+\begin{table}[htb!]\centering
+\caption{Is the toss random? \label{table: balance}}
+\begin{tabular}{llll}
+\hline
+ & Toss Wins & Matches & $p$ value\\ \hline
+Home team & 3240 & 6377 & 0.201\\
+Last toss winners & 17189 & 34283 & 0.611\\
+Underdog & 2001 & 4054 & 0.423\\
+Home Umpire &  2075	& 4044	& 0.098 \\ \hline
+\end{tabular}
+\end{table}
+
+\clearpage
+
+\begin{table}[htb!]
+\centering
+\caption{The effect of winning the toss on the probability of winning.}
+\input{output/reduced_form_table.tex}
+\label{table:reduced_form}
+\end{table}
+
+\clearpage
+
+\begin{table}[htb!]
+\scriptsize
+\centering
+\caption{The choice to bat first and the probability of winning.}
+\input{output/iv_table.tex}
+\label{table:iv_table}
+\end{table}
 
@@
   \includegraphics[width=0.6\textwidth,keepaspectratio]{output/matchcounts.pdf}
-  \caption{Number of matches by format over time}
+  \caption{Number of matches by format over time.}
   \label{fig:timeseries_decomp}
@@
 
-\begin{table}[b!]\centering
-\begin{tabular}{llll}
- & Toss Wins & Matches & $p$ value\\ \hline
-Home team & 3240 & 6377 & 0.201\\
-Last toss winners & 17189 & 34283 & 0.611\\
-Underdog & 2001 & 4054 & 0.423\\
-Home Umpire &  2075	& 4044	& 0.098 \\
-\end{tabular}
-\caption{Balance \label{table: balance}}
-\end{table}
-
-% \clearpage
-
-\begin{table}
-\centering
-\caption{Reduced Form Effect of Toss on win probability}
-\input{output/reduced_form_table.tex}
-\label{table:reduced_form}
-\end{table}
+\clearpage
 
@@
   \centering
-  \figinc{output/reduced_form_by_matchtype.pdf}
-  \caption{Effect of winning the toss on win probability decomposed across different match types, over time, and by continent.}
+  \includegraphics[scale=.5]{output/reduced_form_by_matchtype.pdf}
+  \caption{The effect of winning the toss on win probability decomposed across different match types, over time, and by continent.}
   \label{fig:rf_het_TE}
@@
 
-% \clearpage
+\clearpage
 
@@
   \centering
-  \figinc{output/reduced_form_by_format_overtime.pdf}
-  \caption{Effect of winning the toss on win probability decomposed by format over time}
+  \includegraphics[scale=.5]{output/reduced_form_by_format_overtime.pdf}
+  \caption{The effect of winning the toss on the probability of winning by format over time.}
   \label{fig:rf_het_TE2}
@@
   \centering
-  \figinc{output/reduced_form_by_rank_dl_season.pdf}
-  \caption{Effect of winning the toss on win probability decomposed by competitiveness, use of DL, and Seasonality}
+  \includegraphics[scale=.5]{output/reduced_form_by_rank_dl_season.pdf}
+  \caption{The effect of winning the toss on the probability of winning by competitiveness, use of DL, and seasonality.}
   \label{fig:rf_het_TE3}
@@
 
-\begin{table}
-\scriptsize
+\begin{figure}
  \centering
-\caption{Batting first and win probability}
-\input{output/iv_table.tex}
-\label{table:iv_table}
-\end{table}
+ \includegraphics[scale=.5]{output/first_stage_by_matchtype.pdf}
+ \caption{The effect of winning the toss on the choice to bat first across international and domestic matches, formats, and continents.}
+ \label{fig:fs_het_TE1}
+\end{figure}
+
+\clearpage
 
@@
  \centering
- \figinc{output/first_stage_by_matchtype.pdf}
- \caption{Toss effects on choice to bat decomposed across different match
- types, timing, and continent.}
- \label{fig:fs_het_TE1}
+ \includegraphics[scale=.5]{output/first_stage_by_format_overtime.pdf}
+ \caption{The effect of winning the toss on the choice to bat first across different formats over time.}
+ \label{fig:fs_het_TE2}
 \end{figure}
+
+\clearpage
 
@@
  \centering
- \includegraphics[width=\textwidth,keepaspectratio]{output/first_stage_by_format_overtime.pdf}
- \caption{Toss effects on choice to bat decomposed across different formats
- over time}
- \label{fig:fs_het_TE2}
-\end{figure}
-
-\begin{figure}
- \centering
- \includegraphics[width=0.9\textwidth,keepaspectratio]{output/first_stage_by_rank_dl_season.pdf}
- \caption{Toss effects on choice to bat decomposed by ranking differences, rain-interruptions, and season (in England)}
+ \includegraphics[scale=.5]{output/first_stage_by_rank_dl_season.pdf}
+ \caption{The effect of winning the toss on the choice to bat first by rating differences, rain-interruptions, and season (in England)}
  \label{fig:fs_het_TE3}

### 2023-09-04 scripts/04_replication_by_gaurav_dplyr.R

@@ -1,0 +1,312 @@
+library(tidyverse)
+library(readr)
+library(ggplot2)
+library(magrittr)
+library(dplyr)
+library(stringr)
+library(lubridate)
+library(broom)
+
+# data
+odi_ranks  <- read_csv("data/rankings_odi.csv")
+test_ranks <- read_csv("data/rankings_test.csv")
+ranks <- rbind(odi_ranks, test_ranks)
+ranks$format <- toupper(ranks$format)
+
+# Grounds
+grounds <- read_csv("data/grounds.csv")
+
+# Match data
+match 		<- read_csv("data/final_output.csv")
+
+# %% drop matches with no toss info - mainly abandoned matches, with some weird county matches thrown in
+match %>%
+  filter(bat_or_bowl %in% c(0, NA)) %>%
+  group_by(outcome) %>%
+  summarise(N = n()) %>%
+  arrange(desc(N))
+
+match = match %>%
+  filter(!(bat_or_bowl %in% c(0, NA)))
+
+## Dates
+
+# Lubridate
+date_objects <- ymd(str_trim(match$date))
+
+# Extract day, month, year, and day of the week using lubridate functions
+result <- data.frame(
+  date = match$date,
+  day = day(date_objects),
+  month = month(date_objects),
+  year = year(date_objects),
+  day_of_week = weekdays(date_objects)
+)
+
+# Dedupe before left joining
+result <- result %>% distinct(date, .keep_all = TRUE)
+match <- left_join(match, result, by = "date")
+
+# %%# ## Match Types
+# International can be split by men, women, youth
+# Domestic matches apparently cannot be as we don't have that info.
+# See: https://github.com/dwillis/cricket-stats/issues/15
+
+match <- match %>%
+  mutate(
+    women = as.integer(grepl("Women", type_of_match)),
+    youth = as.integer(grepl("Youth", type_of_match)),
+    unofficial = as.integer(grepl("Unofficial", type_of_match))
+  )
+
+
+match <- match %>%
+  mutate(
+    type_of_match = case_when(
+      grepl("Test", type_of_match)        ~ "TEST",
+      grepl("ODI",  type_of_match)        ~ "ODI",
+      grepl("T20I", type_of_match)        ~ "T20I",
+      grepl("First-class", type_of_match) ~ "FC",
+      grepl("List A", type_of_match)      ~ "LISTA",
+      grepl("Twenty20", type_of_match)    ~ "T20",
[…]

### 2023-09-05 scripts/04_replication_by_gaurav_dplyr.R
add elo analysis

@@ -13,6 +13,25 @@ test_ranks <- read_csv("data/rankings_test.csv")
 ranks <- rbind(odi_ranks, test_ranks)
 ranks$format <- toupper(ranks$format)
 
+odi_elo_ranks <- read_csv("data/elo/odi_ratings_1972_2021.csv") %>% mutate(format = "ODI")
+test_elo_ranks <- read_csv("data/elo/test_ratings_1881_2021.csv") %>% mutate(format = "TEST")
+t20_elo_ranks <- read_csv("data/elo/t20i_ratings_2006_2021.csv") %>% mutate(format = "T20I")
+elo_ranks <- rbind(odi_elo_ranks, test_elo_ranks, t20_elo_ranks)
+
+date_objects <- ymd(str_trim(elo_ranks$date))
+
+# Extract day, month, year, and day of the week using lubridate functions
+result <- data.frame(
+  date = elo_ranks$date,
+  day = day(date_objects),
+  month = month(date_objects),
+  year = year(date_objects)
+)
+
+# Dedupe before left joining
+result <- result %>% distinct(date, .keep_all = TRUE)
+elo_ranks <- left_join(elo_ranks, result, by = "date")
+
 # Grounds
 grounds <- read_csv("data/grounds.csv")
 
@@ -121,9 +140,17 @@ match <- match %>%
 ranks <- ranks %>%
   mutate(
     unique = tolower(paste0(format, country, month, year))
-  )
+  ) %>%
+  distinct(unique, .keep_all = TRUE)
 
-# merge - the sood way
+
+elo_ranks <- elo_ranks %>%
+  mutate(
+    unique = tolower(paste0(format, teams, month, year))
+  ) %>%
+  distinct(unique, .keep_all = TRUE)
+
+# merge to rank
 match <- match %>%
   left_join(ranks[, c("rank", "rating", "unique")], by = c("team1_spid" = "unique")) %>%
   rename(team1_rank = rank, team1_rating = rating)
@@ -131,6 +158,14 @@ match <- match %>%
 match <- match %>%
   left_join(ranks[, c("rank", "rating", "unique")], by = c("team2_spid" = "unique")) %>%
   rename(team2_rank = rank, team2_rating = rating)
+
+match <- match %>%
+  left_join(elo_ranks[, c("m_ratings", "unique")], by = c("team1_spid" = "unique")) %>%
+  rename(elo_team1_rating = m_ratings)
+
+match <- match %>%
[…]

### 2023-09-05 ms/ms.tex
updated paper

@@
 
-\title{Asymmetry in Cricket: The Effect of Winning the Toss on Winning the Match\thanks{Data and scripts needed to replicate the results are available at: https://github.com/outside-edge/asymmetry/}}
+\title{Fairly Random: The Effect of Winning the Toss on Winning the Match\thanks{Data and scripts needed to replicate the results are available at: https://github.com/outside-edge/asymmetry/}}
 
@@
 Apoorva Lal\thanks{Data Scientist, Netflix;  \href{mailto:lal.apoorva@gmail.com}{lal.apoorva@gmail.com }.} \;\;
-Derek Willis\thanks{Lecturer in Data and Computational Journalism, University of Maryland;  \href{mailto:dwillis@gmail.com}{dwillis@gmail.com}.}  \;\;
-Gaurav Sood\thanks{Independent researcher;  \href{mailto:gsood07@gmail.com}{gsood07@gmail.com}; Corresponding author. Phone: 650 269 7946. Address: 156th Place NE, Redmond, WA, 98052.}\;\;
+Derek Willis\thanks{Lecturer in Data and Computational Journalism, Philip Merrill College of Journalism, University of Maryland;  \href{mailto:dwillis@gmail.com}{dwillis@gmail.com}.}  \;\;
+Gaurav Sood\thanks{Independent researcher;  \href{mailto:gsood07@gmail.com}{gsood07@gmail.com}.}\;\;
 Avidit Acharya\thanks{Department of Political Science and Graduate School of Business, Stanford University; \href{mailto:avidit@stanford.edu}{avidit@stanford.edu}.}
@@
 Many sports, however, leave some important things to chance.
-In cricket, a coin toss is used to decide which team bats first.
+In cricket, for instance, a coin toss is used to decide which team bats first.
 The choice is widely seen as consequential.
@@
 In this paper, we measure the advantage conferred by winning the toss.
-To answer the question, we collected data on over 35,000 cricket matches played over 150 years---the largest ever dataset assembled for the question, and more than 50 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}.
-In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into.
-To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on post coin-toss decisions.
+To answer the question, we collected data on over 35,000 cricket matches played over 150 years---the largest ever dataset assembled for the question, and more than 50 times larger than used before \citep[see,][]{dawson2009bat, de1998winning}.
+In analyzing these data, we avoid a common but critical pitfall that some other studies on the topic fall into.
+To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on decisions made after the toss, e.g., whether the team elects to bat or bowl first.
+(We tackle this issue at length later in the paper.)
 
@@
 We then analyze if the data are consistent with the toss being random.
-Next, we present the results.
+Next, we present the results on the effect of winning the toss on the probability of winning.
+Given that some previous papers condition on decisions made after the toss, we discuss what teams decide to do when they win the toss and why conditioning on post-treatment decisions could bias the inference.
 We end with discussing the policy implications.
@@
 
-\section{Batting Vs. Fielding First}
+\section{The Decision to Bat or Bowl First}
 
@@
 
-\subsection{The Difficulty in Estimating the Advantage to Electing to Bat/Field First}
+\subsection{The Difficulty in Estimating the Advantage of Choosing to Bat or Bowl First}
 
@@
 The instrumented effect of batting first would then be estimated as 0.9/6.5 = 13.9 percentage points.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sign of this number gives the estimated instrumented effect of electing to field first.} Can we interpret this as the average rate at which the complier teams that won the toss and elected to bat generated the excess wins?
-No, because these excess wins could have been generated by the defiers making good decisions to field.
+No, because these excess wins could have been generated by the defiers making good decisions to field first.
 Since we saw that teams that win the toss and elect to field first have a higher win-loss differential than those that elect to bat first, it would be irresponsible to rule out this possibility.

### 2023-10-23 ms/fin_submission/cricket.tex
final submission for jsa

--- 2023-09-05 ms/ms.tex
@@
 
-\author{
-Apoorva Lal\thanks{Data Scientist, Netflix;  \href{mailto:lal.apoorva@gmail.com}{lal.apoorva@gmail.com }.} \;\;
-Derek Willis\thanks{Lecturer in Data and Computational Journalism, Philip Merrill College of Journalism, University of Maryland;  \href{mailto:dwillis@gmail.com}{dwillis@gmail.com}.}  \;\;
-Gaurav Sood\thanks{Independent researcher;  \href{mailto:gsood07@gmail.com}{gsood07@gmail.com}.}\;\;
-Avidit Acharya\thanks{Department of Political Science and Graduate School of Business, Stanford University; \href{mailto:avidit@stanford.edu}{avidit@stanford.edu}.}
-}
+%\author{Apoorva Lal\thanks{Data Scientist, Netflix;  \href{mailto:lal.apoorva@gmail.com}{lal.apoorva@gmail.com }.} \;\; Derek Willis\thanks{Lecturer in Data and Computational Journalism, Philip Merrill College of Journalism, University of Maryland;  \href{mailto:dwillis@gmail.com}{dwillis@gmail.com}.}  \;\; Gaurav Sood\thanks{Independent researcher;  \href{mailto:gsood07@gmail.com}{gsood07@gmail.com}.}\;\; Avidit Acharya\thanks{Department of Political Science and Graduate School of Business, Stanford University; \href{mailto:avidit@stanford.edu}{avidit@stanford.edu}.}}
+
 \date{\today}
@@
  \centering
- \includegraphics[scale=.5]{output/first_stage_by_format_overtime.pdf}
+ \includegraphics[scale=.45]{output/first_stage_by_format_overtime.pdf}
  \caption{The effect of winning the toss on the choice to bat first across different formats over time.}

### 2023-10-23 ms/fin_submission/title_page.tex
final submission for jsa

--- 2023-09-05 ms/ms.tex
@@
 \date{\today}
-
 \begin{document}
@@
 \maketitle
-\doublespacing
-
-\begin{abstract}
-
-In cricket, which team bats first is determined by a toss.
-Does the team that wins the toss do better?
-We find that, on average, the effect of winning the toss on winning the match is essentially nil in international limited-overs daytime matches but substantial (3.3 percentage points) in day/night matches.
-In Test cricket, the advantage is yet larger (4.9 percentage points).
-Surprisingly, the effect of winning the toss does not vary substantially across matches between evenly matched and mismatched teams.
-Lastly, and perhaps most surprisingly, in rain-delayed matches in which the chasing team's total is adjusted by the Duckworth-Lewis formula, we find that the effect of winning the toss is statistically indistinguishable from zero.
-
-\smallskip
-
-\textbf{Key words:} cricket, fairness, randomization
-
-\end{abstract}
-
-\section{Introduction}
-
-In competitive sports, teams fight for every little advantage.
-Many sports, however, leave some important things to chance.
-In cricket, for instance, a coin toss is used to decide which team bats first.
-The choice is widely seen as consequential.
-In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}.} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview.
-
-In this paper, we measure the advantage conferred by winning the toss.
-To answer the question, we collected data on over 35,000 cricket matches played over 150 years---the largest ever dataset assembled for the question, and more than 50 times larger than used before \citep[see,][]{dawson2009bat, de1998winning}.
-In analyzing these data, we avoid a common but critical pitfall that some other studies on the topic fall into.
-To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on decisions made after the toss, e.g., whether the team elects to bat or bowl first.
-(We tackle this issue at length later in the paper.)
-
-We find that winning the toss increases the chances of winning the match by 1.8 percentage points: toss winners prevail 39.2\% of the time and lose 37.4\% of the time; the rest of the matches are drawn or tied.
-The effect of winning the toss is the largest in Tests, where it is 4.9 percentage points, and the smallest in ODIs and T20s, where it is essentially zero on average.
-That said, in day/night ODIs and T20s, the effect of winning the toss is 3.3 percentage points.
-
-Contrary to conventional wisdom, the effect of winning the toss does not vary dramatically across continents, though the toss appears less critical on European and African wickets than elsewhere.
-The effect on Aussie and Kiwi wickets (3 percentage points) is not substantially different from its effect on South Asian wickets (2.8 percentage points) but is higher than on European and African wickets (1.1 and 0 percentage points respectively).
-
-With this, we turn to what teams choose after winning the toss.
-When pitch and weather conditions are hard to read, many cricket commentators suggest that batting first is the appropriate default choice: get runs on the board and put the opposing team under psychological pressure by having them chase the target.
-That advice is borne out in the data.
-Toss winners elect to bat somewhat more often (56.5\% of the time) than they elect to field.
-However, of the toss winners that elect to bat, only 38.1\% win the match compared to 37.9\% that lose.
-On the other hand, of the toss winners that opt to field, 40.8\% win the match compared to 36.9\% that lose.
-This means that 94\% of the 1.8 percentage point effect of winning the toss can be attributed to instances in which the toss-winning team elected to field, and just 6\% of the effect can be attributed to the toss-winning team electing to bat.
-
-Lastly, a caveat about our findings.
-Our empirical approach estimates the extent to which teams have capitalized on the advantage granted by winning the toss and not the theoretical advantage granted by the toss.
-The toss may matter a lot in principle, but teams may decide poorly.
-
-The rest of the paper is organized as follows.
-We start by describing the data.
-We then analyze if the data are consistent with the toss being random.
-Next, we present the results on the effect of winning the toss on the probability of winning.
-Given that some previous papers condition on decisions made after the toss, we discuss what teams decide to do when they win the toss and why conditioning on post-treatment decisions could bias the inference.
-We end with discussing the policy implications.
-
-\section{Data}
-
-We collected data on cricket matches from ESPNCricinfo using a purpose-built library \citep{Willis_python-espncricinfo_A_Python_2022} in June 2020.
-From the data we collected, we discarded List A matches that were not ODIs or T20s because they are considered a less professional category (with worse data recording).
-This leaves us with data on all the men's and women's cricket matches in four formats: T20, ODI, Test, and First Class.
-From this census of the relevant population, we discard matches that were abandoned without the toss being conducted, and a small number of limited-overs (ODI and T20) matches in which the minimum number of overs that must be bowled to establish a result were not bowled.
-(In ODIs, for example, each side must bat at least 20 overs for a result to be declared.) In all, we dropped 426 matches with ``no result,'' 3 matches that were ``abandoned'' or ``canceled,'' and another three where we have no record for who batted first.
-The data also suggest little reason to be concerned about biased attrition.
-The ``no result'' matches, which constitute the overwhelming majority of matches we discard, comprise of matches that were canceled for idiosyncratic reasons, ranging from the \href{https://www.espncricinfo.com/series/shell-tri-series-1991-92-61203/australia-women-vs-england-women-final-66975/full-scorecard}{a team being unable to bat because of the weather} to cancellations because of \href{https://www.espncricinfo.com/series/australia-tour-of-pakistan-1982-83-61392/pakistan-vs-australia-3rd-odi-64196/full-scorecard}{unwanted audience participation}.
-
-Our final sample comprises 35,357 cricket matches across four formats played over 150 years.
-We report the distribution of matches over time and by type in Figure \ref{fig:timeseries_decomp}.
-As the figure shows, the number of matches has exploded over the last two decades.
-This is largely due to the introduction of the enormously popular T20 format of the game; T20 matches constitute over a quarter of all matches played in the last decade.
-
-\section{Is The Toss Random?}
-
-Since our analysis rests on the assumption that the coin toss is random, we start by examining this assumption.
-Is there any evidence that the coin toss is rigged?
-For example, does the home side enjoy the rub of the green more often?
-To study this, we compare the rates at which home teams, teams that won the toss in their last match, and underdogs (defined as the lower-ranked team whenever rankings are available for both teams) win the toss relative to their opponents.
-We compute the odds of seeing as many successes in the same number of Bernoulli trials with a probability of 0.5.
-For example, home teams have won 50.8\% of tosses in the 6,377 matches for which we have data.
-The likelihood of tossing a fair coin 6,377 times and getting as large a percentage of heads is 20.1\%.
-We find that toss wins are even more balanced for teams that won (lost) the last toss, and for teams that are underdogs and favorites.
-Table \ref{table: balance} reports these balance results.
-
-That said, we do see one potential source of concern.
-We find that teams that share their nationality with at least two of the three umpires of the match win 51.3\% of the tosses in the 4,044 international matches for which we have this data.
-The likelihood of tossing a fair coin 4,044 times and getting this large a share of heads is only 10\%.
-This number is certainly low enough to raise some eyebrows.
-That said, the practice of allowing home-team umpires in international matches ended in 1992, and so these issues are less relevant in recent decades which comprise the bulk of our sample.
-In addition, a rigged toss in favor of the home team would bias estimates of the advantage of winning the coin toss to the extent that it is correlated with ability.
-If stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards.
-And vice versa, if otherwise.
-Given that home teams win more matches and are seen as stronger, it is possible that the results are biased upwards.
-For now, we proceed as if the toss was random.
-
-\section{The Effect of Winning the Toss on the Probability of Winning}
-
-To estimate the advantage of winning the toss, we run regressions on a team $\times$ match-level dataset.
-Let $i$ index teams, and $j$ index matches.
-$Y_{ij}$ is a dummy for team $i$ winning match $j$, and takes a value of $0.5$ for draws.
-$Z_{ij}$ is a dummy for team $i$ winning the toss in match $j$.
-For the main specifications of interest, we estimate regression equations of the form
-
-\begin{equation}\label{eqn:reduced_form}
-Y_{ij} = \alpha_0 + \tau Z_{ij} + \gamma_i + \delta_t + \psi_k + \varepsilon_{ij}
-\end{equation}
-
-which include fixed effects for teams $\gamma_i$, time $\delta_t$, and match-type $\psi_k$. We cluster the standard errors at the match level.
-
-We report results in Table \ref{table:reduced_form}.
-The first column gives the baseline estimate of the toss effect.
-We progressively add team, match-type, and time period fixed effects in columns 2, 3, and 4, and find that the results are remarkably stable.
-The baseline estimate of 1.8 percentage points reflects the fact that toss winners win 39.2\% of the time and lose 37.4\% of the time; the remaining matches are drawn.
-The estimated effect shuffles to 1.7 percentage points when we add various fixed effects.
-
-We next examine the effect of the toss across various subgroups.
-We report the estimates graphically in Figures \ref{fig:rf_het_TE} - \ref{fig:rf_het_TE3} with point estimates and sample sizes as annotations.
-We summarize the results below:
-\begin{itemize}
-\item \textbf{Format.} Toss effects are most pronounced in Test matches followed by first-class matches, and are statistically indistinguishable from zero in ODIs and T20s.
-The effect of the toss on winning a Test match is 4.7 percentage points and 2.3 percentage points in first-class matches.
-These results support the conventional wisdom among lay cricket followers that the toss grants the greatest advantage in multi-day affairs.
-Pitches invariably deteriorate over the course of these matches, and the fourth innings of a Test match is often the most challenging time to bat.
-The pitch deteriorates far less over the course of a day or in the case of T20s, a few hours.
-Our null results for limited over matches also extend \citet{de1998winning}, who, based on analysis of data from 427 international one-day matches conclude that ``winning the toss at the outset of a match provides no competitive advantage.''
-
-\item \textbf{International vs.~Domestic matches.} Toss effects are virtually the same for international and domestic matches (point estimates are 1.7 percentage points in the former and 1.9 percentage points in the latter).
-
-\item \textbf{Effects over time.} Toss effects appear to have declined over time from a high of 9.9 percentage points in the second half of the 19th century to 1.4 percentage points in the last decade.
-This trend may be a result of a changing composition of matches so we also looked at trends within formats.
-T20s have not been played long enough for us to draw any conclusions about how toss effects have changed over time.
-But in the other formats, except for Test cricket, where there has been an uptick in the last decade, we find a similar pattern as above.
-The potential reasons for the decline are numerous.
-For example, cricketers may have become fitter, and more able to play for long periods without tiring as much as they once did, reducing the impact of the order in which they batted.
-Pitches may have become more resilient to wear than they once were.
-Teams may have become better informed about how to play more effectively whether they bat first or second, etc.
-
-\item \textbf{Relative strength.} We expect the toss to be less important in matches where one team is clearly stronger than the other.
-We expect the stronger team to win, whether it bats first or second.
-To test this hypothesis, we use team ranking data published by cricket's official governing body, the International Cricket Council (ICC).
-The median difference between teams is 17 points.
-We use that as an approximate splitting point, splitting the matches into teams where the gap in the ICC ratings of the two teams is less than 15 points and where it is more.
-In this case, we believe the standard errors are too large for us to draw substantively meaningful conclusions.
-
-\item \textbf{Continents.} Toss effects vary somewhat across continents, hovering around 2.5 percentage points for matches played in Asia, Oceania, and the Americas (though the estimates for the Americas are noisier in part due to the smaller number of matches played there).
-In Europe (effectively England) the toss effect is only 1.1 percentage points (though we cannot rule out a null effect), and in Africa, it is essentially zero.
-
-\item \textbf{Day/Night matches.} It is often claimed that the toss is more crucial in day-and-night matches.
-Part of the claim rests on the theory that batting is easier when there is dew, which generally comes in the evening.
-Another popular theory is that the ball used for day/night limited overs cricket is harder to see at night.
-We do have the data to adjudicate between the theories but have data that shows that the toss is more consequential in day/night limited over games.
-In the limited overs formats where day/night matches take place (ODIs and T20s), the advantage of winning the toss is 3.3 percentage points, compared to 0 for limited overs matches during the day.
-
-\item \textbf{Seasonality.} Weather is thought to have a large impact on the playing conditions in cricket.
-Cool overcast weather, for example, is thought to aid swing bowling, especially on English pitches.
-More generally, the advantage of winning the toss likely varies by weather.
-Although we do not have data on the weather, we can proxy it with seasons.
-In particular, students of the game suspect that the advantage of winning the toss in the early English season is especially significant.
-There is some evidence of a seasonal pattern for matches played in England, with the advantage of winning the toss growing over the summer, though the trend is weak and broken in the month of July.
-
-\item \textbf{Rain delays.} Finally, we look at the differential effects of winning the toss in ODI matches that were rain-delayed, causing the number of overs bowled in the second innings to be adjusted down.
-We examine the subset of matches where rain delays result in revised targets computed using the \citet{duckworth1998} method.
-This method has been the subject of much controversy, with many cricket fans suggesting that it is rigged against the chasing team.
-To our surprise, we find no evidence that toss winners are able to exploit the prediction of rain delays to their advantage.
-If anything, we see that the toss-winners have a 2.4 percentage points \emph{disadvantage} in matches subject to the Duckworth-Lewis adjustment, though this point estimate is statistically indistinguishable from zero.\footnote{Toss-winners won 160 out of 333 ODIs with the Duckworth-Lewis adjustment (i.e.~48\%) and 3,085 out of 6,187 ODIs without the Duckworth-Lewis adjustment (49.9\%).}
-
-\end{itemize}
-
-\section{The Decision to Bat or Bowl First}
-
-56.5\% of teams that won the toss elected to bat.
-This is not surprising given the conventional wisdom is that when you are not sure about whether it is advantageous to bat or field first, bat first.
-The first column of Table \ref{table:iv_table} shows that 49.2\% of teams that batted first won the match, while 50.8\% of teams that fielded first won.
-The second column shows that among teams that won the toss and batted first the ``win-loss differential'' is small: these teams won only 0.2 percentage points more frequently than they lost, winning 38.1\% of matches and losing 37.9\% of matches, as reported in the introduction.
-This win-loss differential is smaller than it is for teams that won the toss and elected to field.
-The third column of the table shows that these teams won 3.9 percentage points more frequently than they lost: they won 40.8\% of their matches and lost 36.9\% of them.
-Combining this information with the fact that 56.5\% of teams that won the toss elected to bat (and only 43.5\% elected to field, as implied by the fourth column of the table), we learn that 94\% ($= .435 \times .039/.018$) of the 1.8 percentage point advantage to winning the toss that we reported in the previous section can be attributed to the success of toss-winning teams that elected to field.
-To further understand what explains the preference to bat or bowl first, we see how winning the toss affects the choice in various subsets:
-
-\begin{itemize}
-\item \textbf{Format.} Teams prefer batting first in longer match formats.
-In Test matches 72.8\% of teams that win the toss bat first, whereas in T20s and ODIs this share is only about 52.9\%.
-In first-class matches it is 56.9\%.
-
-\item \textbf{International vs. Domestic matches.} We see no big differences in the propensity to bat first across international versus domestic matches.
-
-\item \textbf{Effects over time.} Historically, toss-winning teams tilted heavily in favor of batting first, but this declined over the course of the 20th century, and now they show no preference (on average) to bat first.
-Even in test matches, toss-winning teams tend to be tilting less toward batting first than they once did.
-Pre-1950, more than 90\% of the teams that won the toss elected to bat, whereas in the last two decades that number has been closer to two-thirds.
-Perhaps players have gotten fitter and are willing to trade off expecting to have to bat on the final day of the match in favor of taking advantage of playing conditions that are favorable to fielding first.
-As for ODIs, teams interestingly heavily preferred fielding first in the early decades of the format, but in the last thirty years, they have tilted more towards batting first, with just over half of the toss-winning teams electing to bat.
-
-\item \textbf{Relative strength.} We see no notable differences in the toss-winning captain's choice to bat first across matches that are lopsided and those in which the teams are more evenly matched.
-This is especially true in Tests but also appears to be true in ODIs, though a slightly larger share of toss-winning teams elect to bat in matches that are lopsided (54.6\%) in comparison to ones that are less lopsided (51.9\%).
-
-\item \textbf{Continents.} Toss-winning captains show a preference to bat first in matches played in Oceania and Europe, more so than on other pitches.
-For example, in Europe (mainly England) $64.9\%$ of toss-winning teams elect to bat, and in Oceania, this number is $59.9\%$.
-By contrast, on Asian wickets, it is only $50.1\%$.
-
-\item \textbf{Day/Night matches.} The ball being harder to see at night afflicts both batting and fielding.
-But the difficulty for batsmen is probably greater than it is for fielders.
-Indeed, within the shorter formats, the variation between day matches and day/night matches is greatest when it comes to the choice to bat first.
-In day/night matches, $59\%$ of toss-winning teams elect to bat first, whereas in day matches just about half of them do.
-
-\item \textbf{Seasonality.} The preference to bat first in England follows a broadly similar trend to the toss effect over the course of the year-- rising steadily over the summer and dropping in September.
-At its peak in August, top-winning captains choose to bat first $71.9\%$ of the time.
-At its lowest in April, the teams that win the toss choose to bat $57.4\%$ of the time.
-
-\item \textbf{Rain delays.} Remarkably, there is an overall preference to field first in matches that ended up using the Duckworth-Lewis adjustment: only $43.5\%$ of toss-winning teams elected to bat in these matches.
-If captains have the ability to predict rain (and other) delays, they seem to prefer to bat fewer overs and chase the adjusted target.
-
-\end{itemize}
-
-\subsection{The Difficulty in Estimating the Advantage of Choosing to Bat or Bowl First}
-
-Despite reporting the `reduced form' and `first stage' effects of the toss, we cannot interpret the ratio of these effects as the causal effect of batting first.
-If we used the toss as an instrument for batting first, say, we could at best hope to estimate the local average treatment effect (LATE) of electing to bat on winning the match \citep{angrist1996identification}.
-This is because in our setting treatment effects likely vary across groups: some teams in some circumstances will be better able to capitalize on batting first.
-
-However, even the conditions under which the LATE is identified are not satisfied in our setting.
-To see why, first note that our setting has two-sided noncompliance: there are teams that win the toss that bat first and those that lose the toss that bat first as well.
-Then, recall that under two-sided noncompliance, the LATE is identified under a monotonicity assumption that rules out defiers.
-However, defiers are likely to be present in our setting: they correspond to dyads of opposing teams that walk onto the pitch intending to field first if they win the toss.
-In such a setting, the team that wins the toss will choose to field first and the team that loses the toss will be compelled to bat first.
-If these defiers are present, which they likely are, the monotonicity assumption does not hold, so we cannot assert that the IV estimates are giving us the LATE.
-
-We make these points more concrete.
-Recall that we saw that 56.5\% of teams that won the toss elect to bat, which is 6.5 percentage points more than what you would expect if the toss had no effect on the batting/fielding decision.
-In addition, the reduced form estimated effect of winning the toss on winning the match was 1.8 percentage points, implying that toss winners win an excess of 0.9 percent of matches over parity.
-The instrumented effect of batting first would then be estimated as 0.9/6.5 = 13.9 percentage points.\footnote{Since the choice to field first is the perfectly colinear inverse of the choice to bat first, flipping the sign of this number gives the estimated instrumented effect of electing to field first.} Can we interpret this as the average rate at which the complier teams that won the toss and elected to bat generated the excess wins?
-No, because these excess wins could have been generated by the defiers making good decisions to field first.
-Since we saw that teams that win the toss and elect to field first have a higher win-loss differential than those that elect to bat first, it would be irresponsible to rule out this possibility.
-Consequently, it is unlikely the LATE can be identified in our setting.
-
-\section{Policy Proposals}
-
-The data suggest that in certain kinds of cricket matches, e.g., Day/Night matches and Tests, the team that wins the toss wins the match significantly more often.
-However, from one perspective, this is not a problem.
-Given that the toss is random, no team has an advantage in expectation.
-But this view has at least two problems.
-First, for the advantage to average out, we need a series of games.
-This isn't always the case, especially in big tournaments.
-Second, in each match, the toss grants an advantage to one of the teams before the match starts.
-This makes matches less interesting when the decision of who bats first is pivotal to the outcome.
-Furthermore, elsewhere, the rules of cricket already try to correct the asymmetry between teams that bat first and those that bat second.
-In rain-delayed one-day matches, the Duckworth-Lewis formula is used to set a new target for the chasing team.
-The argument that no side was advantaged \emph{ex-ante} could have been used to justify not doing any correction in such rain-delayed matches, and simply declaring that the team that batted first won the match because the chasing team was unable to reach its target.
-But the designers of one-day cricket were clearly unwilling to allow the bad luck of weather to have this much influence on the game.
-If addressing the misfortunes of weather is important, it may also be important to address the misfortune of losing the toss.
-
-For those who take this latter view, there are a variety of policy levers.
-For instance, to address the large advantage in day/night ODIs and T20s, one way institute a policy of not holding day/night matches.
-Or, as some English counties have done, the right to choose whether to bat or bowl first can be given to the visiting team.
-Or, we may want to build wickets that do not deteriorate appreciably.
-Or, institute a baseball-like innings system where short innings are interspersed so that no one team benefits from the conditions.
-All of these solutions likely have deleterious side effects.
-For instance, making pitches less prone to deterioration means that certain kinds of bowling (spin) will have less of an impact robbing the game of an interesting aspect.
-A better solution may be to auction off the ability to choose whether to bat or bowl first \citep{sood2021}.
-Ask teams to use their knowledge and experience to price the value of getting to pick whether to bowl or bat first.
-We propose giving teams time to inspect the playing surface and then asking the teams to place bids for how many runs should be added to the other team's total.
-The team with the highest bid wins the right to choose whether to bowl or bat.
-(If the bids are tied, which suggests that teams price the ability to choose the same, an actual coin toss can be used to decide which team chooses and the runs would be added to the other team's score.) Such ways of resolving who has the right to make the choice to bat or field first could make games more competitive, and thus more enjoyable to many viewers.
-
-\section*{Acknowledgements}
-This paper incorporates some of the text and results from Sood \& Willis's (2016) working paper \href{https://arxiv.org/abs/1605.08753}{Fairly Random: The Impact of Winning the Toss on the Probability of Winning} \citep{sood2016fairly}.
-The authors thank Matt Blackwell, Andrew Gelman, Donald Green, Jens Hainmueller, and Daniel Stone for their comments.
-\newpage
-\bibliographystyle{agsm}
-\bibliography{biblio.bib}
-
-\clearpage
-
-\begin{table}[htb!]\centering
-\caption{Is the toss random? \label{table: balance}}
-\begin{tabular}{llll}
-\hline
- & Toss Wins & Matches & $p$ value\\ \hline
-Home team & 3240 & 6377 & 0.201\\
-Last toss winners & 17189 & 34283 & 0.611\\
-Underdog & 2001 & 4054 & 0.423\\
-Home Umpire &  2075	& 4044	& 0.098 \\ \hline
-\end{tabular}
-\end{table}
-
-\clearpage
-
-\begin{table}[htb!]
-\centering
-\caption{The effect of winning the toss on the probability of winning.}
-\input{output/reduced_form_table.tex}
-\label{table:reduced_form}
-\end{table}
-
-\clearpage
-
-\begin{table}[htb!]
-\scriptsize
-\centering
-\caption{The choice to bat first and the probability of winning.}
-\input{output/iv_table.tex}
-\label{table:iv_table}
-\end{table}
-
-\clearpage
-
-\begin{figure}[tbh]
-  \centering
-  \includegraphics[width=0.6\textwidth,keepaspectratio]{output/matchcounts.pdf}
-  \caption{Number of matches by format over time.}
-  \label{fig:timeseries_decomp}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[b]
-  \centering
-  \includegraphics[scale=.5]{output/reduced_form_by_matchtype.pdf}
-  \caption{The effect of winning the toss on win probability decomposed across different match types, over time, and by continent.}
-  \label{fig:rf_het_TE}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[b]
-  \centering
-  \includegraphics[scale=.5]{output/reduced_form_by_format_overtime.pdf}
-  \caption{The effect of winning the toss on the probability of winning by format over time.}
-  \label{fig:rf_het_TE2}
-\end{figure}
-
-\begin{figure}[b]
-  \centering
-  \includegraphics[scale=.5]{output/reduced_form_by_rank_dl_season.pdf}
-  \caption{The effect of winning the toss on the probability of winning by competitiveness, use of DL, and seasonality.}
-  \label{fig:rf_het_TE3}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
- \centering
- \includegraphics[scale=.5]{output/first_stage_by_matchtype.pdf}
- \caption{The effect of winning the toss on the choice to bat first across international and domestic matches, formats, and continents.}
- \label{fig:fs_het_TE1}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
- \centering
- \includegraphics[scale=.5]{output/first_stage_by_format_overtime.pdf}
- \caption{The effect of winning the toss on the choice to bat first across different formats over time.}
- \label{fig:fs_het_TE2}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
- \centering
- \includegraphics[scale=.5]{output/first_stage_by_rank_dl_season.pdf}
- \caption{The effect of winning the toss on the choice to bat first by rating differences, rain-interruptions, and season (in England)}
- \label{fig:fs_het_TE3}
-\end{figure}
 
